---
title: "Deeper-Dive Analysis Framework: Non-College Graduates Without Meaningful Workforce Participation"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(openxlsx)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/Master-report-dataset.csv") 

kable(names(master.original))

kable(head(master.original))
```

# ¸Research Questions: Workforce Trajectories of Non-College Graduates in Central Minnesota

This phase of the analysis examines workforce participation trajectories among high school graduates in Central Minnesota who did not complete a postsecondary credential. Rather than relying on a single point-in-time workforce outcome, the analysis focuses on patterns of labor market attachment during the first five years after exiting formal education, with particular attention to persistence, instability, and regional attachment.

The research is guided by the following questions:

1.  Among non-college graduates, how persistent is being disconnected from sustained workforce participation either in the region or state?

2.  For non-college graduates who experience non-sustained workforce participation, where does employment occur when it is present?

3.  Among individuals categorized as having no Minnesota employment record during the first five years after high school, how persistent is this status?

4.  How are individual characteristics associated with workforce disconnection and subsequent employment trajectories?

<br>

# Study Population and Group Definitions

**Study Population**

The analytic population consists of individuals who:

- Graduated from a high school in Central Minnesota, and
- Did not complete a postsecondary credential (`ps.grad = No`), and,
- Have workforce participation records available for at least the first five years after high school graduation OR the first five years after exiting formal post-secondary education.

<br>

**Definition of Workforce Disconnection**

Workforce participation is measured annually during the first five years after high school or existing post-secondary attendance. Individuals are classified as **disconnected** in a given year if they are categorized as either:

- Non-sustained workforce participation, or
- No Minnesota employment record.

Using this annual classification, individuals are grouped based on the persistence of disconnection during years 1 through 5:

- **Persistently Disconnected:**\
  Disconnected in four or five of the first five years after graduation.

- **Disconnected (Broad Definition):**\
  Disconnected in three or more of the first five years after graduation.

- **Intermittently Disconnected:**\
  Disconnected in one or two of the first five years after graduation.

- **Consistently attached:**\
  Have sustained workforce participation in all five years.

These definitions emphasize sustained labor market detachment rather than short-term transitions.

<br>

**Regional Attachment**

For years in which individuals are employed but not meaningfully attached to the workforce, employment location is classified as:

- Within Central Minnesota, or
- Outside Central Minnesota (elsewhere in Minnesota).

This distinction is used to assess whether workforce disconnection reflects regional underutilization or disengagement from the regional labor market.

<br>

**Framing Consideration**

Together, these definitions allow the analysis to move beyond point-in-time workforce outcomes and toward a trajectory-based understanding of early-career labor market experiences. This approach supports clearer interpretation of workforce instability, recovery, and regional retention among non-college graduates in Central Minnesota.

<br>

# Methodology and Analytic Steps

This analysis examines workforce participation trajectories among non-college graduates in Central Minnesota, with a focus on identifying patterns of workforce disconnection, regional attachment, and employment stability during the first five years after high school.

## Step 1: Data Preparation and Analytic Population

**1.1 Analytic Population**

The analytic population is restricted to individuals who:

- Graduated from a high school in Central Minnesota, and
- Did not complete a postsecondary credential (`ps.grad = No`), and
- Have complete workforce participation categorizations for years 1 through 5 after high school or,
- first 5 years after exiting attendance in a post-secondary institution.

**1.2 Annual Workforce Participation Classification**

For each individual, annual workforce participation status is identified for: - `grad.year.1` - `grad.year.2` - `grad.year.3` - `grad.year.4` - `grad.year.5`

Each year is classified into one of the following categories:

- Meaningful workforce participation
- Not meaningful workforce participation
- No Minnesota employment record.

It should be noted that "grad.year.x" here means years after exiting formal education, not years after graduating high school.

**1.3 Workforce Disconnection Indicators**

For each annual observation, a binary **disconnection indicator** is created:

- **Disconnected (year-level)** = Non-sustained WF - Central or MN, **or** No Minnesota emp record

Using these annual indicators, individuals are grouped based on persistence of disconnection across years 1–5:

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Disconnected (Broad):** Disconnected in 3 of 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

These categories emphasize sustained labor market attachment or detachment rather than point-in-time outcomes.

**1.4 Regional Location of Not Meaningful Employment**

For years in which individuals are categorized as **Non-sustained WF**, employment location is further classified as:

- Within Central Minnesota
- Outside Central Minnesota (elsewhere in Minnesota)
- Both (simultaneous employment inside and outside the region)

This classification allows identification of regional underemployment versus regional disengagement.

**1.5 Employment Intensity and Job Holding**

For each individual and each year (years 1–5), the following employment intensity measures are constructed:

- Number of distinct employers during the year
- Indicator for multiple job holding (two or more employers)

These measures are used to distinguish underemployment from precarious or fragmented employment.

<br>

## Step 2: Descriptive and Summary Statistics

**2.1 Distribution of Workforce Disconnection**

Summary statistics are produced to describe:

- The proportion of non-college graduates in each workforce disconnection category
- Differences in persistence and intermittency of disconnection

This establishes the overall scope and severity of workforce disconnection in early adulthood.

**2.2 Regional Attachment and Job Structure**

Among individuals experiencing workforce disconnection, summary statistics examine:

- The location of not meaningful employment (within region, outside region, both)
- Average and median number of employers per year
- Prevalence of multiple job holding

These summaries clarify whether disconnection reflects local underemployment, out-of-region work, or fragmented employment patterns.

<br>

## Step 3: Relationship Between Individual Characteristics and Disconnection Trajectories

Step 3 will focus on answering the question of which characteristics are associated with the disconnection persistence categories. This will be a bi-variate analysis.

**Independent Variables**

We will use the following variables that were selected based on our earlier multiple correspondence analysis.

- Demographics
  - `Gender`
  - `RaceEthnicity`
  - `economic.status`
  - `SpecialEdStatus`
  - `english.learner`
  - `non.english.home`
- High school characteristics
  - `Dem_Desc`
  - `edr`
  - `avg.wages.pct.state`
  - `avg.unemp.rate`
- High school academics
  - `pseo.participant`
  - `ap.exam`
  - `took.ACT`
  - `ACTCompositeScore_filled`
  - `sat.taken`
  - `MCA.M`
  - `MCA.R`
  - `MCA.S`
  - `cte.achievement`
  - `total.cte.courses.taken`
- Post-secondary
  - `attended.ps`

**3.2 Analytic Approach**

- Categorical variables are examined using cross-tabulations and measures of association (e.g., Cramer’s V).
- Continuous variables are examined using ANOVA.
- Interpretation emphasizes substantive differences and patterns rather than statistical significance alone.

**3.3 Stratified and Comparative Analyses**

Where sample sizes permit, additional analyses compare:

- Persistently disconnected versus intermittently disconnected individuals
- Individuals disconnected within the region versus outside the region
- Individuals who transition into meaningful workforce participation versus those who do not

These comparisons help identify which characteristics are associated with stability, recovery, and regional retention.

<br>

## Framing and Interpretation

This methodology prioritizes a trajectory-based understanding of workforce participation, recognizing that early-career labor market experiences are dynamic rather than static. By combining persistence measures, regional employment location, and job-holding intensity, the analysis aims to provide actionable insights for workforce development and labor force planning in Central Minnesota.

<br>

# Data prep

## Step 1: Analytic population

The first step is filtering for our population;

- `ps.grad` = "No"

```{r}
analytic.pop <- master.original %>%
  filter(ps.grad == "No",
         grad.year.5 != "After 2023") 

kable(head(analytic.pop))
kable(names(analytic.pop))
```

## Step 2: Identify Workforce State for 5 Years After Formal Education

Next, we need to pull out all the individuals in which we have five years of workforce categories after they either graduate from high school or the last year in which they exited post-secondary attendance.

After that, we will categorize their workforce participation state into the following;

- Sustained WF - Central
- Sustained WF - MN
- Non-sustained WF
- No MN emp record.

I already created the annual workforce participation states for every individuals one year after graduation through 15 years after graduation. I will bring that list in and filter it by our analytic population.

```{r}
annual.wf.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  select(PersonID, grad.year, states) %>%
  mutate(grad.year = fct_relevel(grad.year, "grad.year.0", "grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5", "grad.year.6", "grad.year.7", "grad.year.8", "grad.year.9", "grad.year.10", "grad.year.11", "grad.year.12", "grad.year.13", "grad.year.14", "grad.year.15")) 

no_ps_attendance <- analytic.pop %>%
  filter(attended.ps == "No") %>%
  select(PersonID) %>%
  left_join(annual.wf.states, by = "PersonID") %>%
  filter(states != "After 2023",
         states != "Attending ps",
         grad.year %in% c("grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5")) %>%
  group_by(PersonID) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 5) %>%
  mutate(states = as_factor(states))

last_ps_attendance <- analytic.pop %>%
  filter(attended.ps == "Yes") %>%
  select(PersonID) %>%
  left_join(annual.wf.states, by = "PersonID") %>%
  filter(states != "After 2023") %>%
  mutate(numeric.grad.year = str_sub(grad.year, -2, -1),
         numeric.grad.year = str_replace(numeric.grad.year, "\\.", ""),
         numeric.grad.year = as.integer(numeric.grad.year)) %>%
  filter(states == "Attending ps") %>%
  group_by(PersonID) %>%
  summarize(last_ps_year = max(numeric.grad.year, na.rm = TRUE)) %>%
  ungroup() 

ps_attendance <- annual.wf.states %>%
  right_join(last_ps_attendance, by = "PersonID") %>%
  filter(states != "After 2023") %>%
  mutate(numeric.grad.year = str_sub(grad.year, -2, -1),
         numeric.grad.year = str_replace(numeric.grad.year, "\\.", ""),
         numeric.grad.year = as.integer(numeric.grad.year),
         final.year = last_ps_year + 5) %>%
  group_by(PersonID) %>%
  filter(numeric.grad.year <= final.year,
         numeric.grad.year > last_ps_year) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 5) %>%
  select(PersonID, grad.year, states, n) 

master.annual.wf.five.year.states <- ps_attendance %>%
  rbind(no_ps_attendance) %>%
  mutate(states = str_replace(states, "Not meaningful, not attending ps", "Non-sustained WF"),
         states = str_replace(states, "No MN emp record, not attending ps", "No MN emp record"),
         states = ifelse(states %in% c("Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region"), "Sustained WF - Central", states),
         states = str_replace(states, "Meaningful emp MN", "Sustained WF - MN"),
         states = fct_relevel(states, "Sustained WF - Central", "Sustained WF - MN", "Non-sustained WF", "No MN emp record")) %>%
  select(-n) %>%
  group_by(PersonID) %>%
  arrange(grad.year) %>%
  mutate(year.after.exit = seq(n())) %>%
  ungroup()


kable(head(master.annual.wf.five.year.states))

kable(names(master.annual.wf.five.year.states))

```

We now have a dataset consisting of `r comma(nrow(master.annual.wf.five.year.states))` individuals that meet the criteria that we want to examine along with their workforce categories from 1 to 5 years after exiting formal education.

<br>

## Step 3: Workforce Disconnection Indicators

Next we add another category to determine the level of disconnection, if any, for each individual. Disconnected is defined as any individual that was categorized as not having meaningful workforce participation or no Minnesota employment record in a given year one through five years after exiting formal education.

The level of disconnection is defined as the following;

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Broadly Disconnected:** Disconnected in 3 of the first 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

```{r}
disconnected.level <- master.annual.wf.five.year.states %>%
  mutate(disconnected = ifelse(states %in% c("Non-sustained WF", "No MN emp record"), "Disconnected", "Attached"),
         disconnected = as.factor(disconnected)) %>%
  group_by(PersonID, disconnected) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, disconnected, fill = list(n = 0)) %>%
  filter(disconnected == "Disconnected") %>%
  mutate(disconnected.cat = ifelse(n == 0, "Consistently attached",
                                   ifelse(n %in% c(1,2), "Intermittently disconnected",
                                          ifelse(n == 3, "Broadly disconnected",
                                                 ifelse(n %in% c(4,5), "Persistently disconnected", "A")))),
         disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected", "Persistently disconnected"))

kable(names(disconnected.level))
kable(head(disconnected.level))
```

The dataset now has `r comma(nrow(disconnected.level))` individuals that did not graduate college AND have 5 years of workforce categories immediately after exiting formal education categorized by workforce disconnection levels.

## **Step 4 & 5: Regional Location of Not Meaningful Employment**

For years in which individuals are categorized as **not meaningfully attached to the workforce**, employment location is further classified as: - Within Central Minnesota - Outside Central Minnesota (elsewhere in Minnesota) - Both (simultaneous employment inside and outside the region)

This classification allows identification of regional underemployment versus regional disengagement.

The population of individuals we need are the following;

- Did not graduate college
- For each individual categorize whether they had not meaningful workforce participation in the region, not in the region, or both.

We will use the `master.annual.wf.five.year.states` dataset as the starting point. We will have to have the year they graduated high school so when I match up employment records I'm only using employment records for the first 5 years after exiting formal education.

```{r}
hs.grad.year <- read_csv("Data/SLEDS/K12 Enrollment/K12Enrollment-master.csv")

analytic.pop.annual.wf.states.hs.grad.year <- master.annual.wf.five.year.states %>%
  left_join(hs.grad.year[,c(1,8)], by = "PersonID")  %>%
  rename(after.ed.year = grad.year.x,
         hs.grad.year = grad.year.y) %>%
  mutate(numeric.after.ed.year = str_sub(after.ed.year, -2, -1),
         numeric.after.ed.year = str_replace(numeric.after.ed.year, "\\.", ""),
         numeric.after.ed.year = as.integer(numeric.after.ed.year),
         emp.year = hs.grad.year + numeric.after.ed.year) 
  
kable(head(analytic.pop.annual.wf.states.hs.grad.year))

kable(names(analytic.pop.annual.wf.states.hs.grad.year))
```

Now we need to bring in the employment data for Central graduates and then filter for PersonID that are in the newly created dataset. The columns we are going to keep from the employment dataset are the following;

- PersonID: Identifies the graduate from Central Minnesota
- TimeID: Last day of fiscal quarter
- EmploymentOrganizationID: An internal identification for each employer.
- EmployeeQuarterlyHoursWorked: The total number of hours worked by an employee at an employer’s location during a specific calendar quarter
- EmployeeQuarterlyWages: The total wages earned by an employee at an employer’s location during a specific calendar quarter
- EmployeeTenure: The total number of quarters a worker has been employeed by an organization at a specific location, up to an including the current calendar quarter
- CountyName: The name of the county corresponding to the County and MNCounty Codes for this Location.

```{r}
employment.original <- read_csv("Data/SLEDS/Employment/Employment-Central.csv") %>%
  drop_na(PersonID)

emp.year <- employment.original %>%
  filter(!is.na(TimeID)) %>%
  mutate(emp.year = str_sub(TimeID, 1, 4),
         emp.year = as.integer(emp.year),
         quarter = str_sub(TimeID, 5,6),
         quarter = ifelse(quarter == "03", 1,
                          ifelse(quarter == "06", 2,
                                 ifelse(quarter == "09", 3,
                                        ifelse(quarter == "12", 4, 0))))) %>%
  select(PersonID, EmploymentOrganizationID, EmployeeQuarterlyHoursWorked, EmployeeTenure, CountyName, emp.year, quarter) %>%
  mutate(CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, " County", ""),
         CountyName = str_replace(CountyName, "Saint Louis", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Duluth", "St. Louis"),
         CountyName = str_replace(CountyName, "City of Minneapolis", "Hennepin"),
         CountyName = str_replace(CountyName, "City of St. Paul", "Ramsey"),
         CountyName = str_replace(CountyName, "Shakopee Mdewakanton Community", "Scott"),
         CountyName = str_replace(CountyName, "White Earth Tribal Counci", "Mahnomen")) %>%
  left_join(counties.regions.1[,c(1,2,10)], by = c("CountyName" = "Name"))

not.meaningful.pop.emp.year <- analytic.pop.annual.wf.states.hs.grad.year %>%
  filter(states == "Non-sustained WF") %>%
  select(PersonID, emp.year)

not.meaningful.emp.quarter <- not.meaningful.pop.emp.year %>%
  left_join(emp.year, by = c("PersonID", "emp.year")) %>%
  mutate(project.pr = ifelse(is.na(project.pr), "Unclassified", as.character(project.pr)),
         in_region = (project.pr == "Central"),
         unclassified_region = (project.pr == "Unclassified"),
         out_region = c(!in_region & !unclassified_region)) %>%
  group_by(PersonID, emp.year, quarter) %>%
  summarise(n.emp = n(),
         any_in_quarter = any(in_region, na.rm = TRUE),
         any_out_quarter = any(out_region, na.rm = TRUE),
         any_unclassified_quarter = any(unclassified_region, na.rm = TRUE),
         any_mixed_qtr =  (sum(c(any_in_quarter, any_out_quarter, any_unclassified_quarter), na.rm = TRUE) >= 2)) %>%
  ungroup()

not.meaningful.emp.annual <- not.meaningful.emp.quarter %>%
  group_by(PersonID, emp.year) %>%
  summarize(multiple_emp_same_quarter = any(n.emp > 1),
            max_emp_in_qtr = max(n.emp),
            any_in_region = any(any_in_quarter, na.rm = TRUE),
            any_out_region = any(any_out_quarter, na.rm = TRUE),
            any_unclassified_region = any(any_unclassified_quarter, na.rm = TRUE),
            any_mixed_region = any(any_mixed_qtr, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(region_location_year = case_when(any_in_region & !any_out_region & !any_unclassified_region ~ "in_only", !any_in_region & any_out_region & !any_unclassified_region ~ "out_only", !any_in_region & !any_out_region & any_unclassified_region  ~ "unclassified_only", TRUE ~ "mixed" )) %>%
  mutate(
    mixed_type = case_when(
      region_location_year != "mixed" ~ NA_character_,
      any_in_region & any_out_region & !any_unclassified_region ~ "in_out",
      any_in_region & any_unclassified_region & !any_out_region ~ "in_unclassified",
      any_out_region & any_unclassified_region & !any_in_region ~ "out_unclassified",
      any_in_region & any_out_region & any_unclassified_region  ~ "in_out_unclassified"
    ),
    both_structure = case_when(region_location_year != "mixed" ~ NA_character_,
                               any_mixed_region ~ "simultaneous",
                               !any_mixed_region ~ "sequential",
                               TRUE ~ NA_character_)) %>%
  select(-any_in_region, -any_out_region, -any_unclassified_region, - any_mixed_region)


kable(head(not.meaningful.emp.annual))

kable(names(not.meaningful.emp.annual))
```

We have created a dataset with `r comma(nrow(not.meaningful.emp.annual))` individuals that do not have meaningful workforce participation for any given year up to 5 years after graduating high school that also provides if they had multiple employers in one quarter, the max number of different employers in one quarter for that employment year, and whether the employer was located in the region, outside the region, unclassified, or a mixture of these categories. Another column further refines the "mixed" category into whether it was sequential - meaning the mix occurred in separate quarters - or simultaneous - meaning there was a mix within the same quarter.

Next, we will piece this back together with the entire analytic population and create a complete annual workforce state for all individuals that did not complete post-secondary and has 5 years of workforce categories after graduating high school.

```{r}
Master.not.meaningful <- not.meaningful.emp.annual %>%
  left_join(hs.grad.year[,c(1, 8)], by = "PersonID") %>%
  rename(hs.grad.year = grad.year) %>%
  mutate(grad.year = emp.year - hs.grad.year,
         grad.year = paste("grad.year.", grad.year, sep = "")) %>%
  complete(PersonID, grad.year) %>%
  mutate(grad.year = fct_relevel(grad.year, "grad.year.0", "grad.year.1", "grad.year.2", "grad.year.3", "grad.year.4", "grad.year.5", "grad.year.6", "grad.year.7", "grad.year.8", "grad.year.9", "grad.year.10", "grad.year.11", "grad.year.12", "grad.year.13", "grad.year.14", "grad.year.15"))


Master.annual.wf.states.disconnected <- master.annual.wf.five.year.states %>%
  left_join(Master.not.meaningful, by = c("PersonID", "grad.year")) %>%
  unite("wf.states.refined",
        states, region_location_year, mixed_type, both_structure,
        sep = "_",
        na.rm = TRUE) %>%
  mutate(year.after.exit = paste("after.ed.", year.after.exit, sep = "")) %>%
  select(PersonID, year.after.exit, wf.states.refined) %>%
  pivot_wider(names_from = year.after.exit, values_from = wf.states.refined) %>%
  left_join(disconnected.level[,c(1,4)], by = "PersonID")

kable(head(Master.annual.wf.states.disconnected)) 

kable(names(Master.annual.wf.states.disconnected))
```

Okay, now I need to bring this all together.

```{r}
master <- analytic.pop %>%
  right_join(Master.annual.wf.states.disconnected, by = "PersonID") %>%
  mutate(disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected", "Persistently disconnected"))

kable(head(master))

kable(names(master))

write_csv(master, "Data/Claude/Non-college deeper/Master-dataset.csv")
```

The data set has `r comma(nrow(master))` individuals in it. These individuals did not complete post-secondary and had workforce categories for 5 years after exiting formal education.

<br>

# Descriptive analysis

## **Distribution of Workforce Disconnection**

**Interpretation**

Workforce disconnection is the norm rather than the exception among non-college graduates in this cohort, not a marginal phenomenon. **Persistently disconnected is the single largest category by a wide margin**, accounting for nearly 44% of the population — meaning more than 2 in 5 non-completers were disconnected from meaningful workforce participation in 4 or 5 of their first 5 years post-exit. Combined with the "broadly disconnected" category, **56.8% of non-college graduates experienced disconnection in 3 or more of their first 5 years** — a clear majority.

By contrast, **only 15.6% show consistent workforce attachment across all 5 years** — the smallest of the four categories. This is a striking asymmetry: sustained, uninterrupted attachment is the least common trajectory in this population, while sustained disconnection is the most common one.

The "intermittently disconnected" group (27.6%, disconnected in only 1–2 years) represents the second-largest category, suggesting a meaningful subgroup experiences short-term instability rather than either full attachment or entrenched disconnection. Combined with the consistently attached group, **43.2% of non-completers experienced disconnection in 0–2 years** — i.e., relatively stable trajectories — versus the 56.8% who experienced substantial (3+ year) disconnection.

**Substantive takeaway:** for this population, disconnection is not a transitional or one-time event for most individuals — it's a persistent condition. This sets up the central question for Step 3's bivariate analysis: given that persistent disconnection is the modal outcome rather than an outlier one, which individual, high school, and academic characteristics differentiate the roughly 44% who are persistently disconnected from the roughly 16% who maintain full attachment.

```{r}
data <- master %>%
  group_by(disconnected.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         data_id = seq(n()),
         disconnected.cat = fct_relevel(disconnected.cat, "Consistently attached", "Intermittently disconnected", "Broadly disconnected",  "Persistently disconnected"))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-cat.csv")

plot <- ggplot(data, aes(disconnected.cat, pct, fill = disconnected.cat)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(disconnected.cat, "\nN: ", comma(n), "\n%: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "black", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of non-college grads,\n2008 - 2018 high school grads")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 5, "RdYlBu"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 8, height_svg = 8) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      


```

<br>

## **Regional Attachment and Job Structure**

Among individuals experiencing workforce disconnection, summary statistics examine: - The location of not meaningful employment (within region, outside region, both) - Average and median number of employers per year - Prevalence of multiple job holding

These summaries clarify whether disconnection reflects local underemployment, out-of-region work, or fragmented employment patterns.

Let's start with the distribution of the various workforce states with the newly defined categories within "Not meaningful" but not all of them because it will get too messy. Let's first do the following categories;

1.  Attending ps
2.  Meaningful emp region
3.  Meaningful emp MN
4.  Not meaningful
    1.  In only
    2.  Out only
    3.  Unclassified only
    4.  Mixed
5.  No MN emp record

**Interpretation**

This table tracks how the same population of non-completers is distributed across seven workforce states in each of the first five years after exiting formal education, and the shape of the change over time is the main story here — not any single year's snapshot.

**Sustained employment rises steadily, but its regional composition shifts.** Combined sustained employment (Central + broader MN) climbs from 35.1% in Year 1 to 48.1% by Year 5 — a substantively large gain, consistent with typical early-career labor market settling. But the two sustained categories don't grow at the same rate: **Sustained WF - Central's growth decelerates sharply and nearly flattens by Year 5** (year-over-year gains of +2.72, +1.29, +0.35, then just +0.11 percentage points), while **Sustained WF - MN keeps growing at a fairly steady \~2 points per year through Year 5** with no clear sign of leveling off. As a direct consequence, the *share* of all sustained workers who are specifically attached to the Central region — as opposed to elsewhere in Minnesota — declines steadily across the window, from 59.4% in Year 1 down to 52.6% by Year 5. In other words, later-arriving sustained employment among non-completers increasingly happens outside the region rather than within it, even as overall attachment improves. This is a meaningful regional-retention concern worth flagging directly: the longer a non-completer has been out of school, the more of their eventual "success" (sustained employment) is happening somewhere other than Central Minnesota.

**"No Minnesota employment record" rises every single year, with no reversal.** This category grows monotonically from 21.8% in Year 1 to 27.1% by Year 5, and unlike Sustained Central, shows no deceleration late in the window (the Year 4→5 change, +1.51 points, is actually larger than the Year 3→4 change). This is a notable pattern to highlight: one might expect this category to shrink over time as people eventually surface in the workforce data, but instead it accumulates steadily — by Year 5, more than 1 in 4 non-completers has no traceable Minnesota employment record at all, the single largest of the seven categories that year.

**The non-sustained/fragmented categories shrink consistently, and the "mixed" category shrinks fastest of all.** Every non-sustained category (in-only, out-only, unclassified, and mixed) declines every year. "Non-sustained WF - mixed" — people with simultaneous in- and out-of-region non-sustained employment — falls the furthest in relative terms, from 14.7% in Year 1 to 6.9% by Year 5, roughly halving. Combined, all non-sustained categories together fall from 43.2% of the population in Year 1 to just 24.8% by Year 5.

**Substantive takeaway:** Taken together, these three trends describe a genuine sorting dynamic rather than a uniform "settling down" story. As the fragmented, non-sustained employment categories shrink over the five-year window, that population doesn't move in only one direction — it splits between two diverging poles: rising sustained employment (increasingly located outside the region) **and** rising absence from Minnesota's employment record entirely, with the latter never leveling off. This directly motivates the trajectory-based framing of this analysis: a single point-in-time snapshot at Year 5 alone would show 48.1% sustained and 27.1% with no record, but would miss that the sustained group is quietly de-regionalizing over time, and that the "no record" group is still actively growing rather than stabilizing.

```{r}
wf.states.factors <- c("Attending ps", "Sustained WF - Central", "Sustained WF - MN", "Non-sustained WF_in_only", "Non-sustained WF_out_only", "Non-sustained WF_unclassified_only", "Non-sustained WF - mixed", "No MN emp record")

data <- Master.annual.wf.states.disconnected %>%
  select(PersonID:after.ed.5) %>%
  pivot_longer(names_to = "grad.year", values_to = "wf.states", 2:ncol(.)) %>%
  mutate(wf.states = ifelse(wf.states %in% c("Non-sustained WF_mixed_in_out_simultaneous", "Non-sustained WF_mixed_in_out_sequential", "Non-sustained WF_mixed_in_unclassified_simultaneous" , "Non-sustained WF_mixed_in_out_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_sequential", "Non-sustained WF_mixed_in_unclassified_sequential", "Non-sustained WF_mixed_in_out_unclassified_sequential"), "Non-sustained WF - mixed", wf.states),
         wf.states = fct_relevel(wf.states, wf.states.factors)) %>%
  group_by(grad.year, wf.states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-pattern.csv")

plot <- ggplot(data, aes(wf.states, pct, fill = wf.states)) +
  facet_wrap(~grad.year, ncol = 2) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nWorkforce category: ", wf.states, "\nN: ", comma(n), "\nPercent: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of total building construction establishments")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = c("#8c510a", "#bf812d", "#dfc27d", "#c7eae5", "#80cdc1", "#35978f", "#01665e", "black"),
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "none",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

```{r}
refined.disconnected.factors <- c( "Non-sustained WF_in_only",  "Non-sustained WF_out_only", "Non-sustained WF_unclassified_only", "Non-sustained WF_mixed_in_out_simultaneous", "Non-sustained WF_mixed_in_unclassified_simultaneous", "Non-sustained WF_mixed_out_unclassified_simultaneous", "Non-sustained WF_mixed_in_out_unclassified_simultaneous","Non-sustained WF_mixed_in_out_sequential", "Non-sustained WF_mixed_in_unclassified_sequential", "Non-sustained WF_mixed_out_unclassified_sequential", "Non-sustained WF_mixed_in_out_unclassified_sequential", "No MN emp record")

data <- Master.annual.wf.states.disconnected %>%
  select(PersonID:after.ed.5) %>%
  pivot_longer(names_to = "grad.year", values_to = "wf.state", 2:ncol(.)) %>%
  filter(!wf.state %in% c("Sustained WF - Central", "Sustained WF - MN", "Attending ps")) %>%
  mutate(wf.state = fct_relevel(wf.state, refined.disconnected.factors)) %>%
  group_by(grad.year, wf.state) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

write_csv(data, "Data/Claude/Non-college deeper/Descriptive stats/disconnected-refined-patterns.csv")

plot <- ggplot(data, aes(grad.year, pct, fill = wf.state)) +
  facet_wrap(~wf.state, ncol = 2) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nWorkforce category: ", wf.state, "\nN: ", comma(n), "\nPercent: ", percent(pct, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(pct, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Percent share of disconnected workforce states")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 16),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 8, height_svg = 12) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      
```

<br>

## Summary: What the Disconnection Analysis Tells Us

**Assessment: this table mostly confirms rather than adds to the prior findings, with one modest exception worth keeping.**

This table decomposes the "Non-sustained WF - mixed" category from the prior table into 8 sub-patterns (simultaneous vs. sequential, and which locations — in-region, out-of-region, unclassified — were combined). Note the denominator here is restricted to the non-sustained + no-record population each year (64.9% of the full cohort in Year 1, declining to 51.9% by Year 5), not the full population — worth flagging so the percentages in this table aren't compared directly to the full-population percentages in the earlier table without adjusting for that.

**What holds essentially constant across all 5 years — and is the one real finding here:** within the "mixed" category, the composition by pattern type barely moves. Simultaneous in-region + out-of-region employment consistently accounts for \~58–60% of all mixed cases every year, sequential in/out consistently accounts for \~21–24%, and every combination involving an "unclassified" location consistently accounts for a shrinking but always-small \~16–19% combined (and each of those four unclassified sub-patterns individually is under 1.3% of the *full* population in every year). This stability is worth stating once: when non-completers have fragmented employment spanning both inside and outside the region, it is roughly three times more likely to be simultaneous (holding multiple jobs across the regional boundary at once) than sequential (moving between in-region and out-of-region jobs within the year) — and that ratio doesn't shift across the five-year window even as the overall size of the mixed category shrinks by more than half.

**What doesn't add new substance:** the six individual "unclassified" combination sub-patterns are each small, declining, and structurally similar to each other — none tells a distinct story worth reporting individually. These could reasonably be collapsed into a single "other mixed" catch-all in your write-up without losing anything, since none of them exceeds roughly 1% of the full population in any year and their relative shares don't diverge from one another over time.

**Recommendation:** if you want one sentence from this table for the final write-up, it's the simultaneous-vs-sequential ratio holding steady at roughly 3:1 across all five years — everything else here is consistent with, rather than additive to, the year-over-year shrinkage of the mixed category already established in the prior table.

<br>

# Separating Non-sustained and No MN Emp Record

It's clear from the descriptive statistics that it's necessary to separate the Non-sustained from the MN Employment records to get a sense of how these two overlap or not. We are going to categorize just like we did before, except we will have to separate categories of "disconnection";

1.  non-sustained.persistence.cat:
    - 0 years = none;
    - 1-2 = intermittent;
    - 3 = broad;
    - 4-5 = persistent
2.  no_record.persistence.cat
    - same four buckets, applied to the no-record count instead

## Data Prep

```{r}
non_sustained.level <- master.annual.wf.five.year.states %>%
  mutate(non_sustained = ifelse(states == "Non-sustained WF", "Non-sustained", "Other"),
         non_sustained = as.factor(non_sustained)) %>%
  group_by(PersonID, non_sustained) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, non_sustained, fill = list(n = 0)) %>%
  filter(non_sustained == "Non-sustained") %>%
  mutate(non_sustained.cat = ifelse(n == 0, "Never non-sustained",
                                    ifelse(n %in% c(1,2), "Intermittently non-sustained",
                                           ifelse(n == 3, "Broadly non-sustained",
                                                  ifelse(n %in% c(4,5), "Persistently non-sustained", "A")))),
         non_sustained.cat = fct_relevel(non_sustained.cat, "Never non-sustained", "Intermittently non-sustained", "Broadly non-sustained", "Persistently non-sustained"))

no_record.level <- master.annual.wf.five.year.states %>%
  mutate(no_record = ifelse(states == "No MN emp record", "No record", "Other"),
         no_record = as.factor(no_record)) %>%
  group_by(PersonID, no_record) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  complete(PersonID, no_record, fill = list(n = 0)) %>%
  filter(no_record == "No record") %>%
  mutate(no_record.cat = ifelse(n == 0, "Never no-record",
                                ifelse(n %in% c(1,2), "Intermittently no-record",
                                       ifelse(n == 3, "Broadly no-record",
                                              ifelse(n %in% c(4,5), "Persistently no-record", "A")))),
         no_record.cat = fct_relevel(no_record.cat, "Never no-record", "Intermittently no-record", "Broadly no-record", "Persistently no-record"))

master.sep <- Master.annual.wf.states.disconnected %>%
  left_join(non_sustained.level, by = "PersonID") %>%
  left_join(no_record.level, by = "PersonID") %>%
  rename(non.sustained.n = n.x,
         no.record.n = n.y)

write_csv(master.sep, "Data/Claude/Non-college deeper/master-separated.csv")

```

**Non-Sustained Employment Persistence, Isolated from No-Record Status (N = 41,279)**

**Interpretation**

Separating non-sustained employment from no-MN-record status reshapes the picture considerably compared to the combined "disconnected" categorization from the earlier table, and this reshaping is itself the headline finding here.

**Once no-record status is pulled out, "persistent" disconnection largely disappears as the dominant category.** In the earlier combined categorization, "persistently disconnected" was the single largest group at 43.8%. Here, looking at non-sustained employment alone, "persistently non-sustained" drops all the way down to 14.2% — roughly a third of its former size. This is a substantively important signal: **a large share of what looked like persistent disconnection in the combined analysis appears to have been driven by persistent absence from Minnesota's employment record, not by repeated cycles of unstable in-state employment.** This can't be confirmed definitively until the no-record breakdown is in hand for direct comparison, but the magnitude of the shift is large enough to flag now.

**"Intermittently non-sustained" is now the largest single category by a wide margin (41.2%),** more than double its share in the combined table (27.6%). This suggests that when non-completers do experience unstable employment specifically (as opposed to leaving the employment record altogether), it's most commonly a short-term, 1–2 year experience rather than a chronic one — a meaningfully different story than the combined table implied.

**"Never non-sustained" — meaning years were either sustained employment or no-record, but never specifically non-sustained — accounts for 30.5%,** roughly double the "consistently attached" share (15.6%) from the combined table. Since "consistently attached" required sustained employment in *all five* years, while "never non-sustained" only requires the *absence* of non-sustained years (allowing some years to be no-record instead), this gap is expected — but it's a useful reminder that these two categories aren't measuring the same thing, and shouldn't be read as directly comparable figures.

**What this sets up:** the real value of this reframing will come from directly comparing this table to the equivalent no-record persistence breakdown. If persistent non-sustained employment (14.2%) turns out to be much smaller than persistent no-record status, that would support the interpretation that the earlier "persistently disconnected" finding was substantially a no-record phenomenon rather than an unstable-employment phenomenon — with real implications for how this population's core challenge should be characterized (data/record attrition vs. genuine employment instability).

```{r}
non_sustained.dist <- master.sep %>%
  group_by(non_sustained.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n),
         pct = scales::percent(pct))

kable(non_sustained.dist, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

**No-MN-Employment-Record Persistence, Isolated from Non-Sustained Employment (N = 41,279)**

**Interpretation**

This table, read alongside the non-sustained persistence table, resolves the question raised there — and the answer is more interesting than a simple "one type dominates" finding.

**Persistent no-record status (16.83%) is actually slightly *larger* than persistent non-sustained employment (14.20%)** — not smaller, which runs counter to the hypothesis raised after the non-sustained table alone. Both are far below the combined "persistently disconnected" figure of 43.76%, and it's that gap that turns out to be the real story here, not which of the two types is individually larger.

**The two "never" categories reveal that a substantial share of this population experiences *both* types of disconnection at different points, not predominantly one or the other.** Since "consistently attached" (0 combined disconnected years) is logically the exact overlap of "never non-sustained" and "never no-record," these three figures can be combined to decompose the full population into four mutually exclusive disconnection *patterns*:

- **15.6%** were consistently attached — zero non-sustained years and zero no-record years.
- **14.9%** had a "pure no-record" pattern — no non-sustained years, but at least one no-record year.
- **43.7%** had a "pure non-sustained" pattern — no no-record years, but at least one non-sustained year.
- **25.8%** — roughly **1 in 4 non-completers** — experienced *both* non-sustained employment and no-record status at different points across their first 5 years.

**This directly explains why "persistently disconnected" (43.8%) so far exceeds either persistence measure taken alone (14.2% and 16.8%).** A person doesn't need 4–5 years of *one* disconnection type to land in the combined "persistently disconnected" category — 2 non-sustained years plus 2 no-record years reaches that same 4-year threshold in the combined measure, while counting as only "intermittent" in *both* of the separated tables. The size of that gap (43.8% combined vs. \~31% if you simply added the two persistent categories together) is consistent with a meaningful share of the persistently disconnected population cycling between the two disconnection types rather than experiencing either one chronically.

**Substantive takeaway:** the original combined "persistently disconnected" finding wasn't primarily driven by either chronic unstable employment or chronic absence from Minnesota's records specifically — it's substantially a story of **compound instability**, where a quarter of non-completers move between both forms of disconnection over their first five years. This reframes the earlier finding in an important way: the policy-relevant group isn't cleanly "people who can't hold a job" versus "people who leave the state" — a large share are experiencing some mix of both, which likely calls for a different kind of intervention than either single-cause story would suggest on its own.

```{r}
no_record.dist <- master.sep %>%
  group_by(no_record.cat) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

kable(no_record.dist, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

**Decomposing "Persistently Disconnected" by Driver (N = 18,063)**

The table below showsonly the people who fall into "Persistently disconnected" under our *original* combined measure — meaning across their 5 years, they were disconnected (non-sustained OR no-record, in any mix) in 4 or 5 of those years.

The categories are defined as;

- **"Non-sustained driven"** = this person's disconnected years were mostly/entirely *unstable employment* (non-sustained), not disappearing from the record. Example: someone with 4 years of non-sustained work and 0 years of no-record. They were disconnected in the combined sense, but they were always at least working, just not steadily.
- **"No-record driven"** = the opposite — this person's disconnected years were mostly/entirely *no employment record at all*, not instability. Example: someone with 0 years non-sustained and 4 years no-record. They weren't cycling through shaky jobs; they mostly just weren't showing up in Minnesota's system at all.
- **"Mixed (neither individually persistent)"** = this person doesn't have 4+ years in *either* single category, but their two counts still *add up* to 4 or 5 disconnected years total. Example: 2 years non-sustained + 2 years no-record = 4 disconnected years total (enough to be "persistently disconnected" overall), but only 2 in each — not enough on its own to count as "persistent" in either individual measure. This person alternated between the two problems rather than being stuck in one.

**Interpretation**

This table restricts to the 18,063 people from the very first table who were persistently disconnected overall (4–5 combined disconnected years) and asks *why* — which of the two underlying problems actually drove their disconnection. It's worth noting explicitly that this table cross-validates the two persistence tables from earlier: "No-record driven" (n = 6,949) is the exact same group as "persistent no-record" from that table, and "Non-sustained driven" (n = 5,863) is the exact same group as "persistent non-sustained" — both match to the person. This makes sense given the definitions (having 4+ years of one type leaves at most 1 year for the other, so the two "driven" categories are mechanically mutually exclusive), and it confirms the two sets of tables are internally consistent with each other.

**No single story dominates the persistently disconnected population — it's a genuine three-way split.** "No-record driven" is the largest of the three at 38.5%, followed by "non-sustained driven" at 32.5%, with "mixed" close behind at 29.1%. No driver accounts for even 40% of this group on its own, and the gap between the largest and smallest category is under 10 points. This means the persistently disconnected population isn't one problem wearing different clothes — it's genuinely three distinguishable subpopulations of roughly comparable size.

**This sharpens (and slightly revises) the earlier framing.** The previous summary characterized the compound/mixed pattern and the two single-cause patterns as comparably important at the population level. Restricting specifically to the *worst-off* group — those who are persistently disconnected overall — shows that **no-record driven is actually the single most common explanation, not the mixed pattern.** Nearly 4 in 10 of the most disconnected people have essentially vanished from Minnesota's employment data for 4 or 5 of their first 5 years, without primarily cycling through unstable jobs. Just under a third (32.5%) are the opposite: consistently working, but never stably. And just under 3 in 10 (29.1%) don't fit cleanly into either single-cause story — they alternate between both problems, with neither one alone reaching the persistent threshold.

**Substantive and policy takeaway:** even within the single most concerning trajectory category (persistent disconnection), there isn't a one-size-fits-all explanation, and the three groups likely need genuinely different responses. The "non-sustained driven" third looks like a job-quality-and-stability problem — these are people who are engaged with the labor market but can't find steady footing. The "no-record driven" plurality looks more like a reconnection or data-visibility problem — people who may have left the state, entered informal/unreported work, or disengaged from the labor market entirely. The "mixed" group — nearly 3 in 10 — doesn't fit either framing cleanly and may represent the most complex cases: people cycling between unstable employment and periods of disappearing from the record altogether, suggesting a more fundamentally unstable trajectory rather than one specific, addressable barrier.

```{r}
persistent_decomp <- master.sep %>%
  filter(disconnected.cat == "Persistently disconnected") %>%
  mutate(driver = case_when(
    non_sustained.cat == "Persistently non-sustained" & no_record.cat != "Persistently no-record" ~ "Non-sustained driven",
    no_record.cat == "Persistently no-record" & non_sustained.cat != "Persistently non-sustained" ~ "No-record driven",
    non_sustained.cat == "Persistently non-sustained" & no_record.cat == "Persistently no-record" ~ "Both persistent",
    TRUE ~ "Mixed (neither individually persistent)"
  )) %>%
  group_by(driver) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(pct = n / sum(n))

kable(persistent_decomp, 
      format = "html",
      escape = FALSE,
      align = c("lcc")) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) 
```

## What This Table Tells Us: The Story Behind "Persistent Disconnection"

**The scope of the problem**

Disconnection from meaningful workforce participation is the modal experience for non-completers in this cohort, not a marginal one. Nearly 44% were persistently disconnected (4–5 of their first 5 years), while only 15.6% — the smallest of the four trajectory categories — maintained full attachment across all five years. A majority (56.8%) experienced disconnection in 3 or more years. Any framing of this population needs to start here: instability is the default trajectory, and sustained attachment is the outlier.

**A sorting process, not a uniform recovery**

Tracking the same population year by year shows genuine improvement over time, but not a simple "things get better" story. Combined sustained employment rises steadily from 35.1% (Year 1) to 48.1% (Year 5), and the fragmented non-sustained categories shrink correspondingly. But the population isn't converging on stability — it's splitting toward two increasingly distinct poles. Alongside rising sustained employment, "no Minnesota employment record" also climbs every single year with no deceleration, becoming the single largest annual category by Year 5 (27.1%). And of those who do reach sustained employment, a growing share of that success happens *outside* the region: sustained regional (Central) attachment plateaus sharply after Year 3, while sustained attachment elsewhere in Minnesota keeps climbing steadily through Year 5 — meaning the longer someone is out of school, the less of their eventual stability accrues specifically to this region.

**The critical refinement: two fundamentally different "shapes" of risk**

Separating non-sustained employment from no-record status — rather than treating "disconnection" as one undifferentiated category — reveals that these are not variations on the same problem. They have almost opposite risk profiles:

- **Non-sustained employment is common but usually temporary.** Nearly 70% of non-completers experience it at some point, but given that exposure, only about 1 in 5 (20.4%) ends up persistently non-sustained — most resolve to a short-term, intermittent pattern instead.
- **No-record status is rarer, but far stickier.** Only about 41% ever experience it, but given that exposure, roughly 2 in 5 (41.3%) become persistently no-record — more than double the conditional persistence rate of non-sustained employment, and nearly a coin flip between resolving quickly and becoming chronic.

This distinction matters substantively: non-sustained employment looks like a transitional, job-quality problem that most people age out of on their own, while no-record status looks more like a retention or reconnection problem, where the default outcome — once it starts — is much less likely to self-correct.

**A genuine third group: compound instability**

Neither single-cause story captures everyone. By decomposing the full population using both persistence measures together, roughly a quarter of non-completers (25.8%) experience *both* non-sustained employment and no-record status at different points across their first 5 years, rather than persistently experiencing just one. This directly explains why the combined "persistently disconnected" figure (43.8%) so far exceeds either individual persistence measure (14.2% non-sustained, 16.8% no-record) — a meaningful share of the most disconnected people are cycling between both problems rather than being stuck in one.

**Within the worst-off group, no single explanation dominates**

Restricting specifically to the 18,063 people who were persistently disconnected overall, the population splits into three comparably sized subgroups: **no-record driven (38.5%)**, **non-sustained driven (32.5%)**, and **mixed — neither individually persistent (29.1%)**. No driver reaches even 40% on its own, and the gap between the largest and smallest group is under 10 points. Even within the single most concerning trajectory category, there are genuinely three distinguishable profiles rather than one dominant story: people who kept working but never stably, people who largely vanished from Minnesota's employment data, and people who alternated between both problems without settling into either.

**What this means going into Step 3**

Taken together, these analyses establish that "disconnection" among non-college graduates is not one phenomenon with one likely cause — it's at least three distinguishable trajectories (chronic instability, chronic record-absence, and alternation between the two), layered on top of a regional-retention trend where even the "success" cases increasingly locate outside Central Minnesota over time. This has a direct implication for the bivariate analysis ahead: characteristics associated with *overall* disconnection may mask very different relationships with each of these three underlying drivers. A characteristic that predicts non-sustained-driven disconnection (e.g., a job-stability or job-quality factor) may look very different from one that predicts no-record-driven disconnection (e.g., an out-migration or labor-force-exit factor), and both may differ again from what predicts the mixed/compound pattern. Step 3's cross-tabulations and ANOVA should ideally examine these three driver categories separately wherever sample size allows, rather than relying solely on the single combined disconnection measure, to avoid averaging over what appear to be structurally different phenomena.

<br>

# Relationship Between Individual Characteristics and Disconnection Categories

The next sections focus on exploring what relationships exist between individual characteristics and the different ways in which disconnection from the workforce is measured;

- **Persistently Disconnected:** Disconnected in 4 or 5 of the first 5 years
- **Disconnected (Broad):** Disconnected in 3 of 5 years
- **Intermittently Disconnected:** Disconnected in 1–2 of the first 5 years
- **Consistently Attached:** Disconnected in 0 of the first 5 years

```{r}
updated.master <- master %>%
  left_join(master.sep[,c(1,8:13)], by = "PersonID")

write_csv(updated.master, "Data/Claude/Non-college deeper/updated-master.csv")
```

<br>

## Demographic variables

**Overall pattern across all six variables**

All six relationships are statistically significant (p \< 0.001 in every case), but this reflects the very large sample size (N ≈ 41,279) rather than strong associations — by Cohen's conventional benchmarks for Cramér's V, every one of these six relationships falls at or below the "small effect" threshold. This is worth stating plainly before the variable-by-variable detail: **at the bivariate stage, no single demographic characteristic shows more than a weak standardized association with overall disconnection category.** That said, "weak" as a standardized effect size doesn't mean the raw group differences are negligible in substantive terms — several of these variables show large percentage-point gaps in persistent disconnection specifically, even though their Cramér's V is modest. This distinction matters and is addressed directly below.

**Economic Status** (χ² = 228.76, df = 3, Cramér's V = 0.074)

Economically disadvantaged students show lower consistent attachment (13.1% vs. 17.6%) and higher persistent disconnection (47.1% vs. 41.2%) than their non-disadvantaged peers — a 5.9 percentage-point gap in persistent disconnection. This is the expected direction, consistent with economic disadvantage findings throughout the earlier DV1–DV5 work, though the smallest raw gap of the six variables examined here.

**English Learner Status** (χ² = 277.62, df = 3, Cramér's V = 0.082)

English learners show a substantially higher persistent disconnection rate (56.4% vs. 42.5%, a 13.9-point gap) and lower consistent attachment (11.3% vs. 16.1%) than non-English learners. This is a considerably larger raw gap than economic status, despite a similar Cramér's V — a first hint that standardized effect size and raw group disparity don't always move together in this data.

**Gender** (χ² = 371.53, df = 3, Cramér's V = 0.095)

Female graduates show notably lower consistent attachment (12.0% vs. 18.2%) and higher persistent disconnection (47.5% vs. 41.1%) than male graduates — a 6.4-point gap. This is the **highest Cramér's V of the six variables among those with 2-level predictors**, and it's worth flagging as a genuinely different pattern from the earlier DV1–DV5 analysis: there, female graduates were *more* likely to complete a postsecondary credential and, among completers, showed a real employment-attachment gap that partly attenuated once accomplishments were controlled. Here, restricted specifically to **non-completers**, women again show worse workforce disconnection outcomes than men — suggesting the gender disadvantage in workforce attachment isn't confined to the completer population; if anything, it shows one of the stronger bivariate signals in this specific non-completer analysis.

**Non-English Home** (χ² = 147.24, df = 3, Cramér's V = 0.060)

Students from non-English-speaking homes show the **second-largest raw percentage-point gap of any variable examined** — 57.6% persistently disconnected vs. 43.1% for English-speaking homes, a 14.5-point difference — yet this variable has the **lowest Cramér's V outside of RaceEthnicity** (0.060). This apparent contradiction is explained by sample composition: only 1,762 of the 41,279 non-completers (4.3%) come from non-English-speaking homes, so even a substantively large gap for that small subgroup contributes relatively little to the overall standardized association. Worth noting this variable likely overlaps considerably with English learner status, and the two probably aren't independent signals.

**RaceEthnicity** (χ² = 368.49, df = 12, Cramér's V = 0.055)

This shows the **largest raw disparity of any variable in this set, and the most consistent pattern across categories**. Every non-White group shows dramatically higher persistent disconnection and lower consistent attachment than White graduates:

- American Indian: 60.6% persistently disconnected, only 7.0% consistently attached — the most severe disparity of any group.
- Black: 57.8% persistently disconnected, 7.7% consistently attached.
- Asian/Pacific Islander: 53.4% persistently disconnected, 13.3% consistently attached.
- Hispanic: 52.5% persistently disconnected, 12.7% consistently attached.
- White: 42.3% persistently disconnected, 16.3% consistently attached.

The gap between American Indian and White graduates (18.3 points) is the single largest disparity found in any table here — yet RaceEthnicity has the **lowest Cramér's V of all six variables** (0.055). As with non-English home, this reflects sample composition: White graduates make up 89.8% of this non-completer population (37,071 of 41,279), so the overall association measure is diluted by the size imbalance even though every individual non-White subgroup shows a large, consistent disparity. **This is worth flagging as the single most important finding in this set of tables** — not because of its Cramér's V, but because it directly parallels the persistent, unexplained Black-White gap that ran through the DV1, DV2, and DV5 multivariate analyses of *completers*. Here, in a population that didn't complete any postsecondary credential at all, a comparable (and for American Indian graduates, even larger) racial disparity in workforce disconnection is visible at the purely bivariate level, suggesting this disparity is not specific to the postsecondary pathway — it shows up independent of completion status entirely.

**SpecialEdStatus** (χ² = 447.21, df = 3, Cramér's V = 0.104)

Students with SpecialEd status show lower consistent attachment (11.1% vs. 16.9%) and higher persistent disconnection (52.9% vs. 41.1%, an 11.8-point gap). This is the **only variable in the set whose Cramér's V crosses the conventional "small effect" threshold** (0.10 for a 2-level predictor), making it the strongest single bivariate predictor of disconnection category by that standardized measure, even though its raw percentage-point gap is smaller than English learner, non-English home, or RaceEthnicity.

**Synthesis: two different ways to rank these six variables, and both are legitimate**

| Ranked by Cramér's V (standardized strength) | Ranked by raw persistent-disconnection gap (substantive size) |
|------------------------------------|------------------------------------|
| 1\. SpecialEdStatus (0.104) | 1\. RaceEthnicity — AI vs. White (18.3 pts) |
| 2\. Gender (0.095) | 2\. Non-English home (14.5 pts) |
| 3\. English learner (0.082) | 3\. English learner (13.9 pts) |
| 4\. Economic status (0.074) | 4\. SpecialEdStatus (11.8 pts) |
| 5\. Non-English home (0.060) | 5\. Gender (6.4 pts) |
| 6\. RaceEthnicity (0.055) | 6\. Economic status (5.9 pts) |

These rankings diverge sharply for RaceEthnicity and non-English home specifically, both of which involve a numerically small subgroup with a very large disparity — Cramér's V, as an overall association measure, is diluted by how few people fall into the affected subgroup, even though the individuals in that subgroup face a substantially elevated risk. **Neither ranking is "more correct"** — Cramér's V answers "how much does this variable structure the outcome overall," while the raw gap answers "how much worse off is this specific group." For a report focused on identifying which populations face the greatest workforce disconnection risk, the raw-gap ranking (led by RaceEthnicity, especially American Indian graduates) is likely the more policy-relevant framing, even though it wouldn't be the lead finding if effect size alone were used to prioritize which variables to discuss.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/english-learner.html")
```

<br>

#### `non.english.home`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Demographics/non-english-home.html")
```

<br>

## High school characteristics

**Dem_Desc** (χ² = 55.01, df = 6, Cramér's V = 0.026)

This is the **weakest bivariate relationship found anywhere in this analysis so far** — smaller than every demographic variable interpreted previously, including RaceEthnicity's already-modest 0.055. The three district classifications show only modest differences: Town/rural mix shows the highest persistent disconnection (46.6%) and lowest consistent attachment (14.7%), while Entirely urban (41.8% persistent, 15.5% attached) and Urban/town/rural mix (43.2% persistent, 16.0% attached) sit closer to the overall population averages (43.8%, 15.6%). Given the near-total overlap with `edr` noted above, this table is largely re-describing the same regional pattern that the `edr` table below captures more directly — worth deciding whether both variables need separate treatment in the write-up, or whether `edr` alone (the variable used consistently throughout the DV1–DV5 multivariate work) is the more useful one to lead with.

**edr** (χ² = 95.92, df = 6, Cramér's V = 0.034)

EDR 7W shows meaningfully better outcomes than the other two regions: 16.9% consistently attached and only 41.9% persistently disconnected, compared to EDR 5 (14.7% attached, 46.6% persistent) and EDR 7E (14.0% attached, 44.8% persistent). This directly **echoes the pattern found throughout the entire DV1–DV5 multivariate project**, where EDR 7W consistently showed better regional employment and completion outcomes relative to EDR 5, while EDR 7E consistently showed worse outcomes. Seeing the same regional ordering reappear here, in an entirely different population (non-completers) and a completely different outcome measure (disconnection trajectory), is a useful cross-validation of that earlier finding — it suggests EDR 7W's relative advantage isn't specific to the completer population or to employment-attachment outcomes specifically, but reflects a more general regional pattern.

**avg.unemp.rate** (ANOVA: F(3, 41275) = 19.28, p \< 0.001, η² = 0.0014)

Despite high statistical significance, this variable explains **only about 0.14% of the variance** in local unemployment rate across disconnection categories — a negligible effect size in practical terms, even more so than the weakest categorical variables above. The Tukey HSD comparisons clarify the *shape* of this weak relationship, and it's worth noting explicitly: this is not a graded, dose-response pattern. Consistently attached individuals live in areas with significantly *lower* average unemployment than all three disconnected categories (p \< 0.001 for each comparison), but the three disconnected categories themselves are **not significantly different from each other** — Broadly disconnected vs. Intermittently disconnected (p = 0.498) and Persistently disconnected vs. Broadly disconnected (p = 0.902) both fail to reach significance, and Persistently disconnected vs. Intermittently disconnected is only barely significant (p = 0.024). In short: local unemployment rate distinguishes "any disconnection" from "full attachment," but doesn't meaningfully differentiate *degree* of disconnection once someone experiences any of it.

**avg.wages.pct.state** (ANOVA: F(3, 41275) = 15.80, p \< 0.001, η² = 0.00115)

An even smaller effect size than unemployment rate — about 0.12% of variance explained. Consistently attached individuals live in relatively higher-local-wage areas than the three disconnected categories (all three Tukey comparisons vs. Consistently attached are negative, meaning disconnected groups have lower local wages), which is the expected direction — better local labor markets associate with less disconnection. The pattern here is slightly more graded than unemployment rate: Persistently disconnected is significantly different from *both* Consistently attached (p \< 0.001) and Intermittently disconnected (p \< 0.001), though not from Broadly disconnected (p = 0.244). As with unemployment rate, Intermittently disconnected is not clearly distinguishable from Consistently attached (p = 0.076, borderline) — again suggesting the strongest distinction this variable draws is between the most severe disconnection category and everything else, rather than a clean stepwise gradient across all four categories.

**Synthesis for this group of four variables**

All four are the weakest relationships found in the bivariate analysis so far — every effect size here (Cramér's V of 0.026–0.034, η² of \~0.001) falls well below even the modest associations seen for the demographic variables in the previous set. The two continuous labor-market variables in particular explain a vanishingly small share of variance in disconnection category, and their significant Tukey comparisons consistently isolate the *extremes* (Consistently attached vs. everyone else, or Persistently disconnected vs. everyone else) rather than distinguishing the middle categories from one another. The one substantively interesting finding in this set is `edr`'s replication of the EDR 7W advantage / EDR 7E disadvantage pattern from the earlier multivariate work — a genuine cross-validation, even though the overall association remains weak. Given `Dem_Desc`'s near-total redundancy with `edr` in this data, it likely doesn't need to be carried forward as a separate variable in later stages of this analysis.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `edr`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/edr.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**The headline finding in this batch: two opposite stories depending on which "college-track" signal is examined, plus one clean, consistent story from CTE.**

**took.ACT** (χ² = 225.37, df = 3, Cramér's V = 0.074) — the strongest categorical relationship in this batch, tied with economic status from the demographic set. ACT-takers show **better** outcomes than non-takers: 17.7% consistently attached vs. 14.2%, and only 39.6% persistently disconnected vs. 46.7%. Simply engaging with the ACT process — regardless of score — is associated with meaningfully better workforce trajectories even among people who never completed a postsecondary credential.

**ACTCompositeScore** (ANOVA restricted to the 17,155 who took the ACT; η² = 0.0088, F = 50.70, p \< 0.001) — and this is where the story reverses. Among ACT-takers specifically, **higher scores are associated with *more* disconnection, not less.** The Tukey comparisons show a genuinely graded pattern: Persistently disconnected individuals scored on average 0.99 points higher than Consistently attached individuals (p \< 0.001), and nearly every pairwise comparison across all four categories is significant — this is a real stepwise gradient, not just an extremes-only effect. **This is the same threshold structure that ran through the DV1, DV4, and DV5 multivariate models** (`took.ACT` positive, `ACTCompositeScore` negative once combined) showing up again here in an entirely different population and analytic method. Read together with `took.ACT`, the story is: engaging with the ACT process helps, but among those who do, scoring higher — while never completing a postsecondary credential — associates with worse workforce outcomes, consistent with a "thwarted college-track" or overqualified-but-underemployed pattern specific to non-completers.

**ap.exam** (χ² = 19.18, df = 3, Cramér's V = 0.022 — the weakest categorical relationship in this batch) and **sat.taken** (χ² = 40.55, df = 3, Cramér's V = 0.031) reinforce the same reversed direction, and SAT shows it in a dramatic and important way. AP exam takers show slightly *worse* outcomes than non-takers (46.2% vs. 43.5% persistently disconnected). SAT-takers — a very small group (n = 221, just 0.5% of this population) — show the **single largest raw disparity found anywhere in this entire bivariate analysis**: 64.3% persistently disconnected vs. 43.6% for non-SAT-takers (a 20.7-point gap), and only 7.2% consistently attached vs. 15.7% (less than half the rate). This exceeds even the largest demographic gap found earlier (American Indian vs. White, 18.3 points). As with RaceEthnicity and non-English home in the earlier batch, the small subgroup size keeps the overall Cramér's V modest despite the size of the raw gap — but the substantive finding is striking: non-completers who took the SAT (a clear college-bound signal) show by far the worst workforce trajectories of any group examined in this project's bivariate analysis, worse even than students who showed no college-track ambition at all. This is worth treating as a genuinely important, if statistically fragile (small-n), finding — students who pursued a clear four-year-college pathway but never completed appear to struggle the most with workforce reconnection.

**cte.achievement** (χ² = 389.89, df = 6, Cramér's V = 0.069) shows a clean, fully graded relationship in the **opposite** direction from the college-track variables: No CTE shows the worst outcomes (50.3% persistently disconnected, only 12.7% consistently attached), CTE Participant sits in the middle (42.2% persistent, 15.9% attached), and CTE Concentrator/Completer shows the best outcomes of the three (39.7% persistent, 17.8% attached). This directly **replicates the CTE-as-regional-workforce-pathway finding from the DV1/DV4 multivariate work**, now shown to hold even among people who never completed a postsecondary credential at all — suggesting CTE's benefit operates through direct workforce readiness rather than exclusively through the credentialing pathway.

**total.cte.courses.taken** (η² = 0.0111, F = 153.66, p \< 0.001) is the **strongest continuous relationship found anywhere in this entire step 3 analysis** — every single pairwise Tukey comparison is significant, describing a fully monotonic dose-response gradient (Consistently attached \> Intermittently disconnected \> Broadly disconnected \> Persistently disconnected, differing from Consistent to Persistent by a full course on average). Combined with the categorical CTE finding above, this is the single most reliable protective signal identified in this non-completer population — and notably, it's a genuine dose-response relationship, unlike most of the other variables examined so far, which mainly distinguish extremes (e.g., "any disconnection" vs. none) rather than differentiating smoothly across all four severity levels.

**pseo.participant** (χ² = 3.87, df = 3, Cramér's V = 0.012, **p = 0.276 — not statistically significant**). This is a striking null result given PSEO's central, consistently positive role throughout the entire DV1–DV5 multivariate project. Among non-completers specifically, PSEO participation shows no meaningful relationship with disconnection trajectory at all. One caveat worth flagging before treating this as settled: **the N here is only 28,186, well below the full population of 41,279** — meaning roughly 13,000 non-completers are missing or excluded from this particular table. Worth confirming why before concluding PSEO genuinely has no relationship in this population, since a large missingness gap could be masking something (e.g., if missingness is systematically related to disconnection status).

**MCA-M / Math** (η² = 0.0029, F = 36.86, p \< 0.001) shows the expected direction — higher Math scores associate with *less* disconnection, in a mostly graded pattern (Consistent \> Intermittent \> Broad \> Persistent), with only Broad-vs-Intermittent falling short of significance (p = 0.074). This is worth contrasting directly with the ACT score finding above: MCA Math, taken by virtually the entire student population regardless of college ambition, shows the intuitive "higher score, better outcome" direction, while ACT score — taken only by a self-selected, college-bound subgroup — shows the opposite direction once restricted to people who never completed. This contrast reinforces that the reversed ACT/AP/SAT pattern isn't a general "achievement is bad" finding — it's specific to the mismatch between showing college-bound signals and not completing college.

**MCA-R / Reading** (η² = 0.0009, F = 11.44, p \< 0.001) shows the same direction as Math but far weaker, and — like the labor-market variables from the previous batch — only distinguishes Persistently disconnected from the other three categories (all three comparisons against Persistent are significant; none of the comparisons among Consistent/Intermittent/Broad are). A threshold effect rather than a graded one.

**MCA-S / Science** (η² = 0.0003, F = 2.70, p = 0.044) is the **weakest relationship found in this entire step 3 analysis, barely crossing conventional significance**, and only one Tukey comparison (Persistent vs. Consistent, p = 0.032) reaches significance at all. This variable shows essentially no meaningful bivariate relationship with disconnection category and is unlikely to be worth emphasizing in the write-up.

**Cross-variable synthesis for this batch**

Ranking these five categorical variables by Cramér's V: took.ACT (0.074) \> cte.achievement (0.069) \> sat.taken (0.031) \> ap.exam (0.022) \> pseo.participant (0.012, ns). The real story here, though, isn't the ranking — it's the **directional split**: `took.ACT`, `cte.achievement`, and `total.cte.courses.taken` all point toward better outcomes with more engagement, matching intuition and replicating patterns already established in the completer-focused DV1–DV5 work. `ACTCompositeScore`, `ap.exam`, and `sat.taken` all point the opposite way — worse outcomes associated with stronger college-bound signals, specifically among people who never completed a credential. This split is likely the most important finding to carry forward from this entire batch: it suggests two genuinely different profiles within the non-completer population — one where vocational/CTE engagement supports workforce reconnection as intended, and another where unfulfilled college ambition (especially SAT-taking) marks the group facing the steepest workforce disconnection risk of anyone examined in this project.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `sat.taken`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/sat-taken.html")
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ disconnected.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Disconnected cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

**attended.ps** (χ² = 837.54, df = 3, Cramér's V = 0.142)

**This is the strongest bivariate relationship found anywhere in the entire Step 3 analysis** — larger than SpecialEdStatus (0.104), the previous largest, and by a clear margin larger than every other demographic, regional, and academic-accomplishment variable examined so far. It's also the only relationship in this analysis to move meaningfully past the "small effect" threshold on Cohen's benchmarks (small = 0.10 for a 2-level predictor), landing solidly between small and medium.

Among non-completers, those who **attended some postsecondary education without ever finishing** show substantially better workforce outcomes than those who never attended at all: 20.5% consistently attached vs. 13.0% (a 7.5-point gap), and only 35.1% persistently disconnected vs. 48.5% — a **13.4-point gap**, one of the largest raw disparities found in this entire analysis. Even incomplete postsecondary exposure appears to carry real workforce value in this population.

**The shape of the gap is worth noting specifically, not just its size.** The "broadly disconnected" category is nearly identical between the two groups (13.1% vs. 13.0% — essentially no difference), while the gain among those who attended PS is concentrated at both the top and the middle of the distribution: consistent attachment rises (13.0% → 20.5%) and intermittent disconnection also rises somewhat (25.6% → 31.4%), while persistent disconnection drops sharply (48.5% → 35.1%). In other words, attending postsecondary education without completing doesn't uniformly shift people toward better outcomes across the board — it specifically appears to pull people *out of the worst category* (persistent disconnection) and into either full attachment or a more moderate, intermittent pattern, while leaving the "broadly disconnected" group essentially untouched.

**This connects directly to two earlier findings in this analysis and sharpens both.** First, it stands in stark contrast to `pseo.participant`, which showed no significant relationship with disconnection category at all (p = 0.276). Read together, these two results suggest it's specifically **postsecondary enrollment itself** — not PSEO's high-school dual-enrollment mechanism — that carries the workforce benefit here; PSEO participation alone, without any subsequent postsecondary attendance, doesn't appear to move the needle for this population. Second, it extends the DV3 finding from the completer-focused multivariate work (where PSEO was the single strongest predictor of full completion) by showing that even the "attempted but didn't finish" pathway — the population DV3 didn't examine, since it was restricted to completion as a binary outcome — still carries meaningful, measurable workforce value relative to never having attended at all.

**Substantive takeaway:** of every variable examined in this bivariate stage — demographic, regional, and academic — whether someone ever set foot in a postsecondary institution, independent of whether they finished, is the single strongest predictor of workforce disconnection trajectory found so far. This is likely the headline finding to lead with when this batch of results is written up.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$disconnected.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Disconnected cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of Overall Disconnection Persistence (`disconnected.cat`)

**The headline: one variable stands well above everything else**

Of every characteristic examined — demographic, regional, academic, and postsecondary — **whether someone ever attended a postsecondary institution, independent of whether they completed anything, is by far the strongest predictor of workforce disconnection trajectory** (Cramér's V = 0.142, more than 35% larger than the next-strongest relationship found anywhere in this analysis). Non-completers who attended some postsecondary program show a 13.4-point lower rate of persistent disconnection than those who never attended at all, and the gap is concentrated specifically in pulling people out of the worst outcome category rather than lifting everyone uniformly. If this bivariate stage has one lead finding, this is it.

**Most relationships are statistically significant but individually weak — and that's expected, not a problem**

Across the demographic variables (Cramér's V 0.055–0.104), the regional and labor-market variables (0.026–0.034, and eta-squared values near 0.001 for the two continuous measures), and most of the academic-accomplishment variables (0.012–0.074), effect sizes sit at or below the conventional "small effect" threshold. Given a sample size north of 41,000, this is exactly what you'd expect from a screening-stage bivariate analysis — statistical significance here confirms these relationships aren't due to chance, not that any single variable strongly determines the outcome on its own. `attended.ps` and, to a lesser degree, `SpecialEdStatus` (0.104) are the only two variables that clear even a modest bar.

**But raw percentage-point gaps tell a different, arguably more policy-relevant story**

Several variables with unremarkable Cramér's V values are attached to some of the largest raw disparities in the whole analysis, because they involve numerically small subgroups experiencing very different outcomes. SAT-taking (a 20.7-point gap in persistent disconnection, the largest of any variable examined, based on just 221 people), American Indian graduates (an 18.3-point gap vs. White graduates), and non-English-home students (a 14.5-point gap) all show this pattern: a standardized effect size that looks modest, sitting on top of a subgroup disparity that is anything but. This is worth stating as a general finding in itself — Cramér's V and raw group disparity answer different questions, and for a project aimed at identifying who is most at risk, the raw-gap ranking is usually the more useful one to lead with.

**The racial disparity seen throughout DV1, DV2, and DV5 shows up again here — in a population that never even reached those DVs**

Every non-White racial/ethnic group shows meaningfully worse disconnection outcomes than White graduates, with American Indian graduates facing the most severe gap (60.6% persistently disconnected vs. 42.3% for White graduates, and just 7.0% consistently attached vs. 16.3%). This is a direct echo of the persistent, unexplained Black-White gap that ran through the earlier multivariate completer analysis — except here it appears in an entirely different population (people who never completed any postsecondary credential) using a completely different outcome measure (year-by-year workforce disconnection rather than employment attachment or credential level). Together, these findings suggest the racial disparity in this region's workforce outcomes isn't specific to the postsecondary pathway — it shows up independent of whether someone ever pursued or completed a credential at all.

**Regional geography replicates cleanly, local labor market conditions barely register**

EDR 7W shows meaningfully better outcomes than EDR 5 or EDR 7E, directly mirroring the regional pattern found throughout the entire DV1–DV5 project — a useful cross-validation from an independent population and outcome measure. By contrast, local wage levels and unemployment rates — despite being statistically significant given the sample size — explain a genuinely negligible share of variance in disconnection trajectory (both under 0.15% of variance), and mainly distinguish "any disconnection" from full attachment rather than differentiating degree of disconnection. (Also worth remembering: `Dem_Desc`'s "Town/rural mix" and `edr`'s "EDR 5" are the exact same people in this data, replicating the complete nesting relationship discovered during the earlier multivariate modeling — these two variables aren't independent here either.)

**The most intellectually interesting finding: two college-track signals point in opposite directions**

CTE engagement and general academic proficiency behave exactly as intuition would suggest: more CTE course-taking and higher CTE achievement level both associate cleanly and consistently with less disconnection (CTE course count is in fact the single strongest continuous relationship in the whole analysis), and higher MCA Math scores likewise associate with better outcomes. `took.ACT` also points this direction — simply having taken the test associates with better outcomes than not taking it at all.

But conditional on taking the ACT, a **higher composite score predicts *more* disconnection, not less** — and AP exam participation and, most dramatically, SAT-taking (the 20.7-point gap noted above) show the same reversed direction. This is the same `took.ACT`-positive / `ACTCompositeScore`-negative threshold pattern that ran through the DV1, DV4, and DV5 multivariate models, reappearing here in an entirely different population and analytic method — and it sharpens into something more specific: among people who never completed a postsecondary credential, the strongest college-bound signals (a high ACT score, AP participation, and especially SAT-taking) mark the group facing the *steepest* workforce disconnection risk of anyone examined in this project, worse even than students who showed no college ambition at all. This reads as a "thwarted ambition" or mismatch story — people who were clearly positioned for a four-year college track, but for whom that track never resulted in a credential, appear to struggle the most with workforce reconnection.

**PSEO's absence here is as informative as its presence in the earlier work**

PSEO participation — the single strongest predictor of postsecondary completion found anywhere in the DV1–DV5 project — shows **no significant relationship at all** with disconnection trajectory in this non-completer population (p = 0.276). Read alongside `attended.ps` being the strongest predictor in this entire analysis, this suggests PSEO's benefit operates specifically by helping people complete a credential — once completion doesn't happen, the specific mechanism PSEO provides in high school doesn't carry an independent workforce benefit on its own. It's postsecondary attendance itself, not the dual-enrollment pathway that leads to it, that seems to matter for this population.

**The story in one paragraph**

Among non-college graduates, the single most powerful lever identified is simply having set foot in a postsecondary institution — even without finishing. Beyond that, this population shows the same racial disparity and the same regional pattern found throughout the broader project's completer-focused analysis, confirming those aren't artifacts of studying people who went on to earn a credential. And a genuinely new finding emerges specific to this population: general academic proficiency and vocational (CTE) engagement protect against disconnection as expected, but the clearest signals of college ambition — a strong ACT score, AP participation, and especially SAT-taking — mark the group at *greatest* risk of persistent disconnection, once that ambition didn't result in a completed credential. Any intervention aimed at this population should recognize these as two different groups needing different approaches: one where CTE-style, workforce-connected reengagement is the more natural next step, and another — likely smaller, but facing worse outcomes — of people who pursued a four-year-college path that didn't pan out, and who may need a fundamentally different kind of support to reconnect with the regional workforce.

<br>

# Relationship Between Individual Characteristics and Non-sustaining Categories

The next sections focus on exploring what relationships exist between individual characteristics and the frequency of having non-sustained workforce participation in the first 5 years after formal education.

- **Never non-sustained** — 0 of the first 5 years (post-high-school or post-postsecondary-exit) were classified as "Non-sustained WF"
- **Intermittently non-sustained** — 1-2 of the first 5 years were "Non-sustained WF"
- **Broadly non-sustained** — exactly 3 of the first 5 years were "Non-sustained WF"
- **Persistently non-sustained** — 4-5 of the first 5 years were "Non-sustained WF"

<br>

## Demographic variables

**Economic Status** (χ² = 229.70, df = 3, Cramér's V = 0.075 — essentially identical to disconnected.cat's 0.074)

Economically disadvantaged students show lower "never non-sustained" (27.4% vs. 33.0%) and higher persistent non-sustained employment (16.2% vs. 12.6%). This relationship carries through to the isolated non-sustained measure at almost exactly the same strength as it did for the combined disconnection measure — economic disadvantage appears to genuinely drive unstable-but-employed patterns, not just no-record status.

**Gender** (χ² = 404.23, df = 3, Cramér's V = 0.099 — slightly *stronger* than disconnected.cat's 0.095)

Female graduates show lower "never non-sustained" (26.4% vs. 33.5%) and higher persistent non-sustained employment (16.9% vs. 12.3%). This is the **strongest relationship in this entire batch**, and notably it's actually a bit stronger here than in the combined measure — meaning gender's disadvantage in overall disconnection is substantially, perhaps even primarily, a story about cycling through unstable jobs rather than disappearing from the record.

**SpecialEdStatus** (χ² = 204.42, df = 3, Cramér's V = 0.070 — notably weaker than disconnected.cat's 0.104)

SpecialEd students show higher persistent non-sustained employment (18.6% vs. 13.0%, a 5.6-point gap), still a real and meaningful relationship, but appreciably attenuated compared to the combined measure. This suggests SpecialEdStatus's disadvantage in overall disconnection is only partly about non-sustained employment — a real share likely runs through the no-record dimension as well, worth confirming once that breakdown is available for this variable.

**Three variables show a genuinely different — even partially reversed — story here than they did for combined disconnection, and this is the most important finding in this batch:**

**English Learner** (χ² = 79.09, df = 3, Cramér's V = 0.044 — much weaker than disconnected.cat's 0.082). English learners actually show a *higher* "never non-sustained" rate than non-English-learners (34.8% vs. 30.1%) — the opposite direction from what the combined measure showed — while still showing somewhat higher persistent non-sustained employment (16.7% vs. 13.9%). Given that English learners showed one of the largest gaps in the *combined* disconnection table (56.4% vs. 42.5% persistently disconnected), this much weaker and partially reversed relationship here strongly suggests **their disadvantage in overall disconnection is concentrated in the no-record dimension, not non-sustained employment** — when they are employed, it's not notably less stable than their peers; the larger risk appears to be dropping out of Minnesota's employment record entirely.

**Non-English Home** (χ² = 129.10, df = 3, Cramér's V = 0.056, but **direction fully reversed** from the combined table). Non-English-home individuals show a *considerably higher* "never non-sustained" rate (42.7% vs. 30.0%) and *lower* persistent non-sustained employment (11.1% vs. 14.3%) than English-home individuals — meaning on this specific dimension, they actually look *better off*, not worse. This is a striking contrast to the combined disconnection table, where this same group showed one of the largest disadvantage gaps in the entire analysis (57.6% vs. 43.1% persistently disconnected). Taken together, this makes clear that non-English-home's large overall disconnection disadvantage has essentially nothing to do with job instability — it must be driven almost entirely by no-record status, plausibly reflecting migration patterns (e.g., leaving Minnesota, or never fully entering its formal employment system) rather than difficulty holding steady employment once engaged with it.

**RaceEthnicity** (χ² = 220.12, df = 12, Cramér's V = 0.042, and the group **splits into two opposing patterns** rather than moving together as it did in the combined table):

- **American Indian and Black graduates show *worse* non-sustained outcomes than White graduates** — Black shows the highest persistent non-sustained rate of any group in this table (22.2% vs. White's 13.9%), and American Indian is close behind (18.8%).
- **Asian/Pacific Islander and Hispanic graduates show *better* non-sustained outcomes than White graduates** — both show notably higher "never non-sustained" rates (40.6% and 39.1%, respectively, vs. White's 30.6%) and lower persistent rates (10.7% and 12.6%, vs. White's 13.9%).

This is a genuine reversal of the pattern from the combined disconnection table, where **every** non-White group showed uniformly worse outcomes than White graduates. Splitting the outcome into non-sustained-specific terms reveals that the earlier "uniform racial disadvantage" finding was masking two different underlying mechanisms: for American Indian and Black graduates, unstable employment itself is a real, independent contributor to their overall disconnection disadvantage; for Asian/Pacific Islander and Hispanic graduates, their large disadvantage in the combined measure must be driven almost entirely by the no-record dimension, since on the non-sustained dimension specifically they actually outperform White graduates.

**Synthesis: this batch fundamentally reframes several of the earlier findings**

Comparing Cramér's V between the combined and non-sustained-specific versions side by side:

| Variable | disconnected.cat V | non_sustained.cat V | Change |
|------------------|------------------|------------------|------------------|
| Gender | 0.095 | 0.099 | Slightly stronger |
| Economic status | 0.074 | 0.075 | Essentially unchanged |
| SpecialEdStatus | 0.104 | 0.070 | Notably weaker |
| RaceEthnicity | 0.055 | 0.042 | Weaker, **and direction splits** |
| Non-English home | 0.060 | 0.056 | Similar magnitude, **direction reverses** |
| English learner | 0.082 | 0.044 | Much weaker, partial reversal |

**Gender and economic status are the two variables whose disadvantage in overall disconnection appears to be genuinely about job instability** — their relationship with non-sustained employment specifically is as strong as (or stronger than) their relationship with the combined measure. **English learner, non-English home, and RaceEthnicity, by contrast, all weaken substantially — and two of them partially or fully reverse direction** — once no-record status is set aside. This strongly suggests that for these three variables, the large disparities seen in the earlier combined-disconnection analysis are driven predominantly by disappearing from Minnesota's employment record (out-migration, informal work, or true labor-market exit) rather than by cycling through unstable jobs. This is exactly the kind of variable-specific driver analysis the project's earlier summary called for, and it meaningfully changes the policy framing for these three variables: an intervention aimed at "job stability" would likely help with gender and economic-status-related disconnection, but would probably miss the actual mechanism driving disconnection for English learners, non-English-home students, and (for two of its four subgroups) racial/ethnic disparities.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/english-learner.html")
```

<br>

#### `non.english.home`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Demographics/non-english-home.html")
```

<br>

## High school characteristics

**The overall pattern in this batch: everything gets weaker, and two variables become essentially negligible.** Every relationship in this set — already the weakest category of variables against the combined disconnection measure — attenuates further once non-sustained employment is isolated from no-record status. This is a useful, if less exciting, confirmatory finding: regional and local labor-market context appear to matter more for the combined disconnection experience (which includes disappearing from the record) than for cycling through unstable-but-present employment specifically.

**Dem_Desc** (χ² = 13.31, df = 6, Cramér's V = 0.013, p = 0.038) — **the weakest relationship found anywhere in this entire Step 3 bivariate analysis**, barely clearing conventional significance. The three district classifications are nearly indistinguishable from each other and from the population average (Never non-sustained ranges only 29.4%–31.5% across all three groups). This variable shows essentially no meaningful relationship with non-sustained employment specifically, even weaker than its already-modest 0.026 relationship with the combined disconnection measure.

**edr** (χ² = 48.64, df = 6, Cramér's V = 0.024 — notably weaker than disconnected.cat's 0.034). The pattern here is also a bit different in shape, not just weaker in magnitude: EDR 7E stands out as the one region with worse outcomes (28.0% never non-sustained, 15.3% persistently non-sustained), while EDR 5 and EDR 7W look quite similar to each other (31.5% and 31.3% never non-sustained, respectively). This is worth flagging as a nuance to the earlier finding — in the combined disconnection table, EDR 7W showed a clear advantage over *both* EDR 5 and EDR 7E; here, isolating non-sustained employment specifically, EDR 7W's advantage largely disappears relative to EDR 5, suggesting EDR 7W's edge in the combined measure is more about lower no-record rates than about more stable employment once someone is working. (The same Dem_Desc/edr complete-nesting relationship flagged earlier — "Town/rural mix" and "EDR 5" being the identical group — holds again here, confirmed by matching N and percentages.)

**avg.unemp.rate** (ANOVA: F(3, 41275) = 6.10, p \< 0.001, η² = 0.00044 — about a third the size of its disconnected.cat relationship). The Tukey comparisons reveal something worth stating explicitly: this is **not a monotonic pattern at all**. Intermittently and Broadly non-sustained individuals live in areas with modestly but significantly higher unemployment than the "Never" group (p = 0.004 and p = 0.002), but **Persistently non-sustained is *not* significantly different from Never non-sustained** (p = 0.777) — the opposite of what the combined disconnection table showed, where the starkest, most reliable distinction was between the most severe category and everyone else. Local unemployment rate, in other words, relates to *some* non-sustained employment experience but doesn't track its most severe, persistent form.

**avg.wages.pct.state** (ANOVA: F(3, 41275) = 3.02, p = 0.029, η² = 0.00022 — **the weakest continuous relationship found anywhere in this entire analysis**, weaker even than MCA Science from the earlier academic batch). Only one of the six Tukey comparisons reaches significance (Broadly non-sustained vs. Never, p = 0.022); every other comparison, including Persistent vs. Never (p = 0.359), is non-significant. This variable shows essentially no meaningful, reliable relationship with non-sustained employment persistence.

**Synthesis for this batch**

Every regional and labor-market variable weakens once non-sustained employment is isolated from no-record status, and two (`Dem_Desc` and `avg.wages.pct.state`) become close to negligible. This reinforces the broader pattern established in the demographic batch: local geography and labor-market context appear to matter considerably more for whether someone disappears from Minnesota's employment record than for whether their employment, once present, is stable. If a future analysis examines the no-record-specific version of these same four variables, it would be reasonable to expect `edr` and the two continuous labor-market measures to show *stronger* relationships there than they do here — since the combined-measure associations (already modest) appear to be getting diluted, not concentrated, once non-sustained employment is considered on its own.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `edr`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/edr.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**This batch delivers the clearest confirmation yet of the driver-separation logic established earlier in this project. Several variables that were significant for combined disconnection lose significance entirely once non-sustained employment is isolated — strongly suggesting their earlier relationship was really about no-record status. Others strengthen. This is worth reading side-by-side with the earlier disconnected.cat interpretation rather than as a standalone table.**

**ACTCompositeScore** (ANOVA restricted to the 17,155 ACT-takers; F = 0.998, **p = 0.393 — not significant**, η² = 0.00017, essentially negligible). This is a striking reversal: for combined disconnection, ACT score showed a clear, significant, graded relationship (higher score → more disconnection, η² = 0.0088). Isolated to non-sustained employment specifically, that relationship **disappears entirely**. This strongly confirms that the earlier "higher ACT score predicts worse outcomes among non-completers" finding was driven almost entirely by the no-record dimension — a high-scoring non-completer isn't more likely to cycle through unstable jobs, but does appear more likely to leave Minnesota's employment record altogether (consistent with an out-migration or "left for opportunity elsewhere" interpretation).

**ap.exam** (χ² = 6.34, df = 3, Cramér's V = 0.012, **p = 0.096 — not significant**). Same pattern as ACT score: this was significant (if weak) for combined disconnection; it loses significance entirely once non-sustained employment is isolated.

**sat.taken** (χ² = 3.04, df = 3, Cramér's V = 0.009, **p = 0.386 — not significant**). This is the most dramatic reversal in the whole analysis. Recall that SAT-taking showed the **single largest raw disparity found anywhere** in the combined disconnection table (a 20.7-point gap in persistent disconnection). Here, isolated to non-sustained employment, that relationship vanishes completely (p = 0.386). This confirms decisively that SAT-takers' dramatic disadvantage in the earlier analysis has essentially nothing to do with job instability — it must be entirely a no-record phenomenon. Combined with the ACT and AP results above, there's now a consistent, three-variable pattern: **every college-bound academic signal that showed a "worse outcomes for non-completers" relationship in the combined measure loses that relationship entirely once non-sustained employment is considered on its own.** The "thwarted college ambition" story from the earlier summary should be reframed specifically as a no-record/disengagement story, not a job-instability story.

**took.ACT** (χ² = 64.21, df = 3, Cramér's V = 0.039 — about half the strength of disconnected.cat's 0.074, but still clearly significant). ACT-takers still show modestly better non-sustained outcomes (13.0% persistently non-sustained vs. 15.1% for non-takers), so unlike the three variables above, `took.ACT`'s positive relationship survives the split — attenuated, but real.

**cte.achievement** (χ² = 87.50, df = 6, Cramér's V = 0.033 — less than half of disconnected.cat's 0.069) and **total.cte.courses.taken** (η² = 0.0016 — a large drop from disconnected.cat's 0.0111, the strongest continuous relationship found in that earlier batch). Both CTE variables retain their direction (more CTE engagement, less non-sustained employment) but weaken substantially. Notably, `total.cte.courses.taken`'s Tukey comparisons show the "Intermittently non-sustained vs. Never" contrast is not significant here (p = 0.463) and even points the "wrong" direction (positive, though not significant) — the clean, fully monotonic dose-response pattern found for combined disconnection does **not** fully carry over to non-sustained employment alone. This suggests CTE's strong protective relationship in the combined measure was substantially, though not entirely, about reducing no-record status rather than reducing job instability specifically.

**pseo.participant** (χ² = 15.84, df = 3, Cramér's V = 0.024, **p = 0.001 — now significant**, reversing the null result from the combined measure, p = 0.276). PSEO participants show modestly lower persistent non-sustained employment (13.5% vs. 14.7%). This is the mirror image of the SAT/ACT/AP pattern above: a variable with **no** relationship to combined disconnection turns out to have a small but real, specifically non-sustained-employment-focused relationship once isolated. (Same N caveat as before applies — this table covers only 28,186 of the 41,279 non-completers.)

**MCA-M / Math** (η² = 0.00286 — essentially unchanged from disconnected.cat's 0.00285). This relationship is stable across both DVs, and here it's **fully monotonic** — every single Tukey pairwise comparison is significant, including Broadly vs. Intermittently non-sustained (p = 0.0048), which had fallen just short of significance in the combined table. Higher Math scores relate cleanly and consistently to less non-sustained employment at every step of severity.

**MCA-R / Reading** (η² = 0.00157 — nearly double disconnected.cat's 0.00090). A modest strengthening, though the Tukey pattern remains a threshold effect rather than a graded one (only comparisons involving Persistently non-sustained reach significance).

**MCA-S / Science** (η² = 0.00403, F = 40.17, **p = 6.8 × 10⁻²⁶** — a dramatic strengthening from disconnected.cat's barely-significant η² = 0.00027, p = 0.044). This is the most striking reversal in this entire academic-variable batch: MCA Science showed essentially no meaningful relationship with combined disconnection, but shows a strong, clearly graded relationship with non-sustained employment specifically, with four of six Tukey comparisons significant at p \< 0.001. This strongly suggests MCA-S's true relationship is specifically with employment stability, and that relationship was almost completely masked in the combined measure by no-record status adding noise.

**Synthesis: this batch is the clearest evidence yet for treating "disconnection" as (at least) two distinct phenomena**

| Variable | Cramér's V / η² (disconnected.cat) | Cramér's V / η² (non_sustained.cat) | What changed |
|------------------|------------------|------------------|------------------|
| sat.taken | 0.031, sig. | 0.009, **not sig.** | Relationship vanishes entirely |
| ap.exam | 0.022, sig. | 0.012, **not sig.** | Relationship vanishes entirely |
| ACTCompositeScore | η²=0.0088, sig. | η²=0.0002, **not sig.** | Relationship vanishes entirely |
| total.cte.courses.taken | η²=0.0111 (strongest in batch) | η²=0.0016 | Substantially weaker |
| cte.achievement | 0.069 | 0.033 | Weaker, direction holds |
| took.ACT | 0.074 | 0.039 | Weaker, direction holds |
| MCA-M | η²=0.00285 | η²=0.00286 | Unchanged, now fully monotonic |
| MCA-R | η²=0.00090 | η²=0.00157 | Stronger |
| MCA-S | η²=0.00027, barely sig. | η²=0.00403, highly sig. | **Dramatically stronger** |
| pseo.participant | 0.012, **not sig.** | 0.024, sig. | Relationship emerges |

The pattern is remarkably clean: **college-application-specific signals (SAT, AP, ACT score) that predicted worse combined outcomes for non-completers turn out to have nothing to do with job instability** — that entire "thwarted ambition" story belongs to the no-record dimension. Meanwhile, **general academic proficiency (MCA, especially Science) and PSEO participation show relationships that were being diluted or hidden by the combined measure**, and emerge more clearly once non-sustained employment is considered on its own. CTE's story is more mixed — real in both dimensions, but weaker for non-sustained employment specifically than it appeared in the combined measure. This is a strong argument for treating the no-record breakdown of these same academic variables as a high priority next step, since the pattern here suggests that breakdown will likely show the SAT/AP/ACT-score relationships reappearing there in full force.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `sat.taken`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/sat-taken.html")
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ non_sustained.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/Non-sustained cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

χ² = 270.52, df = 3, Cramér's V = 0.081

**This remains a real, meaningful relationship — the strongest of any variable examined in the entire non_sustained.cat batch except Gender — but it's substantially weaker than its relationship with combined disconnection.** Recall `attended.ps` was the single strongest bivariate relationship found anywhere in the whole Step 3 analysis against the combined measure (Cramér's V = 0.142). Here, isolated to non-sustained employment specifically, its strength drops to about 57% of that (0.081).

Non-completers who attended some postsecondary program show meaningfully lower persistent non-sustained employment (10.9% vs. 16.0% for those who never attended — a 5.1-point gap) and higher "never non-sustained" rates (32.8% vs. 29.3%). Both gaps are real and clearly significant, but both are considerably smaller than the equivalent gaps in the combined disconnection table (13.4 points and 7.5 points, respectively).

**This tells a clean, quantifiable story about where `attended.ps`'s benefit actually comes from.** Of the 13.4-point combined-disconnection advantage that postsecondary attendance provides, only about 5.1 points (38%) shows up specifically in reduced non-sustained employment — meaning the majority of `attended.ps`'s protective effect operates through the no-record dimension rather than through job stability once employed. This is directionally consistent with the pattern seen throughout this entire comparison exercise: several of the strongest-looking findings in the combined disconnection analysis (SAT-taking, AP exam, ACT score, and now a majority of `attended.ps`'s effect) turn out to be substantially or primarily about whether someone stays visible in Minnesota's employment record at all, not about whether their employment, once present, is stable.

**Substantive takeaway:** even after accounting for this split, attending postsecondary education (regardless of completion) still meaningfully reduces the risk of cycling through genuinely unstable employment — this isn't a case like SAT-taking or ACT score, where the non-sustained relationship vanished entirely. But the bulk of why `attended.ps` was the strongest predictor in the whole project appears to be about keeping people connected to Minnesota's formal labor market at all, rather than about the quality or stability of the jobs they hold once they're in it. This reinforces that the no-record breakdown of `attended.ps` (once available) is likely to show an even larger gap than either of the two measures examined so far.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$non_sustained.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/Non-sustained cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of Non-Sustained Employment Persistence (`non_sustained.cat`)

**The single biggest finding: three "college ambition" variables completely disappear**

The most important result in this entire round is what *stopped* being significant. ACT composite score (p = 0.393), AP exam participation (p = 0.096), and SAT-taking (p = 0.386) — all three of which showed real relationships with combined disconnection, with SAT-taking producing the single largest raw disparity found anywhere in that earlier analysis (a 20.7-point gap) — **show no relationship at all with non-sustained employment specifically.** This settles the "thwarted college ambition" interpretation raised earlier: it isn't that academically ambitious non-completers end up cycling through unstable jobs. It's that they appear to leave Minnesota's formal employment record altogether — through out-migration, informal work, or genuine labor-force exit — rather than struggling with job stability once employed. Anyone drafting policy language around this population should now describe this specifically as a *reconnection/visibility* problem for this subgroup, not a *job-quality* problem.

**Gender and economic status are the two variables that are genuinely about job instability**

Unlike almost everything else in this analysis, Gender's relationship with non-sustained employment (Cramér's V = 0.099) is actually *stronger* than its relationship with combined disconnection (0.095), and economic status is essentially unchanged (0.075 vs. 0.074). These are the clearest cases in the whole project where a demographic disadvantage is substantially, specifically about cycling through unstable jobs rather than disappearing from the record. If any single intervention were designed around "job stability support," these are the two populations it would most directly serve.

**Race/ethnicity and non-English home don't just weaken — they reverse or split**

This is the most conceptually important refinement in the whole comparison. In the combined measure, every non-White group showed uniformly worse outcomes than White graduates, and non-English-home students showed one of the largest disadvantage gaps in the entire project. Isolating non-sustained employment specifically reveals this uniformity was an illusion: **American Indian and Black graduates genuinely do show worse non-sustained outcomes**, but **Asian/Pacific Islander and Hispanic graduates actually show *better* outcomes than White graduates** on this specific dimension — and non-English-home students, similarly, show *better* non-sustained outcomes despite their large combined-measure disadvantage. This means the earlier "racial disparity" and "non-English-home disparity" findings were each masking two different underlying stories: real job-instability disadvantage for some groups, and a no-record-driven disadvantage for others that has nothing to do with employment stability once someone is working.

**Everything else weakens, several things substantially — with one dramatic exception**

`SpecialEdStatus` (0.104 → 0.070), `edr` (0.034 → 0.024), `cte.achievement` (0.069 → 0.033), `total.cte.courses.taken` (η² 0.0111 → 0.0016), and `attended.ps` (0.142 → 0.081) all retain their original direction but weaken substantially — meaning a real share of each of their combined-disconnection benefits or costs operates through the no-record dimension, even though a genuine non-sustained-specific relationship survives in every case. `Dem_Desc` and `avg.wages.pct.state`, already the weakest variables in the combined analysis, become close to negligible here — the weakest relationships found anywhere in this entire project.

The one dramatic exception is **MCA Science**, which went from barely clearing significance in the combined measure (η² = 0.00027, p = 0.044) to a strong, clearly graded relationship here (η² = 0.00403, p \< 10⁻²⁵). MCA Math held essentially steady and became fully monotonic (every pairwise Tukey comparison significant), and MCA Reading strengthened modestly. This suggests general academic proficiency's true relationship is specifically with employment stability, and that relationship was being diluted by no-record noise in the combined measure. PSEO participation shows the same kind of "hidden until isolated" pattern — null for combined disconnection, but modestly and genuinely significant here (p = 0.001).

**The story in one paragraph**

Splitting non-sustained employment out from no-record status doesn't just weaken the earlier findings uniformly — it sorts every variable examined into two genuinely different categories. Gender, economic status, and general academic proficiency (MCA scores) are substantially or primarily about job instability itself, and their relationships hold up or even strengthen once isolated. Everything associated with clear college-bound ambition — ACT score, AP participation, and especially SAT-taking — turns out to have essentially nothing to do with job stability at all; their large apparent disadvantage was a no-record phenomenon in disguise. And racial/ethnic disparities and non-English-home status split into two different mechanisms depending on subgroup, undermining the idea of a single uniform disadvantage pattern. The clear next step, strongly implied by this entire comparison, is running this same full battery of variables against the no-record-specific measure — which should reveal the mirror image of what's been established here: the SAT/AP/ACT-score relationships reappearing in force, most of `attended.ps`'s and the CTE variables' remaining unexplained benefit surfacing, and the Asian/Pacific Islander, Hispanic, and non-English-home disadvantages showing up clearly for the first time.

<br>

# Relationship Between Individual Characteristics and having No Minnesota Employment Categories

The next sections focus on exploring what relationships exist between individual characteristics and the frequency of having no Minnesota employment record in the first 5 years after formal education.

- **Never no-record** — 0 of the first 5 years (post-high-school or post-postsecondary-exit) had no Minnesota employment record
- **Intermittently no-record** — 1-2 of the first 5 years had no Minnesota employment record
- **Broadly no-record** — exactly 3 of the first 5 years had no Minnesota employment record
- **Persistently no-record** — 4-5 of the first 5 years had no Minnesota employment record

<br>

## Demographic variables

**Gender** (χ² = 8.21, df = 3, Cramér's V = 0.014, p = 0.042 — barely significant, the weakest relationship in this batch). Female and male non-completers show almost identical no-record patterns (16.7% vs. 16.9% persistently no-record — a negligible 0.2-point gap). This is a clean, decisive confirmation of what the non_sustained.cat analysis predicted: Gender's disadvantage in overall disconnection is essentially *entirely* about non-sustained employment (where it was the strongest relationship in that batch, actually strengthening once isolated) and has almost nothing to do with disappearing from the record.

**Economic Status** (χ² = 72.81, df = 3, Cramér's V = 0.042 — the weakest of the significant demographic relationships here, and notably weaker than its 0.075 relationship with non-sustained employment). Persistent no-record is essentially flat between groups (16.6% vs. 17.0% — a negligible, slightly *reversed* gap). Combined with economic status's strong, stable relationship with non-sustained employment across both prior analyses, this confirms economic disadvantage operates almost entirely through job instability, not through leaving the employment record.

**English Learner** (χ² = 297.30, df = 3, Cramér's V = 0.085 — nearly double its non_sustained.cat relationship of 0.044, and now comparable in strength to its original combined-measure relationship of 0.082). This confirms the earlier prediction directly: English learners show a massive 10.6-point gap in persistent no-record status (26.4% vs. 15.8%) and a matching 10.5-point gap in "never no-record" (49.8% vs. 60.3%). Their large combined-disconnection disadvantage is substantially a no-record phenomenon, not a job-instability one — consistent with unequal access to the formal labor market, out-migration, or disengagement from Minnesota's system entirely, rather than difficulty holding a stable job once employed.

**Non-English Home** (χ² = 392.42, df = 3, Cramér's V = 0.098 — **the strongest relationship in this entire batch**, and nearly double its 0.056 relationship with non-sustained employment). This is the most dramatic confirmation of the predicted pattern anywhere in this analysis: a **17.6-point gap** in persistent no-record status (33.7% vs. 16.1%) — the largest raw gap found in any no-record table so far — alongside a matching 17-point gap in "never no-record" (43.0% vs. 60.0%). Recall this same group showed *better* non-sustained outcomes than the reference group in the prior analysis. Put together across all three measures, the story is now unambiguous: non-English-home students are not more likely to hold unstable jobs — if anything, slightly less likely — but are dramatically more likely to disappear from Minnesota's employment record entirely, strongly consistent with migration-related disengagement (leaving the state, working informally, or never fully entering the formal system) rather than any job-quality problem.

**RaceEthnicity** (χ² = 346.80, df = 12, Cramér's V = 0.053) — and this table resolves the race/ethnicity split in exactly the way the prior analysis anticipated, cleanly separating two different mechanisms:

- **Asian/Pacific Islander (29.2%) and Hispanic (28.9%) show the *worst* persistent no-record rates of any group** — nearly double White graduates' rate (16.1%), and notably worse even than American Indian (21.0%) or Black (18.2%) graduates. Recall these same two groups showed *better*-than-White non-sustained outcomes in the prior analysis. This confirms their large disadvantage in the original combined measure was driven almost entirely by no-record status, not job instability.
- **American Indian (21.0%) and Black (18.2%) graduates also show elevated persistent no-record rates relative to White graduates, though less extreme than Asian/PI or Hispanic.** Since these two groups *also* showed real, elevated non-sustained-employment disadvantage in the prior analysis, their overall disconnection risk appears to operate through **both** channels simultaneously — unlike Asian/Pacific Islander and Hispanic graduates, whose disadvantage is concentrated almost entirely in the no-record dimension.

**SpecialEdStatus** (χ² = 146.58, df = 3, Cramér's V = 0.060 — somewhat weaker than its 0.070 relationship with non-sustained employment). SpecialEd students show a real but more modest gap here (20.4% vs. 15.8% persistently no-record, a 4.6-point gap) compared to non-English home or English learner. Unlike those two variables, SpecialEdStatus's disadvantage appears to split fairly evenly between the two mechanisms, with perhaps a slight lean toward non-sustained employment rather than no-record status specifically.

**Synthesis: the three-way comparison now fully resolves the driver question for every demographic variable**

| Variable | Combined (disconnected.cat) | Non-sustained-specific | No-record-specific | Primary mechanism |
|---------------|---------------|---------------|---------------|---------------|
| Gender | 0.095 | **0.099** | 0.014 | Almost entirely non-sustained employment |
| Economic status | 0.074 | **0.075** | 0.042 | Predominantly non-sustained employment |
| SpecialEdStatus | 0.104 | 0.070 | 0.060 | Both, fairly evenly split |
| RaceEthnicity | 0.055 | 0.042 | 0.053 | **Splits by subgroup** — AI/Black: both; Asian-PI/Hispanic: almost entirely no-record |
| English learner | 0.082 | 0.044 | **0.085** | Predominantly no-record status |
| Non-English home | 0.060 | 0.056 | **0.098** | Almost entirely no-record status |

This table is the cleanest possible summary of what this whole three-part comparison exercise has established: every demographic variable examined in this project has a clearly identifiable primary mechanism, and they split roughly into two camps. Gender, economic status, and (partly) SpecialEdStatus drive disconnection mainly through genuine employment instability. English learner status, non-English-home status, and — for two of its four disadvantaged subgroups specifically — race/ethnicity drive disconnection mainly through disengagement from Minnesota's formal employment record altogether. This is a substantively important distinction for any policy response: interventions aimed at job stability and job quality would meaningfully help the first group, but would likely miss the actual barrier facing the second — for whom reconnection, outreach, or addressing structural/migratory barriers to formal employment access is probably the more relevant lever.

#### `Gender`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = master$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/Gender.html")
```

<br>

#### `RaceEthnicity`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/RaceEthnicity.html")
```

<br>

#### `economic.status`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/economic-status.html")
```

<br>

#### `SpecialEdStatus`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/SpecialEdStatus.html")
```

<br>

#### `english.learner`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/english-learner.html")
```

<br>

#### `non.english.home`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Demographics/non-english-home.html")
```

<br>

## High school characteristics

**Every one of these four variables is now strongest here, of the three DVs examined across this whole project — confirming the pattern already established for the demographic batch.** Regional and local labor-market context appear to matter considerably more for whether someone disappears from Minnesota's employment record than for whether their employment, once present, is unstable.

**avg.unemp.rate** (F(3, 41275) = 31.23, p \< 0.001, η² = 0.00227 — the strongest of the three DV versions, more than 5x the non_sustained.cat relationship). The Tukey pattern here is worth describing carefully, because it's genuinely different in shape from what's been seen before: **it isn't monotonic.** Intermittently, Broadly, and Persistently no-record individuals all live in significantly higher-unemployment areas than the "Never" group — but Intermittently no-record shows the *largest* gap (+0.211), larger than Broadly (+0.193) or even Persistently no-record (+0.117), and Persistent is actually significantly *lower* than Intermittent (p = 0.013). In other words, local unemployment relates most strongly to *some* no-record exposure, not to the most severe, chronic form of it. This suggests high-unemployment areas may be associated with people cycling in and out of Minnesota's employment record intermittently — plausibly tied to seasonal or unstable local labor demand — rather than with the kind of sustained absence that characterizes the persistently no-record group, which may be driven more by factors like out-migration that aren't strongly tied to local unemployment at all.

**avg.wages.pct.state** (F(3, 41275) = 16.71, p \< 0.001, η² = 0.00121 — comparable to its combined-measure relationship, and roughly 5.5x its non_sustained.cat relationship). Same non-monotonic shape as unemployment rate: all three no-record categories differ significantly from "Never," but not from each other (Broadly no-record actually shows the single largest wage gap, not Persistently no-record). As with unemployment, this points to a "some no-record exposure vs. none" threshold relationship with local wages, not a graded severity relationship.

**Dem_Desc** (χ² = 97.75, df = 6, Cramér's V = 0.034 — its strongest relationship of the three DVs, more than double its non_sustained.cat value of 0.013). Town/rural mix shows meaningfully worse outcomes (55.3% never no-record, 19.1% persistent) than Entirely urban (61.9%, 16.0%) or Urban/town/rural mix (60.1%, 16.2%), which are similar to each other. Given the established complete-nesting relationship between "Town/rural mix" and "EDR 5," this finding is best read together with the `edr` table below rather than as an independent signal.

**edr** (χ² = 121.11, df = 6, Cramér's V = 0.038 — also its strongest relationship of the three DVs). **This table reveals a genuinely different regional ranking than either of the previous two DVs showed**, worth highlighting explicitly: **EDR 5 is now the worst-performing region** (55.3% never no-record, 19.1% persistent), **EDR 7W is the best** (61.3% never, only 16.3% persistent), and **EDR 7E sits in between** (58.9% never, 15.9% persistent — actually the *lowest* persistent rate of the three, though not the highest "never" rate). Compare this to the earlier findings: in the combined measure, EDR 7W was clearly best and EDR 5/7E were both worse and similar to each other; in the non-sustained-specific measure, EDR 7E stood out as uniquely worse while EDR 5 and 7W looked similar. Here, isolating no-record status specifically, the ranking reshuffles again, with EDR 5 emerging as the distinct laggard. **Each EDR region appears to carry a different specific disconnection risk profile** — EDR 7E's disadvantage in the combined and non-sustained measures is apparently more about job instability, while EDR 5's disadvantage is more specifically about disappearing from the employment record altogether.

**Synthesis for this batch**

| Variable | Combined (disconnected.cat) | Non-sustained-specific | No-record-specific |
|------------------|------------------|------------------|------------------|
| avg.unemp.rate (η²) | 0.0014 | 0.0004 | **0.0023** |
| avg.wages.pct.state (η²) | 0.0012 | 0.0002 | **0.0012** |
| Dem_Desc (V) | 0.026 | 0.013 | **0.034** |
| edr (V) | 0.034 | 0.024 | **0.038** |

Every regional and labor-market variable follows the same pattern already seen in the demographic batch: weakest for non-sustained employment, strongest for no-record status. Combined with the demographic findings, this reinforces a clear structural picture — geography and local labor-market conditions primarily predict whether someone stays visible in Minnesota's formal employment system, while genuine job instability (once someone is employed) is driven more by individual-level factors like gender and economic status than by where someone's high school was located or local economic conditions. The one new wrinkle worth carrying forward is that `edr`'s regional ranking isn't consistent across the three DVs — EDR 5, 7E, and 7W each show a different "signature" disadvantage depending on which disconnection mechanism is being examined, which is a more nuanced regional story than any single DV version would suggest on its own.

#### `Dem_Desc`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school characteristics/Dem_Desc.html")
```

<br>

#### `edr`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school characteristics/edr.html")
```

<br>

#### `avg.unemp.rate`

```{r}
summary <- aov(avg.unemp.rate ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school characteristics/avg-unemp-rate.xlsx", overwrite = TRUE)
```

<br>

#### `avg.wages.pct.state`

```{r}
summary <- aov(avg.wages.pct.state ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school characteristics/avg-wages-pct-state.xlsx", overwrite = TRUE)
```

<br>

## High school academics

**This table resolves every open question left by the earlier non-sustained-specific analysis. Every "college ambition" variable that vanished for non-sustained employment reappears strongly here — confirming their combined-measure disadvantage was a no-record phenomenon all along.**

**ACTCompositeScore** (F(3,17151) = 69.36, p ≈ 1.4 × 10⁻⁴¹, η² = 0.01199 — **now the strongest continuous relationship found anywhere across all three DVs in this entire project**, more than 70x its non-sustained-specific value of 0.00017, and even stronger than its combined-measure value of 0.0088). This is a complete, decisive confirmation: ACT score's relationship with disconnection is essentially entirely about no-record status. The Tukey pattern again shows a non-monotonic, threshold-like shape — Broadly no-record shows the single largest gap from Never (+1.33 points), even larger than Persistently no-record (+1.10) — consistent with the "some no-record exposure vs. none" pattern seen for the regional/labor-market variables in the prior batch.

**sat.taken** (χ² = 43.83, df = 3, Cramér's V = 0.033 — comparable to its combined-measure relationship of 0.031, and a complete reversal of its null non-sustained relationship, p = 0.386). This is the clean resolution of the single most dramatic finding in the whole non-completer analysis. SAT-takers show a **14.9-point gap in persistent no-record status** (31.7% vs. 16.8%) and an **18.7-point gap in "never no-record"** (40.7% vs. 59.4%). Combined with the complete absence of any SAT relationship with non-sustained employment, this confirms unambiguously: SAT-takers who never completed a postsecondary credential are not more likely to hold unstable jobs — they are dramatically more likely to disappear from Minnesota's formal employment record entirely. This is almost certainly the strongest single "mechanism-specific" finding in this entire project.

**ap.exam** (χ² = 36.18, df = 3, Cramér's V = 0.030 — stronger than its combined-measure relationship of 0.022, and a full reversal of its null non-sustained result, p = 0.096). Worth noting the *shape* here is distinct from SAT: AP exam takers show essentially identical persistent no-record rates to non-takers (16.7% vs. 16.8% — no gap at all), but a real gap concentrated in the intermittent (20.8% vs. 17.9%) and broadly no-record (7.2% vs. 5.6%) categories. AP participation relates to *some* no-record exposure, but not to the most chronic form of it — a milder version of the SAT pattern.

**took.ACT** (χ² = 213.17, df = 3, Cramér's V = 0.072 — nearly as strong as its combined-measure relationship of 0.074, and notably stronger than its non-sustained-specific value of 0.039). ACT-takers show a real 5.2-point advantage in persistent no-record status (13.8% vs. 19.0%). Unlike the three variables above, `took.ACT` retains a meaningful relationship with *both* mechanisms — its combined-measure strength reflects genuine contributions from both non-sustained employment and no-record status, though the no-record channel is clearly the larger of the two.

**cte.achievement** (χ² = 240.31, df = 6, Cramér's V = 0.054) and **total.cte.courses.taken** (η² = 0.00652) both retain meaningful relationships with no-record status as well as non-sustained employment — confirming CTE engagement is a genuine two-channel protective factor, not one masquerading as the other. Worth noting a shape difference for `cte.achievement`: here, CTE Concentrator/Completer (15.1% persistent no-record) and CTE Participant (15.2%) are nearly identical to each other, both clearly better than No CTE (20.5%) — a simpler "any CTE vs. none" pattern, rather than the graded concentrator \> participant \> none ordering seen for non-sustained employment.

**pseo.participant** (χ² = 32.84, df = 3, Cramér's V = 0.034 — **its strongest relationship of all three DVs**, and the clearest case in this project of a variable whose real effect was completely masked by combining two mechanisms). PSEO participants show a modest but real advantage in persistent no-record status (15.3% vs. 16.6%). Given PSEO showed no relationship at all with the naive combined measure, this confirms that variable had two small, real, and apparently offsetting-enough relationships (with non-sustained employment and no-record status) that neither surfaced clearly until examined separately.

**MCA-M** (η² = 0.00058) and **MCA-R** (η² = 0.00018, **not significant**, p = 0.077) both weaken sharply here compared to their non-sustained-specific relationships (0.00286 and 0.00157, respectively) — confirming these two variables' true signal is specifically about job stability, not no-record status, the mirror image of the ACT/AP/SAT pattern.

**MCA-S** (η² = 0.00165) sits between its dramatic non-sustained-specific relationship (0.00403) and its negligible combined-measure relationship (0.00027) — a real, moderate relationship with no-record status exists, but this variable's primary signal remains its strong relationship with non-sustained employment established in the prior analysis.

**Complete three-way synthesis for every academic variable examined in this project**

| Variable | Combined | Non-sustained-specific | No-record-specific | Primary mechanism |
|---------------|---------------|---------------|---------------|---------------|
| ACTCompositeScore | 0.0088 | 0.0002 (ns) | **0.0120** | Almost entirely no-record |
| sat.taken | 0.031 | 0.009 (ns) | **0.033** | Almost entirely no-record |
| ap.exam | 0.022 | 0.012 (ns) | **0.030** | Almost entirely no-record |
| took.ACT | 0.074 | 0.039 | **0.072** | Both, no-record somewhat stronger |
| cte.achievement | 0.069 | 0.033 | 0.054 | Both, no-record somewhat stronger |
| total.cte.courses.taken | **0.0111** | 0.0016 | 0.0065 | Both channels contribute |
| pseo.participant | 0.012 (ns) | 0.024 | **0.034** | Both, masked when combined |
| MCA-M | 0.0029 | **0.0029** | 0.0006 | Almost entirely non-sustained |
| MCA-R | 0.0009 | **0.0016** | 0.0002 (ns) | Almost entirely non-sustained |
| MCA-S | 0.0003 (barely) | **0.0040** | 0.0017 | Mostly non-sustained, some no-record |

This completes the full picture for academic variables. The pattern that has now emerged consistently across every batch of variables in this three-way comparison — demographic, regional, and academic — is the same: **every "college-ambition" signal (ACT score, AP, SAT) is fundamentally about disengagement from Minnesota's formal employment record, while general academic proficiency (MCA scores) is fundamentally about employment stability itself.** CTE engagement and PSEO participation are the two variables in this entire project that operate through both mechanisms simultaneously, which is itself a notable and consistent finding — these are the two accomplishment variables whose benefit doesn't reduce to a single underlying story.

#### `pseo.participant`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/pseo-participant.html")
```

<br>

#### `ap.exam`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/ap-exam.html")
```

<br>

#### `took.ACT`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/took-ACT.html")
```

<br>

#### `ACTCompositeScore`

```{r}
summary <- aov(ACTCompositeScore ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/ACTCompositeScore.xlsx", overwrite = TRUE)
```

<br>

#### `sat.taken`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/sat-taken.html")
```

<br>

#### `MCA.M`

```{r}
summary <- aov(MCA.M ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-M.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.R`

```{r}
summary <- aov(MCA.R ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-R.xlsx", overwrite = TRUE)
```

<br>

#### `MCA.S`

```{r}
summary <- aov(MCA.S ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/MCA-S.xlsx", overwrite = TRUE)
```

<br>

#### `cte.achievement`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/High school accomplishments/cte-achievement.html")
```

<br>

#### `total.cte.courses.taken`

```{r}
summary <- aov(total.cte.courses.taken ~ no_record.cat, data = updated.master)

summary

anova_df <- broom::tidy(summary)

TukeyHSD(summary)

tukey_df <- broom::tidy(TukeyHSD(summary))

wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
addWorksheet(wb, "TukeyHSD")
writeData(wb, "ANOVA", anova_df)
writeData(wb, "TukeyHSD", tukey_df)
saveWorkbook(wb, "Data/Claude/Non-college deeper/No record cat/High school accomplishments/total-cte-courses-taken.xlsx", overwrite = TRUE)
```

<br>

## Post-secondary

χ² = 321.45, df = 3, Cramér's V = 0.088 — slightly stronger than its non-sustained-specific relationship (0.081), and, like every other variable in this project, weaker than its combined-measure relationship (0.142, still the strongest single relationship found anywhere in this entire analysis).

Non-completers who attended some postsecondary education show a real advantage here too: 13.2% persistently no-record vs. 18.8% for those who never attended (a 5.6-point gap), and 64.7% never no-record vs. 56.3% (an 8.4-point gap).

**This is the cleanest "both channels contribute roughly evenly" case found anywhere in the entire three-way comparison.** The non-sustained-specific gap (5.1 points) and the no-record-specific gap (5.6 points) are almost identical in size, and together they account for nearly all of the combined-measure gap (10.7 of 13.4 points, about 80%). Unlike ACT score, AP exam, and SAT-taking — all overwhelmingly no-record phenomena — or MCA scores — overwhelmingly non-sustained phenomena — `attended.ps` is genuinely a dual-mechanism variable: postsecondary attendance, even without completion, measurably reduces both the risk of cycling through unstable jobs *and* the risk of disappearing from Minnesota's employment record, in roughly equal measure.

**This is a fitting final result for the whole comparison exercise, since it confirms `attended.ps` earns its status as the single strongest predictor in this project honestly** — not because it happens to align with one dominant mechanism the way several other variables do, but because it provides a genuine, balanced benefit across both pathways to disconnection. That makes it arguably the most robust single finding in the entire non-completer analysis: whatever the reason someone is at risk of disconnection — job instability or disengagement from the formal labor market — postsecondary attendance, even incomplete, appears to help on both fronts.

#### `attended.ps`

```{r}
tab_xtab(var.row = updated.master$no_record.cat, var.col = updated.master$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE,
         use.viewer = TRUE,
         file = "Data/Claude/Non-college deeper/No record cat/Post-secondary/attended-ps.html")
```

<br>

## Summary: Bivariate Predictors of No-Record Persistence (`no_record.cat`)

**The overarching finding: this measure is where the "hidden" relationships from the whole project finally surface**

Across every batch examined — demographic, regional, academic, and postsecondary attendance — the pattern has been remarkably consistent: variables that showed weak, null, or even reversed relationships with non-sustained employment specifically turn out to have their strongest, clearest relationships here. No-record status is where several of this project's most important findings were actually hiding.

**The three college-ambition variables resolve completely**

ACT composite score, AP exam participation, and SAT-taking all showed **no relationship whatsoever** with non-sustained employment in isolation, despite meaningful (and in SAT's case, dramatic) relationships with the combined disconnection measure. Here, all three reappear in force — ACT score's relationship (η² = 0.012) is in fact **the single strongest continuous relationship found anywhere in this entire three-DV comparison**, and SAT-taking shows a **14.9-point gap in persistent no-record status**, confirming that the largest raw disparity found anywhere in the whole non-completer analysis is fundamentally a story about disengagement from Minnesota's formal employment record, not job instability. This settles the "thwarted college ambition" interpretation raised early in this analysis: these are not people who struggle to hold jobs — they are people who, having pursued (but not completed) a four-year-college pathway, appear considerably more likely to disappear from the state's employment data entirely, plausibly reflecting relocation, informal work, or a genuinely different post-high-school trajectory than the CTE-oriented non-completers this project has otherwise focused on.

**General academic proficiency (MCA) tells the opposite story**

MCA Math and Reading both weaken sharply here after showing strong, stable relationships with non-sustained employment — Reading, in fact, is not even statistically significant for no-record status (p = 0.077). MCA Science sits in between, retaining a moderate relationship here but a much stronger one for non-sustained employment. Together with the ACT/AP/SAT findings above, this draws about as clean a line as this project has found: **academic proficiency (as measured by MCA) is fundamentally about employment stability once someone is working; college-application-specific signals (ACT, AP, SAT) are fundamentally about whether someone stays visible in the workforce at all.**

**Demographic disparities split into three clear patterns, not one**

- **Gender and economic status** are almost entirely non-sustained-employment stories (their no-record relationships are negligible-to-weak), confirming these two variables' disadvantage is genuinely about job instability.
- **English learner status and non-English-home status** are overwhelmingly no-record stories — non-English-home shows the **largest raw gap of any demographic variable in the no-record analysis** (17.6 points), a complete reversal of the *better*-than-average outcomes this same group showed for non-sustained employment specifically.
- **Race/ethnicity splits by subgroup**, and this table resolves that split definitively: American Indian and Black graduates show real disadvantage through *both* mechanisms, while Asian/Pacific Islander and Hispanic graduates — who showed *better*-than-White non-sustained outcomes — show the *worst* no-record outcomes of any group examined (29.2% and 28.9% persistently no-record, respectively). Their large disadvantage in the original combined measure was concentrated almost entirely in disappearing from the employment record, not job instability.

**CTE, PSEO, and postsecondary attendance are the project's genuine "both-mechanism" variables**

Unlike the variables above, `took.ACT`, `cte.achievement`, `total.cte.courses.taken`, `pseo.participant`, and `attended.ps` all show meaningful relationships with *both* non-sustained employment and no-record status. `pseo.participant` is a particularly notable case: it showed literally no relationship with the combined disconnection measure, but real, independent relationships with each individual mechanism — the clearest example in this project of an effect that was completely masked by combining two offsetting-strength signals. `attended.ps` closes out the analysis fittingly: its two component gaps (5.1 points for non-sustained, 5.6 points for no-record) are nearly identical and together account for about 80% of its combined-measure advantage — confirming it earns its status as the strongest single predictor in this entire project by genuinely helping through both pathways, rather than by riding one dominant mechanism the way most other variables do.

**Regional geography tells a genuinely different story here than anywhere else in the project**

`edr`'s ranking reshuffles for the third time across the three DVs: EDR 7W was best for combined disconnection, EDR 7E was uniquely disadvantaged for non-sustained employment, and here **EDR 5 emerges as the clear laggard** for no-record status specifically, with EDR 7W again the strongest performer. Each region appears to carry a genuinely different disconnection "signature" depending on which mechanism is examined — a nuance that would be completely invisible using the combined measure alone. Local unemployment and wages also both show their strongest relationships here (as with the demographic and academic batches), reinforcing that regional and local labor-market context primarily shapes whether someone stays connected to Minnesota's formal employment system at all, more than it shapes job stability once employed.

**The complete picture across all three DVs, now finished**

With this table, every variable examined in this project's bivariate Step 3 has now been decomposed across all three measures of disconnection. The consistent finding across nearly thirty variables is that **"disconnection" was never one phenomenon** — it is at minimum two distinguishable trajectories (chronic job instability and chronic disappearance from the employment record), each with its own distinct set of predictors, and a genuine third category of people who experience both. Any final report or subsequent multivariate confirmatory stage built on this bivariate work should treat these as separate outcomes to the greatest extent the analysis allows, since combining them — as the original `disconnected.cat` measure does — reliably obscures real, policy-relevant relationships in both directions.

<br>

## Full Synthesis: Workforce Disconnection Among Non-College Graduates in Central Minnesota

### The scope of the problem

Disconnection from meaningful workforce participation is the modal experience for non-completers in this cohort, not a marginal one. Nearly 44% of non-completers were persistently disconnected (disconnected in 4 or 5 of their first 5 years), while only 15.6% — the smallest of four trajectory categories — maintained full attachment throughout. Tracking this same population year by year revealed a genuine sorting dynamic rather than a uniform recovery: combined sustained employment rose steadily (35.1% → 48.1% across years 1–5), but so did "no Minnesota employment record" (21.8% → 27.1%, with no deceleration by year 5). And among those who did reach sustained employment, a growing share of that success occurred *outside* the region — sustained regional attachment plateaus sharply after year 3, while sustained attachment elsewhere in Minnesota kept climbing. Any framing of this population needs to hold two things at once: instability is the default trajectory, and even the "success" cases increasingly locate outside Central Minnesota over time.

### The single most important methodological finding: "disconnection" is not one phenomenon

The decision to separate non-sustained employment from no-Minnesota-record status turned out to be the pivotal analytic choice in this entire project. These two forms of disconnection have almost opposite shapes:

- **Non-sustained employment is common but usually temporary.** Nearly 70% of non-completers experience it at some point, but conditional on that exposure, only about 1 in 5 becomes persistent — most resolve within a year or two.
- **No-record status is rarer but far stickier.** Only about 41% of non-completers ever experience it, but conditional on exposure, roughly 2 in 5 become persistently no-record — more than double the conditional persistence rate of non-sustained employment.
- **A genuine third group — roughly a quarter of all non-completers — experiences both** at different points, rather than being stuck in one. Within the persistently disconnected population specifically, no single driver dominates: no-record-driven (38.5%), non-sustained-driven (32.5%), and mixed (29.1%) are all comparably sized. This is why the combined "persistently disconnected" figure (43.8%) so far exceeds either individual persistence measure (14.2% and 16.8%) — a meaningful share of the worst-off group is cycling between both problems, not stuck in one.

This distinction reshaped the interpretation of nearly every subsequent finding in the bivariate analysis, and is the organizing idea the rest of this synthesis builds on.

### The bivariate analysis: three DVs, one variable at a time, and a consistent decomposition logic

Running the same battery of demographic, regional, and academic-accomplishment variables against all three measures of disconnection (the combined `disconnected.cat`, and the two isolated measures `non_sustained.cat` and `no_record.cat`) revealed that almost every variable examined has an identifiable *primary mechanism* — even when its combined-measure relationship looked ambiguous, weak, or even null.

**Variables genuinely about job instability** (their relationship holds or strengthens when non-sustained employment is isolated, and weakens or disappears for no-record status): - **Gender** — the cleanest case in the whole project. Its no-record relationship is negligible (Cramér's V = 0.014, barely significant), while its non-sustained relationship is, if anything, slightly *stronger* than its combined-measure relationship. - **Economic status** — nearly identical pattern, weaker no-record relationship (0.042) than non-sustained (0.075). - **MCA Math and Reading** — general academic proficiency relates to employment stability specifically; MCA Reading isn't even significant for no-record status.

**Variables genuinely about disengagement from Minnesota's formal employment record** (the opposite pattern — weak or null for non-sustained employment, strong for no-record status): - **ACT composite score, AP exam participation, and SAT-taking** — all three showed **no relationship whatsoever** with non-sustained employment in isolation, despite meaningful (and for SAT, the single largest raw disparity in the entire project — a 20.7-point gap) relationships with the combined measure. Isolated to no-record status, all three reappear in force; ACT score's no-record relationship (η² = 0.012) is in fact the strongest continuous relationship found anywhere across all three DVs. This resolves what looked like a "thwarted college ambition" paradox into something more specific: non-completers who show the clearest college-bound academic signals are not more likely to struggle with job stability — they are dramatically more likely to disappear from Minnesota's employment data entirely, consistent with relocation, informal work, or a genuinely different post-high-school trajectory. - **English learner status and non-English-home status** — both showed large disadvantage in the combined measure, but that disadvantage is concentrated overwhelmingly in the no-record dimension. Non-English-home, remarkably, showed *better*-than-average non-sustained outcomes but the *worst* no-record gap of any demographic variable (17.6 points) — plausibly reflecting migration-related disengagement from the formal labor market rather than any job-quality problem. - **RaceEthnicity splits cleanly by subgroup**, resolving the uniform "all non-White groups disadvantaged" appearance of the combined measure into two distinct stories: American Indian and Black graduates show real disadvantage through *both* mechanisms, while Asian/Pacific Islander and Hispanic graduates — who actually showed *better*-than-White non-sustained outcomes — show the worst no-record outcomes of any group examined (29.2% and 28.9% persistently no-record, nearly double White graduates' rate).

**Variables that are genuinely dual-mechanism** (real, independent relationships with both non-sustained employment and no-record status): - **took.ACT, cte.achievement, total.cte.courses.taken**, and especially **pseo.participant** — the last of which showed literally no relationship with the combined measure (p = 0.276) but real, independent relationships with each isolated mechanism, the clearest example in this project of a true effect completely masked by combining two separate signals. - **attended.ps** — the single strongest bivariate relationship found anywhere in the entire analysis (Cramér's V = 0.142 for the combined measure), and notably the cleanest "both channels contribute equally" case: its non-sustained-specific gap (5.1 points) and no-record-specific gap (5.6 points) are nearly identical, together accounting for about 80% of its combined-measure advantage. Postsecondary attendance, even without completion, appears to genuinely help through both pathways to disconnection — which is likely why it emerged as the strongest single finding in the whole project.

**Regional geography (`edr`) reshuffles depending on which mechanism is examined** — EDR 7W was strongest for the combined measure, EDR 7E was the uniquely disadvantaged region for non-sustained employment specifically, and EDR 5 emerged as the clear laggard for no-record status specifically. Each region appears to carry a genuinely different disconnection "signature," a nuance invisible in the combined measure alone. Local unemployment and wages both showed their strongest relationships with no-record status, reinforcing that geography and local labor-market conditions shape *whether someone stays visible in the formal employment system* more than they shape job stability once someone is working.

### How this connects to the earlier completer-focused project (DV1–DV5)

Several threads from the original multivariate confirmation analysis reappear here, in an entirely different population, which strengthens confidence in both:

- **The unexplained Black-White gap** that persisted (and widened) across DV1, DV2, and DV5 shows up again in this non-completer population — American Indian graduates in particular show the most severe disconnection disparity of any group (60.6% persistently disconnected combined, 7.0% consistently attached). This disparity appears independent of postsecondary completion status entirely.
- **The regional pattern** (EDR 7W advantage, EDR 7E and EDR 5 disadvantage on different measures) echoes the DV1–DV5 finding that regional context consistently mattered for workforce outcomes — though this analysis reveals that pattern is more mechanism-specific than the original project's single-outcome framing could show.
- **PSEO and CTE's dual role as regional-anchoring, workforce-connected pathways** continues here, now shown to matter even among people who never completed a credential — reinforcing that these variables operate through direct workforce-readiness channels, not just through the credentialing pathway examined in DV1–DV5.
- **The "college-track mobility" pattern** — where ACT score, AP, and SAT predicted leaving the region or not completing regionally in DV1/DV4/DV5 — reappears here in a sharper, more specific form: among people who never completed anything, those same signals predict disappearing from the state's employment record altogether, the logical extreme of the "mobility" pattern first identified in the completer-focused work.
- **The `Dem_Desc`/`edr` complete-nesting relationship** ("Town/rural mix" and "EDR 5" being the identical group) discovered during the original DV1 multivariate modeling was confirmed to hold in this new population as well — worth remembering if this analysis ever proceeds to a multivariate confirmatory stage.

### Policy and reporting implications

Three genuinely different populations of at-risk non-completers emerge from this analysis, and each likely needs a different kind of intervention:

1.  **A job-instability population** (disproportionately identified by gender and economic status): engaged with the labor market but unable to find steady footing. Job-quality and job-stability support is the natural intervention here.
2.  **A disengagement population** (disproportionately identified by English learner status, non-English-home status, high ACT/AP/SAT scores among non-completers, and — for two of its four subgroups — race/ethnicity): not struggling with job stability so much as disappearing from the formal system entirely. This population likely needs reconnection and outreach, and for some subgroups, attention to migration-related or structural barriers to formal labor-market access, not job-quality fixes.
3.  **A compound population** (roughly a quarter of all non-completers, and about 29% of the persistently disconnected specifically): cycling between both problems rather than experiencing either one chronically, likely representing the most complex cases requiring more comprehensive support.

Cutting across all three groups, **postsecondary attendance — even without completion — is the single most consistent, actionable lever identified in this entire analysis**, helping through both mechanisms roughly equally. CTE engagement and PSEO participation are the next most consistently protective factors, also operating through both channels.

```{r}
write_csv(updated.master, "Data/SLEDS/Masters/Master-non-college-grads-disconnection.csv")
```
