---
title: "Post-secondary pathways"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-after-ct.csv") %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "1", "Public, 4-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "4", "Public, 2-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "11", "Other", as.character(ps.grad.InstitutionSector)),
         ps.grad.location = ifelse(ps.grad.location == "PS grad EDR", "PS grad region", ps.grad.location))

kable(names(master.original))

kable(head(master.original))
```

# Dependent Variable

The cross-tab and ANOVA analysis exploring relationships between demographic, high school, and post-secondary variables with employment outcomes highlighted the importance of post-secondary variables in an individuals journey to meaningful employment inside and outside the region. Due to this, it makes sense to explore if some of the other variables play a role in post-secondary outcomes - whether an individual attends, graduates, the type of institution they attend and the location.

We will explore this here. We will explore this in three sections;

[**Tier 1: Core Postsecondary Access and Attainment**]{.underline}

These outcomes are most foundational and closely tied to employment.

1\. `attended.ps.within.first.year.hsgrad`

**Question:** Who transitions directly to postsecondary? **Why it matters:** Strongly linked to long-term success and minimizes "disconnection."

2\. `attended.ps` (ever attended postsecondary)

**Question:** Who attends postsecondary at all? **Why it matters:** Broader view than immediate enrollment.

3\. `ps.grad` (graduated from postsecondary)

**Question:** Who completes a credential? **Why it matters:** Credential completion is often the real driver of employment outcomes.

[**🔹 Tier 2: Type of Postsecondary Experience**]{.underline}

These outcomes differentiate the **nature** of the experience, which may shape career pathways and economic outcomes.

4\. `ps.grad.InstitutionSector`

**Question:** Who graduates from 2-year vs. 4-year, public vs. private schools? **Why it matters:** This affects cost, prestige, job access, and transfer opportunities.

5\. `highest.cred.level`

**Question:** Who earns certificates, associate degrees, bachelor’s degrees? **Why it matters:** Different credentials yield different labor market returns.

[**Tier 3: Location of Postsecondary Experience**]{.underline}

These outcomes reveal **mobility**, local engagement, or place-based equity patterns.

6\. `ps.grad.location`

**Question:** Who stays in the region for school? **Why it matters:** Tied to affordability and regional workforce planning.

Each dependent variable has the following categories;

```{r dependent variable categories}
data <- master.original %>%
  select(attended.ps.within.first.year.hsgrad, attended.ps, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location) %>%
  mutate(across(c("attended.ps.within.first.year.hsgrad":"ps.grad.location"), factor))
```

**attended.ps.within.first.yer.hsgrad**

```{r echo = FALSE}
levels(data$attended.ps.within.first.year.hsgrad)

```

**attended.ps**

```{r echo = FALSE}
levels(data$attended.ps)
```

**ps.grad**

```{r echo = FALSE}
levels(data$ps.grad)
```

**ps.grad.InstitutionSector**

```{r echo = FALSE}
levels(data$ps.grad.InstitutionSector)
```

**highest.cred.level**

```{r echo = FALSE}
levels(data$highest.cred.level)
```

**ps.grad.location**

```{r echo = FALSE}
levels(data$ps.grad.location)
```

<br>

# Predictors

For all of these, we will use the predictors that were in Tier 1 and Tier 2 of the previous cross-tab analysis.

-   Demographics
    -   economic.status
    -   RaceEthnicity
    -   SpecialEdStatus
-   High school characteristics
    -   Dem_Desc
    -   avg.unemp.rate
    -   avg.wage.pct.state
-   High school accomplishments
    -   ACTCompositeScore
    -   pseo.participant
    -   cte.achievement
    -   total.cte.courses.taken
    -   MCA.M
    -   MCA.R
    -   MCA.S

# attended.ps.within.first.year.hsgrad

## Interpretation of findings

Immediate postsecondary enrollment within one year of high school graduation is shaped by a combination of **individual**, **school-level**, and **academic preparation** factors.

From a **demographic perspective**, the strongest influences were special education status and economic status. Students who had received special education services and those who were economically disadvantaged were substantially less likely to attend immediately, suggesting barriers related to academic support needs and financial resources. Race/ethnicity also played a role, with Asian/PI and White students having the highest rates and American Indian students the lowest, reflecting broader systemic disparities.

**High school characteristics** added another layer of influence. Economic conditions in the community—particularly lower unemployment rates and higher wages relative to the state average—were more strongly associated with immediate enrollment than rurality alone. While urban students had slightly higher attendance rates than their rural counterparts, community economic health appeared to be the more critical driver, potentially reflecting both the local opportunity structure and the resources available to schools and families.

**High school accomplishments** showed the most direct and consistent relationship with postsecondary entry. Strong academic performance (ACT and MCA scores) and early college exposure through PSEO were powerful predictors of immediate enrollment, with PSEO standing out as the single strongest individual factor. In contrast, greater engagement in Career and Technical Education was linked to lower immediate enrollment, likely reflecting alternative workforce-oriented pathways.

Taken together, these findings suggest that **academic readiness, early exposure to college, and favorable economic contexts all facilitate rapid postsecondary transition**, while financial, academic support, and systemic equity barriers can slow or divert that path. At the same time, CTE participation highlights that not all students aim for immediate college enrollment—emphasizing the importance of supporting multiple, equally viable pathways into adulthood.

## Rank of variable importance

1.  **PSEO participation**
    -   W**hy:** Strongest effect size (Cramer's V=0.356) and largest differences in immediate postsecondary attendance (81.4% vs. 46.8% for non-participants).
    -   **Interpretation:** Clear indicator of college readiness and momentum toward rapid postsecondary entry.
2.  **Special education status**
    -   **Why:** Second-highest effect size (Cramer's V=0.282) with a large gap in attendance (67.9% vs. 29.5%).
    -   **Interpretation:** Indicates substantial barriers for students with special education needs.
3.  **ACT Composite Score**
    -   **Why:** Very large F-value (2827.0) and substantial mean score differences between groups (up to 3 points).
    -   **Interpretation:** Strong measure of academic readiness and predictor of immediate enrollment.
4.  **Economic status**
    -   **Why:** Moderate effect size (Cramer's V=0.225) and large gap (70% vs. 46.7%).
    -   **Interpretation:** Reflects the role of financial resources in enabling postsecondary transition.
5.  **MCA scores (Math, Reading, Science)**
    -   **Why:** All had very large F-values (4,000–6,000+) with consistent gaps favoring completers.
    -   **Interpretation:** Reinforces that academic performance across core subjects correlates with postsecondary entry.
6.  **Average county unemployment rate**
    -   **Why:** Largest F-value among school characteristics (575.6) and clear differences in economic contexts between completers and non-completers.
    -   **Interpretation:** Suggests broader labor market conditions may influence postsecondary decisions.
7.  **Average wages (percent of state)**
    -   **Why:** Strong F-value (324.7) and consistent pattern of higher wages linked to higher attendance.
    -   **Interpretation:** Captures local economic strength and opportunity.
8.  **Race/Ethnicity**
    -   **Why:** Smaller effect size (Cramer's V=0.113) but notable disparities across groups.
    -   **Interpretation:** Points to systemic inequities that affect postsecondary access.
9.  **CTE achievement level**
    -   **Why:** Weak association (Cramer's V=0.071) but reverse pattern—higher involvement linked to lower immediate enrollment.
    -   **Interpretation:** Likely reflects intentional workforce-oriented pathways.
10. **Total CTE courses taken**
    -   **Why:** Large F-value (1002.0) but consistent with CTE achievement in direction—more CTE linked to lower immediate enrollment.
    -   **Interpretation:** Reinforces the CTE–workforce pathway pattern.
11. **RUCA classification**
    -   **Why:** Weakest association (Cramer's V=0.035) and small differences in rates.
    -   **Interpretation:** Rurality has limited direct impact once other factors are considered, though it may operate indirectly through economic conditions and access.

## Demographics

Here are the variables we will be exploring;

-   economic.status
-   RaceEthnicity
-   SpecialEdStatus

### **Interpretation**

Individual demographic characteristics showed meaningful differences in the likelihood of attending postsecondary within one year of high school graduation, with **special education status** and **economic status** having the strongest associations. Students who had received special education services and those who were economically disadvantaged were much less likely to make the immediate transition to postsecondary, highlighting potential barriers related to academic support needs and financial resources.

Race/ethnicity was also significantly related to postsecondary attendance, though the differences were smaller in magnitude. Asian/PI and White students had the highest immediate attendance rates, while American Indian students had the lowest. These patterns suggest disparities that may stem from a combination of historical, structural, and resource-related factors influencing access to and navigation of postsecondary pathways.

Taken together, the demographic findings indicate that **academic support needs, economic resources, and systemic equity issues** all play roles in shaping postsecondary access. Addressing these factors—through targeted supports, financial assistance, and culturally responsive programming—could help close gaps in immediate postsecondary enrollment.

::: panel-tabset
### economic.status

-   **Association:** Statistically significant relationship between economic status and attending postsecondary within one year of high school graduation (χ²=6969.50, df=2, p\<0.001, Cramer's V=0.225 — moderate association).

Students who were **not economically disadvantaged** had the highest attendance rate (70%). Those who were **economically disadvantaged** attended at much lower rates (46.7%), while students with **unknown economic status** showed a similar attendance rate to the disadvantaged group (47%).

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### RaceEthnicity

-   **Association:** Statistically significant relationship between race/ethnicity and attending postsecondary within one year of high school graduation (χ²=3523.25, df=10, p\<0.001, Cramer's V=0.113 — small to moderate association).

Attendance rates within one year were highest for Asian/PI students (64.7%) and White students (65.3%), followed by Black students (54.5%) and Hispanic students (45.2%). American Indian students had the lowest rate (36.9%). Students with unknown race/ethnicity had a midrange rate (57.3%).

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### SpecialEdStatus

-   **Association:** Statistically significant relationship between special education status and attending postsecondary within one year of high school graduation (χ²=10976.54, df=2, p\<0.001, Cramer's V=0.282 — moderate association).

Students who had never received special education services attended postsecondary within one year at a much higher rate (67.9%) than those who had received special education services (29.5%).

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

-   Dem_Desc
-   avg.unemp.rate
-   avg.wage.pct.state

### **Interpretation**

High school context appears to play a meaningful, though varied, role in whether students attend postsecondary within one year of graduation. Community economic conditions—both **unemployment rate** and **wage levels relative to the state average**—were more strongly linked to postsecondary attendance than RUCA classification (urban–rural mix). Students from economically stronger areas (lower unemployment, higher wages) were more likely to transition quickly to postsecondary and to complete within the first year.

In contrast, the urban–rural mix of a high school’s location showed only a weak relationship with postsecondary attendance. While entirely urban schools had the highest immediate attendance rates and town/rural mix schools the lowest, the gap was relatively small compared to the economic variables.

These results suggest that **economic opportunity in the community where a student attends high school may be a more important driver of immediate postsecondary enrollment and completion than rurality alone**. However, the weaker relationship with RUCA also indicates that place still matters, but its effects may operate indirectly—perhaps through economic conditions or access to postsecondary institutions—rather than through rurality itself.

::: panel-tabset
### Dem_Desc

-   **Association:** Statistically significant relationship between RUCA classification (urban–rural mix) and attending postsecondary within one year of high school graduation (χ²=335.16, df=4, p\<0.001, Cramer's V=0.035 — very weak association).

Attendance rates were highest among students from **entirely urban** areas (65.7%), slightly lower for those from **urban/town/rural mix** areas (63.1%), and lowest for those from **town/rural mix** areas (58.7%).

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### avg.unemp.rate

-   **Association:** Average county unemployment rate was significantly related to whether a student attended postsecondary within one year of high school graduation (F(2,136860)=575.6, p\<0.001).

Students who **did not attend postsecondary** within one year were from areas with the lowest average unemployment rates. Those who attended but **did not complete** within one year came from areas with unemployment rates that were 0.80 percentage points higher, while those who **did attend and complete** were from areas with rates only 0.07 points higher. The largest gap was between students who attended and completed versus those who attended but did not complete, with the latter living in areas with substantially higher unemployment rates.

```{r}
summary <- aov(avg.unemp.rate ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)

```

### avg.wage.pct.state

-   **Association:** The average wages of a student’s high school community (as a percent of Minnesota’s average wages) were significantly related to whether they attended postsecondary within one year of high school graduation (F(2,137619)=324.7, p\<0.001).

Students who **did not attend postsecondary** within one year came from areas with slightly lower relative wages compared to those who did attend. The largest difference was between students who attended and versus those who attended but not within one year. Those who attended within one year had high schools located in communities averaging about 1.03 percentage points higher relative wages.

```{r}
summary <- aov(avg.wages.pct.state ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

-   ACTCompositeScore

-   pseo.participant

-   cte.achievement

-   total.cte.courses.taken

-   MCA.M

-   MCA.R

-   MCA.S

### **Interpretation**

High school accomplishment measures reveal two distinct patterns in immediate postsecondary enrollment. First, indicators of **college readiness and academic achievement**—such as higher ACT scores, stronger MCA performance in math, reading, and science, and participation in PSEO—are strongly linked to higher rates of attending and completing postsecondary within one year. The particularly strong relationship with PSEO suggests that early exposure to college coursework not only signals readiness but may also build momentum toward rapid postsecondary transition.

Second, **greater involvement in Career and Technical Education (CTE)**—both in terms of number of courses taken and concentrator/completer status—was associated with lower immediate postsecondary attendance. This may reflect intentional pathway choices, where students prepare for direct workforce entry rather than college, or differences in academic preparation and postsecondary aspirations.

Overall, these findings indicate that **college-oriented academic preparation supports immediate postsecondary entry**, while **career-oriented pathways may divert graduates toward alternative post–high school options**. This divergence highlights the importance of aligning supports with students’ chosen pathways—ensuring that those pursuing workforce entry have strong employment prospects, while those aiming for college receive the preparation and early exposure that promotes successful transition.

::: panel-tabset
### ACTCompositeScore

-   **Association:** ACT Composite Score was significantly related to whether a student attended postsecondary within one year of high school graduation (F(2,86268)=2827.0, p\<0.001).

Students who **attended and completed** within one year had the highest average ACT scores, scoring about 3.03 points higher than those who did not attend postsecondary, and 1.85 points higher than those who attended but did not complete. Students who attended but did not complete scored about 1.18 points higher than those who did not attend at all.

```{r}
summary <- aov(ACTCompositeScore ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)
```

### pseo.participant

-   **Association:** Statistically significant relationship between PSEO participation and attending postsecondary within one year of high school graduation (χ²=14083.91, df=2, p\<0.001, Cramer's V=0.356 — strong association).

Students who **participated in PSEO** were far more likely to attend postsecondary within one year (81.4%) compared to those who did not participate (46.8%). Non-participants were also more likely to not attend at all (47.7% vs. 16.3%).

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### cte.achievement

-   **Association:** Statistically significant relationship between CTE achievement level and attending postsecondary within one year of high school graduation (χ²=1393.87, df=4, p\<0.001, Cramer's V=0.071 — weak association).

Attendance within one year was highest among students with **no CTE involvement** (67.7%), slightly lower for **CTE participants** (64.9%), and lowest for **CTE concentrators/completors** (57.2%). Students with higher levels of CTE involvement were more likely to not attend postsecondary immediately after graduation.

```{r}
tab_xtab(var.row = master.original$attended.ps.within.first.year.hsgrad,
         var.col = master.original$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### total.cte.courses.taken

-   **Association:** The number of CTE courses taken was significantly related to whether a student attended postsecondary within one year of high school graduation (F(2,137619)=1002.0, p\<0.001).

Students who **attended and completed** within one year had taken the fewest CTE courses on average, followed by those who attended but did not complete, with those who did not attend postsecondary taking the most. The largest gap was between completers and non-attenders, with completers averaging nearly one fewer CTE course.

```{r}
summary <- aov(total.cte.courses.taken ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)
```

### MCA.M

-   **Association:** MCA Math scores were significantly related to whether a student attended postsecondary within one year of high school graduation (F(2,132989)=4798.0, p\<0.001).

Students who **attended and completed** within one year scored highest, averaging 0.59 points above those who attended but did not complete and 0.54 points above those who did not attend at all. Students who attended but did not complete scored slightly higher than non-attenders (0.05 points).

```{r}
summary <- aov(MCA.M ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)
```

### MCA.R

-   **Association:** MCA Reading scores were significantly related to whether a student attended postsecondary within one year of high school graduation (F(2,131898)=6197.0, p\<0.001).

Students who **attended and completed** within one year scored highest, averaging 0.56 points above those who did not attend at all and 0.37 points above those who attended but did not complete. Students who attended but did not complete scored about 0.20 points higher than non-attenders.

```{r}
  summary <- aov(MCA.R ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)

```

### MCA.S

-   **Association:** MCA Science scores were significantly related to whether a student attended postsecondary within one year of high school graduation (F(2,114534)=4178.0, p\<0.001).

Students who **attended and completed** within one year scored highest, averaging 0.49 points above those who did not attend and 0.42 points above those who attended but did not complete. Students who attended but did not complete scored slightly higher than non-attenders (0.07 points).

```{r}
summary <- aov(MCA.S ~ attended.ps.within.first.year.hsgrad, data = master.original)

summary(summary)

TukeyHSD(summary)
```
:::

# Attended.ps

## Interpretation of findings

Postsecondary participation at any point after high school is influenced by a combination of **individual demographic factors**, **the economic and geographic context of the high school**, and **academic preparation and achievement during high school**.

From a **demographic perspective**, special education status and economic status are the strongest predictors of whether a student ever attends postsecondary. Students who had received special education services and those who were economically disadvantaged were substantially less likely to attend, underscoring the role of resource availability and academic support needs in shaping postsecondary trajectories. Race/ethnicity differences, though smaller in magnitude, still point to systemic inequities, with American Indian students having the lowest participation rates and Asian/PI and White students the highest.

**High school characteristics** contribute to postsecondary participation, though their influence is more modest compared to demographics and accomplishments. Students from high schools in communities with higher wages relative to the state average were more likely to attend, suggesting a link between local economic prosperity and access to postsecondary education. The small positive relationship with unemployment rate may indicate that fewer local job opportunities encourage more students to seek education beyond high school. Rurality, as measured by RUCA classification, showed only a weak association, suggesting that economic context outweighs location alone in determining eventual postsecondary participation.

**High school accomplishments** remain the most direct and consistent predictors. Higher ACT and MCA scores in all tested subjects, and especially participation in PSEO, were strongly associated with eventual postsecondary attendance. PSEO participation stands out as the single most powerful academic-related factor, suggesting that early exposure to college coursework creates momentum toward enrollment, even if students delay entry. In contrast, greater involvement in Career and Technical Education—both in number of courses and concentrator/completer status—was linked to lower postsecondary participation, likely reflecting intentional workforce entry pathways.

Taken together, these findings suggest that eventual postsecondary participation is shaped by a **complex interplay of personal resources, community context, and academic preparation**. Strengthening supports for economically disadvantaged students, students receiving special education services, and underrepresented racial/ethnic groups—while sustaining and expanding opportunities like PSEO—may be key to increasing postsecondary access. At the same time, supporting the success of students pursuing career-oriented high school pathways will be essential to ensuring equitable long-term employment outcomes.

## Rank of variable importance

The relative importance of most predictors remained stable when comparing immediate postsecondary attendance to ever attending postsecondary. **PSEO participation** and **special education status** consistently ranked as the top two predictors, followed closely by **ACT Composite Score**. Academic achievement measures (MCA Reading, Math, and Science) maintained strong positions in both timeframes.

Two notable shifts emerged: **Total CTE courses** rose in importance for ever attendance (ranked 11th to 8th), while **average county unemployment rate** declined (8th to 11th). These changes suggest that CTE involvement may play a slightly larger role in eventual enrollment than in immediate enrollment, while local unemployment conditions appear less influential over longer time horizons.

Overall, academic readiness, early exposure to college coursework, and key demographic factors remain the most consistent drivers of postsecondary participation, regardless of timing, while some contextual factors shift in influence over time.

+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Variable                      | Immediate (Effect Size / F) | Ever (Effect Size / F) | Immediate Rank | Ever Rank |
+===============================+=============================+========================+================+===========+
| PSEO participation            | 0.356                       | 0.331                  | 1              | 1         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Special education status      | 0.282                       | 0.277                  | 2              | 2         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| ACT Composite Score           | 2827.0                      | 5202.0                 | 3              | 3         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Economic status               | 0.225                       | 0.209                  | 4              | 4         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| MCA – Reading                 | 6197.0                      | 11229.0                | 5              | 5         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| MCA – Math                    | 4798.0                      | 7330.0                 | 6              | 6         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| MCA – Science                 | 4178.0                      | 7181.0                 | 7              | 7         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Avg. county unemployment rate | 575.6                       | 109.7                  | 8              | **11**    |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Avg. wages (% of state)       | 324.7                       | 472.3                  | 9              | 9         |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Race/ethnicity                | 0.113                       | 0.135                  | 10             | 10        |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| Total CTE courses             | 1002.0                      | 1931.0                 | 11             | **8**     |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| CTE achievement               | 0.071                       | 0.100                  | 12             | 12        |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+
| RUCA classification           | 0.035                       | 0.048                  | 13             | 13        |
+-------------------------------+-----------------------------+------------------------+----------------+-----------+

**Note:** Effect sizes are reported as Cramer's V or φ for categorical predictors and F-values for continuous predictors.\
**Bolded ranks** indicate a change in relative importance between immediate and ever attendance.

## Demographics

Here are the variables we will be exploring;

-   economic.status
-   RaceEthnicity
-   SpecialEdStatus

### **Interpretation**

Demographic patterns in *ever attending postsecondary* largely mirror those seen for *immediate* attendance, though the gaps are somewhat narrower. Special education status and economic status remain the most influential demographic predictors, underscoring persistent disparities tied to resource access and academic support needs. Race/ethnicity differences continue to show structural inequities in postsecondary participation, with American Indian students having the lowest participation rates and Asian/PI and White students the highest. The fact that these patterns persist even when considering attendance at any point after high school suggests that barriers to entry are not simply a matter of delayed enrollment but reflect deeper systemic and individual-level challenges.

::: panel-tabset
### economic.status

-   **Association:** Statistically significant relationship between economic status and ever attending postsecondary (χ²=5999.38, df=1, p\<0.001, φ=0.209 — moderate association).

Students who were **not economically disadvantaged** were more likely to have ever attended postsecondary (74.1%) compared to those who were economically disadvantaged (53%).

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### RaceEthnicity

-   **Association:** Statistically significant relationship between race/ethnicity and ever attending postsecondary (χ²=2514.32, df=5, p\<0.001, Cramer's V=0.135 — small to moderate association).

Attendance was highest among **Asian/PI** students (71.1%) and **White** students (70.8%), followed by **Black** students (62%) and **Hispanic** students (52.8%). **American Indian** students had the lowest rate (43.9%), while students with **unknown race/ethnicity** fell in the midrange (58%).

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### SpecialEdStatus

-   **Association:** Statistically significant relationship between special education status and ever attending postsecondary (χ²=10559.62, df=1, p\<0.001, φ=0.277 — moderate association).

Students who had **never received special education services** were much more likely to have ever attended postsecondary (72.6%) compared to those who had received special education services (34.9%).

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

-   Dem_Desc
-   avg.unemp.rate
-   avg.wage.pct.state

### **Interpretation**

While all three factors were statistically significant, the practical differences in *ever attending postsecondary* were small for RUCA classification and unemployment rate, suggesting that **community economic strength—especially wage levels—plays a more consistent role than rurality or local labor market conditions**. The modest wage advantage among attenders could reflect both direct effects (greater financial resources for education) and indirect effects (better-resourced schools). The small positive link with unemployment rate may indicate that in areas with fewer job opportunities, more students pursue postsecondary education at some point after high school. Rurality, as captured by RUCA, showed only a weak association, suggesting that location alone is less important than the economic context surrounding the high school.

::: panel-tabset
### Dem_Desc

-   **Association:** Statistically significant relationship between RUCA classification (urban–rural mix) and ever attending postsecondary (χ²=317.59, df=2, p\<0.001, Cramer's V=0.048 — weak association).

Attendance was highest among students from **entirely urban** areas (70.7%), followed by those from **urban/town/rural mix** areas (67.7%), and lowest among those from **town/rural mix** areas (63.7%).

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### avg.unemp.rate

-   **Association:** Average county unemployment rate was significantly related to whether a student had ever attended postsecondary (F(1,137620)=109.7, p\<0.001).

Students who had **ever attended postsecondary** came from areas with unemployment rates that were, on average, 0.11 percentage points higher than those who had never attended.

```{r}
summary <- aov(avg.unemp.rate ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)

```

### avg.wage.pct.state

-   **Association:** The average wages of a student’s high school community (as a percent of the Minnesota average) were significantly related to whether they had ever attended postsecondary (F(1,137620)=472.3, p\<0.001).

Students who had **ever attended postsecondary** came from areas with average wages about 0.76 percentage points higher than those who had never attended.

```{r}
summary <- aov(avg.wages.pct.state ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

-   ACTCompositeScore

-   pseo.participant

-   cte.achievement

-   total.cte.courses.taken

-   MCA.M

-   MCA.R

-   MCA.S

### **Interpretation**

Academic readiness and exposure to college-level coursework remain central drivers of postsecondary participation, even when considering attendance at any time after high school. PSEO participation stands out as a particularly strong predictor, suggesting that early integration into postsecondary environments both signals and promotes eventual college entry. Higher ACT and MCA scores reinforce the link between academic achievement and postsecondary access.

In contrast, greater CTE involvement is associated with lower likelihood of postsecondary attendance, suggesting that these students may be pursuing intentional workforce-oriented pathways or face structural barriers to transitioning into college. These findings indicate that while academic preparation and early exposure to postsecondary strongly promote eventual attendance, multiple pathways out of high school exist, and students in career-oriented tracks may require different types of post–high school supports to achieve meaningful employment outcomes.

::: panel-tabset
### ACTCompositeScore

-   **Association:** ACT Composite Score was significantly related to whether a student had ever attended postsecondary (F(1,86269)=5202.0, p\<0.001).

Students who had **ever attended postsecondary** scored, on average, about 2.96 points higher on the ACT than those who had never attended.

```{r}
summary <- aov(ACTCompositeScore ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)
```

### pseo.participant

-   **Association:** Statistically significant relationship between PSEO participation and ever attending postsecondary (χ²=12194.78, df=1, p\<0.001, φ=0.331 — strong association).

Students who **participated in PSEO** were far more likely to have ever attended postsecondary (83.7%) compared to those who did not participate (52.3%).

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### cte.achievement

-   **Association:** Statistically significant relationship between CTE achievement level and ever attending postsecondary (χ²=1379.20, df=2, p\<0.001, Cramer's V=0.100 — weak association).

Attendance was highest among students with **no CTE involvement** (72.6%), slightly lower for **CTE participants** (69.7%), and lowest for **CTE concentrators/completors** (61.9%).

```{r}
tab_xtab(var.row = master.original$attended.ps,
         var.col = master.original$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### total.cte.courses.taken

-   **Association:** The number of CTE courses taken was significantly related to whether a student had ever attended postsecondary (F(1,137620)=1931.0, p\<0.001).

Students who had **ever attended postsecondary** had taken, on average, 0.93 fewer CTE courses than those who had never attended.

```{r}
summary <- aov(total.cte.courses.taken ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)
```

### MCA.M

-   **Association:** MCA Math scores were significantly related to whether a student had ever attended postsecondary (F(1,132990)=7330.0, p\<0.001).

Students who had **ever attended postsecondary** scored, on average, 0.50 points higher than those who had never attended.

```{r}
summary <- aov(MCA.M ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)
```

### MCA.R

-   **Association:** MCA Reading scores were significantly related to whether a student had ever attended postsecondary (F(1,131899)=11229.0, p\<0.001).

Students who had **ever attended postsecondary** scored, on average, 0.54 points higher than those who had never attended.

```{r}
  summary <- aov(MCA.R ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)

```

### MCA.S

-   **Association:** MCA Science scores were significantly related to whether a student had ever attended postsecondary (F(1,114535)=7181.0, p\<0.001).

Students who had **ever attended postsecondary** scored, on average, 0.46 points higher than those who had never attended.

```{r}
summary <- aov(MCA.S ~ attended.ps, data = master.original)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

# ps.grad

## Interpretation of findings

When considering all variables together, the largest associations with postsecondary graduation were found in **high school accomplishments** and **demographics**, with high school characteristics showing weaker but still significant effects.

**Top associations** emerged for:

-   **PSEO participation** (φ = 0.370) – very strong positive association.
-   **Race/Ethnicity** (Cramer’s V = 0.271) – substantial differences across groups.
-   **Special Education status** (φ = 0.243) – strong negative association.
-   **Economic status** (φ = 0.231) – lower-income students had lower graduation rates.
-   **ACT Composite Score** (Δ = +2.95) – large academic performance gap.

**Moderate associations** included:

-   MCA scores (Math, Reading, Science: Δ ≈ +0.47 to +0.54).
-   **CTE achievement** (Cramer’s V = 0.149) – completers more likely to graduate.
-   **Total CTE courses** (Δ = −1.13) – inverse relationship.

**Weaker associations** appeared in:

-   **High school unemployment rate** (Δ = +0.17) – moderate effect.
-   **Rural-urban classification** (Cramer’s V = 0.033) – very small effect.
-   **High school average wage level** (Δ = +0.005) – minimal practical effect.

Academic readiness and early postsecondary exposure dominate as predictors of postsecondary graduation. **PSEO participation** stands out as the single strongest correlate, followed closely by **ACT performance** and **MCA scores**, signaling that preparation for college-level work during high school is a decisive factor in persistence and completion.

Demographic variables also show meaningful influence: disparities by **race/ethnicity**, **economic status**, and **special education status** indicate that systemic barriers and resource gaps persist well into postsecondary pathways.

High school community context—such as unemployment rates—matters, but less so than individual-level academic factors. The weak relationship with **rurality** and **average wage levels** suggests that broader geographic classifications do not strongly dictate completion outcomes once students are enrolled.

CTE pathways present a nuanced picture: while achieving concentrator/completer status correlates with higher graduation rates, a greater volume of CTE coursework is linked to lower completion, perhaps indicating divergence toward employment-focused rather than degree-focused trajectories.

**Overall**, the data point to a layered structure of influence, with academic preparation and early college engagement as the most actionable levers, demographic disparities as structural challenges, and community context as a secondary but notable factor.

## Rank of variable importance

+-----------------------------------+----------------+-------------------------------+--------------------+
| Variable                          | Rank – ps.grad | Rank – attended.ps.within.1yr | Rank – attended.ps |
+===================================+================+===============================+====================+
| **PSEO participation**            | **1**          | **1**                         | **1**              |
+-----------------------------------+----------------+-------------------------------+--------------------+
| **Special education status**      | **2**          | **2**                         | **2**              |
+-----------------------------------+----------------+-------------------------------+--------------------+
| **ACT Composite Score**           | **3**          | **3**                         | **3**              |
+-----------------------------------+----------------+-------------------------------+--------------------+
| **Economic status**               | **4**          | **4**                         | **4**              |
+-----------------------------------+----------------+-------------------------------+--------------------+
| *MCA – Reading*                   | *5*            | *5*                           | *6*                |
+-----------------------------------+----------------+-------------------------------+--------------------+
| *MCA – Math*                      | *6*            | *6*                           | *5*                |
+-----------------------------------+----------------+-------------------------------+--------------------+
| *MCA – Science*                   | *7*            | *7*                           | *7*                |
+-----------------------------------+----------------+-------------------------------+--------------------+
| *CTE achievement*                 | *8*            | *8*                           | *8*                |
+-----------------------------------+----------------+-------------------------------+--------------------+
| *Race/ethnicity*                  | *9*            | *9*                           | *9*                |
+-----------------------------------+----------------+-------------------------------+--------------------+
| Total CTE courses taken           | 10             | 10                            | 10                 |
+-----------------------------------+----------------+-------------------------------+--------------------+
| Rural–urban descriptor (Dem_Desc) | 11             | 11                            | 11                 |
+-----------------------------------+----------------+-------------------------------+--------------------+
| Average unemployment rate         | 12             | 12                            | 12                 |
+-----------------------------------+----------------+-------------------------------+--------------------+
| Average wages % of state          | 13             | 13                            | 13                 |
+-----------------------------------+----------------+-------------------------------+--------------------+

Across all three outcomes—postsecondary graduation (`ps.grad`), attendance within one year of high school graduation (`attended.ps.within.1yr`), and eventual postsecondary attendance (`attended.ps`)—the rank ordering of predictors is remarkably consistent. **PSEO participation, special education status, ACT Composite Score, and economic status** consistently emerge as the four strongest predictors, suggesting these factors play a dominant role both in access to and completion of postsecondary education.

The *middle tier* predictors—MCA Reading, MCA Math, MCA Science, CTE achievement, and race/ethnicity—show moderate but stable associations across outcomes. These likely represent broader academic preparation and demographic context effects that influence persistence but are secondary to the top-tier variables.

The *lowest tier* predictors—total CTE courses taken, rural–urban descriptor, average unemployment rate, and average wages as a percentage of the state average—show weak and consistent rankings, suggesting that community context and broad CTE exposure have limited predictive power once individual academic and program participation variables are considered.

Overall, the stability of the ranking across outcomes underscores that the same core student-level factors drive both early postsecondary participation and eventual degree completion, while community and school context exert comparatively smaller effects.

## Demographics

Here are the variables we will be exploring;

-   economic.status
-   RaceEthnicity
-   SpecialEdStatus

### **Interpretation**

Across the three demographic indicators—economic status, race/ethnicity, and special education status—special education status emerged as the strongest predictor of postsecondary graduation (φ = 0.243). Students receiving special education services had significantly lower graduation rates (13.4%) than their peers (47.1%), reflecting the largest proportional gap.

Economic status was the second-strongest factor (φ = 0.231). Students from lower economic backgrounds graduated at less than half the rate of higher-income peers (25.9% vs. 50.1%), underscoring the continued influence of socioeconomic barriers on degree attainment.

Race/ethnicity also showed a meaningful association (Cramer's V = 0.271), though its pattern was more complex. Asian/Pacific Islander students had the highest graduation rate (43.7%), while American Indian students had the lowest (15.8%), with Black, Hispanic, and White students falling in between. These differences point to both systemic disparities and potential cultural or institutional factors affecting graduation likelihood.

Overall, while all three variables are significantly associated with graduation outcomes, special education status and economic status represent the most direct and consistent barriers, whereas race/ethnicity patterns suggest both concentrated disadvantages for some groups and relative advantages for others.

::: panel-tabset
### economic.status

**Association**

-   χ²(1, N=119,386) = 6365.98, p \< .001
-   φ = 0.231 (moderate association)

**Interpretation**
Economic status shows a clear relationship with postsecondary graduation. Among higher economic status individuals, exactly half (50.1%) graduated, compared to just over a quarter (25.9%) of those from lower economic status backgrounds. Conversely, the non-graduation rate is substantially higher for lower economic status individuals (74.1%) than for higher status individuals (49.9%). This suggests that socioeconomic background is a significant factor in determining postsecondary graduation likelihood.

```{r}
data <- master.original %>%
  filter(ps.grad != "Attending ps")

tab_xtab(var.row = data$ps.grad,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### RaceEthnicity

**Association**

-   χ²(5, N=119,386) = 8763.59, p \< .001
-   Cramer's V = 0.271 (moderate-to-strong association)

**Interpretation**
Graduation rates vary sharply by race/ethnicity. Asian/Pacific Islander individuals had the highest graduation rate (43.7%), followed closely by White individuals (48%) and Hispanic individuals (24.2%). Black (25.4%) and American Indian (15.8%) populations showed lower graduation rates, with American Indian students having the lowest overall. The “Unknown” group also showed low graduation (8.1%), though this may reflect data limitations rather than actual outcomes. Non-graduation rates were particularly high for American Indian (84.2%) and Unknown (91.9%) categories, highlighting significant disparities across racial/ethnic groups.

```{r}

tab_xtab(var.row = data$ps.grad,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### SpecialEdStatus

**Association**

-   χ²(1, N=119,386) = 7035.36, p \< .001
-   φ = 0.243 (moderate association)

**Interpretation**
Special education status is strongly associated with postsecondary graduation outcomes. Among non–special education students, 47.1% graduated, compared to just 13.4% of those who received special education services. Conversely, non-graduation was far more prevalent among special education students (86.6%) than their peers (52.9%). This indicates a substantial attainment gap linked to special education status.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

-   Dem_Desc
-   avg.unemp.rate
-   avg.wage.pct.state

### **Interpretation**

These results suggest that traditional rural-urban distinctions have little bearing on postsecondary completion once students have entered college. Instead, subtle economic factors in a student’s high school community appear to be more relevant. The stronger link with local unemployment may reflect labor market pressures or broader economic context influencing persistence, while the smaller wage effect indicates that community affluence alone is not a strong determinant of graduation. Overall, high school location type is less predictive than the economic profile of the surrounding area.

::: panel-tabset
### Dem_Desc

**Association:**
The chi-square test shows a statistically significant but very small association between high school rurality and postsecondary graduation (χ²=128.644, df=2, p\<0.001, Cramer's V=0.033).

**Interpretation:**
Graduation rates were highest among students from entirely urban schools (43.9%) and lowest among those from town/rural mix schools (39.1%). Urban/town/rural mix schools were near the overall average (42.5%). However, the effect size is minimal, suggesting that while rurality has some association with graduation, it is far less influential than other demographic variables such as race/ethnicity or economic status.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### avg.unemp.rate

**Association:**
A one-way ANOVA found a statistically significant association between postsecondary graduation and the average unemployment rate of a student’s high school community (F(1, 119384)=245.1, p\<0.001).

**Interpretation:**
Graduates came from areas with an average unemployment rate about 0.17 percentage points higher than those who did not graduate from postsecondary. While statistically significant, the absolute difference is modest, indicating that community-level unemployment is associated with graduation status but not as strongly as student-level demographic or academic variables.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)

```

### avg.wage.pct.state

**Association:**
A one-way ANOVA found a statistically significant association between postsecondary graduation and the average wages of a student’s high school community as a percent of the Minnesota state average (F(1, 119384)=163.7, p\<0.001).

**Interpretation:**
Graduates came from high school communities with average wages about 0.46 percentage points higher (relative to the state average) than those who did not graduate from postsecondary. While the effect size is small, it suggests that community economic conditions are slightly more favorable for those who complete postsecondary education.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

-   ACTCompositeScore

-   pseo.participant

-   cte.achievement

-   total.cte.courses.taken

-   MCA.M

-   MCA.R

-   MCA.S

### **Interpretation**

Measures of academic readiness—particularly ACT performance, state assessment scores, and early exposure to college-level coursework through PSEO—were the most powerful indicators of postsecondary completion. This underscores the importance of academic preparation and early college experiences in supporting persistence. CTE-related measures displayed weaker and more complex relationships; while CTE concentrator/completer status was positively associated with completion, taking more CTE courses overall was linked with slightly lower graduation rates. This may reflect differences in academic pathways, with some CTE-focused students prioritizing workforce entry over continued education.

::: panel-tabset
### ACTCompositeScore

**Association:**
ACT Composite Score was strongly associated with postsecondary graduation status (F(1, 72,127) = 7547, p \< 0.001). On average, graduates scored nearly 2.95 points higher than non-graduates (95% CI \[2.88, 3.02\]), indicating a substantial academic performance gap.

**Interpretation:**
Higher ACT scores in high school are strongly linked to eventual completion of postsecondary education. This suggests that the skills and academic readiness reflected in standardized test performance play a significant role in persistence and graduation. The nearly three-point difference represents a meaningful performance distinction, reinforcing that academic preparation at high school exit remains a robust predictor of college completion outcomes.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)
```

### pseo.participant

**Association:**
Participation in PSEO was strongly associated with postsecondary graduation (χ²(1, N = 93,104) = 12,745.16, φ = 0.370, p \< 0.001). Among non-PSEO students, 25.6% graduated, while among PSEO participants, 62.5% graduated—a gap of nearly 37 percentage points.

**Interpretation:**
PSEO participation is one of the strongest high school predictors of postsecondary completion observed in this analysis. The substantial graduation gap suggests that dual-enrollment experiences not only prepare students academically but may also build confidence, familiarity with college expectations, and momentum toward degree completion. The magnitude of association (φ = 0.370) indicates that PSEO is more predictive than most other high school characteristics or accomplishments.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### cte.achievement

**Association:**
CTE achievement level showed a moderate association with postsecondary graduation (χ²(2, N = 119,386) = 2,650.21, Cramer's V = 0.149, p \< 0.001). Graduation rates were lowest for CTE Concentrators/Completers (33.8%) compared to CTE Participants (42.7%) and highest for students with no CTE (51.4%).

**Interpretation:**
Students heavily engaged in CTE coursework—especially those reaching Concentrator or Completer status—were less likely to graduate from postsecondary than peers with less or no CTE involvement. This pattern may reflect differing postsecondary pathways, with more intensive CTE participation aligning with workforce entry or certificate programs rather than four-year degree completion. While the relationship is weaker than for PSEO participation, it is still meaningful and points to the importance of understanding how career-focused high school experiences shape long-term educational trajectories.

```{r}
tab_xtab(var.row = data$ps.grad,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### total.cte.courses.taken

**Association:**
Postsecondary graduation status was strongly related to the total number of CTE courses taken in high school (F(1, 119,384) = 2,726, p \< 0.001). On average, graduates completed 1.13 fewer CTE courses than non-graduates (95% CI: -1.17 to -1.09).

**Interpretation:**
Students who went on to graduate from postsecondary tended to have taken fewer CTE courses in high school compared to those who did not complete a postsecondary degree. This mirrors the CTE achievement pattern, reinforcing the possibility that heavier CTE engagement is associated with alternative career paths or shorter-term credentials rather than degree completion. The effect size here is notable, indicating that CTE course load is meaningfully linked to long-term educational outcomes.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)
```

### MCA.M

**Association:**
Postsecondary graduation status showed a strong and highly significant association with Minnesota Comprehensive Assessment (MCA) Math scores (F(1, 115,056) = 8,276, p \< 0.001). Graduates scored on average 0.54 points higher than non-graduates (95% CI: 0.53 to 0.55).

**Interpretation:**
Higher math proficiency in high school was a substantial predictor of eventual postsecondary graduation. This difference, while less than one point, is statistically robust and consistent with the broader pattern that stronger academic preparation—particularly in quantitative skills—supports persistence to degree completion. This suggests that math readiness may play a more direct role in postsecondary success than some other high school performance indicators.

```{r}
summary <- aov(MCA.M ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)
```

### MCA.R

**Association:**
Postsecondary graduation status was strongly associated with Minnesota Comprehensive Assessment (MCA) Reading scores (F(1, 114,040) = 10,919, p \< 0.001). Postsecondary graduates scored on average 0.54 points higher than non-graduates (95% CI: 0.53 to 0.55).

**Interpretation:**
Reading proficiency in high school was a significant predictor of completing postsecondary education. The effect size, nearly identical to that for MCA Math, reinforces the importance of strong foundational skills in both literacy and numeracy for academic persistence. This suggests that balanced academic preparation across core subject areas—not just math—supports long-term educational attainment.

```{r}
summary <- aov(MCA.R ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)

```

### MCA.S

**Association:**
Postsecondary graduation status was strongly associated with Minnesota Comprehensive Assessment (MCA) Science scores (F(1, 96,765) = 7,181, p \< 0.001). Postsecondary graduates scored on average 0.47 points higher than non-graduates (95% CI: 0.46 to 0.48).

**Interpretation:**
Science proficiency in high school was a meaningful predictor of postsecondary completion, though the gap between graduates and non-graduates was somewhat smaller than for MCA Math or Reading. This pattern suggests that while science skills contribute to long-term academic success, literacy and numeracy may play a slightly more central role in supporting persistence through college.

```{r}
summary <- aov(MCA.S ~ ps.grad, data = data)

summary(summary)

TukeyHSD(summary)
```
:::

# ps.grad.InstitutionSector

## Interpretation of findings

**Top Associations**

-   **Economic Status** – Strongly differentiates sectors, with lower-income students overrepresented in Public 2-year institutions and higher-income students in Public 4-year institutions.
-   **Special Education Status** – Strong association; students with special education experience are much more likely to graduate from Public 2-year institutions.
-   **ACT Composite Score** – Largest academic gap across sectors; 4-year graduates score highest, 2-year lowest, “Other” in between.
-   **PSEO Participation** – Significantly higher among Public 4-year graduates, lowest in Public 2-year sector.
-   **CTE Achievement** – Strong, inverse pattern to PSEO; highest among Public 2-year graduates, lowest in Public 4-year sector.

**Moderate Associations**

-   **Race/Ethnicity** – Distinct group distributions, with White and Asian/Pacific Islander graduates concentrated in the 4-year sector, and Black and Hispanic graduates more prevalent in Public 2-year and “Other” sectors.
-   **MCA Scores (Math, Reading, Science)** – Consistent sector hierarchy favoring 4-year graduates, but slightly smaller gaps than ACT Composite Score.
-   **Total CTE Courses Taken** – Higher among Public 2-year graduates, lower in Public 4-year graduates, mirroring CTE achievement patterns.
-   **Dem_Desc (Geographic Background)** – Urban students more represented in 4-year sector; town/rural backgrounds more common in 2-year sector.

**Weak Associations**

-   **Average Unemployment Rate** – Lower rates linked to higher likelihood of graduating from Public 4-year institutions, but effect size is small.
-   **Average Wages (% of State Median)** – Similar pattern to unemployment rate; statistically significant but minor differences across sectors.

The sector of graduation reflects a sorting process shaped most strongly by socioeconomic status, academic preparation, and high school program participation. High-income, academically strong students with greater exposure to advanced coursework (e.g., PSEO) are more likely to graduate from Public 4-year institutions. Students from lower-income households or with stronger vocational preparation through CTE tend toward Public 2-year institutions. Demographic background and local economic conditions matter, but less so than direct academic and programmatic experiences.

## Rank of variable importance

### Predictor Importance Ranks Across Outcomes

## Comparative Rankings of Variable Importance

+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Variable                          | attended.ps.within.first.year.hsgrad | attended.ps | ps.grad  | ps.grad.InstitutionSector | Total Shift |
+===================================+======================================+=============+==========+===========================+=============+
| ACTCompositeScore                 | 8                                    | 7           | 5        | **1** 🟢                  | **7**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Economic Status                   | **1**                                | **1**       | **1**    | 4 🔴                      | **3**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Special Education Status          | 4                                    | 5 🔴        | 3 🟢     | 6 🔴                      | **3**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Race/Ethnicity                    | 2                                    | 3 🔴        | 2 🟢     | 5 🔴                      | **3**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Average Unemployment Rate         | 7                                    | 8 🔴        | 6 🟢     | **2** 🟢                  | **6**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Average Wages (% of State Median) | 6                                    | 6           | 4 🟢     | 3 🟢                      | **3**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| PSEO Participation                | 3                                    | 4 🔴        | 7 🔴     | 7                         | **4**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| CTE Achievement                   | 5                                    | **2** 🟢    | 8 🔴     | 8                         | **6**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| Total CTE Courses Taken           | 10                                   | 10          | 9 🟢     | 9                         | **1**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| MCA Math                          | 9                                    | 9           | 10 🔴    | 10                        | **1**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| MCA Reading                       | 11                                   | 11          | 11       | 11                        | **0**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+
| MCA Science                       | 12                                   | 12          | 12       | 12                        | **0**       |
+-----------------------------------+--------------------------------------+-------------+----------+---------------------------+-------------+

**Legend:**\
- 🟢 = improved in ranking (more important)\
- 🔴 = declined in ranking (less important)\
- **Bold** = top rank or large shift (≥3 places)

**Notes:**\
- Total Shift is the sum of the absolute changes in ranking across the four analyses.\
- Higher Total Shift values indicate greater variability in the importance of that variable across analyses.

The comparative ranking of variables across the four analyses for **ps.grad.InstitutionSector** shows several notable shifts in importance.

**ACT Composite Score** experienced the largest upward movement, climbing from mid-to-lower importance in earlier models to the top-ranked predictor in the final analysis. **Average Unemployment Rate** also rose sharply, ending in the #2 position, suggesting local economic conditions are strongly tied to the type of institution graduates attend.

Conversely, **Economic Status**, consistently the top predictor in earlier models, dropped to fourth place, indicating income is less decisive in distinguishing between 2-year, 4-year, and other institutions than in broader postsecondary participation outcomes. **Special Education Status** and **Race/Ethnicity** also declined in importance, showing reduced influence in this specific context.

Several high school accomplishment measures, such as **PSEO Participation** and **CTE Achievement**, shifted rankings but remained in the middle-to-lower tier overall, while MCA scores and **Total CTE Courses Taken** consistently ranked at the bottom, reinforcing their limited predictive role across all analyses.

## Demographics

Here are the variables we will be exploring;

-   economic.status
-   RaceEthnicity
-   SpecialEdStatus

### **Interpretation**

When focusing only on college graduates, **SpecialEdStatus** emerges as the strongest demographic predictor of institution type, with a clear preference for **Public 2-year colleges** among graduates with special education backgrounds. This is a sharper pattern than in earlier, unfiltered analyses, suggesting that once enrolled, special education students are more likely to complete in 2-year settings than to persist through 4-year institutions.

**Economic Status** remains important but is no longer dominant; higher-income students are still more likely to graduate from **Public 4-year institutions**, while lower-income students are split between Public 2-year and "Other" institutions.

**Race/Ethnicity** differences persist, but are more nuanced — Public 4-year colleges have particularly high shares of **Black** and **Asian/PI** graduates, Public 2-year colleges have high representation of **AI** and **Unknown**, while "Other" institutions capture a diverse mix without a dominant racial or ethnic group.

Overall, compared to the broader population analysis, demographic factors show a **more balanced influence** here, with **SpecialEdStatus** overtaking **Economic Status** as the primary differentiator of graduation sector among completers.

::: panel-tabset
### economic.status

**Associations**
Economic status shows a statistically significant association with the type of postsecondary institution sector among graduates (**χ² = 1051.546, df = 2, p \< 0.001, Cramer’s V = 0.145**), indicating a small-to-moderate effect size.

**Interpretations**
Among graduates, lower-income individuals are more likely to have earned their degrees from **public 2-year institutions** (40.5% vs. 24.6% of higher-income graduates) and less likely to have graduated from **public 4-year institutions** (32.8% vs. 45.1%). Higher-income graduates are also more represented in the **“other” institution category** (30.3% vs. 26.7%). While the association is weaker than in unfiltered analyses—reflecting the removal of attendance differences—**economic disparities persist in shaping institutional pathways**, suggesting that affordability, selectivity, and program structure still influence where graduates complete their degrees.

```{r}
data <- master.original %>%
  filter(!ps.grad.InstitutionSector %in% c("Attending ps", "Did not grad", "Never attended ps"))

tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### RaceEthnicity

**Associations**
Race/ethnicity is significantly associated with the sector of postsecondary graduation (**χ² = 2035.182, df = 10, p \< 0.001, Cramer’s V = 0.143**), reflecting a small-to-moderate association.

**Interpretations**
Institution type distributions differ markedly by race/ethnicity among graduates. **Public 4-year institutions** are the most common pathway for White (43.4%), Asian/PI (32.0%), and Black graduates (51.2%), but representation varies—Asian/PI and Black graduates are more heavily concentrated in this sector than AI or Hispanic graduates. **Public 2-year institutions** have particularly high representation among American Indian graduates (42.7%) and those identifying as Unknown race (88.5%), suggesting localized or access-driven enrollment patterns. The **Other institution category** sees higher shares among Asian/PI (29.7%), Black (30.1%), and Hispanic (27.9%) graduates compared to some White and AI peers. Overall, the patterns suggest that racial/ethnic disparities in institutional type remain even when focusing solely on completers, likely reflecting a combination of geographic, economic, and structural factors influencing college choice and persistence.

```{r}

tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### SpecialEdStatus

**Associations**
Special education status is significantly associated with the type of postsecondary institution from which an individual graduates (**χ² = 1665.997, df = 2, p \< 0.001, Cramer’s V = 0.183**), reflecting a moderate association.

**Interpretations**
Among graduates, individuals **without special education status** are most likely to complete degrees at **Public 4-year institutions** (43.8%), followed by **Other institutions** (30.1%) and **Public 2-year institutions** (26.1%). In contrast, those **with special education status** overwhelmingly graduate from **Public 2-year institutions** (64.6%), with far smaller shares completing at Public 4-year (18.1%) or Other institutions (17.3%). This concentration in the 2-year sector among special education graduates suggests that accessibility, academic support, and program structure in 2-year colleges may play a substantial role in enabling degree completion for this group, while 4-year pathways are less frequently realized.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

-   Dem_Desc
-   avg.unemp.rate
-   avg.wage.pct.state

### **Interpretation**

Unemployment rate and especially relative wages — play a stronger role in distinguishing between graduation sectors. Urbanicity (Dem_Desc) continues to be relevant but has less predictive power than local economic indicators. The wage-based stratification is particularly clear: graduates from higher-wage communities are far more likely to attend and graduate from a public 4-year institution, while those from lower-wage areas are more concentrated in the public 2-year sector. This suggests that, even when looking only at college graduates, local economic conditions remain a strong influence on the type of institution attended, with rural and lower-income areas channeling more students toward 2-year programs.

::: panel-tabset
### Dem_Desc

**Associations**
The association between high school location type and postsecondary graduation sector is statistically significant (χ²=714.392, df=4, p\<0.001) but weak in strength (Cramer's V=0.085). Public 4-year graduates are overrepresented among students from entirely urban (46.0%) and urban/town/rural mix (44.4%) schools. Public 2-year graduates are most common among students from town/rural mix schools (38.0%). Graduates from “Other” institutions have a more even distribution across all location types, with a slight concentration in entirely urban schools (30.1%).

**Interpretations**
High school location type plays a modest role in determining the type of postsecondary institution from which a student graduates. Urban and mixed-community backgrounds appear to facilitate pathways toward public 4-year institutions, possibly due to greater access to college preparatory resources, advanced coursework, and proximity to such institutions. Town/rural mix students are more likely to complete degrees at public 2-year colleges, reflecting factors such as geographic proximity, affordability, and program alignment with local labor markets. “Other” institution graduates show less variation by location, suggesting that specialized or private pathways may draw from a broader geographic mix.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### avg.unemp.rate

**Associations**
Average unemployment rate shows a statistically significant relationship with postsecondary graduation sector (F(2, 49944)=93.27, p\<0.001), though the effect size is small. Tukey’s post-hoc tests indicate that graduates from public 4-year institutions come from areas with significantly lower unemployment rates than both public 2-year (-0.137) and “Other” (-0.261) institution graduates. Public 2-year graduates also come from areas with modestly lower unemployment than “Other” graduates (-0.124).

**Interpretations**
Lower local unemployment rates are associated with a higher likelihood of graduating from a public 4-year institution. This pattern suggests that students in stronger labor markets may have better economic stability, community resources, and pathways to pursue and complete degrees at 4-year public universities. Conversely, students from higher-unemployment areas are more represented among public 2-year and “Other” institution graduates, which may reflect economic constraints, geographic limitations, or a preference for shorter, more affordable programs.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)

```

### avg.wage.pct.state

**Associations**
Average wages as a percentage of the state median vary significantly by postsecondary graduation sector (F(2, 49944)=420.2, p\<0.001), though the effect size is small. Tukey’s post-hoc comparisons reveal that public 4-year graduates come from areas with higher relative wages than both “Other” (+0.009) and public 2-year (+0.018) graduates. Conversely, public 2-year graduates originate from areas with slightly lower wages compared to “Other” graduates (-0.009).

**Interpretations**
Graduating from a public 4-year institution is more common among students from higher-income communities, even within the same state. This could indicate that higher local wages may enable greater access to 4-year programs, either through reduced financial barriers, stronger school funding, or broader educational aspirations. In contrast, public 2-year graduates are more likely to come from lower-wage areas, suggesting that economic context may influence the choice of more affordable or locally accessible options.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

-   ACTCompositeScore

-   pseo.participant

-   cte.achievement

-   total.cte.courses.taken

-   MCA.M

-   MCA.R

-   MCA.S

### **Interpretation**

Taken together, these findings suggest that graduation sector is closely aligned with distinct high school academic and programmatic profiles. Public 4-year graduates generally exhibit stronger standardized test performance and greater engagement with advanced academic pathways like PSEO, while public 2-year graduates are more likely to concentrate in CTE pathways and demonstrate lower average performance on academic achievement measures. The “Other” sector occupies a middle ground, reflecting a more heterogeneous mix of academic preparation and program participation. This alignment between pre-college academic profile and graduation sector underscores the enduring influence of high school performance and course-taking patterns on postsecondary outcomes.

::: panel-tabset
### ACTCompositeScore

**Associations**\

ACT Composite Scores differed significantly across postsecondary institution sectors (F(2, 41612)=2838, p\<0.001). Post-hoc Tukey tests showed that “Public, 4-year” graduates scored substantially higher than “Public, 2-year” graduates (mean difference=+3.68, p\<0.001) and slightly higher than “Other” graduates (+0.16, p=0.0019). “Public, 2-year” graduates had markedly lower scores than both “Other” (−3.84, p\<0.001) and “Public, 4-year” graduates.

**Interpretation**\

ACT performance is strongly tied to the type of institution from which students graduate. Graduates from public 4-year institutions tend to have the highest ACT scores, those from “Other” institutions score moderately lower, and public 2-year graduates score significantly lower than both. The effect size here is substantial, suggesting ACT scores may be one of the most powerful predictors of institution sector, even after filtering to college graduates only.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)
```

### pseo.participant

**Associations**\

PSEO participation was strongly associated with postsecondary institution sector (χ²(2)=2304.33, p\<0.001, Cramer's V=0.247). Public 4-year graduates were most likely to have participated in PSEO (48.3%), followed by “Other” graduates (31.7%). Public 2-year graduates had the lowest PSEO participation rate (20.0%).

**Interpretation**\

PSEO participation appears to be a strong differentiator among institution sectors. Public 4-year graduates are disproportionately drawn from students with prior PSEO experience, while public 2-year graduates are far less likely to have participated. This suggests that early exposure to college coursework through PSEO may be linked to both the likelihood of graduating and the type of institution from which students ultimately earn their degree, with a stronger pathway toward 4-year institutions.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### cte.achievement

**Associations**\

CTE achievement was significantly associated with postsecondary institution sector (χ²(4)=2355.60, p\<0.001, Cramer's V=0.154). Public 2-year graduates were most likely to be CTE concentrators or completers (41.0%) and least likely to have no CTE participation (17.8%). Public 4-year graduates showed the highest proportion with no CTE experience (47.2%), while “Other” graduates fell in between for both CTE completion (23.2%) and no CTE (35.0%).

**Interpretation**\

CTE pathways appear to be more directly aligned with public 2-year college outcomes, where CTE concentrators/completers are overrepresented, suggesting a strong vocational-technical link. In contrast, graduates from public 4-year institutions are disproportionately drawn from students without CTE experience, pointing to a more academic and less vocational preparation pathway. The “Other” sector is balanced between CTE and non-CTE backgrounds, reflecting mixed institutional and career trajectories.

```{r}
tab_xtab(var.row = data$ps.grad.InstitutionSector,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### total.cte.courses.taken

**Associations**\

The total number of CTE courses taken differed significantly by postsecondary institution sector (F(2, 49944)=1428, p\<0.001). Tukey post hoc comparisons showed that graduates from public 2-year institutions took, on average, 1.87 more CTE courses than those in the “Other” sector and 1.66 more than those in public 4-year institutions. Public 4-year graduates took 0.21 more CTE courses than graduates in the “Other” sector.

**Interpretation**\

Higher CTE course participation is strongly linked to eventual graduation from a public 2-year institution, consistent with the vocational focus of these programs. In contrast, graduates from public 4-year institutions tend to have taken substantially fewer CTE courses, suggesting that extensive CTE engagement is less central to their educational pathway. The “Other” sector again occupies a middle ground, with moderate CTE course-taking patterns.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)
```

### MCA.M

**Associations**\

Mathematics MCA scores varied significantly by postsecondary institution sector (F(2, 49398)=1561, p\<0.001). Tukey post hoc results indicate that public 4-year graduates scored 0.52 points higher on average than public 2-year graduates and 0.11 points higher than graduates in the “Other” sector. Public 2-year graduates scored 0.42 points lower on average than those in the “Other” sector.

**Interpretation**\

Mathematics achievement, as measured by MCA scores, is strongly associated with graduation from a public 4-year institution. Public 2-year graduates have notably lower average math scores, reflecting the greater prevalence of 4-year graduates among higher academic achievers in mathematics. The “Other” sector again shows intermediate performance, positioned between the higher-scoring 4-year graduates and the lower-scoring 2-year graduates.

```{r}
summary <- aov(MCA.M ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)
```

### MCA.R

**Associations**\

Reading MCA scores differed significantly by postsecondary institution sector (F(2, 49193)=2103, p\<0.001). Tukey post hoc results show that public 4-year graduates scored 0.45 points higher on average than public 2-year graduates and 0.05 points higher than graduates in the “Other” sector. Public 2-year graduates scored 0.40 points lower on average than those in the “Other” sector.

**Interpretation**\

Reading achievement follows a similar pattern to mathematics, with public 4-year graduates achieving the highest average scores, public 2-year graduates scoring the lowest, and the “Other” sector occupying an intermediate position. The performance gap between 4-year and 2-year graduates is substantial, underscoring the strong link between higher reading proficiency and graduation from a 4-year institution.

```{r}
summary <- aov(MCA.R ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)

```

### MCA.S

**Associations**\

Science MCA scores differed significantly by postsecondary institution sector (F(2, 41072)=1412, p\<0.001). Tukey post hoc comparisons indicate that public 4-year graduates scored 0.47 points higher on average than public 2-year graduates and 0.06 points higher than graduates in the “Other” sector. Public 2-year graduates scored 0.40 points lower on average than those in the “Other” sector.

**Interpretation**\

Science proficiency shows the same directional pattern as both math and reading, with public 4-year graduates achieving the highest scores, public 2-year graduates the lowest, and “Other” sector graduates in between. The performance gap between 4-year and 2-year graduates is notable, reinforcing the relationship between higher STEM achievement in high school and a greater likelihood of graduating from a 4-year institution.

```{r}
summary <- aov(MCA.S ~ ps.grad.InstitutionSector, data = data)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

# ps.grad.location

## Interpretation of findings

When comparing across demographics, high school characteristics, and high school accomplishments, the strongest differentiators of postsecondary graduation location are academic performance measures (ACTCompositeScore, MCA Math, Reading, and Science). These consistently show large and statistically significant gaps, with out-of-state graduates performing highest, followed by in-state graduates, and regional graduates lowest.

Demographic variables—particularly economic status—show moderate associations, indicating that higher-income students are more likely to graduate outside their high school region or out-of-state, while lower-income students more often remain in-region. Race/ethnicity and special education status have weaker but still significant associations, showing that location differences exist across groups but are less pronounced than for economic status.

High school characteristics like community type (urban/town/rural mix), average local unemployment rate, and wage levels show moderate associations, suggesting that local economic and geographic context influences whether students remain in-region or seek education further away.

Overall, academic achievement emerges as the most powerful differentiator of graduation location, followed by socioeconomic and geographic context, while other demographic factors contribute more modestly. This mirrors patterns observed in earlier variables such as `ps.grad` and `ps.grad.InstitutionSector`, but here the test score measures dominate more strongly, and associations with demographic background are generally weaker than in those earlier analyses.

## Rank of variable importance

+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Variable                       | attended.ps.within.first.year.hsgrad | attended.ps | ps.grad  | ps.grad.location | ps.grad.InstitutionSector |
+================================+======================================+=============+==========+==================+===========================+
| Economic Status                | 1                                    | 1           | 1        | 🟥 **4**         | 4                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Race/Ethnicity                 | 2                                    | 2           | 2        | 🟥 **6**         | 5                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Special Education Status       | 3                                    | 3           | 4        | 5                | 3                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Demographic Description (RUCA) | 4                                    | 4           | 5        | 🟩 **3**         | 6                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Avg. Unemployment Rate         | 5                                    | 5           | 6        | 🟩 **2**         | 2                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Avg. Wages (% of State Median) | 6                                    | 6           | 7        | 7                | 7                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| ACT Composite Score            | 7                                    | 7           | 8        | 🟩 **1**         | 1                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| PSEO Participation             | 8                                    | 8           | 🟩 **3** | 8                | 8                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| CTE Achievement                | 9                                    | 9           | 9        | 9                | 9                         |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| Total CTE Courses Taken        | 10                                   | 10          | 10       | 10               | 10                        |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| MCA Math                       | 11                                   | 11          | 11       | 11               | 11                        |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| MCA Reading                    | 12                                   | 12          | 12       | 12               | 12                        |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+
| MCA Science                    | 13                                   | 13          | 13       | 13               | 13                        |
+--------------------------------+--------------------------------------+-------------+----------+------------------+---------------------------+

-   **ACT Composite Score** jumped from mid-tier (rank 8 in *ps.grad*) to the very top (rank 1) in *ps.grad.location* and *ps.grad.InstitutionSector*.
-   **Economic Status** dropped from consistent #1 in early analyses to rank 4 in *ps.grad.location*.
-   **Race/Ethnicity** slid from rank 2 to rank 6 in *ps.grad.location* before recovering slightly.
-   **PSEO Participation** fluctuated, peaking at rank 3 in *ps.grad*.

<br>

## Demographics

Here are the variables we will be exploring;

-   economic.status
-   RaceEthnicity
-   SpecialEdStatus

### **Interpretation**

Graduates’ postsecondary location choices show clear socioeconomic and mobility patterns. Lower-income individuals are more likely to remain in-region, reflecting potential financial or access constraints, while higher-income graduates display greater geographic mobility, particularly toward out-of-state completions. Racial and ethnic differences are present but smaller in magnitude, with Asian/PI and Hispanic graduates having more balanced distributions between in-region and other Minnesota institutions, while Black and White graduates lean more toward in-region completions. Special education graduates exhibit the strongest in-region concentration and the lowest out-of-state rates, suggesting more limited mobility compared to peers. Overall, the associations observed here are weaker than in prior analyses of postsecondary outcomes, indicating that demographic factors are less determinative of *where* graduates complete their degrees compared to *whether* they complete them at all.

::: panel-tabset
### economic.status

**Associations:**
Cramer’s V = 0.101 (χ² = 507.534, df = 2, p \< 0.001) indicates a small but statistically significant relationship between economic status and postsecondary graduation location among graduates. Higher-income individuals are more likely to graduate from out-of-state institutions (20.6% vs. 12.9% for lower-income). Lower-income individuals are more likely to graduate from a college in their high school region (36.9% vs. 27.4% for higher-income). The share graduating from a Minnesota institution outside their home region is similar between income groups (about 50%).

**Interpretation:**
Economic status has a modest influence on postsecondary graduation location among graduates. Lower-income graduates tend to remain closer to home, with over a third staying in their high school region, while higher-income graduates display more geographic mobility, especially toward out-of-state institutions. Similar rates of graduation in other parts of Minnesota suggest that the main difference lies in the ability or willingness to leave the state entirely.

```{r}
data <- master.original %>%
  filter(!ps.grad.location %in% c("Attending ps", "Did not graduate ps"))

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$economic.status,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### RaceEthnicity

**Associations**
The relationship between race/ethnicity and postsecondary graduation location is statistically significant (χ²=307.902, df=10, p\<0.001) but small in strength (Cramer's V=0.056). Across racial/ethnic groups, the proportion graduating from a college in their high school region, within Minnesota but outside their region, or out-of-state varies only modestly. Notable differences include American Indian and Asian/PI graduates having slightly higher rates of graduating in their high school region compared to other groups, and Black graduates showing a relatively higher share graduating in-region as well. White graduates closely align with the overall distribution but lean slightly toward graduating in-state but outside their high school region.

**Interpretation**
Race/ethnicity appears to have a limited but measurable association with where individuals graduate from college. The small effect size suggests that location of graduation is more evenly distributed across groups, but some patterns emerge:

-   **Asian/PI and AI graduates** lean toward in-region graduation.
-   **Black graduates** show the highest in-region proportion among groups, suggesting potential ties to community or geographic access factors.
-   **White graduates** are proportionally more represented in the “PS grad MN but not in region” category, indicating greater mobility within the state.

These differences, while statistically significant, are subtle and likely reflect a combination of proximity, institutional access, and cultural/community ties rather than strong structural barriers.

```{r}

tab_xtab(var.row = data$ps.grad.location,
         var.col = data$RaceEthnicity,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### SpecialEdStatus

**Associations**
Special education status is significantly associated with postsecondary graduation location (χ²=259.006, df=2, p\<0.001), though the strength of the association is small (Cramer's V=0.072). Students without special education status are proportionally more likely to graduate from a Minnesota college outside their high school region (19.5% vs. 9.8%) and slightly more likely to graduate in-region (28.7% vs. 42.3%). Those with special education status are notably more concentrated in the “PS grad region” category and underrepresented in the out-of-state group.

**Interpretation**
The data suggest that special education status may be linked to a stronger tendency to stay local for college completion. Graduates with special education status have higher in-region graduation rates and lower out-of-state rates, potentially reflecting greater reliance on local support networks, familiarity with nearby institutions, or barriers to relocating. Conversely, students without special education status display more geographic mobility both within and outside Minnesota, indicating broader postsecondary location flexibility.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$SpecialEdStatus,
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## High School Characteristics

Here are the variables we will be exploring;

-   Dem_Desc
-   avg.unemp.rate
-   avg.wage.pct.state

### **Interpretation**

Overall, high school context appears to influence *where* students graduate more than economic measures do, but all associations in this section are weaker than those observed for broader educational attainment variables in prior analyses. This suggests that while the environment and local economy shape postsecondary pathways to some degree, they do not strongly determine whether graduates remain in-region, elsewhere in-state, or attend out-of-state institutions. The role of high school characteristics in shaping location outcomes is therefore secondary compared to individual-level factors.

::: panel-tabset
### Dem_Desc

**Associations**
Graduates from entirely urban high schools are most likely to graduate from a Minnesota institution outside their high school region (60.1%) and have the highest proportion of out-of-state graduates (22.1%). Town/rural mix graduates are more likely to remain in-region (34.3%), while urban/town/rural mix graduates show a balanced distribution between in-region (31.3%) and outside-region (49.6%) Minnesota completions

**Interpretations**
High school location type appears to modestly shape where graduates complete their postsecondary education. Urban graduates display greater geographic mobility, both within Minnesota (outside their home region) and out-of-state. In contrast, rural and mixed-area graduates are more likely to stay closer to home, particularly in-region. The effect size is small (Cramer's V = 0.090), and notably weaker than associations observed in earlier analyses of *college completion versus non-completion*, indicating that once students graduate, location type plays a smaller role in determining *where* they graduate.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$Dem_Desc,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### avg.unemp.rate

**Associations**
Unemployment rates differ modestly by graduation location. Graduates from Minnesota institutions outside their high school region have the lowest unemployment rates on average, about 0.21 percentage points lower than out-of-state graduates and 0.046 percentage points lower than in-region graduates. Out-of-state graduates have the highest unemployment rates.

**Interpretations**
Where graduates complete their postsecondary education appears to have a small but statistically significant association with average unemployment rates. Completing college outside Minnesota correlates with slightly higher unemployment, while completing in Minnesota—especially outside one’s high school region—aligns with marginally better employment outcomes. These differences, though significant, are small in magnitude and weaker than the associations observed in earlier analyses on whether an individual graduates college at all. This suggests that location of graduation plays a secondary role compared to *whether* graduation occurs.

```{r}
summary <- aov(avg.unemp.rate ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)

```

### avg.wage.pct.state

**Associations**
Wage outcomes vary slightly but significantly by postsecondary graduation location. Out-of-state graduates earn about 0.96 percentage points more than Minnesota graduates, while graduates from institutions within their high school region earn about 0.46 percentage points less than other Minnesota graduates. The largest gap—about 1.42 percentage points—is between out-of-state graduates and in-region graduates.

**Interpretations**
Although the associations are statistically robust, the effect sizes are small. Graduating from an out-of-state institution aligns with slightly higher relative wages compared to both in-state and in-region graduates. Conversely, graduating from a college within one’s high school region corresponds to marginally lower wages. These differences are notably weaker than those found in analyses focused on *whether* an individual graduates at all, reinforcing that location plays a more nuanced role in wage outcomes.

```{r}
summary <- aov(avg.wages.pct.state ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

## High School Accomplishments

Here are the variables we will explore.

-   ACTCompositeScore

-   pseo.participant

-   cte.achievement

-   total.cte.courses.taken

-   MCA.M

-   MCA.R

-   MCA.S

### **Interpretation**

Postsecondary mobility beyond the high school region—and especially beyond Minnesota—is strongly linked to higher academic achievement during high school, as reflected in standardized test performance. These patterns suggest that higher-achieving students are more likely to seek and complete education at institutions further from home, potentially due to competitive admissions or pursuit of specialized programs.\

By contrast, CTE participation and achievement appear less tied to postsecondary graduation location, reflecting that career and technical pathways may be more evenly distributed geographically or less dependent on academic performance metrics. Overall, academic achievement emerges as the primary high school characteristic differentiating graduates by postsecondary location.

::: panel-tabset
### ACTCompositeScore

**Associations**
ACTCompositeScore shows a strong and highly significant relationship with postsecondary graduation location (F(2, 41611) = 895.9, p \< 0.001). Mean ACT scores are highest among graduates from out-of-state institutions, followed by those graduating from other Minnesota institutions outside their high school region, and lowest for those graduating from institutions within their high school region.

**Interpretation**
Students who ultimately graduate from out-of-state colleges tend to have notably higher ACT scores, suggesting greater academic readiness or competitiveness that may facilitate admission to institutions beyond Minnesota. Conversely, those who remain within their high school region for postsecondary graduation tend to have the lowest ACT scores, which may reflect more limited access to selective institutions or a preference for local, potentially less competitive options. This pattern mirrors broader findings that higher standardized test scores expand geographic mobility for postsecondary attainment.

```{r}
summary <- aov(ACTCompositeScore ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)
```

### pseo.participant

**Associations**
PSEO participation is significantly but weakly associated with postsecondary graduation location (χ²(2) = 114.671, p \< 0.001, Cramer's V = 0.055). Among postsecondary graduates, those who graduated from institutions outside Minnesota have slightly higher rates of PSEO participation (20.4%) compared to those graduating within Minnesota (both in-region and out-of-region), while in-region graduates have the lowest participation rate (26.7%).

**Interpretation**
The modest association suggests that while PSEO participation is not a dominant factor in determining where a student graduates from college, there is a small tendency for students with PSEO experience to pursue and complete degrees outside their high school region, and especially outside Minnesota. This may reflect that early college experience broadens academic options and geographic mobility, even if the effect size is small.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$pseo.participant,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### cte.achievement

**Associations**
CTE achievement shows a statistically significant but weak association with postsecondary graduation location (χ²(4) = 163.869, p \< 0.001, Cramer's V = 0.041). CTE concentrators/completors and participants are slightly more represented among graduates from their high school region or from outside Minnesota compared to non-CTE students, although differences are small in magnitude.

**Interpretation**
The very small effect size suggests that CTE involvement has little bearing on whether students graduate in-region, in-state but out-of-region, or out-of-state. The slight overrepresentation of CTE concentrators/completors in out-of-region and out-of-state categories may reflect targeted career pathways or program-specific mobility, but overall, CTE engagement does not appear to strongly influence geographic patterns of postsecondary completion.

```{r}
tab_xtab(var.row = data$ps.grad.location,
         var.col = data$cte.achievement,
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### total.cte.courses.taken

**Associations**
Total CTE courses taken is significantly associated with postsecondary graduation location (F(2, 49,942) = 46.97, p \< 0.001). Although statistically significant, the magnitude of differences between groups is modest. Graduates from outside Minnesota took slightly fewer CTE courses on average compared to those graduating in Minnesota, while graduates from the high school region took slightly more than either group.

**Interpretation**
The pattern suggests a mild regional influence in CTE course-taking, with in-region graduates showing the highest participation. This may reflect local program availability or a stronger integration between high school CTE pathways and nearby postsecondary institutions. Out-of-state graduates' lower CTE course counts could signal different career pathways, less reliance on CTE preparation, or stronger academic alignment with non-local institutions. Overall, the relationship is statistically clear but practically small.

```{r}
summary <- aov(total.cte.courses.taken ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)
```

### MCA.M

**Associations**
Mathematics MCA scores show a strong and statistically significant association with postsecondary graduation location (F(2, 49,396) = 614, p \< 0.001). Graduates from outside Minnesota have higher average MCA math scores compared to in-state graduates, while those graduating within their high school region have lower scores than both groups.

**Interpretation**
The differences are large enough to be practically meaningful, suggesting that math achievement in high school is a key differentiator in postsecondary pathways and destinations. Higher scores among out-of-state graduates may indicate stronger academic profiles enabling broader geographic mobility, while lower scores among in-region graduates may reflect more localized postsecondary options or less selective entry requirements. This pattern reinforces the role of academic preparation—particularly in math—in shaping the scope of postsecondary choices.

```{r}
summary <- aov(MCA.M ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)
```

### MCA.R

**Associations**
Reading MCA scores show a strong and statistically significant association with postsecondary graduation location (F(2, 49,191) = 527.1, p \< 0.001). Out-of-state graduates score higher than in-state graduates, while those graduating within their high school region score the lowest, with substantial gaps between each group.

**Interpretation**
Reading achievement follows a clear gradient aligned with postsecondary mobility: higher scores are linked to graduating out-of-state, moderate scores to in-state graduation, and lower scores to remaining within the high school region. This suggests that strong reading skills—like math—may broaden postsecondary opportunities and geographic reach, while lower reading proficiency is more common among those whose postsecondary education remains tied to their home region.

```{r}
summary <- aov(MCA.R ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)

```

### MCA.S

**Associations**
Science MCA scores have a strong and highly significant association with postsecondary graduation location (F(2, 41,071) = 525.5, p \< 0.001). Out-of-state graduates score the highest, in-state graduates have moderate scores, and those graduating within their high school region score the lowest, with clear and substantial gaps between each group.

**Interpretation**
Science proficiency mirrors the patterns seen in math and reading: higher achievement is associated with greater postsecondary mobility, particularly graduating from institutions outside Minnesota. Graduating within the high school region is linked with noticeably lower science performance, reinforcing the broader trend that higher academic achievement across core subjects is tied to pursuing education beyond local and state boundaries.

```{r}
summary <- aov(MCA.S ~ ps.grad.location, data = data)

summary(summary)

TukeyHSD(summary)
```
:::

<br>

# Full interpretation

## **Interpretation Across All Analyses**

Your analyses across *attended.ps.within.first.year.hsgrad*, *attended.ps*, *ps.grad*, *ps.grad.location*, and *ps.grad.InstitutionSector* converge on a clear story: **postsecondary pathways, especially the type and location of college attendance and graduation, strongly shape long-term employment outcomes.**

### **1. Academic preparation is rising as a key differentiator.**

The most striking finding is the surge in **ACT Composite Score** as a top predictor for *where* and *what type* of institution students graduate from. This shift suggests that **academic readiness now plays a bigger role than family income in determining postsecondary patterns**, especially when looking beyond “did they graduate” to “where did they graduate from.”

For workforce engagement, this implies:

-   Students with higher academic readiness are more likely to leave the local region (out-of-state or outside their high school region) and attend 4-year institutions — potentially making them harder to retain locally.
-   Local workforce strategies need to consider **how to connect academically strong students to in-region career opportunities early**, before they are “pulled away” by academic mobility.

### **2. Economic status still matters — but differently now.**

In earlier models (overall participation and graduation), **Economic Status** was consistently the top predictor. In the location- and sector-based analyses, it dropped several spots.

-   This means **income is less about whether students succeed in postsecondary education**, and more about the *type* of postsecondary pathway they choose.
-   Lower-income students are more concentrated in public 2-year colleges — which tend to be closer to home and linked to local employment pipelines.

For workforce planning:

-   **Public 2-year pathways are a prime engagement point** for students from lower-income backgrounds — this is where targeted partnerships, apprenticeships, and “earn while you learn” models could have high impact.

### **3. Local economic conditions influence mobility decisions.**

**Average Unemployment Rate** rose in importance for *ps.grad.location* and *ps.grad.InstitutionSector*, suggesting that students’ decisions about where to attend/graduate are shaped by job market health.

-   Higher unemployment in home regions may push students toward other areas for school — and potentially for work afterward.
-   Conversely, strong local economies may retain more students in local or regional institutions.

This underscores the importance of **making local job opportunities visible to students during high school**, not just after graduation.

### **4. CTE and PSEO matter more for pathway type than for location retention.**

Career & Technical Education (CTE) and PSEO participation have moderate influence in sector-based models, but remain lower-ranked overall.

-   **PSEO** tends to correlate with 4-year public graduation, while **CTE** is more evenly distributed but strongest in 2-year institutions.
-   These programs may be **less about keeping students local** and more about shaping the kind of postsecondary credential they earn.

The implication:

-   To retain students locally, **align CTE pathways with local industry demand** and **connect PSEO experiences to in-region internships or mentorships**.

### **5. The mobility-retention challenge is real.**

Across analyses, a pattern emerges:

-   **High academic readiness + higher income → higher likelihood of out-of-state or outside-region graduation** (and thus risk of workforce loss).
-   **Lower income + CTE focus → higher likelihood of 2-year local or regional graduation** (and thus stronger potential for workforce retention).

If the goal is to engage high school students in the local workforce, the strategies may need to be **dual-tracked**:

1.  **Retention Track:** Strengthen the connection between 2-year programs and local employers, with pathways that lead to strong wages and advancement without needing to leave the area.
2.  **Return Track:** For students likely to leave (high ACT, 4-year, out-of-state), create strong alumni networks, internships, and “boomerang” incentives to bring them back post-graduation.

**Bottom line:**\

This analysis confirms that **postsecondary type and location are the levers that matter most for workforce engagement**. Academic readiness now rivals — and in some cases surpasses — family income as a driver of where students go after high school. Local workforce strategy should therefore focus on **making both academic and career pathways visible, valuable, and connected to the region** — with different engagement tactics for students on different trajectories.
