---
title: "Local Employment Cross-tabs"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.factor.names <- master.original %>%
  select(PersonID, Dem_Desc, edr, grad.year.covid, Gender, economic.status, pseo.participant, SpecialEdStatus, non.english.home, RaceEthnicity, took.ACT, ap.exam, cte.achievement, english.learner, sat.taken, attended.ps, attended.ps.within.first.year.hsgrad, ps.grad, ps.grad.InstitutionSector, highest.cred.level) 

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.states %>%
  left_join(master.factor.names, by = "PersonID")

kable(head(master))

kable(names(master))
```

```{r independent and dependent variables}
updated.factor.var.names <-  master.factor.names %>%
  select(-PersonID) %>%
  names() %>%
  c("ps.grad.in.same.ruca", "ps.grad.in.same.edr", "ps.grad.in.same.pr", "ps.grad.in.MN")

numeric.var.names <- master.original %>%
  select(ACTCompositeScore, total.cte.courses.taken, `ACTE.SPED`:Youth.Apprenticeship,  MCA.M:MCA.S, n.institutions, avg.unemp.rate, wages.3year.avg, avg.wages.pct.state) %>%
  names()
  
updated.independent.var.names <- c(updated.factor.var.names, numeric.var.names)

cte.cc.var.names <- master.original %>%
  select(ACTE.SPED:Youth.Apprenticeship) %>%
  names()

dep.var.names <- master %>%
  select(PersonID, grad.year.1, grad.year.5, grad.year.10)

updated.master <- master.original %>%
  select(PersonID, all_of(updated.independent.var.names)) %>%
  left_join(dep.var.names, by = "PersonID") %>%
  mutate_at(vars(Dem_Desc:ps.grad.in.MN), as.factor) %>%
  mutate_at(vars(grad.year.1:grad.year.10), as.factor) %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Did not grad", "Some college", as.character(highest.cred.level)),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Never attended ps", "No college", as.character(highest.cred.level)),
         highest.cred.level = as.factor(highest.cred.level),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix", "Urban/town/rural mix", "Entirely urban")) %>%
  droplevels()

kable(head(updated.master))

kable(names(updated.master))
```

<br>

There are a number of "states" in which an individual in this dataset are assigned;

1.  Has meaningful employment
    1.  In the same county as their high school.
    2.  In the same EDR as their high school, but not county.
    3.  In Minnesota, but not in Central Minnesota.
2.  Is attending post-secondary.
3.  Has a MN employment record but it's not meaningful, and not attending post-secondary.
4.  Does not have a MN employment record and is not attending post-secondary.

There are three "times" in which we checked to see if the individual had meaningful employment within these geographies;

1.  one year after graduating high school
2.  five years after graduation, and
3.  ten years after graduation.

Meaningful employment is determined by whether an individual worked 1,000 hours for an employer during time X. In addition, if the individual is working for an employer but not meaningful at time X but worked 1,000 hours for that employer during another year it's still considered "meaningful".

Due to time x potentially being passed the date of the latest data (2023) for some individuals, the analysis below will filter out all individuals where time x is after 2023.

From the multiple correspondence analysis and the principle component analysis we determined that the following independent variables will be included in the analysis.

```{r list of ind var}
updated.master %>%
  select(-PersonID, -grad.year.1, -grad.year.5, -grad.year.10) %>%
  lapply(class) 

```

# Methodology

For this analysis we are keeping it simple. For categorical variables I’ll be using the Cramer’s v-square measure of association. The V\^2 value is from 0 to 1 with 0 meaning there’s no association and a value closer to 1 indicates a strong association.

| **Cramér’s V Value** | **Interpretation**               |
|----------------------|----------------------------------|
| 0.00–0.10            | Negligible association           |
| 0.10–0.20            | Weak                             |
| 0.20–0.40            | Moderate                         |
| 0.40–0.60            | Relatively strong                |
| 0.60–0.80            | Strong                           |
| 0.80–1.00            | Very strong (possibly redundant) |

For the variables with continuous values I’ll be using the Wilcoxon two paired groups test. P-value will be used to determine whether we can reject the NULL hypothesis at a .01 level. I’ll also include boxplots to see in which direction there might be influence from the variables.

We are going to split up the analysis into two different datasets. The first dataset will combine 3 of the previous states into one.

-   Has meaningful employment anywhere in Central Minnesota, which combines the following states;
    -   Meaningful emp County
    -   Meaningful emp EDR
    -   Meaningful emp region

The remaining states will stay the same.

-   Has meaningful employment in Minnesota, but not in Central Minnesota
-   Is attending a post-secondary institution.
-   Has a MN employment record but it's not meaningful, and they are not attending post-secondary.
-   Does not have a MN employment record and is not attending post-secondary.

We have quite a few variables remaining, even after MCA analysis. So let's take a look at how we are going to proceed and group all of this together to see how these relationships work.

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner
-   High school characteristics
    -   Dem_Desc
    -   edr
    -   avg.unemp.rate
    -   avg.wage.pct.state
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ACTCompositeScore
    -   ap.exam
    -   sat.taken
    -   MCA scores
        -   MCA.M
        -   MCA.R
        -   MCA.S
    -   cte.achievement
    -   total.cte.courses.taken
-   Post-secondary
    -   attended.ps.within.first.year.hsgrad
    -   attended.ps
    -   ps.grad.
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   Combine post-secondary grad location
        -   ps.grad.in.same.ruca
        -   ps.grad.in.same.edr
        -   ps.grad.in.same.pr
        -   ps.grad.in.MN

<br>

# Meaningful employment in Central Minnesota

```{r prep meaningful employment in Central MN}
meaningful.emp.central <- updated.master %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.1)),
         grad.year.1 = as_factor(grad.year.1),
         grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.5)),
         grad.year.5 = as_factor(grad.year.5),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.10)),
         grad.year.10 = as_factor(grad.year.10),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         highest.cred.level = fct_relevel(highest.cred.level, "No college", "Some college", "Less than Associate Degree", "Associate degree", "Bachelor degree or higher"))

write_csv(meaningful.emp.central, "Data/SLEDS/Masters/After analysis/Master-meaningful-emp-central.csv")
```

Let's first take a look at proportion of states in each of the time.X. The chart below shows that half of the graduates are attending post-secondary (50%) one year after graduating high school. The further away from high school graduation results in a bit of an increase in the percentage of individuals who have meaningful employment in the region and in Minnesota, while about a quarter of graduates do not have a MN employment record and are not going to post-secondary.

```{r proportion of states meaningful emp SW}
prop.states.meaningful.emp.central <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10) %>%
  pivot_longer(names_to = "grad.year", values_to = "states", 1:3) %>%
  group_by(grad.year, states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  filter(states != "After 2023") %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         grad.year = fct_relevel(grad.year, "grad.year.1", "grad.year.5", "grad.year.10"))

prop.states.meaningful.emp.central.plot <- ggplot(prop.states.meaningful.emp.central, aes(grad.year, pct, fill = states, group = states)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nState of individuals: ", states, "\nNumber of people in states: ", comma(n), "\nPercent of people in state: ", percent(pct, accuracy = .1), sep = ""))) +
   labs(x="", y = "", color="", title = "Percent of people in each state")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu"),
                    guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = prop.states.meaningful.emp.central.plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

Let's examine the variables to see if there are any significant differences in the proportion of individuals in these states across the variable categories or values.

<br>

# Demographics

In the demographics section we will examine the relationship between that paths taken by individuals and the following variables;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner

## **Summary**

+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Demographic      | Association Strength   | Highlights                                                                                                                                                                                                   |
+==================+:======================:+==============================================================================================================================================================================================================+
| Gender           | Weak to negligible     | Females start with higher PS enrollment; males steadily overtake in employment. Disconnection grows for both.                                                                                                |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| grad.year.covid  | Weak                   | Post-COVID grads attend post-secondary at a lower rate.                                                                                                                                                      |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| economic.status  | Moderate to negligible | Low-income students are less likely to pursue PS, and consistently more likely to lack MN employment or be disconnected.                                                                                     |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SpecialEdStatus  | Moderate to negligible | SPED students consistently show lower engagement in PS and work, and higher disconnection.                                                                                                                   |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| RaceEthnicity    | Negligible             | White and Asian/PI grads show strongest MN employment; Black, Hispanic, and AI groups show consistently higher disconnection risk.                                                                           |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| non.english.home | Negligible             | Students from non-English-speaking homes have **lower postsecondary enrollment** and are increasingly underrepresented in the MN workforce by year 10, with a high percentage lacking MN employment records. |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| english.learner  | Weak to negligible     | English learners start with significantly lower PS participation and grow increasingly disconnected from MN employment systems over time.                                                                    |
+------------------+------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

<br>

## Gender

There is a statistically significant association between whether an individual qualifies for free or reduced lunch and their employment state, but the association starts weak and becomes negligible by five years after graduating high school.

The most consistent difference is that females attend post-secondary at a significantly higher rate.

<br>

::: panel-tabset
### One year after graduating

There is a statistically significant, weak association between gender and the employment state.

Females;

-   Have a lower rate of meaningful employment in the region.
-   Attend post-secondary at a significantly higher rate.

```{r demographics gender one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Gender) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between gender and the employment state.

Females;

-   Have a lower rate of meaningful employment in the region.
-   Attend post-secondary at a significantly higher rate.

```{r demographics gender five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Gender) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between economic status and the employment state.

Barely any differences here.

```{r demographics gender ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Gender) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## Grad.year.covid

There is a statistically significant, weak association between whether an individual graduated before or after COVID and the employment state. The most striking difference is that post-covid grads were attending post-secondary at a significantly lower rate than individuals who graduated before COVID.

<br>

::: panel-tabset
### One year after graduating

```{r demographics grad.year.covid one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.covid) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$grad.year.covid, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## economic.status

There is a statistically significant association between whether an individual qualifies for free or reduced lunch and their employment state, but the association starts at moderate and becomes negligible by ten years after graduating high school.

The most consistent difference through all the years was that individuals who qualify for free and reduced lunch attend post-secondary at a lower rate, and have a lower rate of meaningful employment in Minnesota.

<br>

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a higher rate of meaningful employment in the region.
-   Have a higher rate of meaningful employment in Minnesota.
-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, economic.status) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, economic.status) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a lower rate of having meaningful employment in Minnesota.
-   Have a higher rate of not having meaningful employment.

```{r demographics economic.status ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, economic.status) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## SpecialEdStatus

There is a statistically significant association between whether an individual is categorized as needing special education and their employment state, but the association starts at moderate and becomes negligible by ten years after graduating high school.

The most consistent difference through all the years was that individuals who are categorized as needing special education attend post-secondary at a SIGFNICANTLY lower rate, and have a higher rate of not having meaningful employment.

Having no MN employment record is significantly higher at first, but evens out by ten years after high school.

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, SpecialEdStatus) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   slightly lower rate of having meaningful employment in Minnesota
-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, SpecialEdStatus) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   SIGNIFICANTLY lower rate of having meaningful employment in Minnesota
-   SIGNFICANTLY higher rate of not having meaningful employment

```{r demographics SpecialEdStatus ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, SpecialEdStatus) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## RaceEthnicity

There is a statistically significant association between race/ethnicity and their employment state, but the association is negligible.

The most consistent difference through all the years were White and Asian/PI have a higher rate of attending post-secondary, and American Indian and Black individuals had a significantly higher rate of not having meaningful employment nor attending post-secondary.

Non-white individuals had an increasing rate of not having a MN employment record the further out from high school graduation.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between whether race/ethnicity and the employment state.

Individuals who are;

-   Hispanic have a higher rate of meaningful employment in the region.
-   White or Asian/PI attended post-secondary at a SIGNIFICANTLY higher rate
-   American Indian, Black or Hispanic had SIGNFICANTLY higher rate of not having meaningful employment
-   American Indian had a SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics RaceEthnicity one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, RaceEthnicity) %>%
  filter(grad.year.1 != "After 2023",
         RaceEthnicity != "Unknown") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between whether race/ethnicity and the employment state.

Individuals who are;

-   White or Asian/PI attended post-secondary at a SIGNIFICANTLY higher rate
-   American Indian, Black or Hispanic had SIGNFICANTLY higher rate of not having meaningful employment
-   American Indian had a SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics RaceEthnicity five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, RaceEthnicity) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether race/ethnicity and the employment state.

Individuals who are;

-   Asian/PI and black individuals had a lower rate of meaningful employment in the region.
-   American Indian had a significantly lower rate of meaningful employment in MN.
-   American Indian and Black had SIGNFICANTLY higher rate of not having meaningful employment
-   Asian/PI, Black, and Hispanic had a significantly higher rate of having no MN employment record.

```{r demographics RaceEthnicity ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, RaceEthnicity) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## non.english.home

There is a statistically significant association between whether an individual had a non-English speaker in their household and their employment state, but the association is negligible.

The most biggest trend was the individuals that were confirmed to have a non-English speaking individual in their household had lower rates of meaningful employment in the region or in Minnesota and SIGNIFCANTLY higher rate of not having a MN employment record over time.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between whether ran individual had a non native English speaking person in their household and the employment state.

Individuals who had a non-English speaking person at home;

-   Had a significantly lower rate of attending post-secondary.
-   Had a significantly higher rate of having no MN employment record.

```{r demographics non.english.home one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, non.english.home) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between whether a graduate had a non-English speaking person in their household and the employment state.

Individuals who had a non-English speaking member at home;

-   Had a significantly higher rate of having no MN employment record.

```{r demographics non.english.home five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, non.english.home) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether a graduate had a non-English speaking person in their household and the employment state.

Individuals who had a non-English speaking member at home;

-   Had a significantly lower rate of meaningful employment in the region
-   Had a significantly lower rate of meaningful employment in Minnesota.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics non.english.home ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, non.english.home) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## english.learner

There is a statistically significant association between whether an individual was categorized as an English learner and their employment state, but the association was weak at first and became negligible over time.

Through all the years, graduates categorized as an English learner had SIGNIFICANTLY lower rates of attending post-secondary and had SIGNFICANTLY higher rates of having no MN employment record.

As time away from high school grows, the rate at which these individuals had meaningful employment in the region or in Minnesota became significantly lower.

::: panel-tabset
### One year after graduating

There is a statistically significant, weak association between whether a graduate was categorized as an English learner and the employment state.

Individuals who were classified as an English learner;

-   Had a SIGNIFICANTLY lower rate of attending post-secondary.
-   Had a significantly higher rate of not having meaningful employment.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics english.learner one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, english.learner) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between whether a graduate was categorized as an English learner and the employment state.

Individuals who were classified as an English learner;

-   Had a lower rate of meaningful employment in Minnesota.
-   Had a SIGNIFICANTLY lower rate of attending post-secondary.
-   Had a significantly higher rate of not having meaningful employment.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics english.learner five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, english.learner) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between whether a graduate was categorized as an English learner and the employment state.

Individuals who were classified as an English learner;

-   Had a lower rate of meaningful employment in the region.
-   Had a SIGNIFICANTLY lower rate of meaningful employment in Minnesota.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics english.learner ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, english.learner) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ChatGPT

This is the summary provided by ChatGPT when asked to analyze and interpret.

### Gender Trends Over Time

+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Outcome                     | 1 Year (F) | 5 Year (F) | 10 Year (F) | 1 Year (M) | 5 Year (M) | 10 Year (M) |
+=============================+============+============+=============+============+============+=============+
| Attending PS                | 55.3%      | 17.8%      | \~0.0%      | 43.7%      | 13.4%      | 0.0%        |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Meaningful Emp in MN        | 6.8%       | 24.2%      | 28.6%       | 8.2%       | 24.9%      | 31.8%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Meaningful Emp in Region    | 10.9%      | 19.9%      | 21.5%       | 15.3%      | 24.1%      | 24.8%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| No MN Emp Record, Not in PS | 7.7%       | 17.1%      | 30.3%       | 11.9%      | 19.1%      | 28.4%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Not Meaningful, Not in PS   | 19.3%      | 20.9%      | 19.6%       | 20.9%      | 18.5%      | 15.1%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+

**Trend**: Females start with higher PS enrollment; males steadily overtake in employment. Disconnection grows for both.

------------------------------------------------------------------------

### Economic Status Trends Over Time

+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Outcome                     | 1 Yr (0) | 5 Yr (0) | 10 Yr (0) | 1 Yr (1) | 5 Yr (1) | 10 Yr (1) |
+=============================+==========+==========+===========+==========+==========+===========+
| Attending PS                | 57.4%    | 17.8%    | \~0.0%    | 32.6%    | 11.1%    | \~0.0%    |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in MN        | 6.7%     | 26.2%    | 32.6%     | 9.2%     | 21.1%    | 25.0%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in Region    | 11.7%    | 21.4%    | 22.8%     | 16.2%    | 23.4%    | 23.8%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| No MN Emp Record, Not in PS | 8.1%     | 17.0%    | 28.7%     | 13.5%    | 20.4%    | 30.7%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Not Meaningful, Not in PS   | 16.1%    | 17.6%    | 15.9%     | 28.5%    | 24.1%    | 20.5%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+

**Trend**: Low-income students are less likely to pursue PS, and consistently more likely to lack MN employment or be disconnected.

------------------------------------------------------------------------

### Special Education Trends Over Time

+-----------------------------+----------------+--------+--------+-------------+--------+--------+
| Outcome                     | 1 Yr (No SPED) | 5 Yr   | 10 Yr  | 1 Yr (SPED) | 5 Yr   | 10 Yr  |
+=============================+================+========+========+=============+========+========+
| Attending PS                | 54.2%          | 17.0%  | \~0.0% | 19.8%       | 6.4%   | 0.0%   |
+-----------------------------+----------------+--------+--------+-------------+--------+--------+
| Meaningful Emp in MN        | 7.3%           | 25.6%  | 31.5%  | 8.6%        | 17.6%  | 21.2%  |
+-----------------------------+----------------+--------+--------+-------------+--------+--------+
| Meaningful Emp in Region    | 12.7%          | 21.7%  | 23.0%  | 15.9%       | 24.1%  | 24.0%  |
+-----------------------------+----------------+--------+--------+-------------+--------+--------+
| No MN Emp Record, Not in PS | 8.1%           | 17.2%  | 29.0%  | 21.0%       | 24.5%  | 31.5%  |
+-----------------------------+----------------+--------+--------+-------------+--------+--------+
| Not Meaningful, Not in PS   | 17.8%          | 18.5%  | 16.5%  | 34.8%       | 27.5%  | 23.3%  |
+-----------------------------+----------------+--------+--------+-------------+--------+--------+

**Trend**: SPED students consistently show lower engagement in PS and work, and higher disconnection.

------------------------------------------------------------------------

### Race/Ethnicity Snapshot

+----------------------+--------------------+---------------+---------------------+-----------------------+
| Race                 | Attending PS (1yr) | Emp MN (10yr) | No MN Record (10yr) | Not Meaningful (10yr) |
+======================+====================+===============+=====================+=======================+
| American Indian (AI) | 26.0%              | 21.6%         | 32.1%               | 27.3%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| Asian/PI             | 52.0%              | 30.2%         | 40.2%               | 15.6%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| Black                | 38.7%              | 25.3%         | 36.0%               | 25.2%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| Hispanic             | 31.2%              | 25.1%         | 40.2%               | 17.5%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| White                | 51.1%              | 30.5%         | 28.8%               | 17.0%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+

**Trend**: White and Asian/PI grads show strongest MN employment; Black, Hispanic, and AI groups show consistently higher disconnection risk.

### Non-English speakers at home

+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Outcome                     | 1 Yr (0) | 5 Yr (0) | 10 Yr (0) | 1 Yr (1) | 5 Yr (1) | 10 Yr (1) |
+=============================+==========+==========+===========+==========+==========+===========+
| Attending PS                | 50.0%    | 15.7%    | \~0.0%    | 36.4%    | 13.9%    | 0.0%      |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in MN        | 7.4%     | 24.7%    | 30.4%     | 10.5%    | 19.4%    | 23.2%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in Region    | 13.1%    | 22.2%    | 23.4%     | 14.0%    | 17.0%    | 13.6%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| No MN Emp Record, Not in PS | 9.5%     | 17.7%    | 28.9%     | 18.6%    | 30.4%    | 47.7%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Not Meaningful, Not in PS   | 20.1%    | 19.7%    | 17.4%     | 20.5%    | 19.3%    | 15.5%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+

**Trend**: Students from non-English-speaking homes have **lower postsecondary enrollment** and are increasingly underrepresented in the MN workforce by year 10, with a high percentage lacking MN employment records.

------------------------------------------------------------------------

### English learners

+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Outcome                     | 1 Yr (0) | 5 Yr (0) | 10 Yr (0) | 1 Yr (1) | 5 Yr (1) | 10 Yr (1) |
+=============================+==========+==========+===========+==========+==========+===========+
| Attending PS                | 51.3%    | 16.1%    | \~0.0%    | 23.6%    | 7.2%     | 0.0%      |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in MN        | 7.4%     | 25.0%    | 30.6%     | 9.5%     | 17.9%    | 18.5%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in Region    | 12.9%    | 22.1%    | 23.3%     | 15.7%    | 21.5%    | 17.6%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| No MN Emp Record, Not in PS | 8.9%     | 17.4%    | 28.9%     | 22.1%    | 29.1%    | 43.1%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Not Meaningful, Not in PS   | 19.4%    | 19.4%    | 17.2%     | 29.1%    | 24.3%    | 20.8%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+

**Trend**: English learners start with significantly lower PS participation and grow increasingly disconnected from MN employment systems over time.

------------------------------------------------------------------------

```{r demogrpahics chatgpt}
demographics.chatgpt <- meaningful.emp.central %>%
  select(PersonID, Gender, grad.year.covid, economic.status, SpecialEdStatus, RaceEthnicity, non.english.home, english.learner, grad.year.1, grad.year.5, grad.year.10)

write_csv(demographics.chatgpt, "Data/ChatGPT/demographics-chatgpt.csv")
```

# High school characteristics

We will take a look at the following variables;

-   Dem_Desc
-   edr
-   avg.unemp.rate
-   avg.wage.pct.state

## Summary

+-----------------------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| High School Characteristics | Association Strength | Highlights                                                                                                                                                                                                                                                   |
+=============================+:====================:+==============================================================================================================================================================================================================================================================+
| Dem_Desc                    | Negligible           | Rural graduates had a higher rate of having meaningful employment in the region while urban graduates had a higher rate of meaningful employment in Minnesota.                                                                                               |
+-----------------------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| EDR                         | Negligible           | EDR 5 had a higher rate of meaningful employment in the region and not having a MN employment record.                                                                                                                                                        |
+-----------------------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| avg.unemp.rate              | Moderate             | Higher unemployment was associated with a slight **increase in postsecondary attendance** and a **decrease in meaningful employment**, particularly within Minnesota. While these effects were small, they remained consistent across all three time points. |
+-----------------------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| avg.wages.pct.state         | Most important       | Individuals from **higher-wage counties** were significantly more likely to have **meaningful employment within Minnesota**, even years after graduation. Conversely, they were less likely to attend post-secondary school immediately after graduation.    |
+-----------------------------+----------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

<br>

## Dem_Desc

There is a statistically significant association between an graduate's RUCA category and their employment state, but the association was negligible.

Through all the years, the more rural a graduate was, the higher the rate of having meaningful employment in the region, while urban individuals had a significantly higher rate of meaningful employment in Minnesota.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a higher rate of having meaningful employment in the region.
-   more urban area had a higher rate of having meaningful employment in MN.
-   more rural areas had a lower rate of attending post-secondary.
-   more rural areas had a higher rate of not having a MN employment record.

```{r hs characteristics Dem_Desc one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Dem_Desc) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a higher rate of having meaningful employment in the region.
-   more urban area had a significantly higher rate of having meaningful employment in MN.

```{r hs characteristics Dem_Desc five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Dem_Desc) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a significantly higher rate of having meaningful employment in the region.
-   more urban area had a significantly higher rate of having meaningful employment in MN.

```{r hs characteristics Dem_Desc ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Dem_Desc) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## EDR

There is a statistically significant association between an graduate's EDR and their employment state, but the association was negligible.

Through all the years, graduates from EDR 5 had an increasingly higher rate of meaningful employment in the region and a significantly higher rate of not having a MN employment rate.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between a graduate's EDR and the employment state.

Graduates from;

-   EDR 5 had a slightly higher rate ofhaving meaningful employment in the region.
-   EDR 5 had a slightly lower rate of attending post-secondary.

```{r hs characteristics edr one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, edr) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between a graduate's EDR and the employment state.

Graduates from;

-   EDR 5 had a higher rate of having meaningful employment in the region.
-   EDR 5 had a lower rate of having meaningful employment in Minnesota.
-   EDR 7E had a higher rate of having meaningful employment in Minnesota
-   EDR 5 had a higher rate of not having a MN employment rate.

```{r hs characteristics edr five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, edr) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between a graduate's EDR and the employment state.

Graduates from;

-   EDR 5 had a higher rate of having meaningful employment in the region.
-   EDR 5 had a significantly lower rate of having meaningful employment in Minnesota.

```{r hs characteristics edr ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, edr) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## avg.unemp.rate

There is a statistically significant association between an the unemployment rate of a school district's county and a graduates employment state.

Through all the years, the main trend was that graduates without meaningful employment or no MN employment record came from schools with significantly higher unemployment rates than graduates that have meaningful employment in the region or Minnesota, or attended post-secondary.

::: panel-tabset
The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with 30%+ higher unemployment rates than graduates that meaningful employment in the region or Minnesota.
-   had no meaningful employment or no MN employment record came from districts with 20%+ higher unemployment rates than graduates attending post-secondary.

```{r hs characteristics avg.unemp.rate one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.unemp.rate) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.unemp.rate, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 20))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)
```

### Five years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with over 10% higher unemployment rates.
-   had no meaningful employment came from districts with 12% higher unemployment rates than graduates that are attending post-secondary.

```{r hs characteristics avg.unemp.rate five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.unemp.rate) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.unemp.rate, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or MN employment record came from districts with 18% higher unemployment rates compared to graduates with meaningful employment in the region or in MN.
-   had no MN employment record came from districts with 11% lower unemployment rates than graduates with no meaningful employment.

```{r hs characteristics avg.unemp.rate ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.unemp.rate) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.unemp.rate, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## avg.wages.pct.state

There is a statistically significant association between an the unemployment rate of a school district's county and a graduates employment state.

Through all the years, the main trend was that graduates without meaningful employment or no MN employment record came from schools with significantly higher unemployment rates than graduates that have meaningful employment in the region or Minnesota, or attended post-secondary.

::: panel-tabset
### One year after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with -1% difference between the percent of state wages of graduates that have meaningful employment in the region or Minnesota.
-   had meaningful employment in MN or were attending post-secondary came from districts with a 1.2% difference in the percent of state wages of graduates who had meaningful employment in the region.

```{r hs characteristics avg.wages.pct.state one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.wages.pct.state) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.wages.pct.state, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with -1% difference between the percent of state wages of graduates that have meaningful employment in the region or Minnesota.
-   had meaningful employment in MN or were attending post-secondary came from districts with a 1.2% difference in the percent of state wages of graduates who had meaningful employment in the region.

```{r hs characteristics avg.wages.pct.state five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.wages.pct.state) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.wages.pct.state, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with -5%+ difference between the percent of state wages of graduates that have meaningful employment in Minnesota.
-   had meaningful employment in MN or were attending post-secondary came from districts with a 1.5% difference in the percent of state wages of graduates who had meaningful employment in the region.

```{r hs characteristics avg.wages.pct.state ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.wages.pct.state) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.wages.pct.state, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

## ChatGPT

This is the summary provided by ChatGPT when asked to analyze and interpret.

### Urbanization Level (`dem_desc`)

Rurality of the high school district showed a small but growing influence on employment outcomes over time. Graduates from **town/rural mix** districts were more likely to have **regional employment** and less likely to attend **postsecondary** one year after graduation. Conversely, those from **entirely urban** districts were more likely to be enrolled in postsecondary or employed within Minnesota. The strength of this association increased slightly by year 10.\

### Economic Development Region (`edr`)

The high school’s economic region had a consistent but weak association with employment outcomes. Students from **EDR 7W (Central)** were more likely to attend postsecondary or secure **MN-based employment**, while those from **EDR 5 (North Central)** were more likely to be employed **regionally** or have **no MN employment record**. These patterns remained stable from year 1 through year 10.\

### County Unemployment Rate (`avg.unemp.rate`)

County-level unemployment rates had a modest impact. Higher unemployment was associated with a slight **increase in postsecondary attendance** and a **decrease in meaningful employment**, particularly within Minnesota. While these effects were small, they remained consistent across all three time points.\

### County Wages as % of State Average (`avg.wages.pct.state`)

This variable was the most predictive across the board. Individuals from **higher-wage counties** were significantly more likely to have **meaningful employment within Minnesota**, even years after graduation. Conversely, they were less likely to attend postsecondary school immediately after graduation. This relationship was both strong and persistent across years 1, 5, and 10.\

### Overall Insights

Among the four variables, **county wage levels** stood out as the most influential predictor of long-term employment outcomes. While **urbanization** and **economic region** showed weak but meaningful patterns, their predictive power was notably lower. **Unemployment rate** had a smaller and more stable effect, influencing early decisions such as whether to work or pursue further education.

```{r high school characteristics chatgpt}
hs.characteristics.chatgpt <- meaningful.emp.central %>%
  select(PersonID, Dem_Desc, edr, avg.unemp.rate, avg.wages.pct.state, grad.year.1, grad.year.5, grad.year.10)

write_csv(hs.characteristics.chatgpt, "Data/ChatGPT/hs-characteristics-chatgpt.csv")
```

<br>

# High School Accomplishments

We are going to analyze the following variables;

-   pseo.participant
-   took.ACT
-   ACTCompositeScore
-   ap.exam
-   sat.taken
-   MCA scores
    -   MCA.M
    -   MCA.R
    -   MCS.S
-   cte.achievement
-   total.cte.courses.taken

**Summary**

+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
| High School Accomplishments | Association strength      | Highlights                                                                                                                                      |
+=============================+===========================+=================================================================================================================================================+
| pseo.participant            | Moderate to negligible    | Graduates that participated in PSEO were significantly more likely to attend post-secondary                                                     |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
| took.ACT                    | Relatively strong to weak | Graduates that took the ACT were significantly more likely to attend post-secondary.                                                            |
|                             |                           |                                                                                                                                                 |
|                             |                           | Graduates that did not take the ACT were significantly more likely to have meaningful employment in the region (not in MN), or be disconnected. |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
| ACTCompositeScore           | Moderate                  | **Post-secondary attendees** scored highest, and **those with no MN employment record** scored lowest.                                          |
|                             |                           |                                                                                                                                                 |
|                             |                           | ACT performance appears to strongly predict **educational continuation and labor market engagement**.                                           |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
| ap.exam                     |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
| MCA.                        |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
|                             |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
|                             |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
|                             |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
|                             |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+
|                             |                           |                                                                                                                                                 |
+-----------------------------+---------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------+

<br>

## pseo.participant

There is a statistically significant association between whether a graduate partilcipates in PSEO and their employment state. But the association started as moderate and ended as negligible.

Through all the years, graduates that participated in PSEO were significantly more likely to attend post-secondary.

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between a whether a graduate has taken PSEO and the employment state.

Graduates that took PSEO courses;

-   had lower rates of meaningful employment in the region
-   had SIGNIFICANTLY higher rates of attending post-secondary
-   had significantly lower rates of not having meaningful employment
-   had significantly lower rates of not having a MN employment record.

```{r hs accomplishments pseo.participant one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, pseo.participant) %>%
  filter(grad.year.1 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between a whether a graduate has taken PSEO and the employment state.

Graduates that took PSEO courses;

-   had higher rates of attending post-secondary

```{r hs accomplishments pseo.participant five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, pseo.participant) %>%
  filter(grad.year.5 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between a whether a graduate has taken PSEO and the employment state.

Graduates that took PSEO courses;

-   slightly lower rates of not having meaningful employmernt.

```{r hs accomplishments pseo.participant ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, pseo.participant) %>%
  filter(grad.year.10 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## took.ACT

There is a statistically significant association between whether a graduate takes the ACT and their employment state. The association started as relatively strong and ended as weak.

Through all the years, graduates that took the ACT were significantly more likely to attend post-secondary. Graduates that did not take the ACT were significantly more likely to have meaningful employment in the region (not in MN), or be disconnected.

::: panel-tabset
### One year after graduating

There is a statistically significant, relatively strong association between a whether a graduate has taken the ACT and the employment state.

Graduates that;

-   did not take the ACT had higher rates of meaningful employment in the region
-   took the ACT had SIGNIFICANTLY higher rates of attending post-secondary
-   did not take the ACT had SIGNIFICANTLY higher rates of not having meaningful employment
-   did not take the ACT had significantly higher rates of not having a MN employment record.

```{r hs accomplishments took.ACT one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, took.ACT) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, moderate association between a whether a graduate has taken the ACT and the employment state.

Graduates that;

-   did not take the ACT had higher rates of meaningful employment in the region
-   did not take the ACT had lower rates of meaningful employment in MN
-   did not take the ACT had SIGNIFICANTLY lower rates of attending post-secondary
-   did not take the ACT had SIGNIFICANTLY higher rates of not having meaningful employment
-   did not take the ACT had significantly higher rates of not having a MN employment record.

```{r hs accomplishments took.ACT five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, took.ACT) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between a whether a graduate has taken the ACT and the employment state.

Graduates that;

-   did not take the ACT had higher rates of meaningful employment in the region
-   did not take the ACT had lower rates of meaningful employment in MN

```{r hs accomplishments took.ACT ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, took.ACT) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ACTCompositeScore

There is a statistically significant association between the ACT Compite Score of a graduate and their employment state.

Through all the years, the main trend was;

-   Graduates with meaningful employment in the region had lowest ACT Composite Scores than all other states.
-   Graduates with meaningful employment in MN had second lowest ACT Composite Scores
-   No MN employment record had high ACT Composite Scores as the years progressed.
-   Graduates attending post-secondary had the highest.

::: panel-tabset
### One year after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   attended post-secondary had ACT Composite Scores 300% higher than graduates who had meaningful employment participation in the region or in Minnesota.
-   had meaningful employment had ACT Composite Scores 31% higher than graduates with meaningful employment in the region.
-   Graduates with no meaningful employment had an ACT Composite Score 58% higher than graduates with meaningful employment in the region, 27% higher than graduates with employment in Minnesota, and scores 274% lower than graduates who attended post-secondary.
-   Graduates with no MN employment record has ACT Composite Scores 138% higher than graduates with meaningful employment in the region, 107% higher than graduates with meaningful employment in MN, 194% lower than graduates attending post-secondary, and 80% higher than graduates with no meaningful employment.

```{r hs accomplishments ACTCompositeScore one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ACTCompositeScore) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, ACTCompositeScore, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   attended post-secondary had ACT Composite Scores 250% higher than graduates who had meaningful employment participation in the region, 52% higher than graduates with meaningful employment in Minnesota.
-   had meaningful employment had ACT Composite Scores 197% higher than graduates with meaningful employment in the region.
-   Graduates with no meaningful employment had an ACT Composite Score 108% higher than graduates with meaningful employment in the region, 89% lower than graduates with employment in Minnesota, and scores 141% lower than graduates who attended post-secondary.
-   Graduates with no MN employment record has ACT Composite Scores 200% higher than graduates with meaningful employment in the region, 4% higher than graduates with meaningful employment in MN, 48% lower than graduates attending post-secondary, and 93% higher than graduates with no meaningful employment.

```{r hs accomplishments ACTCompositeScore five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ACTCompositeScore) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, ACTCompositeScore, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had meaningful employment in MN had ACT Composite Scores 144% higher than graduates with meaningful employment in the region.
-   Graduates with no meaningful employment had an ACT Composite Score 66% higher than graduates with meaningful employment in the region, 78% lower than graduates with employment in Minnesota.
-   Graduates with no MN employment record had ACT Composite Scores 174% higher than graduates with meaningful employment in the region, 31% higher than graduates with meaningful employment in MN, and 109% higher than graduates with no meaningful employment.

```{r hs accomplishments ACTCompositeScore ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ACTCompositeScore) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, ACTCompositeScore, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 40))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.R

MCA Reading scores were consistently associated with post-graduation outcomes across one, five, and ten years. Clear academic gradients were observed, though the strength and pattern of these associations shifted over time.

**Year 1**:\

Students attending *post-secondary* had the highest scores, followed by those with *meaningful employment*. Students with no MN employment record or not meaningful employment scored significantly lower.

**Year 5**:\

The top performers were those with *meaningful employment in Minnesota*, followed closely by *post-secondary attendees*. Those disconnected from both education and meaningful employment continued to score lower.

**Year 10**:\

*Meaningful MN employment* remained strongly associated with higher scores. Interestingly, individuals with *no MN employment record* also had high scores—comparable to those in MN employment—suggesting potential out-of-state or untracked employment. Students in *not meaningful employment* remained the lowest-performing group.

**Summary:**\

Higher MCA Reading scores are consistently linked to stronger early and long-term outcomes—particularly *meaningful employment in Minnesota* and *post-secondary enrollment*. The association weakens slightly by Year 10, but academic achievement remains a clear differentiator, especially for those disengaged from both education and the workforce.

::: panel-tabset
### One year after graduating

The p-value indicates a significant difference between one of the groups.

-   **Attending post-secondary** students had the highest reading scores overall. Their scores were **0.28 to 0.64 points higher** than those in any other group.
-   **Students with meaningful employment** (either in Minnesota or the region) had moderately high scores, but still lower than those attending post-secondary.
    -   Those with **meaningful employment in MN** scored **0.04 points higher** than those employed regionally.
-   The **lowest scores** were found among students who:
    -   Had **no MN employment record** and were **not attending post-secondary**, or
    -   Had **not meaningful employment** and were **not attending**.
    -   These groups scored approximately **0.58–0.64 points lower** than post-secondary attendees.
-   All pairwise comparisons were statistically significant. The results show a **clear academic gradient**, where higher MCA Reading scores were associated with more favorable early outcomes.

```{r hs accomplishments MCA.R one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.R) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.R, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

-   **Meaningful employment in Minnesota** was associated with the **highest reading scores** at Year 5, with scores **0.18 points higher** than those with meaningful employment in the region.
-   **Post-secondary attendees** also had relatively high scores, about **0.28 points higher** than those in regional employment, and **0.10 points higher** than those with meaningful MN employment.
-   **Students not meaningfully employed or not in post-secondary** showed **mixed results**:
    -   Their scores were slightly **higher than those in regional employment** (+0.04), but still **lower than those in MN employment** (−0.14) and **post-secondary attendees** (−0.24).
-   **Those with no MN employment record and not in post-secondary** scored **below every group except those not meaningfully employed**, with scores **\~0.13 lower than regional employment**, **\~0.05 lower than MN employment**, and **\~0.15 lower than post-secondary attendees**.
-   The **smallest difference** was between the two lower-performing groups: students with no MN record scored **0.09 points higher** than those with not meaningful employment.

```{r hs accomplishments MCA.R five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.R) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.R, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

-   **Meaningful employment in Minnesota** was associated with the **highest MCA Reading scores** at Year 10.
    -   Scores were **0.37 points higher** than those with meaningful employment in the region.
-   **Students with no MN employment record and not in post-secondary** also had **elevated scores** — nearly **identical** to those with meaningful MN employment.
    -   This group scored **0.37 points higher** than those in regional employment.
-   **Students with not meaningful employment and not in post-secondary** had the **lowest reading scores overall**.
    -   Their scores were **0.36 points lower** than those with no MN record and **0.36 points lower** than those with MN employment.
    -   The difference between them and those with regional employment, however, was **not statistically significant** (+0.01, p = 0.82).
-   The **difference between no MN employment record and meaningful MN employment** was **not statistically significant** either (+0.006, p = 0.92), suggesting similar reading performance between these two groups.

```{r hs accomplishments MCA.R ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.R) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.R, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.S

MCA Science scores consistently aligned with post-graduation outcomes, showing clear academic gradients that persisted over time. While some differences narrowed by Year 10, the overall pattern remained stable.

**Year 1**

-   **Post-secondary attendees** had the **highest Science scores**, followed by those with **meaningful MN employment**.
-   Students with **not meaningful employment** or **no MN employment record** scored significantly lower.
-   Most group differences were statistically significant, indicating a strong early link between academic performance and immediate outcomes.

**Year 5**

-   **Post-secondary** and **meaningful MN employment** continued to be associated with the **highest Science scores**.

-   Students with **no MN record** scored below these top groups but still significantly higher than those not meaningfully employed.

-   The difference between **not meaningful employment** and **regional employment** was not significant, suggesting convergence at lower performance levels.

**Year 10**

-   **Meaningful MN employment** remained tied to the **highest scores**.

<!-- -->

-   Interestingly, those with **no MN record** also had relatively high scores—nearly on par with MN-employed peers—suggesting out-of-state or untracked employment among academically stronger individuals.

-   Students with **not meaningful employment** consistently had the **lowest Science scores**, though differences with other low-engagement groups became statistically insignificant in some cases

**Summary**\

MCA Science scores are strongly predictive of positive post-graduation outcomes—particularly **post-secondary participation** and **meaningful employment in Minnesota**. By Year 10, the distinction between high- and low-performing groups softens slightly, but academic achievement in high school remains a clear marker of long-term engagement in education or the workforce.

::: panel-tabset
### One year after graduating

-   **Post-secondary attendees** had the **highest Science scores**, averaging nearly **0.45 points higher** than those with meaningful regional employment and **\~0.55 points higher** than those with not meaningful employment.
-   **Meaningful employment in Minnesota** was associated with moderately high scores. These were **0.42 points lower than post-secondary attendees** but slightly higher than those with regional employment.
-   **Not meaningful employment** was linked to **significantly lower scores**, trailing post-secondary attendees by over **0.54 points**.
-   **No MN employment record** students had scores slightly below most other groups, but the difference with regional employment was **not statistically significant**.

**Conclusion:**\

Higher MCA Science scores are strongly linked to **post-secondary enrollment** and **meaningful employment**, particularly within Minnesota. The results suggest that Science proficiency in high school aligns closely with early academic and workforce engagement.

```{r hs accomplishments MCA.S one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.S) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.S, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

-   **Post-secondary attendees** and those with **meaningful employment in Minnesota** had the **highest Science scores**.
    -   Post-secondary students scored **0.39 points higher** than those with regional employment and **0.15 points higher** than those with MN employment.
-   **Meaningful employment in Minnesota** was associated with **strong Science scores**, scoring **0.24 points above** regional employment and well above all non-engaged groups.
-   **Students with no MN employment record** scored **below MN-employed peers** but still **0.17 points above** those with regional employment.
-   **Students with not meaningful employment and not attending post-secondary** had the **lowest scores overall**, nearly **0.39 points below** post-secondary attendees and **0.24 points below** those with MN employment.
-   The difference between **not meaningful employment** and **regional employment** was **not statistically significant**, suggesting similar performance levels.

**Conclusion:**\

By Year 5, Science scores continue to track with more favorable outcomes, especially **post-secondary education** and **meaningful MN employment**. Academic engagement in high school appears to differentiate those more securely attached to the labor force or continuing education.

```{r hs accomplishments MCA.S five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.S) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.S, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

-   **Meaningful employment in Minnesota** was associated with the **highest Science scores**, averaging **0.23 points higher** than those in regional employment.
-   **Students with no MN employment record** also had **elevated scores**, about **0.18 points above** those in regional employment and **only 0.05 points below** those with MN employment.
-   The difference between **not meaningful employment** and **regional employment** was **not statistically significant** (p = 0.67), suggesting similar academic profiles.
-   **Students not meaningfully employed and not in post-secondary** had the **lowest scores**, scoring **0.24 points below** MN-employed peers and **0.20 points below** those with no MN record.

**Conclusion:**\

By Year 10, high school Science performance continues to differentiate individuals in **meaningful MN employment** or possibly **out-of-state or untracked employment**. Those disengaged from both the labor force and post-secondary education remain the lowest-performing group. While gaps persist, some distinctions—particularly between lower-performing categories—have narrowed over time.

```{r hs accomplishments MCA.S ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.S) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.S, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.M

MCA Math scores consistently aligned with students' post-graduation pathways across one, five, and ten years. The results show a strong academic gradient, especially in the early years, with slight narrowing of differences by Year 10.

**Year 1**

-   **Post-secondary attendees** had the **highest Math scores**, outperforming all other groups by wide margins—**0.64 points above** regional employment and **0.83 points above** those with no MN employment record.

-   **Meaningful MN employment** was associated with moderately high scores, while **not meaningful employment** and **no MN employment** were linked to the **lowest performance**.

**Year 5**

-   The pattern persisted, with **post-secondary** and **meaningful MN employment** groups maintaining the highest scores.

-   Gaps between top and bottom groups remained large—up to **0.65 points**.

-   Students not meaningfully employed continued to score significantly lower than their peers.

**Year 10**

-   **Meaningful MN employment** remained the top-performing group.

-   Interestingly, those with **no MN employment record** had **relatively high scores**, suggesting potential out-of-state or untracked employment among academically stronger individuals.

-   The difference between **not meaningful employment** and **regional employment** was no longer significant, indicating convergence among lower-performing groups.

**Summary**\

MCA Math scores are the **strongest and most persistent academic predictor** of long-term employment and education outcomes. Higher scores are consistently associated with **post-secondary education** and **meaningful employment**, especially within Minnesota. Although some differences among lower-performing groups narrow over time, Math achievement in high school remains a **clear differentiator** well into adulthood.

::: panel-tabset
### One year after graduating

-   **Post-secondary attendees** had the **highest Math scores by far**, scoring:
    -   **0.64 points higher** than those with meaningful regional employment
    -   **0.77–0.83 points higher** than those not meaningfully employed or with no MN employment record
-   **Meaningful employment in Minnesota** was associated with **moderately high scores**:
    -   Slightly higher than regional employment (**+0.04 points**)

    -   Roughly **0.17–0.23 points higher** than non-engaged groups
-   **Not meaningful employment** and **no MN employment record** were linked to the **lowest Math scores**, with all gaps statistically significant.

**Conclusion:**\

Math scores showed the **strongest early academic differentiation** of any subject so far. Students pursuing **post-secondary education** or engaged in **meaningful employment**—especially in Minnesota—scored significantly higher than those disconnected from both education and stable employment.

```{r hs accomplishments MCA.M one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.M) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.M, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

-   **Post-secondary attendees** continued to have the **highest Math scores**, scoring:

    -   **0.60 points above** those with regional employment
    -   **0.33 points above** those with meaningful MN employment
    -   Over **0.65 points above** those with not meaningful employment

-   **Meaningful MN employment** was again associated with strong academic performance, with scores:

    -   **0.27 points above** regional employment

    -   **0.32 points above** those not meaningfully employed

    -   **0.17 points above** those with no MN employment record

-   **Students not meaningfully employed and not attending post-secondary** had the **lowest Math scores**, significantly below all other groups.

-   **Students with no MN employment record** had **moderate scores**, higher than non-engaged peers but lower than those in meaningful employment.

**Conclusion:**\

By Year 5, MCA Math scores remain a powerful predictor of post-school engagement. The highest scores are concentrated among those in **post-secondary** or **meaningful MN employment**, while the lowest scores continue to cluster among those **disconnected from both education and stable employment**.

```{r hs accomplishments MCA.M five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.M) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.M, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

-   **Meaningful employment in Minnesota** was linked to the **highest Math scores**, scoring:
    -   **0.29 points higher** than regional employment
    -   **0.30 points higher** than those not meaningfully employed
    -   **0.07 points higher** than those with no MN employment record
-   **Students with no MN employment record** also had **elevated scores**, **0.22 points above** regional employment and **0.23 points above** those not meaningfully employed.
-   The difference between **not meaningful employment** and **regional employment** was **not statistically significant**, suggesting convergence among lower-scoring groups.
-   Students who were **not meaningfully employed and not in post-secondary** had the **lowest scores**, consistent with patterns in earlier years.

**Conclusion:**\

At Year 10, MCA Math scores continue to reflect long-term outcomes. Individuals in **meaningful MN employment** or with **no MN record** had the highest scores, while those disengaged from both education and employment consistently showed the lowest academic performance. Gaps between lower-performing groups narrowed, but high-achieving students remained concentrated in stable employment outcomes.

```{r hs accomplishments MCA.M ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.M) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.M, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## cte.achievement

This section summarizes the association between **Career and Technical Education (CTE) achievement** and **employment or education outcomes** at one, five, and ten years after high school. While all three chi-square tests were statistically significant, the **Cramér’s V values** suggest a consistently **very weak relationship**:

-   **Year 1:** Cramér’s V = 0.091
-   **Year 5:** Cramér’s V = 0.069
-   **Year 10:** Cramér’s V = 0.054

**Key trends over time:**

**Year 1:**

-   **CTE Concentrators/Completers** were more likely to enter **regional or meaningful MN employment** early.
-   **Non-CTE students** were more likely to pursue **post-secondary education**, with over half still enrolled one year after graduation.
-   Early workforce engagement was more common among those with higher CTE involvement.

**Year 5:**

-   Differences began to narrow, but **CTE Concentrators/Completers** remained more likely to be in **regional or MN employment**.
-   **Non-CTE students** continued to show higher levels of **post-secondary engagement**, though to a lesser extent than at Year 1.
-   The association between CTE status and outcomes weakened overall.

**Year 10:**

-   Outcome distributions became **similar across all CTE groups**.
-   **Meaningful MN employment** was the most common outcome regardless of CTE status (\~30%).
-   **Non-CTE students** were slightly more likely to have **no MN employment record**, suggesting possible migration or untracked employment.

**Conclusion:**\

CTE achievement is **modestly associated** with earlier post-graduation outcomes, especially around **regional employment and post-secondary enrollment**. However, by ten years out, differences by CTE status largely **diminish**, indicating that **CTE pathways may shape short-term trajectories**, but **long-term outcomes converge** across student groups.

::: panel-tabset
### One year after graduating

This cross-tabulation examines the relationship between **CTE achievement level** and **employment or education outcomes** one year after high school. The association is statistically significant:\

**χ² = 2155.68**, **df = 8**, **p \< 0.001**, **Cramér’s V = 0.091**.

**Key takeaways:**

-   The **Cramér’s V of 0.091** indicates a **very weak** association between CTE achievement and post-graduation outcomes. While statistically significant (due to large sample size), the effect size is small.
-   **CTE Concentrators/Completers** were:
    -   **More likely to be employed regionally** (16.4%) compared to participants (12.6%) and non-CTE students (9.6%)

    -   **Slightly more likely to have meaningful MN employment** (9.2% vs. 7% participants and 5.9% non-CTE)

    -   **Less likely to be attending post-secondary** (42.6%) than participants (51.9%) or non-CTE students (55.8%)
-   **CTE Participants** and **Non-CTE students** were more likely to pursue **post-secondary education**, especially non-CTE students (55.8%).
-   Differences in the lowest engagement categories (not meaningful or no MN employment) were relatively small across CTE groups.

**Conclusion:**\

While CTE achievement is **weakly associated** with post-graduation outcomes, **CTE concentrators/completers appear more likely to enter the workforce early**, especially in **regional or meaningful MN employment**. In contrast, **non-CTE students are more likely to pursue post-secondary education**. These findings support the notion that CTE pathways are more strongly aligned with **early workforce entry**, particularly in local or regional labor markets.

```{r hs accomplishments cte.achievement one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, cte.achievement) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

This table shows the relationship between **CTE achievement level** and **employment or education outcomes** five years after high school graduation. The results are statistically significant:\

**χ² = 900.39**, **df = 8**, **p \< 0.001**, **Cramér’s V = 0.069**.

**Key takeaways:**

-   The **Cramér’s V of 0.069** indicates a **very weak association**, though statistically significant due to the large sample size. This suggests **CTE achievement has limited influence** on outcomes by Year 5.
-   **CTE Concentrators/Completers** remain **more likely to be employed regionally** (25.2%) or in **meaningful MN employment** (25.4%) compared to other groups.
-   **CTE Participants** and **Non-CTE students** had **higher post-secondary attendance rates** (16.3% and 18.5%, respectively) than CTE Concentrators/Completers (12%).
-   Differences across **low-engagement categories** (not meaningful employment or no MN employment record) were **small across all CTE groups**, ranging from 17% to 20%.

**Conclusion:**\

By Year 5, **CTE Concentrators/Completers are more workforce-oriented**, particularly toward **regional or meaningful MN employment**. Meanwhile, **Non-CTE students are more likely to remain engaged in post-secondary education**. However, the overall strength of the relationship is **weak**, indicating that while CTE pathways may steer students toward employment, many other factors likely influence long-term outcomes.

```{r hs accomplishments cte.achievement five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, cte.achievement) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

This table displays the relationship between **CTE achievement level** and **employment outcomes** ten years after high school graduation. The association is statistically significant:\

**χ² = 299.54**, **df = 8**, **Fisher’s p \< 0.001**, **Cramér’s V = 0.054**.

**Key takeaways:**

-   The **Cramér’s V of 0.054** indicates a **very weak association** between CTE achievement and employment outcomes by Year 10.
-   **Meaningful MN employment** was the most common outcome for all groups, with:
    -   30.5% of **CTE Concentrators/Completers**
    -   30.6% of **CTE Participants**
    -   29.6% of **Non-CTE students**
-   **Regional employment** was more common among **CTE Concentrators/Completers** (25.7%) than among **Participants** (23.7%) and **Non-CTE students** (19.9%).
-   The **“No MN employment record”** category was most frequent among **Non-CTE students** (33.3%), compared to 26.6% and 28.1% in the CTE groups.
-   Virtually no one was **still attending post-secondary** at Year 10 (as expected).

**Conclusion:**\

By Year 10, the relationship between **CTE achievement and long-term outcomes** is **minimal**. Employment outcomes are **similar across CTE and non-CTE groups**, though **CTE Concentrators/Completers remain slightly more likely to be employed regionally**, while **Non-CTE students are more likely to have no MN employment record**. This suggests that **initial CTE pathways may influence early employment decisions**, but **long-term patterns converge**, and other factors likely drive employment location and quality over time.

```{r hs accomplishments cte.achievement ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, cte.achievement) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## total.cte.courses.taken

This section summarizes how the **average number of CTE courses taken in high school** varies by employment and education outcomes at one, five, and ten years after graduation. Across all three years, the results show **statistically significant** patterns, though **effect sizes shrink over time**.

**Year 1**

-   **Regional employment** is associated with the **highest CTE course counts**.
-   **Post-secondary attendees** took significantly **fewer** CTE courses than any other group (–1.55 vs. regional employment).
-   Those **not meaningfully employed** and those with **no MN employment record** took **moderately fewer** courses than regional workers, but **more than post-secondary students**.
-   CTE coursework appears to align with **early entry into the workforce**, especially regionally.

**Year 5**

-   **Regional employment** still corresponds to **higher CTE participation**, but differences begin to narrow.
-   **Post-secondary attendees** continue to have the **lowest CTE course counts**.
-   **Meaningful vs. not meaningful employment** groups now show **no significant difference** in CTE participation, suggesting possible convergence.
-   **CTE appears to support earlier labor market engagement**, but distinctions across employment types begin to blur.

**Year 10**

-   **Regional employees** maintain the highest average number of CTE courses.
-   Differences between **meaningful** and **not meaningful employment** are **no longer significant**.
-   Those with **no MN employment record** consistently had the **lowest CTE course counts**, reinforcing patterns of longer-term disconnection.
-   Overall, the relationship between CTE course-taking and employment weakens over time, suggesting that **long-term labor outcomes are shaped by more than high school coursework alone**.

**Summary**

TE course-taking in high school is **strongly linked to early post-graduation employment**, particularly in the **regional labor market**. However, these associations **diminish by Year 10**, with **converging patterns** among most employment categories. The **most consistent pattern** is that individuals **disconnected from the MN employment system** tend to have had the **lowest CTE engagement** in high school.

::: panel-tabset
### One year after graduating

his ANOVA examines differences in the **average number of CTE courses taken** across employment and education outcomes one year after graduation. Most group comparisons are statistically significant (**p \< 0.001**).

**Key takeaways:**

-   Individuals in **regional employment** took the **most CTE courses**, on average.
-   **Post-secondary attendees** took **1.55 fewer CTE courses** than those employed in the region. They also took:
    -   **1.48 fewer** than those with meaningful MN employment
    -   **1.01 fewer** than those not meaningfully employed
    -   **0.64 fewer** than those with no MN employment record
-   Individuals with **no MN employment record** and those **not meaningfully employed** took **moderately fewer CTE courses** than those employed regionally, but **more** than post-secondary attendees.
-   The only non-significant difference was between **regional employment** and **meaningful MN employment**, where CTE course counts were statistically similar (**p = 0.26**)

**Conclusion:**\

CTE course-taking is **highest among those entering the regional workforce** and **lowest among post-secondary students**. This suggests that CTE coursework may be aligned more with **early labor market entry** than with **continued education**, even just one year out from high school.

```{r hs accomplishments total.cte.courses.taken one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, total.cte.courses.taken) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, total.cte.courses.taken, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

This ANOVA analyzes differences in the **average number of CTE courses taken** across employment and education outcomes five years after high school. Most pairwise comparisons are **statistically significant (p \< 0.001)**, with some patterns shifting slightly from Year 1.

**Key takeaways:**

-   **Individuals employed regionally** continued to take the **most CTE courses**, on average.
-   Compared to regional employees:
    -   Those with **meaningful MN employment** took **0.35 fewer CTE courses**
    -   **Post-secondary attendees** took **1.22 fewer**
    -   Those **not meaningfully employed** took **0.38 fewer**
    -   Individuals with **no MN employment record** took **0.94 fewer**
-   The difference between **meaningful MN employment** and **not meaningful employment** was **not statistically significant** (*p = 0.79*), indicating similar CTE course participation between those two groups by Year 5.
-   **Post-secondary attendees** took significantly **fewer CTE courses** than all other groups — consistent with Year 1 patterns.

**Conclusion:**\

At Year 5, CTE course-taking remains **most strongly associated with regional employment**, though the gap between **meaningful** and **not meaningful employment** has narrowed. This suggests that while CTE coursework may continue to support **early regional workforce engagement**, its influence on longer-term employment distinctions may diminish over time. Post-secondary students remain clearly on a different academic track, taking substantially fewer CTE courses in high school.

```{r hs accomplishments total.cte.courses.taken five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, total.cte.courses.taken) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, total.cte.courses.taken, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

This ANOVA explores how the number of **CTE courses taken in high school** relates to employment outcomes ten years post-graduation. Most group comparisons are statistically significant (**p \< 0.001**), though overall patterns suggest **narrowing distinctions** over time.

**Key takeaways:**

-   **Individuals employed regionally** continued to have the **highest CTE course counts**, maintaining the trend observed in earlier years.
-   Compared to regional employees:
    -   Those with **meaningful MN employment** took **0.57 fewer CTE courses**
    -   Those **not meaningfully employed** took **0.52 fewer**
    -   Those with **no MN employment record** took **0.85 fewer**
-   The **difference between meaningful MN employment and not meaningful employment** was **not statistically significant** (*p = 0.615*), suggesting these two groups had **similar CTE participation** by Year 10.
-   Students with **no MN employment record** had the **lowest CTE course participation**, significantly lower than all other groups.

**Conclusion:**\

By Year 10, the **strongest CTE connection remains with regional employment**. However, distinctions between those with and without meaningful MN employment have largely **disappeared** in terms of CTE course-taking. Individuals with **no MN employment record** consistently had the **lowest levels of CTE engagement**, suggesting long-term disengagement from both high school career preparation and in-state employment.

```{r hs accomplishments total.cte.courses.taken ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, total.cte.courses.taken) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, total.cte.courses.taken, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## ChatGPT

```{r high school accomplishments chatgpt}
hs.accomplishments.chatgpt <- meaningful.emp.central %>%
  select(PersonID, pseo.participant, took.ACT, ACTCompositeScore, MCA.R, MCA.S, MCA.M, cte.achievement, total.cte.courses.taken, grad.year.1, grad.year.5, grad.year.10)

write_csv(hs.accomplishments.chatgpt, "Data/ChatGPT/hs-accomplishments-chatgpt.csv")

```

# Post-secondary

We will analyze the following variables;

-   attended.ps.within.first.year.hsgrad
-   attended.ps
-   ps.grad
-   ps.grad.InstitutionSector
-   highest.cred.level
-   ps.grad.location

The analysis provides the following insights;

-   Meaningful employment in the region, long-term
    -   More likely:
        -   Attended or graduated
        -   Graduated from a public, 2-year
        -   Less than associated degree or an associate degree
        -   Graduated in region
    -   Less likely:
        -   Did not attend or graduated
        -   Graduated from any 4-year or any for-profit
        -   Bachelors degree or higher & No college
        -   Outside region, worse if outside Minnesota
-   Meaningful employment in MN, long-term
    -   More likely:
        -   Attended or graduated.
        -   Graduated from any 4-year or any for-profit
        -   Bachelor degree or higher
        -   Graduated in Minnesota
    -   Less likely
        -   Did not attend or graduate
        -   Graduated from a public 2-year
        -   No college or some college, associate degree, less than associate degree
-   Not meaningful
    -   More likely:
        -   Did not attend or graduate
        -   Graduated from a for-profit
    -   Less likely:
        -   Graduated from post-secondary
        -   ASsociate or higher
-   No MN employment record
    -   More likely:
        -   Did not attend or graduate
        -   Graduated from a for-profit, or public 4-year
        -   Master degree or higher
        -   Outside Minnesota
    -   Less likely:
        -   Graduated
        -   Graduated from a 2-year
        -   Associate degree, less than associate degree
        -   In region

<br>

## attended.ps.within.first.year.hsgrad

-   One year after graduating;
    -   Individuals that did not attend college, or attended not immediately after graduating college were significantly more likely to have meaningful employment in the region.
    -   These individuals were significantly more likely to not have meaningful employment.
    -   Significantly more likely to not have a MN employment record.
-   Five years after graduating;
    -   Significantly more likely to not have meaningful employment
    -   Individuals who did not attend post-secondary were significantly more likely to not have a MN employment record
-   Ten years after graduating;
    -   individuals that never attended post-secondary were significantly less likely to have meaningful employment in the region.
    -   Individuals that did not attend post-secondary were significantly more likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## attended.ps

-   One year after graduating; individuals that did not attend post-secondary were;
    -   more likely to have meaningful employment in the region.
    -   Significantly more likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Five years after graduating; individuals that did not attend post-secondary were;
    -   More likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Ten years after graduating; individuals that did not attend post-secondary were;
    -   Less likely to have meaningful employment in region or Minnesota
    -   Significantly more likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad

-   One year after graduating; individuals that graduated from post-secondary were;
    -   less likely to have meaningful employment in the region.
    -   Significantly more likely to attend post-secondary
    -   Significantly less likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Five years after graduating; individuals that did not graduate from post-secondary were;
    -   More likely to have meaningful employment in Minnesota
    -   Significantly less likely to not have meaningful employment
    -   Significantly less likely to not have a MN employment record
-   Ten years after graduating; individuals that did not attend post-secondary were;
    -   Slightly more likely to have meaningful employment in the region or in Minnesota

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad.InstitutionSector

A quick reminder on what the codes mean;

-   1 - Public, 4-year or above
-   2 - Private not-for-profit, 4-year or above
-   4 - Public, 2-year
-   10 - went to multiple sectors
-   11 - Includes all of the following
    -   3 - Private for-profit, 4-year or above
    -   6 - Private for-profit, 2-year
    -   9 - Private for-profit, less-than-2-year

<!-- -->

-   One year after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 2
    -   No MN employment record
        -   More likely: 11 (low numbers though)
-   Five years after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 4
        -   More likely: 1, 11, 2
    -   Not meaningful
        -   Less likely: 10
    -   No MN employment record
        -   Less likely: 10
        -   More likely: 11, 2
-   Ten years after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2, 11,
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 4, 11
        -   More likely: 1, 2
    -   Not meaningful
        -   More likely: 11 (low numbers)
    -   No MN employment record
        -   Less likely:
        -   More likely: 11

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.InstitutionSector one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.InstitutionSector) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.InstitutionSector five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.InstitutionSector) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.InstitutionSector ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.InstitutionSector) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## highest.cred.level

-   One year after graduating;
    -   Have meaningful employment in region
        -   Less likely: Bachelor degree or higher
        -   More likely: No college, Associate degree
    -   Meaningful emp in MN
        -   Less likely: Bachelor or higher
        -   More likely: No college
    -   Not meaningful
        -   Less likely: Less than associate degree or higher
        -   More likely: No college
    -   No MN employment record
        -   Less likely: Some college or higher
        -   More likely: No college
-   Five years after graduating;
    -   Have meaningful employment in region
        -   Less likely: Bachelor degree or higher
        -   More likely: Less than associate degree to associate degree
    -   Meaningful emp in MN
        -   Less likely: Bachelor degree
        -   More likely: Master degree or higher
    -   Not meaningful
        -   Less likely: Master degree or higher
        -   More likely: No college or some college
    -   No MN employment record
        -   Less likely: Associate degree or higher
        -   More likely: No college
-   Ten years after graduating;
    -   Have meaningful employment in region
        -   Less likely: Master degree, No college
        -   More likely: Associate degree, Less than associate degree
    -   Meaningful emp in MN
        -   Less likely: No college, some college, Associate degree, Less than associate degree
        -   More likely: Bachelor degree or higher
    -   Not meaningful
        -   Less likely: Associate degree or higher
        -   More likely: Some college
    -   No MN employment record
        -   Less likely: Less than Associate degree
        -   More likely:

::: panel-tabset
### One year after graduating

```{r post-secondary highest.cred.level one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, highest.cred.level) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary highest.cred.level five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, highest.cred.level) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary highest.cred.level ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, highest.cred.level) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad.location

-   One year after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside MN
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: Outside Minnesota
        -   More likely:
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota
        -   More likely: Outside Minnesota
    -   No MN employment record
        -   Less likely: Inside and outside region
        -   More likely: Inside and outside Minnesota, outside Minnesota

-   Five years after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside MN
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: In region, inside and outside region, inside and outside MN, outside MN
        -   More likely: In Minnesota
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota
        -   More likely:
    -   No MN employment record
        -   Less likely: Inside and outside region, Inside and outside MN
        -   More likely: outside Minnesota

-   Ten years after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside Minnesota
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: In region
        -   More likely: In Minnesota, Inside and outside Minnesota
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota, outside Minnesota
        -   More likely:
    -   No MN employment record
        -   Less likely: In region
        -   More likely: Inside and outside Minnesota, Outside minnesota

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.location one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.1 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.location five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.5 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.location ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.10 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

# Variables to keep

Okay, after this analysis, here is what we will move forward with;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   english.learner
-   High school characteristics
    -   Dem_Desc
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ap.exam
    -   ACTCompositeScore
    -   MCA.M
    -   MCA.R
    -   MCA.S
    -   cte.achievement
    -   n.cte.courses.taken
-   Post-secondary
    -   ps.grad
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   ps.grad.location

```{r create new master after analysis}
master.after.ct <- meaningful.emp.central %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  select(PersonID, grad.year.covid, economic.status, , pseo.participant, SpecialEdStatus, RaceEthnicity, Dem_Desc, took.ACT, cte.achievement, english.learner, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location, ACTCompositeScore, MCA.M, MCA.R, MCA.S, total.cte.courses.taken,grad.year.1, grad.year.5, grad.year.10)
  
write_csv(master.after.ct, "Data/SLEDS/Masters/After analysis/Master-after-ct.csv")
```
