---
title: "Local Employment Cross-tabs"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.factor.names <- master.original %>%
  select(PersonID, Dem_Desc, edr, grad.year.covid, Gender, economic.status, pseo.participant, SpecialEdStatus, non.english.home, RaceEthnicity, took.ACT, ap.exam, cte.achievement, english.learner, sat.taken, attended.ps, attended.ps.within.first.year.hsgrad, ps.grad, ps.grad.InstitutionSector, highest.cred.level) 

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.states %>%
  left_join(master.factor.names, by = "PersonID")

kable(head(master))

kable(names(master))
```

```{r independent and dependent variables}
updated.factor.var.names <-  master.factor.names %>%
  select(-PersonID) %>%
  names() %>%
  c("ps.grad.in.same.ruca", "ps.grad.in.same.edr", "ps.grad.in.same.pr", "ps.grad.in.MN")

numeric.var.names <- master.original %>%
  select(ACTCompositeScore, total.cte.courses.taken, `ACTE.SPED`:Youth.Apprenticeship,  MCA.M:MCA.S, n.institutions, avg.unemp.rate, wages.3year.avg, avg.wages.pct.state) %>%
  names()
  
updated.independent.var.names <- c(updated.factor.var.names, numeric.var.names)

cte.cc.var.names <- master.original %>%
  select(ACTE.SPED:Youth.Apprenticeship) %>%
  names()

dep.var.names <- master %>%
  select(PersonID, grad.year.1, grad.year.5, grad.year.10)

updated.master <- master.original %>%
  select(PersonID, all_of(updated.independent.var.names)) %>%
  left_join(dep.var.names, by = "PersonID") %>%
  mutate_at(vars(Dem_Desc:ps.grad.in.MN), as.factor) %>%
  mutate_at(vars(grad.year.1:grad.year.10), as.factor) %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Did not grad", "Some college", as.character(highest.cred.level)),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Never attended ps", "No college", as.character(highest.cred.level)),
         highest.cred.level = as.factor(highest.cred.level),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix", "Urban/town/rural mix", "Entirely urban")) %>%
  droplevels()

kable(head(updated.master))

kable(names(updated.master))
```

<br>

There are a number of "states" in which an individual in this dataset are assigned;

1.  Has meaningful employment
    1.  In the same county as their high school.
    2.  In the same EDR as their high school, but not county.
    3.  In Minnesota, but not in Central Minnesota.
2.  Is attending post-secondary.
3.  Has a MN employment record but it's not meaningful, and not attending post-secondary.
4.  Does not have a MN employment record and is not attending post-secondary.

There are three "times" in which we checked to see if the individual had meaningful employment within these geographies;

1.  one year after graduating high school
2.  five years after graduation, and
3.  ten years after graduation.

Meaningful employment is determined by whether an individual worked 1,000 hours for an employer during time X. In addition, if the individual is working for an employer but not meaningful at time X but worked 1,000 hours for that employer during another year it's still considered "meaningful".

Due to time x potentially being passed the date of the latest data (2023) for some individuals, the analysis below will filter out all individuals where time x is after 2023.

From the multiple correspondence analysis and the principle component analysis we determined that the following independent variables will be included in the analysis.

```{r list of ind var}
updated.master %>%
  select(-PersonID, -grad.year.1, -grad.year.5, -grad.year.10) %>%
  lapply(class) 

```

# Methodology

For this analysis we are keeping it simple. For categorical variables I’ll be using the Cramer’s v-square measure of association. The V\^2 value is from 0 to 1 with 0 meaning there’s no association and a value closer to 1 indicates a strong association.

| **Cramér’s V Value** | **Interpretation**               |
|----------------------|----------------------------------|
| 0.00–0.10            | Negligible association           |
| 0.10–0.20            | Weak                             |
| 0.20–0.40            | Moderate                         |
| 0.40–0.60            | Relatively strong                |
| 0.60–0.80            | Strong                           |
| 0.80–1.00            | Very strong (possibly redundant) |

For the variables with continuous values I’ll be using the Wilcoxon two paired groups test. P-value will be used to determine whether we can reject the NULL hypothesis at a .01 level. I’ll also include boxplots to see in which direction there might be influence from the variables.

We are going to split up the analysis into two different datasets. The first dataset will combine 3 of the previous states into one.

-   Has meaningful employment anywhere in Central Minnesota, which combines the following states;
    -   Meaningful emp County
    -   Meaningful emp EDR
    -   Meaningful emp region

The remaining states will stay the same.

-   Has meaningful employment in Minnesota, but not in Central Minnesota
-   Is attending a post-secondary institution.
-   Has a MN employment record but it's not meaningful, and they are not attending post-secondary.
-   Does not have a MN employment record and is not attending post-secondary.

We have quite a few variables remaining, even after MCA analysis. So let's take a look at how we are going to proceed and group all of this together to see how these relationships work.

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner
-   High school characteristics
    -   Dem_Desc
    -   edr
    -   avg.unemp.rate
    -   avg.wage.pct.state
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ACTCompositeScore
    -   ap.exam
    -   sat.taken
    -   MCA scores
        -   MCA.M
        -   MCA.R
        -   MCA.S
    -   cte.achievement
    -   total.cte.courses.taken
-   Post-secondary
    -   attended.ps.within.first.year.hsgrad
    -   attended.ps
    -   ps.grad.
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   Combine post-secondary grad location
        -   ps.grad.in.same.ruca
        -   ps.grad.in.same.edr
        -   ps.grad.in.same.pr
        -   ps.grad.in.MN

<br>

# Meaningful employment in Central Minnesota

```{r prep meaningful employment in Central MN}
meaningful.emp.central <- updated.master %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.1)),
         grad.year.1 = as_factor(grad.year.1),
         grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.5)),
         grad.year.5 = as_factor(grad.year.5),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.10)),
         grad.year.10 = as_factor(grad.year.10),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         highest.cred.level = fct_relevel(highest.cred.level, "No college", "Some college", "Less than Associate Degree", "Associate degree", "Bachelor degree or higher"))

write_csv(meaningful.emp.central, "Data/SLEDS/Masters/After analysis/Master-meaningful-emp-central.csv")
```

Let's first take a look at proportion of states in each of the time.X. The chart below shows that half of the graduates are attending post-secondary (50%) one year after graduating high school. The further away from high school graduation results in a bit of an increase in the percentage of individuals who have meaningful employment in the region and in Minnesota, while about a quarter of graduates do not have a MN employment record and are not going to post-secondary.

```{r proportion of states meaningful emp SW}
prop.states.meaningful.emp.central <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10) %>%
  pivot_longer(names_to = "grad.year", values_to = "states", 1:3) %>%
  group_by(grad.year, states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  filter(states != "After 2023") %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         grad.year = fct_relevel(grad.year, "grad.year.1", "grad.year.5", "grad.year.10"))

prop.states.meaningful.emp.central.plot <- ggplot(prop.states.meaningful.emp.central, aes(grad.year, pct, fill = states, group = states)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nState of individuals: ", states, "\nNumber of people in states: ", comma(n), "\nPercent of people in state: ", percent(pct, accuracy = .1), sep = ""))) +
   labs(x="", y = "", color="", title = "Percent of people in each state")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu"),
                    guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = prop.states.meaningful.emp.central.plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

Let's examine the variables to see if there are any significant differences in the proportion of individuals in these states across the variable categories or values.

<br>

# Demographics

In the demographics section we will examine the relationship between that paths taken by individuals and the following variables;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner

## Summary of all variables

**Strength of Association (Cramer's V by Year)**

| Variable | Year 1 V | Year 5 V | Year 10 V | Peak Strength | Pattern of Change |
|--------------|:--------:|:--------:|:-------:|------------|---------------|
| **Gender** | 0.127 | 0.081 | 0.073 | Moderate (Year 1) | Declines over time |
| **Grad.year.covid** | 0.052 | — | — | Weak (Year 1) | Single-year data only |
| **economic.status** | 0.235 | 0.122 | 0.085 | Strong (Year 1) | Gradual decline |
| **SpecialEdStatus** | 0.248 | 0.145 | 0.084 | Strong (Year 1) | Declines but still meaningful |
| **RaceEthnicity** | 0.059 | 0.041 | 0.038 | Weak (Year 1) | Small but persistent |
| **non.english.home** | 0.101 | 0.073 | 0.066 | Weak–Moderate | Modest decline |
| **english.learner** | 0.139 | 0.084 | 0.068 | Moderate (Year 1) | Persistent, narrowing gap |

**Key Patterns & Comparative Insights**

**1. Strongest Predictors of Outcome Differences**

-   **SpecialEdStatus** and **economic.status** had the **largest effect sizes**, especially one year after graduation. Both were associated with **lower postsecondary participation** and **greater disconnection from meaningful work**, with disparities that persisted across a decade.
-   These two variables reflect **structural barriers** — economic and learning support — that strongly shape early and long-term outcomes.

**2. Language-Related Variables**

-   Both **english.learner** and **non.english.home** showed **moderate effects early on**.
-   Students with these language indicators were **less likely to attend postsecondary** and **more likely to appear in disconnected categories**.
-   While **gaps narrowed over time**, persistent differences suggest long-term barriers related to language acquisition and access.

**3. Gender**

-   Gender had a **moderate short-term effect** that decreased over time.
-   Males were **more likely to take local employment**, while females were **more likely to attend postsecondary**.
-   The gap gradually narrowed but **did not fully disappear** even after 10 years.

**4. RaceEthnicity**

-   Despite its importance, race/ethnicity had **weaker statistical associations** than other factors.
-   However, group-level disparities — especially for **AI, Black, and Hispanic** graduates — were **consistent**.
-   **Asian/PI** students showed unique patterns of **early postsecondary participation** and **underrepresentation in local employment**, possibly due to out-of-state or non-traditional career paths.

**5. COVID-era Graduation**

-   Graduation year relative to COVID had a **statistically significant but weak** effect.
-   Post-COVID grads were **less likely to attend PS** and **more likely to work locally**, likely due to **pandemic-era constraints**.

**Conclusion**

Among all the demographic variables, **economic disadvantage and special education status** emerged as the **most powerful and persistent predictors** of postsecondary and employment outcomes. **Language and gender** also played meaningful roles, particularly in the short term, while **race/ethnicity** and **COVID-era graduation** had weaker but still important associations.

These findings suggest that interventions aiming to improve post-high school pathways should prioritize support for **low-income students**, **students receiving special education**, and **multilingual or English learner populations**, especially in the critical first years after graduation.

<br>

## Gender

### **Key Findings**

-   There is a statistically significant relationship between gender and post-graduation outcomes in all three timeframes:
    -   **1 year after graduation**: χ² = 2083.575, df = 4, Cramer's V = **0.127**, *p* \< .001
    -   **5 years after graduation**: χ² = 612.448, df = 4, Cramer's V = **0.081**, *p* \< .001
    -   **10 years after graduation**: χ² = 278.976, df = 4, Cramer's V = **0.073**, *p* \< .001

The strength of association is **moderate in the first year** and diminishes over time to a **weak but still significant** relationship by year ten.

### **Interpretation Over Time**

**1 Year After Graduation**: Females are more likely to be **attending postsecondary education** (55.3% vs. 43.7%), while males are more likely to be employed — especially **within their high school region** (15.3% vs. 10.9%). Males also have higher rates of **no MN employment record** (11.9% vs. 7.7%). These early differences reflect a pattern in which **males enter the local workforce more often**, while **females pursue longer-term educational pathways**.

**5 Years After Graduation**: Postsecondary enrollment declines across the board, but differences remain. Males continue to have higher rates of **regional employment** and are slightly more likely to be employed meaningfully somewhere in MN. The gender gap in disconnected outcomes narrows, suggesting some convergence over time.

**10 Years After Graduation**: Differences persist but are smaller. Males still maintain an edge in meaningful employment — both **in-region and statewide** — while females are slightly more likely to appear in the **"no MN record"** group, possibly due to moving out of state or untracked employment.

### **Conclusion**

Gender-based differences are **most pronounced in the first year**, with males more likely to take **local employment routes** and females more likely to pursue postsecondary education. Over time, the differences **narrow but remain**, especially in terms of where employment is located and whether individuals are tracked in the MN employment system.

<br>

::: panel-tabset
### One year after graduating

```{r demographics gender one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Gender) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)

data1 <- data
var1 <- data$grad.year.1
```

### Five years after graduating

```{r demographics gender five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Gender) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)

data5 <- data
var5 <- data$grad.year.5
```

### Ten years after graduating

```{r demographics gender ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Gender) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)

data10 <- data
var10 <- data$grad.year.10

```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, Gender) %>%
  write_csv(., "Data/ChatGPT/Demographics/Gender.csv")
```

<br>

## Grad.year.covid

### **Key Findings**

There is a statistically significant relationship between graduation timing and outcomes **one year after high school** (χ² = 349.128, df = 4, *p* \< .001). The strength of association is **weak**, with a Cramer's V of **0.052**.

### **Interpretation (1 Year After Graduation)**

Graduates who finished high school **after the onset of COVID-19 (2020 or later)** were **less likely to be attending postsecondary education** (45.5% vs. 50.5%) and **more likely to be employed**, particularly in-region (15.0% vs. 12.6%) or statewide (9.2% vs. 7.1%). They also had slightly higher rates of being untracked or disconnected (10.6% with no MN record vs. 9.6%).

This pattern suggests that **COVID-era graduates may have faced barriers to postsecondary access** or opted more often for immediate employment, perhaps due to economic pressures, remote learning challenges, or changes in job availability during the pandemic.

<br>

::: panel-tabset
### One year after graduating

```{r demographics grad.year.covid one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.covid) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$grad.year.covid, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, grad.year.covid) %>%
  write_csv(., "Data/ChatGPT/Demographics/grad-year-covid.csv")
```

<br>

## economic.status

### **Key Findings**

There is a statistically significant relationship between economic status and post-graduation outcomes:

-   **1 year after graduation**: χ² = 7116.172, df = 4, Cramer's V = **0.235**, *p* \< .001
-   **5 years after graduation**: χ² = 1409.165, df = 4, Cramer's V = **0.122**, *p* \< .001
-   **10 years after graduation**: χ² = 373.783, df = 4, Cramer's V = **0.085**, *p* \< .001

The strength of association is **moderate to strong in the first year** and gradually weakens over time, though differences persist.

### **Interpretation Over Time**

**1 Year After Graduation**: Students not eligible for free/reduced lunch were **far more likely to attend postsecondary education** (57.4% vs. 32.6%). Students from low-income backgrounds were more likely to be employed **locally in their high school region** (16.2% vs. 11.7%) and more likely to be disconnected — either in non-meaningful roles (28.5%) or lacking MN employment records (13.5%).

**5 Years After Graduation**: Gaps narrow slightly but remain. Higher-income students continue to show **higher rates of meaningful MN employment**, while low-income students maintain **elevated disconnection** and are more concentrated in **regional jobs**, likely reflecting limited mobility or economic necessity.

**10 Years After Graduation**: Regional employment rates are similar, but higher-income students are still more likely to have **meaningful jobs elsewhere in MN** and **less likely to be disconnected**. The long-term trend suggests that **early disparities in education and employment compound over time**, even as the gap narrows.

### **Conclusion**

Economic status is a **powerful predictor** of early postsecondary and employment outcomes. Students from low-income backgrounds are more likely to take **local jobs out of necessity**, less likely to pursue postsecondary education, and more vulnerable to **long-term disconnection from the workforce**. While disparities decrease by year ten, they remain meaningful.

<br>

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a higher rate of meaningful employment in the region.
-   Have a higher rate of meaningful employment in Minnesota.
-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, economic.status) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, economic.status) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a lower rate of having meaningful employment in Minnesota.
-   Have a higher rate of not having meaningful employment.

```{r demographics economic.status ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, economic.status) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, economic.status) %>%
  write_csv(., "Data/ChatGPT/Demographics/economic-status.csv")

```

<br>

## SpecialEdStatus

### **Key Findings**

There is a statistically significant relationship between SpecialEdStatus and post-graduation outcomes across all three years:

-   1 **year after graduation**: χ² = 3177.82, df = 4, Cramer's V = **0.248**, *p* \< .001
-   **5 years after graduation**: χ² = 1091.37, df = 4, Cramer's V = **0.145**, *p* \< .001
-   **10 years after graduation**: χ² = 366.90, df = 4, Cramer's V = **0.084**, *p* \< .001

The strength of association is **strongest one year after graduation** and diminishes over time, though remains statistically significant even ten years later.

### **Interpretation Over Time**

**1 Year After Graduation**: Students who received special education services in high school were substantially less likely to be **attending postsecondary education** (only 1,375 compared to 24,769 peers without special ed status) and more likely to be in **non-meaningful or untracked employment categories**, including **“no MN employment record”** and **“not meaningful”** employment. While some were employed locally (838 in-region), the disparities suggest substantial barriers to immediate postsecondary success.

**5 Years After Graduation**: The gap narrows but persists. Special education students still trail their peers in **postsecondary enrollment** and are **less likely to hold meaningful employment in MN** (1,068 vs. 11,204) or **in their local region** (1,529 vs. 9,866). They are overrepresented in **non-meaningful** and **no-record** categories, though many do gain some form of employment by this point.

**10 Years After Graduation**: By year ten, special education students continue to be **underrepresented in meaningful MN employment** (1,373 vs. 14,267) and **more likely to have no MN employment record** or to be in a non-meaningful status. Postsecondary participation is negligible for both groups at this point. The long-term pattern reflects **lasting disparities in workforce outcomes**, though gaps are smaller than in earlier years.

### **Conclusion**

Special education status is **strongly associated with post-graduation outcomes**, especially in the first year. Students with special education histories are consistently **less likely to attend post secondary education** and more likely to experience **disconnection from the workforce** over time. While some individuals do gain meaningful regional or statewide employment, the **long-term opportunity gap remains**, pointing to the need for sustained support well beyond high school.

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, SpecialEdStatus) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   slightly lower rate of having meaningful employment in Minnesota
-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, SpecialEdStatus) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   SIGNIFICANTLY lower rate of having meaningful employment in Minnesota
-   SIGNFICANTLY higher rate of not having meaningful employment

```{r demographics SpecialEdStatus ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, SpecialEdStatus) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, SpecialEdStatus) %>%
  write_csv(., "Data/ChatGPT/Demographics/economic-status.csv")

```

<br>

## RaceEthnicity

### **Key Findings**

There is a statistically significant relationship between race/ethnicity and post-graduation outcomes at all time points:

-   **1 year after graduation**: χ² = 719.74, df = 16, Cramer's V = **0.059**, *p* \< .001
-   **5 years after graduation**: χ² = 352.07, df = 16, Cramer's V = **0.041**, *p* \< .001
-   **10 years after graduation**: χ² = 301.57, df = 16, Cramer's V = **0.038**, *p* \< .001

The association is **statistically significant but weak** at each time point and **decreases slightly over time**.

### **Interpretation Over Time**

**1 Year After Graduation**: White graduates were the most likely to be **attending postsecondary education** (24,849), while other groups — especially American Indian (AI) and Hispanic students — were more likely to be in **non-meaningful employment** or have **no MN employment record**. White students also had the highest counts in **regional meaningful employment** (5,742), while the counts for all other groups were much lower.

**5 Years After Graduation**: The overall pattern persists. White graduates had higher levels of **meaningful MN employment** and **regional employment**, while American Indian, Black, and Hispanic graduates were more concentrated in **non-meaningful** or **disconnected categories**. Postsecondary enrollment declined across all groups, but disparities in employment outcomes remained.

**10 Years After Graduation**: Postsecondary attendance was nearly zero across the board. White graduates continued to lead in **meaningful MN and regional employment**, while American Indian, Black, and Hispanic graduates had **higher rates of disconnection** (no MN record or not meaningful employment). The differences were smaller than in earlier years but still evident.

### **Conclusion**

Racial and ethnic differences in post-graduation outcomes are **present across all time points**, though the **overall effect size is weak**. White graduates are consistently more likely to be enrolled in postsecondary education and employed meaningfully in both **statewide and local contexts**, while graduates identifying as American Indian, Black, or Hispanic are **overrepresented in disconnected and non-meaningful employment categories**. These patterns **persist over time**, albeit with **diminishing strength**, highlighting enduring disparities in opportunity and access.

::: panel-tabset
### One year after graduating

```{r demographics RaceEthnicity one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, RaceEthnicity) %>%
  filter(grad.year.1 != "After 2023",
         RaceEthnicity != "Unknown") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r demographics RaceEthnicity five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, RaceEthnicity) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r demographics RaceEthnicity ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, RaceEthnicity) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, RaceEthnicity) %>%
  write_csv(., "Data/ChatGPT/Demographics/RaceEthnicity.csv")

```

<br>

## non.english.home

### **Key Findings**

There is a statistically significant relationship between having a non-English speaker at home and post-graduation outcomes:

-   **1 year after graduation**: χ² = 529.90, df = 4, Cramer's V = **0.101**, *p* \< .001
-   **5 years after graduation**: χ² = 275.73, df = 4, Cramer's V = **0.073**, *p* \< .001
-   **10 years after graduation**: χ² = 228.43, df = 4, Cramer's V = **0.066**, *p* \< .001

The association is **consistently weak but statistically meaningful**, with the strongest divergence seen one year after graduation.

### **Interpretation Over Time**

**1 Year After Graduation**: Students from households where a non-English language is spoken were significantly **less likely to be attending postsecondary education** (472 vs. 25,672) and **more likely to be in disconnected categories**, such as **no MN employment record** or **non-meaningful employment**. They were also underrepresented in **meaningful employment**, both locally and statewide.

**5 Years After Graduation**: The gap remains but narrows. Most graduates, regardless of home language, have transitioned out of postsecondary education. However, students from non-English-speaking households continue to have **lower representation in meaningful MN and regional employment** and are **slightly more likely to be disconnected**.

**10 Years After Graduation**: Postsecondary enrollment disappears entirely for this group, and differences in employment outcomes persist. Those from English-speaking households are still more likely to be in **meaningful MN employment** and **less likely to be in disconnected categories**.

### **Conclusion**

Having a non-English speaker in the home is **associated with lower postsecondary participation and increased early disconnection** from employment. While the differences shrink over time, disparities persist — particularly in access to **meaningful employment opportunities**, both regionally and statewide. These findings point to **language and cultural barriers** that may limit immediate access to education and career pathways following high school.

::: panel-tabset
### One year after graduating

```{r demographics non.english.home one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, non.english.home) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r demographics non.english.home five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, non.english.home) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r demographics non.english.home ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, non.english.home) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, non.english.home) %>%
  write_csv(., "Data/ChatGPT/Demographics/non-english-home.csv")

```

<br>

## english.learner

### **Key Findings**

There is a statistically significant relationship between being an English Learner in high school and post-graduation outcomes:

-   **1 year after graduation**: χ² = 1006.49, df = 4, Cramer's V = **0.139**, *p* \< .001
-   **5 years after graduation**: χ² = 362.71, df = 4, Cramer's V = **0.084**, *p* \< .001
-   **10 years after graduation**: χ² = 237.28, df = 4, Cramer's V = **0.068**, *p* \< .001

The strength of association is **moderate one year after graduation** and becomes **weaker but still statistically meaningful** over time.

### **Interpretation Over Time**

**1 Year After Graduation**: Students identified as English Learners were **significantly less likely to be attending postsecondary education** (407 vs. 25,737) and **more likely to be in disconnected or non-meaningful employment states**. Only 317 total English Learners were employed meaningfully either in the region or in MN. Over 1,000 were either disconnected or untracked in employment records.

**5 Years After Graduation**: Many English Learners begin to transition into employment, but gaps persist. Only 285 had meaningful employment in MN and 342 in the region — far fewer than their non-EL peers. They remain **underrepresented in meaningful outcomes** and **overrepresented in disconnection**, suggesting ongoing barriers to long-term opportunity.

**10 Years After Graduation**: Disparities shrink but do not disappear. Just over 600 English Learners were meaningfully employed, while more than 1,100 were still in non-meaningful or untracked categories. Postsecondary enrollment had disappeared entirely. In contrast, non-EL students showed **higher levels of workforce engagement**, especially in MN-wide employment.

### **Conclusion**

English Learner status is a **meaningful predictor of post-graduation outcomes**, especially in the short term. These students face **substantial early disadvantages** in both education and employment, with limited access to **meaningful opportunities** in the region or state. Although some gaps close over time, English Learners remain **more disconnected from the workforce** than their peers a full decade after graduation — signaling a persistent equity concern.

::: panel-tabset
### One year after graduating

```{r demographics english.learner one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, english.learner) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r demographics english.learner five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, english.learner) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r demographics english.learner ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, english.learner) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, english.learner) %>%
  write_csv(., "Data/ChatGPT/Demographics/english-learner.csv")

```

<br>

# High school characteristics

We will take a look at the following variables;

-   Dem_Desc
-   edr
-   avg.unemp.rate
-   avg.wage.pct.state

## Summary

<br>

## Dem_Desc

::: panel-tabset
### One year after graduating

```{r hs characteristics Dem_Desc one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Dem_Desc) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs characteristics Dem_Desc five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Dem_Desc) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs characteristics Dem_Desc ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Dem_Desc) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, Dem_Desc) %>%
  write_csv(., "Data/ChatGPT/High school characteristics/Dem-Desc.csv")
```

<br>

## EDR

There is a statistically significant association between an graduate's EDR and their employment state, but the association was negligible.

Through all the years, graduates from EDR 5 had an increasingly higher rate of meaningful employment in the region and a significantly higher rate of not having a MN employment rate.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between a graduate's EDR and the employment state.

Graduates from;

-   EDR 5 had a slightly higher rate ofhaving meaningful employment in the region.
-   EDR 5 had a slightly lower rate of attending post-secondary.

```{r hs characteristics edr one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, edr) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between a graduate's EDR and the employment state.

Graduates from;

-   EDR 5 had a higher rate of having meaningful employment in the region.
-   EDR 5 had a lower rate of having meaningful employment in Minnesota.
-   EDR 7E had a higher rate of having meaningful employment in Minnesota
-   EDR 5 had a higher rate of not having a MN employment rate.

```{r hs characteristics edr five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, edr) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between a graduate's EDR and the employment state.

Graduates from;

-   EDR 5 had a higher rate of having meaningful employment in the region.
-   EDR 5 had a significantly lower rate of having meaningful employment in Minnesota.

```{r hs characteristics edr ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, edr) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## avg.unemp.rate

There is a statistically significant association between an the unemployment rate of a school district's county and a graduates employment state.

Through all the years, the main trend was that graduates without meaningful employment or no MN employment record came from schools with significantly higher unemployment rates than graduates that have meaningful employment in the region or Minnesota, or attended post-secondary.

::: panel-tabset
The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with 30%+ higher unemployment rates than graduates that meaningful employment in the region or Minnesota.
-   had no meaningful employment or no MN employment record came from districts with 20%+ higher unemployment rates than graduates attending post-secondary.

```{r hs characteristics avg.unemp.rate one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.unemp.rate) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.unemp.rate, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 20))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)
```

### Five years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with over 10% higher unemployment rates.
-   had no meaningful employment came from districts with 12% higher unemployment rates than graduates that are attending post-secondary.

```{r hs characteristics avg.unemp.rate five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.unemp.rate) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.unemp.rate, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or MN employment record came from districts with 18% higher unemployment rates compared to graduates with meaningful employment in the region or in MN.
-   had no MN employment record came from districts with 11% lower unemployment rates than graduates with no meaningful employment.

```{r hs characteristics avg.unemp.rate ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.unemp.rate) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.unemp.rate, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## avg.wages.pct.state

There is a statistically significant association between an the unemployment rate of a school district's county and a graduates employment state.

Through all the years, the main trend was that graduates without meaningful employment or no MN employment record came from schools with significantly higher unemployment rates than graduates that have meaningful employment in the region or Minnesota, or attended post-secondary.

::: panel-tabset
### One year after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with -1% difference between the percent of state wages of graduates that have meaningful employment in the region or Minnesota.
-   had meaningful employment in MN or were attending post-secondary came from districts with a 1.2% difference in the percent of state wages of graduates who had meaningful employment in the region.

```{r hs characteristics avg.wages.pct.state one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.wages.pct.state) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.wages.pct.state, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with -1% difference between the percent of state wages of graduates that have meaningful employment in the region or Minnesota.
-   had meaningful employment in MN or were attending post-secondary came from districts with a 1.2% difference in the percent of state wages of graduates who had meaningful employment in the region.

```{r hs characteristics avg.wages.pct.state five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.wages.pct.state) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.wages.pct.state, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

The p-value indicates a significant difference between one of the groups.

Graduates that;

-   had no meaningful employment or no MN employment record came from districts with -5%+ difference between the percent of state wages of graduates that have meaningful employment in Minnesota.
-   had meaningful employment in MN or were attending post-secondary came from districts with a 1.5% difference in the percent of state wages of graduates who had meaningful employment in the region.

```{r hs characteristics avg.wages.pct.state ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.wages.pct.state) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.wages.pct.state, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

## ChatGPT

This is the summary provided by ChatGPT when asked to analyze and interpret.

### Urbanization Level (`dem_desc`)

Rurality of the high school district showed a small but growing influence on employment outcomes over time. Graduates from **town/rural mix** districts were more likely to have **regional employment** and less likely to attend **postsecondary** one year after graduation. Conversely, those from **entirely urban** districts were more likely to be enrolled in postsecondary or employed within Minnesota. The strength of this association increased slightly by year 10.\

### Economic Development Region (`edr`)

The high school’s economic region had a consistent but weak association with employment outcomes. Students from **EDR 7W (Central)** were more likely to attend postsecondary or secure **MN-based employment**, while those from **EDR 5 (North Central)** were more likely to be employed **regionally** or have **no MN employment record**. These patterns remained stable from year 1 through year 10.\

### County Unemployment Rate (`avg.unemp.rate`)

County-level unemployment rates had a modest impact. Higher unemployment was associated with a slight **increase in postsecondary attendance** and a **decrease in meaningful employment**, particularly within Minnesota. While these effects were small, they remained consistent across all three time points.\

### County Wages as % of State Average (`avg.wages.pct.state`)

This variable was the most predictive across the board. Individuals from **higher-wage counties** were significantly more likely to have **meaningful employment within Minnesota**, even years after graduation. Conversely, they were less likely to attend postsecondary school immediately after graduation. This relationship was both strong and persistent across years 1, 5, and 10.\

### Overall Insights

Among the four variables, **county wage levels** stood out as the most influential predictor of long-term employment outcomes. While **urbanization** and **economic region** showed weak but meaningful patterns, their predictive power was notably lower. **Unemployment rate** had a smaller and more stable effect, influencing early decisions such as whether to work or pursue further education.

```{r high school characteristics chatgpt}
hs.characteristics.chatgpt <- meaningful.emp.central %>%
  select(PersonID, Dem_Desc, edr, avg.unemp.rate, avg.wages.pct.state, grad.year.1, grad.year.5, grad.year.10)

write_csv(hs.characteristics.chatgpt, "Data/ChatGPT/hs-characteristics-chatgpt.csv")
```

<br>

# High School Accomplishments

We are going to analyze the following variables;

-   pseo.participant
-   took.ACT
-   ACTCompositeScore
-   ap.exam
-   sat.taken
-   MCA scores
    -   MCA.M
    -   MCA.R
    -   MCS.S
-   cte.achievement
-   total.cte.courses.taken

## **Summary**

### **Categorical Predictors (Cramer's V):**

**1 year after graduation**\
High school factors showed **moderate associations** with immediate outcomes. ACT participation (**V = 0.418**) and PSEO participation (**V = 0.355**) were the strongest predictors, followed by AP exam participation (**V = 0.250**). Most other categorical indicators had **small but statistically significant** associations (V \< 0.2).

**5 years after graduation**\
The strength of associations **generally weakened**, though ACT (**V = 0.215**) and PSEO (**V = 0.149**) continued to relate to outcomes. AP participation dropped further (**V = 0.156**), suggesting a diminishing link to mid-term employment or education status.

**10 years after graduation**\
By 10 years post-graduation, associations were **consistently weak**. ACT (**V = 0.131**), AP (**V = 0.100**), and PSEO (**V = 0.056**) still had statistically significant but **limited predictive power**.

### **Continuous Predictors (ANOVA):**

**ACT Composite Score**\
ACT scores were **strongly predictive** of outcomes 1 year after graduation (**F = 1983**), particularly distinguishing students in postsecondary from other groups. This effect **gradually diminished** over time (5-year **F = 573**, 10-year **F = 244**), though still significant.

**MCA Scores (Reading, Math, Science)**\
All three MCA subject scores were **significantly associated** with postsecondary or employment outcomes across all time points. Math and reading scores showed the **strongest associations** initially, while science followed a similar trend of **gradual decline** in predictive strength over time.

**CTE Course Participation**\
The number of CTE courses taken showed a **moderate association** with outcomes in the short term (**F = 1312** at 1 year). Importantly, students with higher CTE participation were more likely to be **employed in their region**, even if they were not attending college. This pattern was **sustained over time**—with significant differences still present at 5 years (**F = 447**) and 10 years (**F = 208**)—suggesting that CTE coursework supports **earlier and more stable regional employment** pathways, particularly for students who may not pursue a four-year degree.

### **Interpretation over time**

**1 year after graduation**\
Academic preparation and college-aligned behaviors (ACT, AP, PSEO) were **strongly tied** to outcomes, particularly college enrollment. However, students with more CTE coursework were **disproportionately represented** in **meaningful regional employment**, highlighting CTE's role in connecting youth to the local labor market.

**5 years after graduation**\
High school academics remained relevant, though to a lesser extent. CTE participation continued to show a **discernible link** to regional employment, offering **a parallel success pathway** for students not pursuing higher education.

**10 years after graduation**\
While academic indicators had largely faded in influence, CTE participation retained a **residual positive effect**, particularly among those stably employed in their home region. This suggests that CTE may play a role in **long-term regional retention and workforce alignment**.

### **Conclusion**

High school academic indicators—ACT scores, advanced coursework, and MCA assessments—are **most predictive** of early outcomes but **fade over time**. In contrast, **Career and Technical Education courses appear to offer a more durable connection to local employment**, particularly for non-college-bound youth. CTE supports **regional labor market integration**, potentially providing a stable foundation for long-term success without requiring immediate postsecondary enrollment. These findings underscore the importance of maintaining **diverse, high-quality pathways** in high school that support both college and career readiness.

<br>

## pseo.participant

This analysis examines how participation in **Postsecondary Enrollment Options (PSEO)** is associated with high school graduates’ education and employment outcomes at **1, 5, and 10 years** after graduation.

**Key Findings**

-   The association between PSEO participation and outcomes is **statistically significant across all three time points**, though the **strength of the relationship declines over time**.
    -   **Year 1**: Cramér’s V = 0.355 (**strong**)
    -   **Year 5**: Cramér’s V = 0.149 (**moderate**)
    -   **Year 10**: Cramér’s V = 0.056 (**weak**)

**Interpretation Over Time**

-   **Year 1**
    -   PSEO participants were **much more likely** to be enrolled in post-secondary education.
    -   They were **less likely** to be working locally or to be disconnected from both education and work.
    -   Non-participants were more concentrated in **regional employment** or **not meaningfully engaged**.
-   **Year 5**
    -   PSEO students continued to show **higher post-secondary persistence** and **greater geographic mobility** in employment.
    -   Differences in disconnection and local employment persisted but were **less pronounced** than in Year 1.
-   **Year 10**
    -   Differences between participants and non-participants **narrowed considerably**.
    -   PSEO participants remained slightly more likely to be **employed outside their home region** and **less likely to be disconnected**.
    -   The gap in regional employment and disconnection had **mostly leveled out**.

**Conclusion**

PSEO participation is strongly associated with **post-secondary enrollment and broader employment opportunities** in the short and medium term. By ten years after high school, **the influence of PSEO participation diminishes**, but subtle differences in employment patterns and disconnection remain.

::: panel-tabset
### One year after graduating

```{r hs accomplishments pseo.participant one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, pseo.participant) %>%
  filter(grad.year.1 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments pseo.participant five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, pseo.participant) %>%
  filter(grad.year.5 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments pseo.participant ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, pseo.participant) %>%
  filter(grad.year.10 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## took.ACT

This analysis examines how **taking the ACT exam** relates to graduates’ employment and education outcomes across three time points: one, five, and ten years after high school.

**Key Findings**\

The association between ACT participation and outcomes is **statistically significant** at all three time points, with **declining strength over time**:\

**Year 1**: Cramér’s V = 0.418 (**strong**)\

**Year 5**: Cramér’s V = 0.215 (**moderate**)\

**Year 10**: Cramér’s V = 0.131 (**modest**)

**Interpretation Over Time**\

In **Year 1**, ACT takers were far more likely to be enrolled in **post-secondary education**. They were less likely to be **disconnected** from school or work. Non-takers were more likely to be in **regional employment** or **not meaningfully engaged**.

By **Year 5**, ACT takers continued to show higher rates of **post-secondary participation** and were more likely to have **employment outside their home region**. Non-takers were more concentrated in **regional employment** or in disconnected categories. The strength of the association declined but remained meaningful.

In **Year 10**, ACT takers were still more likely to be employed **outside their region**. The difference in **disconnection rates** narrowed, but ACT takers remained somewhat more **engaged overall**. Non-takers continued to be more likely to work **locally** or lack meaningful engagement.

**Conclusion**\

ACT participation is a **strong early indicator** of both **post-secondary enrollment** and **long-term employment mobility**. While the strength of the association decreases over time, ACT takers remain more likely to be **employed beyond their home region** and **less likely to be disconnected**, even a decade after graduation.

::: panel-tabset
### One year after graduating

```{r hs accomplishments took.ACT one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, took.ACT) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments took.ACT five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, took.ACT) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments took.ACT ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, took.ACT) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ACTCompositeScore

**Summary**

This summary interprets the differences in **ACT Composite Scores** across employment outcomes one, five, and ten years after high school graduation, based on ANOVA and Tukey's HSD tests.

**Year 1**\

One year after high school, **graduates attending post-secondary** had the highest ACT scores, averaging **3.32 points higher** than those in meaningful employment in their home region.\

Those with **no Minnesota employment record** scored **1.38 points higher**, while those with **not meaningful employment and not attending post-secondary** scored **0.58 points higher** than the regional employed group.\

Graduates with **meaningful employment outside their home region** scored modestly higher (+0.31).\

Overall, ACT scores were **strongly associated with post-secondary enrollment and mobility**, suggesting that higher academic achievement was linked to educational continuation and broader employment opportunities.

**Year 5**\

By five years after high school, the pattern remained but with **smaller score gaps**.\

Graduates attending post-secondary still had **the highest average ACT scores**, scoring **2.49 points higher** than those in regional employment.\

Those with **no MN employment record** and those with **meaningful employment outside the region** also scored significantly higher (+2.00 and +1.97, respectively).\

Notably, the score differences between local and mobile employment groups persisted, suggesting continued alignment between **higher ACT scores and greater geographic or career mobility**.

**Year 10**\

Ten years after high school, the gap narrowed further but **remained statistically significant**.\

Graduates with **no MN employment record** had the highest ACT scores, **1.75 points above** those employed in their home region.\

Those with **meaningful employment elsewhere in Minnesota** scored **1.44 points higher**, and those **not meaningfully employed and not attending post-secondary** scored **0.66 points higher** than the regional group.\

These results indicate a **long-term association between higher ACT scores and mobility or non-local employment**, though the magnitude of differences diminishes over time.

**Conclusion**\

Across all three time points, **ACT Composite Scores consistently distinguished employment outcomes**.\

Higher scores were associated with **continued education, geographic mobility**, and employment outside one’s high school region.\

Lower scores were most common among those with **local employment**, suggesting that **academic achievement influences both post-secondary trajectories and regional employment patterns well into adulthood**.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ACTCompositeScore one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ACTCompositeScore) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, ACTCompositeScore, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments ACTCompositeScore five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ACTCompositeScore) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, ACTCompositeScore, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments ACTCompositeScore ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ACTCompositeScore) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, ACTCompositeScore, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 40))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## ap.exam

**Key Findings**\

**One year after high school**, there was a **moderate association** between AP exam participation and post-graduation outcomes (Cramer's V = 0.250, p \< .001).\

**Five years after high school**, the association **weakened slightly** but remained **meaningful** (Cramer's V = 0.156, p \< .001).\

**Ten years after high school**, the association **diminished further**, indicating a **weaker but still statistically significant** relationship (Cramer's V = 0.100, p \< .001).

**Interpretation Over Time**\

**One year after graduation**, AP exam takers were far more likely to be enrolled in post-secondary education (75.7%) compared to non-takers (43.5%). They were also underrepresented in all employment categories, especially in regional and meaningful employment.\

**Five years after graduation**, AP exam participants still had higher post-secondary enrollment than non-participants, though the gap narrowed. They were more likely to have meaningful employment in Minnesota (29.7% vs. 23.9%) and were underrepresented in low-engagement categories.\

**Ten years after graduation**, AP exam participants continued to have higher rates of meaningful employment in Minnesota (37.5% vs. 24.5%) and were less likely to fall into the “not meaningful, not attending post-secondary” category. The trend suggests **strong early academic orientation continues to correlate with more favorable employment outcomes** over time.

**Conclusion**\

AP exam participation is associated with **stronger early post-secondary engagement** and **long-term advantages** in meaningful employment, especially within Minnesota. While the strength of the relationship decreases over time, the overall pattern suggests **enduring benefits** for students who participate in AP coursework and exams.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ap.exam one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ap.exam) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments ap.exam five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ap.exam) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments ap.exam ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ap.exam) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## MCA.R

**Key Findings**\

There were statistically significant differences in MCA Reading scores across employment states in all three years after graduation:

**Year 1:** F(4, 119,756) = 3471, p \< .001\

**Year 5:** F(4, 84,266) = 426.6, p \< .001\

**Year 10:** F(3, 47,064) = 750.3, p \< .001

**In all three years, students attending post-secondary had the highest average MCA Reading scores.** Those with no Minnesota employment record or who were not meaningfully employed consistently had the lowest scores. Graduates with meaningful employment in their home region generally had **lower scores** than those employed meaningfully outside their region (elsewhere in Minnesota).

**Interpretation Over Time**\

**One Year After Graduation:** Students attending post-secondary scored **0.33 points higher** than those with meaningful employment in their home region. Those with no MN employment record scored **0.32 points lower**, while the not-meaningfully-employed group scored **0.25 points lower** than the home-region group.

**Five Years After Graduation:** The attending post-secondary group continued to have higher reading scores (**0.28 points above** the home-region group). Meaningfully employed graduates outside their home region scored **0.18 points higher**, while the no MN employment group scored **0.13 points higher**. Not-meaningfully-employed graduates had a small but significant edge (**0.04 points higher**) than the home-region group.

**Ten Years After Graduation:** The trend intensified. Those employed meaningfully outside their region scored **0.37 points higher** than the in-region group, and those with no MN employment record scored similarly (**0.37 points higher**). However, the difference between the in-region and not-meaningfully-employed group was **not significant**, suggesting this group's reading achievement had converged over time.

**Conclusion**\
Across all three post-graduation points, MCA Reading scores were **strongly associated** with future employment states. The most consistent pattern was that **graduates who pursued post-secondary education or secured employment beyond their home region had higher reading scores**, suggesting both academic preparation and geographic mobility play important roles in post-school success. Conversely, **graduates disconnected from education or stable employment tended to have lower scores**, though these differences became less pronounced by year ten.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.R one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.R) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.R, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.R five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.R) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.R, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.R ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.R) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.R, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.S

**Key Findings**\

**Year 1**: ANOVA results revealed significant differences in MCA Science scores across employment outcomes (*F* = 2233, *p* \< .001). Graduates attending post-secondary scored **0.45 points higher** on average than those with meaningful employment in their high school region. Those with no MN employment record and not attending post-secondary scored **0.32 points lower** than the regional employment group, and those not meaningfully employed or attending post-secondary scored **0.25 points lower**.\

**Year 5**: Differences remained statistically significant (*F* = 498.4, *p* \< .001). Post-secondary attendees scored **0.39 points higher** than those employed in the region. Those with no MN employment record scored **0.17 points higher**, while those not meaningfully employed or in post-secondary were **not significantly different** from the regional employment group.\

**Year 10**: Differences continued to be significant (*F* = 168.6, *p* \< .001). Those meaningfully employed elsewhere in Minnesota scored **0.23 points higher**, and those with no MN employment record scored **0.18 points higher** than the regional employment group. Those not meaningfully employed nor attending post-secondary were **not significantly different** from the regional group.

**Interpretation Over Time**\

**One year after high school**, higher MCA Science scores were **strongly associated** with attending post-secondary and **moderately associated** with meaningful employment in other parts of Minnesota.\

**Five years after high school**, this pattern **persisted**, though differences between most groups **narrowed**. Post-secondary students and those with no MN employment record continued to score higher, but the gaps were **less pronounced**.\

**Ten years after high school**, academic achievement as measured by MCA Science remained **predictive**, but differences across employment groups **weakened slightly**, particularly for those not meaningfully employed.

**Conclusion**\

Science achievement in high school, as measured by the MCA, is **consistently associated** with more favorable postsecondary and employment outcomes. While the strength of these associations **diminishes over time**, they remain **statistically and practically significant** well into the first decade after graduation.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.S five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.S) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.S, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.S ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.S) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.S, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.M

**Key Findings**\

Statistically significant differences were found in MCA Math scores across post-graduation employment groups in all three years:\

**1 year after high school**: F(4, 119,783) = 5009, p \< .001\

**5 years after high school**: F(4, 89,361) = 1027, p \< .001\

**10 years after high school**: F(3, 49,527) = 244.9, p \< .001

**Interpretation Over Time**\

**1 Year After Graduation** Graduates attending post-secondary education had the **highest MCA Math scores**, averaging 0.64 points higher than those with meaningful employment in the region. Those not meaningfully employed or without a Minnesota employment record scored significantly **lower**, with differences ranging from -0.13 to -0.19 points. The smallest gap was between those employed in Minnesota outside the region and those employed in the region (+0.04), though still statistically significant.

**5 Years After Graduation**\
Differences persisted, though gaps **narrowed slightly**. Postsecondary attendees remained the highest scorers (+0.60 vs. regional employment), while those not meaningfully employed scored lower (-0.05 to -0.17). Interestingly, those with no MN employment record scored **slightly higher** than those employed in-region (+0.10), a reversal from year one.

**10 Years After Graduation**\
The trend largely held. Postsecondary attendees were no longer a distinct group in the model, but those employed in Minnesota outside their region still scored **higher** (+0.29) than regional workers. Those not meaningfully employed showed **no statistically significant difference** from regional workers (-0.01, p = 0.91), and those with no MN record scored **modestly higher** (+0.22).

**Conclusion** MCA Math scores were **strongly associated** with post-graduation outcomes early on, especially for those pursuing higher education. Over time, score differences **remained**, especially between regional employment and out-of-region employment, but **gaps narrowed**, particularly among those not attending post-secondary or not meaningfully employed.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.M one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.M) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.M, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.M five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.M) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.M, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.M ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.M) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.M, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## cte.achievement

This section summarizes the association between **Career and Technical Education (CTE) achievement** and **employment or education outcomes** at one, five, and ten years after high school. While all three chi-square tests were statistically significant, the **Cramér’s V values** suggest a consistently **very weak relationship**:

-   **Year 1:** Cramér’s V = 0.091
-   **Year 5:** Cramér’s V = 0.069
-   **Year 10:** Cramér’s V = 0.054

**Key trends over time:**

**Year 1:**

-   **CTE Concentrators/Completers** were more likely to enter **regional or meaningful MN employment** early.
-   **Non-CTE students** were more likely to pursue **post-secondary education**, with over half still enrolled one year after graduation.
-   Early workforce engagement was more common among those with higher CTE involvement.

**Year 5:**

-   Differences began to narrow, but **CTE Concentrators/Completers** remained more likely to be in **regional or MN employment**.
-   **Non-CTE students** continued to show higher levels of **post-secondary engagement**, though to a lesser extent than at Year 1.
-   The association between CTE status and outcomes weakened overall.

**Year 10:**

-   Outcome distributions became **similar across all CTE groups**.
-   **Meaningful MN employment** was the most common outcome regardless of CTE status (\~30%).
-   **Non-CTE students** were slightly more likely to have **no MN employment record**, suggesting possible migration or untracked employment.

**Conclusion:**\
CTE achievement is **modestly associated** with earlier post-graduation outcomes, especially around **regional employment and post-secondary enrollment**. However, by ten years out, differences by CTE status largely **diminish**, indicating that **CTE pathways may shape short-term trajectories**, but **long-term outcomes converge** across student groups.

::: panel-tabset
### One year after graduating

```{r hs accomplishments cte.achievement one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, cte.achievement) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments cte.achievement five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, cte.achievement) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments cte.achievement ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, cte.achievement) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## total.cte.courses.taken

**Key Findings**\

There was a **statistically significant relationship** between the number of CTE courses taken in high school and employment outcomes in all three years analyzed.\

Year 1: F(4, 124040) = 1312, p \< .001\

Year 5: F(4, 91140) = 446.8, p \< .001\

Year 10: F(3, 49670) = 208.5, p \< .001

**Interpretation Over Time**

**1 year after graduation:** Students with meaningful employment in the region took slightly more CTE courses on average than most other groups, especially compared to those attending post-secondary institutions. The largest differences appeared between those attending post-secondary (fewer CTE courses) and those with meaningful employment in the region.\

**5 years after graduation:** The trend persisted—graduates with meaningful employment in the region had taken more CTE courses than those in other outcome groups, particularly those attending post-secondary. Notably, students not meaningfully employed or not in education had intermediate CTE course averages.\

**10 years after graduation:** The differences diminished slightly but remained statistically significant. Graduates with meaningful employment in the region continued to show higher average CTE course completion than those not in the workforce or in meaningful employment. However, differences between the non-engaged groups (not employed, not in school) were narrower.

**Conclusion**\
Graduates with **meaningful employment in their home region** consistently took **more CTE courses** than their peers, especially compared to those who pursued post-secondary education. This pattern supports the notion that CTE coursework is more strongly associated with **early and regional employment outcomes**, while those pursuing college took fewer CTE courses overall. The strength of the association appears to **weaken modestly over time**.

::: panel-tabset
### One year after graduating

```{r hs accomplishments total.cte.courses.taken one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, total.cte.courses.taken) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, total.cte.courses.taken, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments total.cte.courses.taken five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, total.cte.courses.taken) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, total.cte.courses.taken, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments total.cte.courses.taken ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, total.cte.courses.taken) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, total.cte.courses.taken, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

# Post-secondary

We will analyze the following variables;

-   attended.ps.within.first.year.hsgrad
-   attended.ps
-   ps.grad
-   ps.grad.InstitutionSector
-   highest.cred.level
-   ps.grad.location

The analysis provides the following insights;

-   Meaningful employment in the region, long-term
    -   More likely:
        -   Attended or graduated
        -   Graduated from a public, 2-year
        -   Less than associated degree or an associate degree
        -   Graduated in region
    -   Less likely:
        -   Did not attend or graduated
        -   Graduated from any 4-year or any for-profit
        -   Bachelors degree or higher & No college
        -   Outside region, worse if outside Minnesota
-   Meaningful employment in MN, long-term
    -   More likely:
        -   Attended or graduated.
        -   Graduated from any 4-year or any for-profit
        -   Bachelor degree or higher
        -   Graduated in Minnesota
    -   Less likely
        -   Did not attend or graduate
        -   Graduated from a public 2-year
        -   No college or some college, associate degree, less than associate degree
-   Not meaningful
    -   More likely:
        -   Did not attend or graduate
        -   Graduated from a for-profit
    -   Less likely:
        -   Graduated from post-secondary
        -   ASsociate or higher
-   No MN employment record
    -   More likely:
        -   Did not attend or graduate
        -   Graduated from a for-profit, or public 4-year
        -   Master degree or higher
        -   Outside Minnesota
    -   Less likely:
        -   Graduated
        -   Graduated from a 2-year
        -   Associate degree, less than associate degree
        -   In region

<br>

## attended.ps.within.first.year.hsgrad

-   One year after graduating;
    -   Individuals that did not attend college, or attended not immediately after graduating college were significantly more likely to have meaningful employment in the region.
    -   These individuals were significantly more likely to not have meaningful employment.
    -   Significantly more likely to not have a MN employment record.
-   Five years after graduating;
    -   Significantly more likely to not have meaningful employment
    -   Individuals who did not attend post-secondary were significantly more likely to not have a MN employment record
-   Ten years after graduating;
    -   individuals that never attended post-secondary were significantly less likely to have meaningful employment in the region.
    -   Individuals that did not attend post-secondary were significantly more likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## attended.ps

-   One year after graduating; individuals that did not attend post-secondary were;
    -   more likely to have meaningful employment in the region.
    -   Significantly more likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Five years after graduating; individuals that did not attend post-secondary were;
    -   More likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Ten years after graduating; individuals that did not attend post-secondary were;
    -   Less likely to have meaningful employment in region or Minnesota
    -   Significantly more likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad

-   One year after graduating; individuals that graduated from post-secondary were;
    -   less likely to have meaningful employment in the region.
    -   Significantly more likely to attend post-secondary
    -   Significantly less likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Five years after graduating; individuals that did not graduate from post-secondary were;
    -   More likely to have meaningful employment in Minnesota
    -   Significantly less likely to not have meaningful employment
    -   Significantly less likely to not have a MN employment record
-   Ten years after graduating; individuals that did not attend post-secondary were;
    -   Slightly more likely to have meaningful employment in the region or in Minnesota

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad.InstitutionSector

A quick reminder on what the codes mean;

-   1 - Public, 4-year or above
-   2 - Private not-for-profit, 4-year or above
-   4 - Public, 2-year
-   10 - went to multiple sectors
-   11 - Includes all of the following
    -   3 - Private for-profit, 4-year or above
    -   6 - Private for-profit, 2-year
    -   9 - Private for-profit, less-than-2-year

<!-- -->

-   One year after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 2
    -   No MN employment record
        -   More likely: 11 (low numbers though)
-   Five years after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 4
        -   More likely: 1, 11, 2
    -   Not meaningful
        -   Less likely: 10
    -   No MN employment record
        -   Less likely: 10
        -   More likely: 11, 2
-   Ten years after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2, 11,
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 4, 11
        -   More likely: 1, 2
    -   Not meaningful
        -   More likely: 11 (low numbers)
    -   No MN employment record
        -   Less likely:
        -   More likely: 11

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.InstitutionSector one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.InstitutionSector) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.InstitutionSector five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.InstitutionSector) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.InstitutionSector ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.InstitutionSector) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## highest.cred.level

-   One year after graduating;
    -   Have meaningful employment in region
        -   Less likely: Bachelor degree or higher
        -   More likely: No college, Associate degree
    -   Meaningful emp in MN
        -   Less likely: Bachelor or higher
        -   More likely: No college
    -   Not meaningful
        -   Less likely: Less than associate degree or higher
        -   More likely: No college
    -   No MN employment record
        -   Less likely: Some college or higher
        -   More likely: No college
-   Five years after graduating;
    -   Have meaningful employment in region
        -   Less likely: Bachelor degree or higher
        -   More likely: Less than associate degree to associate degree
    -   Meaningful emp in MN
        -   Less likely: Bachelor degree
        -   More likely: Master degree or higher
    -   Not meaningful
        -   Less likely: Master degree or higher
        -   More likely: No college or some college
    -   No MN employment record
        -   Less likely: Associate degree or higher
        -   More likely: No college
-   Ten years after graduating;
    -   Have meaningful employment in region
        -   Less likely: Master degree, No college
        -   More likely: Associate degree, Less than associate degree
    -   Meaningful emp in MN
        -   Less likely: No college, some college, Associate degree, Less than associate degree
        -   More likely: Bachelor degree or higher
    -   Not meaningful
        -   Less likely: Associate degree or higher
        -   More likely: Some college
    -   No MN employment record
        -   Less likely: Less than Associate degree
        -   More likely:

::: panel-tabset
### One year after graduating

```{r post-secondary highest.cred.level one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, highest.cred.level) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary highest.cred.level five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, highest.cred.level) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary highest.cred.level ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, highest.cred.level) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad.location

-   One year after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside MN
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: Outside Minnesota
        -   More likely:
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota
        -   More likely: Outside Minnesota
    -   No MN employment record
        -   Less likely: Inside and outside region
        -   More likely: Inside and outside Minnesota, outside Minnesota

-   Five years after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside MN
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: In region, inside and outside region, inside and outside MN, outside MN
        -   More likely: In Minnesota
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota
        -   More likely:
    -   No MN employment record
        -   Less likely: Inside and outside region, Inside and outside MN
        -   More likely: outside Minnesota

-   Ten years after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside Minnesota
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: In region
        -   More likely: In Minnesota, Inside and outside Minnesota
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota, outside Minnesota
        -   More likely:
    -   No MN employment record
        -   Less likely: In region
        -   More likely: Inside and outside Minnesota, Outside minnesota

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.location one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.1 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.location five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.5 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.location ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.10 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

# Variables to keep

Okay, after this analysis, here is what we will move forward with;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   english.learner
-   High school characteristics
    -   Dem_Desc
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ap.exam
    -   ACTCompositeScore
    -   MCA.M
    -   MCA.R
    -   MCA.S
    -   cte.achievement
    -   n.cte.courses.taken
-   Post-secondary
    -   ps.grad
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   ps.grad.location

```{r create new master after analysis}
master.after.ct <- meaningful.emp.central %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  select(PersonID, grad.year.covid, economic.status, , pseo.participant, SpecialEdStatus, RaceEthnicity, Dem_Desc, took.ACT, cte.achievement, english.learner, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location, ACTCompositeScore, MCA.M, MCA.R, MCA.S, total.cte.courses.taken,grad.year.1, grad.year.5, grad.year.10)
  
write_csv(master.after.ct, "Data/SLEDS/Masters/After analysis/Master-after-ct.csv")
```
