---
title: "Local Employment Cross-tabs"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.factor.names <- master.original %>%
  select(PersonID, Dem_Desc, edr, grad.year.covid, Gender, economic.status, pseo.participant, SpecialEdStatus, non.english.home, RaceEthnicity, took.ACT, ap.exam, cte.achievement, english.learner, sat.taken, attended.ps, attended.ps.within.first.year.hsgrad, ps.grad, ps.grad.InstitutionSector, highest.cred.level) 

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.states %>%
  left_join(master.factor.names, by = "PersonID")

kable(head(master))

kable(names(master))
```

```{r independent and dependent variables}
updated.factor.var.names <-  master.factor.names %>%
  select(-PersonID) %>%
  names() %>%
  c("ps.grad.in.same.ruca", "ps.grad.in.same.edr", "ps.grad.in.same.pr", "ps.grad.in.MN")

numeric.var.names <- master.original %>%
  select(ACTCompositeScore, total.cte.courses.taken, `ACTE.SPED`:Youth.Apprenticeship,  MCA.M:MCA.S, n.institutions, avg.unemp.rate, wages.3year.avg, avg.wages.pct.state) %>%
  names()
  
updated.independent.var.names <- c(updated.factor.var.names, numeric.var.names)

cte.cc.var.names <- master.original %>%
  select(ACTE.SPED:Youth.Apprenticeship) %>%
  names()

dep.var.names <- master %>%
  select(PersonID, grad.year.1, grad.year.5, grad.year.10)

updated.master <- master.original %>%
  select(PersonID, all_of(updated.independent.var.names)) %>%
  left_join(dep.var.names, by = "PersonID") %>%
  mutate_at(vars(Dem_Desc:ps.grad.in.MN), as.factor) %>%
  mutate_at(vars(grad.year.1:grad.year.10), as.factor) %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Did not grad", "Some college", as.character(highest.cred.level)),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Never attended ps", "No college", as.character(highest.cred.level)),
         highest.cred.level = as.factor(highest.cred.level),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix", "Urban/town/rural mix", "Entirely urban")) %>%
  droplevels()

kable(head(updated.master))

kable(names(updated.master))
```

<br>

There are a number of "states" in which an individual in this dataset are assigned;

1.  Has meaningful employment
    1.  In the same county as their high school.
    2.  In the same EDR as their high school, but not county.
    3.  In Minnesota, but not in Central Minnesota.
2.  Is attending post-secondary.
3.  Has a MN employment record but it's not meaningful, and not attending post-secondary.
4.  Does not have a MN employment record and is not attending post-secondary.

There are three "times" in which we checked to see if the individual had meaningful employment within these geographies;

1.  one year after graduating high school
2.  five years after graduation, and
3.  ten years after graduation.

Meaningful employment is determined by whether an individual worked 1,000 hours for an employer during time X. In addition, if the individual is working for an employer but not meaningful at time X but worked 1,000 hours for that employer during another year it's still considered "meaningful".

Due to time x potentially being passed the date of the latest data (2023) for some individuals, the analysis below will filter out all individuals where time x is after 2023.

From the multiple correspondence analysis and the principle component analysis we determined that the following independent variables will be included in the analysis.

```{r list of ind var}
updated.master %>%
  select(-PersonID, -grad.year.1, -grad.year.5, -grad.year.10) %>%
  lapply(class) 

```

# Methodology

For this analysis we are keeping it simple. For categorical variables I’ll be using the Cramer’s v-square measure of association. The V\^2 value is from 0 to 1 with 0 meaning there’s no association and a value closer to 1 indicates a strong association.

| **Cramér’s V Value** | **Interpretation**               |
|----------------------|----------------------------------|
| 0.00–0.10            | Negligible association           |
| 0.10–0.20            | Weak                             |
| 0.20–0.40            | Moderate                         |
| 0.40–0.60            | Relatively strong                |
| 0.60–0.80            | Strong                           |
| 0.80–1.00            | Very strong (possibly redundant) |

For the variables with continuous values I’ll be using the Wilcoxon two paired groups test. P-value will be used to determine whether we can reject the NULL hypothesis at a .01 level. I’ll also include boxplots to see in which direction there might be influence from the variables.

We are going to split up the analysis into two different datasets. The first dataset will combine 3 of the previous states into one.

-   Has meaningful employment anywhere in Central Minnesota, which combines the following states;
    -   Meaningful emp County
    -   Meaningful emp EDR
    -   Meaningful emp region

The remaining states will stay the same.

-   Has meaningful employment in Minnesota, but not in Central Minnesota
-   Is attending a post-secondary institution.
-   Has a MN employment record but it's not meaningful, and they are not attending post-secondary.
-   Does not have a MN employment record and is not attending post-secondary.

We have quite a few variables remaining, even after MCA analysis. So let's take a look at how we are going to proceed and group all of this together to see how these relationships work.

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner
-   High school characteristics
    -   Dem_Desc
    -   edr
    -   avg.unemp.rate
    -   avg.wage.pct.state
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ACTCompositeScore
    -   ap.exam
    -   sat.taken
    -   ACTCompositeScore
    -   MCA scores
        -   MCA.M
        -   MCA.R
        -   MCA.S
    -   cte.achievement
    -   total.cte.courses.taken
    -   avg.unemp.rate
-   Post-secondary
    -   attended.ps.within.first.year.hsgrad
    -   attended.ps
    -   ps.grad.
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   Combine post-secondary grad location
        -   ps.grad.in.same.ruca
        -   ps.grad.in.same.edr
        -   ps.grad.in.same.pr
        -   ps.grad.in.MN

<br>

# Meaningful employment in Central Minnesota

```{r prep meaningful employment in Central MN}
meaningful.emp.central <- updated.master %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.1)),
         grad.year.1 = as_factor(grad.year.1),
         grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.5)),
         grad.year.5 = as_factor(grad.year.5),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.10)),
         grad.year.10 = as_factor(grad.year.10),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         highest.cred.level = fct_relevel(highest.cred.level, "No college", "Some college", "Less than Associate Degree", "Associate degree", "Bachelor degree or higher"))

write_csv(meaningful.emp.central, "Data/SLEDS/Masters/After analysis/Master-meaningful-emp-central.csv")
```

Let's first take a look at proportion of states in each of the time.X. The chart below shows that half of the graduates are attending post-secondary (50%) one year after graduating high school. The further away from high school graduation results in a bit of an increase in the percentage of individuals who have meaningful employment in the region and in Minnesota, while about a quarter of graduates do not have a MN employment record and are not going to post-secondary.

```{r proportion of states meaningful emp SW}
prop.states.meaningful.emp.central <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10) %>%
  pivot_longer(names_to = "grad.year", values_to = "states", 1:3) %>%
  group_by(grad.year, states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  filter(states != "After 2023") %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         grad.year = fct_relevel(grad.year, "grad.year.1", "grad.year.5", "grad.year.10"))

prop.states.meaningful.emp.central.plot <- ggplot(prop.states.meaningful.emp.central, aes(grad.year, pct, fill = states, group = states)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nState of individuals: ", states, "\nNumber of people in states: ", comma(n), "\nPercent of people in state: ", percent(pct, accuracy = .1), sep = ""))) +
   labs(x="", y = "", color="", title = "Percent of people in each state")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu"),
                    guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = prop.states.meaningful.emp.central.plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

Let's examine the variables to see if there are any significant differences in the proportion of individuals in these states across the variable categories or values.

<br>

# Demographics

In the demographics section we will examine the relationship between that paths taken by individuals and the following variables;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner

## **Summary**

+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Demographic      | Association Strength   | Highlights                                                                                                                                                                                          |
+==================+:======================:+=====================================================================================================================================================================================================+
| Gender           | Weak to negligible     | Females attend post-secondary at a much higher rate                                                                                                                                                 |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| grad.year.covid  | Weak                   | Post-COVID grads attend post-secondary at a lower rate.                                                                                                                                             |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| economic.status  | Moderate to negligible | Individuals who qualify for free or reduced lunch attend post-secondary at a lower, and have a lower rate of meaningful employment in Minnesota.                                                    |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| SpecialEdStatus  | Moderate to negligible | Special Ed students attend post-secondary at a SIGNIFICANTLY lower rate, and have a higher rate of not having meaningful employment.                                                                |
|                  |                        |                                                                                                                                                                                                     |
|                  |                        | Not having a MN employment record is significantly higher at first.                                                                                                                                 |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| RaceEthnicity    | Negligible             | White and Asian/PI attend post-secondary at a significantly higher rate.                                                                                                                            |
|                  |                        |                                                                                                                                                                                                     |
|                  |                        | American Indian and Black graduates have a significantly higher rate of not having a MN employment record or attending post-secondary.                                                              |
|                  |                        |                                                                                                                                                                                                     |
|                  |                        | The rate of non-white individuals not having a MN employment record increases significantly over time.                                                                                              |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| non.english.home | Negligible             | Individuals with non-English speakers in their household had lower rate of meaningful employment in the region or in Minnesota, and SIGNIFICANTLY higher rate of not having a MN employment record. |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| english.learner  | Weak to negligible     | English learners had a SIGNIFICANTLY lower rates of attending post-secondary.                                                                                                                       |
|                  |                        |                                                                                                                                                                                                     |
|                  |                        | The further from high school graduation saw a significant decrease in the rate of individuals with meaningful employment in the region or Minnesota.                                                |
+------------------+------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

<br>

## Gender

There is a statistically significant association between whether an individual qualifies for free or reduced lunch and their employment state, but the association starts weak and becomes negligible by five years after graduating high school.

The most consistent difference is that females attend post-secondary at a significantly higher rate.

<br>

::: panel-tabset
### One year after graduating

There is a statistically significant, weak association between gender and the employment state.

Females;

-   Have a lower rate of meaningful employment in the region.
-   Attend post-secondary at a significantly higher rate.

```{r demographics gender one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Gender) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between gender and the employment state.

Females;

-   Have a lower rate of meaningful employment in the region.
-   Attend post-secondary at a significantly higher rate.

```{r demographics gender five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Gender) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between economic status and the employment state.

Barely any differences here.

```{r demographics gender ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Gender) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## Grad.year.covid

There is a statistically significant, weak association between whether an individual graduated before or after COVID and the employment state. The most striking difference is that post-covid grads were attending post-secondary at a significantly lower rate than individuals who graduated before COVID.

<br>

::: panel-tabset
### One year after graduating

```{r demographics grad.year.covid one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.covid) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$grad.year.covid, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## economic.status

There is a statistically significant association between whether an individual qualifies for free or reduced lunch and their employment state, but the association starts at moderate and becomes negligible by ten years after graduating high school.

The most consistent difference through all the years was that individuals who qualify for free and reduced lunch attend post-secondary at a lower rate, and have a lower rate of meaningful employment in Minnesota.

<br>

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a higher rate of meaningful employment in the region.
-   Have a higher rate of meaningful employment in Minnesota.
-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, economic.status) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, economic.status) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a lower rate of having meaningful employment in Minnesota.
-   Have a higher rate of not having meaningful employment.

```{r demographics economic.status ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, economic.status) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## SpecialEdStatus

There is a statistically significant association between whether an individual is categorized as needing special education and their employment state, but the association starts at moderate and becomes negligible by ten years after graduating high school.

The most consistent difference through all the years was that individuals who are categorized as needing special education attend post-secondary at a SIGFNICANTLY lower rate, and have a higher rate of not having meaningful employment.

Having no MN employment record is significantly higher at first, but evens out by ten years after high school.

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, SpecialEdStatus) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   slightly lower rate of having meaningful employment in Minnesota
-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, SpecialEdStatus) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   SIGNIFICANTLY lower rate of having meaningful employment in Minnesota
-   SIGNFICANTLY higher rate of not having meaningful employment

```{r demographics SpecialEdStatus ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, SpecialEdStatus) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## RaceEthnicity

There is a statistically significant association between race/ethnicity and their employment state, but the association is negligible.

The most consistent difference through all the years were White and Asian/PI have a higher rate of attending post-secondary, and American Indian and Black individuals had a significantly higher rate of not having meaningful employment nor attending post-secondary.

Non-white individuals had an increasing rate of not having a MN employment record the further out from high school graduation.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between whether race/ethnicity and the employment state.

Individuals who are;

-   Hispanic have a higher rate of meaningful employment in the region.
-   White or Asian/PI attended post-secondary at a SIGNIFICANTLY higher rate
-   American Indian, Black or Hispanic had SIGNFICANTLY higher rate of not having meaningful employment
-   American Indian had a SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics RaceEthnicity one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, RaceEthnicity) %>%
  filter(grad.year.1 != "After 2023",
         RaceEthnicity != "Unknown") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between whether race/ethnicity and the employment state.

Individuals who are;

-   White or Asian/PI attended post-secondary at a SIGNIFICANTLY higher rate
-   American Indian, Black or Hispanic had SIGNFICANTLY higher rate of not having meaningful employment
-   American Indian had a SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics RaceEthnicity five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, RaceEthnicity) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether race/ethnicity and the employment state.

Individuals who are;

-   Asian/PI and black individuals had a lower rate of meaningful employment in the region.
-   American Indian had a significantly lower rate of meaningful employment in MN.
-   American Indian and Black had SIGNFICANTLY higher rate of not having meaningful employment
-   Asian/PI, Black, and Hispanic had a significantly higher rate of having no MN employment record.

```{r demographics RaceEthnicity ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, RaceEthnicity) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## non.english.home

There is a statistically significant association between whether an individual had a non-English speaker in their household and their employment state, but the association is negligible.

The most biggest trend was the individuals that were confirmed to have a non-English speaking individual in their household had lower rates of meaningful employment in the region or in Minnesota and SIGNIFCANTLY higher rate of not having a MN employment record over time.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between whether ran individual had a non native English speaking person in their household and the employment state.

Individuals who had a non-English speaking person at home;

-   Had a significantly lower rate of attending post-secondary.
-   Had a significantly higher rate of having no MN employment record.

```{r demographics non.english.home one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, non.english.home) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between whether a graduate had a non-English speaking person in their household and the employment state.

Individuals who had a non-English speaking member at home;

-   Had a significantly higher rate of having no MN employment record.

```{r demographics non.english.home five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, non.english.home) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether a graduate had a non-English speaking person in their household and the employment state.

Individuals who had a non-English speaking member at home;

-   Had a significantly lower rate of meaningful employment in the region
-   Had a significantly lower rate of meaningful employment in Minnesota.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics non.english.home ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, non.english.home) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## english.learner

There is a statistically significant association between whether an individual was categorized as an English learner and their employment state, but the association was weak at first and became negligible over time.

Through all the years, graduates categorized as an English learner had SIGNIFICANTLY lower rates of attending post-secondary and had SIGNFICANTLY higher rates of having no MN employment record.

As time away from high school grows, the rate at which these individuals had meaningful employment in the region or in Minnesota became significantly lower.

::: panel-tabset
### One year after graduating

There is a statistically significant, weak association between whether a graduate was categorized as an English learner and the employment state.

Individuals who were classified as an English learner;

-   Had a SIGNIFICANTLY lower rate of attending post-secondary.
-   Had a significantly higher rate of not having meaningful employment.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics english.learner one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, english.learner) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between whether a graduate was categorized as an English learner and the employment state.

Individuals who were classified as an English learner;

-   Had a lower rate of meaningful employment in Minnesota.
-   Had a SIGNIFICANTLY lower rate of attending post-secondary.
-   Had a significantly higher rate of not having meaningful employment.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics english.learner five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, english.learner) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between whether a graduate was categorized as an English learner and the employment state.

Individuals who were classified as an English learner;

-   Had a lower rate of meaningful employment in the region.
-   Had a SIGNIFICANTLY lower rate of meaningful employment in Minnesota.
-   Had a SIGNIFICANTLY higher rate of having no MN employment record.

```{r demographics english.learner ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, english.learner) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ChatGPT

This is the summary provided by ChatGPT when asked to analyze and interpret.

### Gender Trends Over Time

+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Outcome                     | 1 Year (F) | 5 Year (F) | 10 Year (F) | 1 Year (M) | 5 Year (M) | 10 Year (M) |
+=============================+============+============+=============+============+============+=============+
| Attending PS                | 55.3%      | 17.8%      | \~0.0%      | 43.7%      | 13.4%      | 0.0%        |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Meaningful Emp in MN        | 6.8%       | 24.2%      | 28.6%       | 8.2%       | 24.9%      | 31.8%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Meaningful Emp in Region    | 10.9%      | 19.9%      | 21.5%       | 15.3%      | 24.1%      | 24.8%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| No MN Emp Record, Not in PS | 7.7%       | 17.1%      | 30.3%       | 11.9%      | 19.1%      | 28.4%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+
| Not Meaningful, Not in PS   | 19.3%      | 20.9%      | 19.6%       | 20.9%      | 18.5%      | 15.1%       |
+-----------------------------+------------+------------+-------------+------------+------------+-------------+

**Trend**: Females start with higher PS enrollment; males steadily overtake in employment. Disconnection grows for both.

------------------------------------------------------------------------

### Economic Status Trends Over Time

+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Outcome                     | 1 Yr (0) | 5 Yr (0) | 10 Yr (0) | 1 Yr (1) | 5 Yr (1) | 10 Yr (1) |
+=============================+==========+==========+===========+==========+==========+===========+
| Attending PS                | 57.4%    | 17.8%    | \~0.0%    | 32.6%    | 11.1%    | \~0.0%    |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in MN        | 6.7%     | 26.2%    | 32.6%     | 9.2%     | 21.1%    | 25.0%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in Region    | 11.7%    | 21.4%    | 22.8%     | 16.2%    | 23.4%    | 23.8%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| No MN Emp Record, Not in PS | 8.1%     | 17.0%    | 28.7%     | 13.5%    | 20.4%    | 30.7%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Not Meaningful, Not in PS   | 16.1%    | 17.6%    | 15.9%     | 28.5%    | 24.1%    | 20.5%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+

**Trend**: Low-income students are less likely to pursue PS, and consistently more likely to lack MN employment or be disconnected.

------------------------------------------------------------------------

### Special Education Trends Over Time

+-----------------------------+----------------+-------+--------+-------------+-------+-------+
| Outcome                     | 1 Yr (No SPED) | 5 Yr  | 10 Yr  | 1 Yr (SPED) | 5 Yr  | 10 Yr |
+=============================+================+=======+========+=============+=======+=======+
| Attending PS                | 54.2%          | 17.0% | \~0.0% | 19.8%       | 6.4%  | 0.0%  |
+-----------------------------+----------------+-------+--------+-------------+-------+-------+
| Meaningful Emp in MN        | 7.3%           | 25.6% | 31.5%  | 8.6%        | 17.6% | 21.2% |
+-----------------------------+----------------+-------+--------+-------------+-------+-------+
| Meaningful Emp in Region    | 12.7%          | 21.7% | 23.0%  | 15.9%       | 24.1% | 24.0% |
+-----------------------------+----------------+-------+--------+-------------+-------+-------+
| No MN Emp Record, Not in PS | 8.1%           | 17.2% | 29.0%  | 21.0%       | 24.5% | 31.5% |
+-----------------------------+----------------+-------+--------+-------------+-------+-------+
| Not Meaningful, Not in PS   | 17.8%          | 18.5% | 16.5%  | 34.8%       | 27.5% | 23.3% |
+-----------------------------+----------------+-------+--------+-------------+-------+-------+

**Trend**: SPED students consistently show lower engagement in PS and work, and higher disconnection.

------------------------------------------------------------------------

### Race/Ethnicity Snapshot

+----------------------+--------------------+---------------+---------------------+-----------------------+
| Race                 | Attending PS (1yr) | Emp MN (10yr) | No MN Record (10yr) | Not Meaningful (10yr) |
+======================+====================+===============+=====================+=======================+
| American Indian (AI) | 26.0%              | 21.6%         | 32.1%               | 27.3%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| Asian/PI             | 52.0%              | 30.2%         | 40.2%               | 15.6%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| Black                | 38.7%              | 25.3%         | 36.0%               | 25.2%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| Hispanic             | 31.2%              | 25.1%         | 40.2%               | 17.5%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+
| White                | 51.1%              | 30.5%         | 28.8%               | 17.0%                 |
+----------------------+--------------------+---------------+---------------------+-----------------------+

**Trend**: White and Asian/PI grads show strongest MN employment; Black, Hispanic, and AI groups show consistently higher disconnection risk.

### Non-English speakers at home

+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Outcome                     | 1 Yr (0) | 5 Yr (0) | 10 Yr (0) | 1 Yr (1) | 5 Yr (1) | 10 Yr (1) |
+=============================+==========+==========+===========+==========+==========+===========+
| Attending PS                | 50.0%    | 15.7%    | \~0.0%    | 36.4%    | 13.9%    | 0.0%      |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in MN        | 7.4%     | 24.7%    | 30.4%     | 10.5%    | 19.4%    | 23.2%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in Region    | 13.1%    | 22.2%    | 23.4%     | 14.0%    | 17.0%    | 13.6%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| No MN Emp Record, Not in PS | 9.5%     | 17.7%    | 28.9%     | 18.6%    | 30.4%    | 47.7%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Not Meaningful, Not in PS   | 20.1%    | 19.7%    | 17.4%     | 20.5%    | 19.3%    | 15.5%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+

**Trend**: Students from non-English-speaking homes have **lower postsecondary enrollment** and are increasingly underrepresented in the MN workforce by year 10, with a high percentage lacking MN employment records.

------------------------------------------------------------------------

### English learners

+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Outcome                     | 1 Yr (0) | 5 Yr (0) | 10 Yr (0) | 1 Yr (1) | 5 Yr (1) | 10 Yr (1) |
+=============================+==========+==========+===========+==========+==========+===========+
| Attending PS                | 51.3%    | 16.1%    | \~0.0%    | 23.6%    | 7.2%     | 0.0%      |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in MN        | 7.4%     | 25.0%    | 30.6%     | 9.5%     | 17.9%    | 18.5%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Meaningful Emp in Region    | 12.9%    | 22.1%    | 23.3%     | 15.7%    | 21.5%    | 17.6%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| No MN Emp Record, Not in PS | 8.9%     | 17.4%    | 28.9%     | 22.1%    | 29.1%    | 43.1%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+
| Not Meaningful, Not in PS   | 19.4%    | 19.4%    | 17.2%     | 29.1%    | 24.3%    | 20.8%     |
+-----------------------------+----------+----------+-----------+----------+----------+-----------+

**Trend**: English learners start with significantly lower PS participation and grow increasingly disconnected from MN employment systems over time.

------------------------------------------------------------------------

```{r demogrpahics chatgpt}
demographics.chatgpt <- meaningful.emp.central %>%
  select(PersonID, Gender, grad.year.covid, economic.status, SpecialEdStatus, RaceEthnicity, non.english.home, english.learner, grad.year.1, grad.year.5, grad.year.10)

write_csv(demographics.chatgpt, "Data/ChatGPT/demographics-chatgpt.csv")
```

# High school characteristics

We will take a look at the following variables;

-   Dem_Desc
-   edr
-   avg.unemp.rate
-   avg.wage.pct.state

## Summary

<br>

## Dem_Desc

There is a statistically significant association between an graduate's RUCA category and their employment state, but the association was negligible.

Through all the years, the more rural a graduate was, the higher the rate of having meaningful employment in the region, while urban individuals had a significantly higher rate of meaningful employment in Minnesota.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a higher rate of having meaningful employment in the region.
-   more urban area had a higher rate of having meaningful employment in MN.
-   more rural areas had a lower rate of attending post-secondary.
-   more rural areas had a higher rate of not having a MN employment record.

```{r hs characteristics Dem_Desc one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Dem_Desc) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a higher rate of having meaningful employment in the region.
-   more urban area had a significantly higher rate of having meaningful employment in MN.

```{r hs characteristics Dem_Desc five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Dem_Desc) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a significantly higher rate of having meaningful employment in the region.
-   more urban area had a significantly higher rate of having meaningful employment in MN.

```{r hs characteristics Dem_Desc ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Dem_Desc) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## avg.unemp.rate

There is a statistically significant association between an graduate's RUCA category and their employment state, but the association was negligible.

Through all the years, the more rural a graduate was, the higher the rate of having meaningful employment in the region, while urban individuals had a significantly higher rate of meaningful employment in Minnesota.

::: panel-tabset
### One year after graduating

There is a statistically significant, negligible association between a graduate's RUCA category and the employment state.

Graduates from;

-   more rural areas had a higher rate of having meaningful employment in the region.
-   more urban area had a higher rate of having meaningful employment in MN.
-   more rural areas had a lower rate of attending post-secondary.
-   more rural areas had a higher rate of not having a MN employment record.

```{r hs characteristics avg.unemp.rate one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.unemp.rate) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.unemp.rate, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 20))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

hs.characteristics.chatgpt <- meaningful.emp.central %>%
  select(PersonID, Dem_Desc, edr, avg.unemp.rate, avg.wages.pct.state, grad.year.1, grad.year.5, grad.year.10)

write_csv(hs.characteristics.chatgpt, "Data/ChatGPT/hs-characteristics-chatgpt.csv")
```

### Five years after graduating

```{r hs characteristics avg.unemp.rate five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.unemp.rate) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.unemp.rate, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs characteristics avg.unemp.rate ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.unemp.rate) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.unemp.rate, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## avg.wages.pct.state

-   One year after graduating
    -   The individuals graduating from districts with slightly lower wages had meaningful employment in MN, but not the region.
-   Five years after graduating
    -   Not much
-   Ten years after graduating
    -   Not much

::: panel-tabset
### One year after graduating

```{r hs characteristics avg.wages.pct.state one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.wages.pct.state) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.wages.pct.state, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs characteristics avg.wages.pct.state five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.wages.pct.state) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.wages.pct.state, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs characteristics avg.wages.pct.state ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.wages.pct.state) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.wages.pct.state, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

# High School Accomplishments

We are going to analyze the following variables;

-   pseo.participant
-   took.ACT
-   ap.exam
-   ACTCompositeScore
-   MCA scores
    -   MCA.M
    -   MCA.R
    -   MCS.S
-   cte.achievement
-   n.cte.courses.taken
-   avg.unemp.rate

The analysis provides the following insights;

-   Higher performing students with all the tests tended to
    -   attend post-secondary at significantly higher rates
    -   have meaningful workforce participation in Minnesota at higher rates.
-   Students that attended PSEO all tended to
    -   attend post-secondary at significantly higher rates
    -   have meaningful workforce participation in Minnesota at higher rates
-   Students involved in CTE tended to;
    -   Have meaningful workforce participation in Minnesota at higher rates
    -   Not attend post-secondary at higher rates.

<br>

## pseo.participant

-   One year after graduating, PSEO participants are;
    -   Less likely to have meaningful employment in the region and Minnesota.
    -   Significantly more likely to attend post-secondary.
    -   Significantly less likely to not have meaningful employment
    -   Significantly less likely to not have a MN employment record.
-   Five years after graduating, PSEO participants are;
    -   Less likely to have meaningful employment in the region.
    -   Slightly more likely to have meaningful employment in Minnesota.
    -   Significantly more likely to attend post-secondary.
    -   Significantly less likely to not have meaningful employment
    -   Significantly less likely to not have a MN employment record.
-   Ten years after graduating
    -   Significantly more likely to have meaningful employment in Minnesota.
    -   Significantly less likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r hs accomplishments pseo.participant one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, pseo.participant) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments pseo.participant five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, pseo.participant) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments pseo.participant ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, pseo.participant) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## took.ACT

-   One year after graduating, individuals that took the ACT are;
    -   Significantly more likely to attend post-secondary.
    -   Significantly less likely to not have meaningful employment
    -   Significantly less likely to not have a MN employment record.
-   Five years after graduating, individuals that took the ACT are;
    -   Slightly more likely to have meaningful employment in Minnesota.
    -   Significantly more likely to attend post-secondary.
    -   Significantly less likely to not have a MN employment record.
-   Ten years after graduating, individuals that took the ACT are;
    -   Significantly more likely to have meaningful employment in Minnesota.

::: panel-tabset
### One year after graduating

```{r hs accomplishments took.ACT one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, took.ACT) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments took.ACT five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, took.ACT) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments took.ACT ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, took.ACT) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ACTCompositeScore

-   One year after graduating
    -   Individuals attending post-secondary had higher scores.
    -   Individuals with meaningful workforce participation in the region had lower scores.
-   Five years after graduating
    -   Individuals with meaningful wf participation in the region had lower scores.
    -   Individuals with meaningful employment in Minnesota had higher scores
    -   Individuals attending post-secondary had higher scores
-   Ten years after graduating
    -   Individuals with meaningful workforce participation in Minnesota had higher scores
    -   Individuals with no MN employment record had higher scores.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ACTCompositeScore one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ACTCompositeScore) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, ACTCompositeScore, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments ACTCompositeScore five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ACTCompositeScore) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, ACTCompositeScore, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments ACTCompositeScore ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ACTCompositeScore) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, ACTCompositeScore, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 40))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.R

-   One year after graduating
    -   Individuals attending post-secondary had higher scores.
-   Five years after graduating
    -   Individuals with meaningful employment in Minnesota had higher scores
    -   Individuals attending post-secondary had higher scores
-   Ten years after graduating
    -   Individuals with no meaningful employment had significantly lower scores.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.R one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.R) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.R, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.R five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.R) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.R, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.R ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.R) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.R, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.S

-   One year after graduating
    -   Individuals attending post-secondary had higher scores.
-   Five years after graduating
    -   Individuals with meaningful employment in Minnesota had higher scores
    -   Individuals attending post-secondary had higher scores
-   Ten years after graduating
    -   Individuals with no meaningful employment had significantly lower scores.
    -   Individuals with meaningful employment in Minnesota had higher scores.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.S one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.S) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.S, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.S five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.S) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.S, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.S ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.S) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.S, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## MCA.M

-   One year after graduating
    -   Individuals attending post-secondary had higher scores.
-   Five years after graduating
    -   Individuals with meaningful employment in Minnesota had higher scores
    -   Individuals attending post-secondary had higher scores
-   Ten years after graduating
    -   Individuals with no meaningful employment had significantly lower scores.
    -   Individuals with meaningful employment in Minnesota had higher scores.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.M one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.M) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.M, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.M five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.M) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.M, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.M ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.M) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.M, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

## cte.achievement

-   One year after graduating, individuals that took more CTE were;
    -   Significantly more likely to have meaningful employment in the region
    -   Significantly less likely to attend post-secondary.
    -   Significantly more likely to not have a MN employment record.
-   Five years after graduating, individuals that took the AP exam are;
    -   Significantly more likely to have meaningful employment in region.
    -   Significantly less likely to attend post-secondary
-   Ten years after graduating, individuals that took the AP exam are;
    -   Significantly more likely to have meaningful employment in region.

::: panel-tabset
### One year after graduating

```{r hs accomplishments cte.achievement one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, cte.achievement) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments cte.achievement five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, cte.achievement) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments cte.achievement ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, cte.achievement) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## total.cte.courses.taken

-   One year after graduating
    -   Individuals with meaningful employment in the region had taken a higher number of CTE courses
    -   Individuals attending post-secondary had taken a lower number of CTE courses.
-   Five years after graduating
    -   Individuals with meaningful employment in Minnesota had taken a higher number of CTE courses
-   Ten years after graduating
    -   Individuals with meaningful employment in Minnesota had taken a higher number of CTE courses

::: panel-tabset
### One year after graduating

```{r hs accomplishments total.cte.courses.taken one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, total.cte.courses.taken) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, total.cte.courses.taken, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments total.cte.courses.taken five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, total.cte.courses.taken) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, total.cte.courses.taken, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments total.cte.courses.taken ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, total.cte.courses.taken) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, total.cte.courses.taken, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

<br>

# Post-secondary

We will analyze the following variables;

-   attended.ps.within.first.year.hsgrad
-   attended.ps
-   ps.grad
-   ps.grad.InstitutionSector
-   highest.cred.level
-   ps.grad.location

The analysis provides the following insights;

-   Meaningful employment in the region, long-term
    -   More likely:
        -   Attended or graduated
        -   Graduated from a public, 2-year
        -   Less than associated degree or an associate degree
        -   Graduated in region
    -   Less likely:
        -   Did not attend or graduated
        -   Graduated from any 4-year or any for-profit
        -   Bachelors degree or higher & No college
        -   Outside region, worse if outside Minnesota
-   Meaningful employment in MN, long-term
    -   More likely:
        -   Attended or graduated.
        -   Graduated from any 4-year or any for-profit
        -   Bachelor degree or higher
        -   Graduated in Minnesota
    -   Less likely
        -   Did not attend or graduate
        -   Graduated from a public 2-year
        -   No college or some college, associate degree, less than associate degree
-   Not meaningful
    -   More likely:
        -   Did not attend or graduate
        -   Graduated from a for-profit
    -   Less likely:
        -   Graduated from post-secondary
        -   ASsociate or higher
-   No MN employment record
    -   More likely:
        -   Did not attend or graduate
        -   Graduated from a for-profit, or public 4-year
        -   Master degree or higher
        -   Outside Minnesota
    -   Less likely:
        -   Graduated
        -   Graduated from a 2-year
        -   Associate degree, less than associate degree
        -   In region

<br>

## attended.ps.within.first.year.hsgrad

-   One year after graduating;
    -   Individuals that did not attend college, or attended not immediately after graduating college were significantly more likely to have meaningful employment in the region.
    -   These individuals were significantly more likely to not have meaningful employment.
    -   Significantly more likely to not have a MN employment record.
-   Five years after graduating;
    -   Significantly more likely to not have meaningful employment
    -   Individuals who did not attend post-secondary were significantly more likely to not have a MN employment record
-   Ten years after graduating;
    -   individuals that never attended post-secondary were significantly less likely to have meaningful employment in the region.
    -   Individuals that did not attend post-secondary were significantly more likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## attended.ps

-   One year after graduating; individuals that did not attend post-secondary were;
    -   more likely to have meaningful employment in the region.
    -   Significantly more likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Five years after graduating; individuals that did not attend post-secondary were;
    -   More likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Ten years after graduating; individuals that did not attend post-secondary were;
    -   Less likely to have meaningful employment in region or Minnesota
    -   Significantly more likely to not have a MN employment record.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad

-   One year after graduating; individuals that graduated from post-secondary were;
    -   less likely to have meaningful employment in the region.
    -   Significantly more likely to attend post-secondary
    -   Significantly less likely to not have meaningful employment
    -   Significantly more likely to not have a MN employment record
-   Five years after graduating; individuals that did not graduate from post-secondary were;
    -   More likely to have meaningful employment in Minnesota
    -   Significantly less likely to not have meaningful employment
    -   Significantly less likely to not have a MN employment record
-   Ten years after graduating; individuals that did not attend post-secondary were;
    -   Slightly more likely to have meaningful employment in the region or in Minnesota

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad.InstitutionSector

A quick reminder on what the codes mean;

-   1 - Public, 4-year or above
-   2 - Private not-for-profit, 4-year or above
-   4 - Public, 2-year
-   10 - went to multiple sectors
-   11 - Includes all of the following
    -   3 - Private for-profit, 4-year or above
    -   6 - Private for-profit, 2-year
    -   9 - Private for-profit, less-than-2-year

<!-- -->

-   One year after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 2
    -   No MN employment record
        -   More likely: 11 (low numbers though)
-   Five years after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 4
        -   More likely: 1, 11, 2
    -   Not meaningful
        -   Less likely: 10
    -   No MN employment record
        -   Less likely: 10
        -   More likely: 11, 2
-   Ten years after graduating;
    -   Have meaningful employment in region
        -   Less likely: 1, 2, 11,
        -   More likely: 4
    -   Meaningful emp in MN
        -   Less likely: 4, 11
        -   More likely: 1, 2
    -   Not meaningful
        -   More likely: 11 (low numbers)
    -   No MN employment record
        -   Less likely:
        -   More likely: 11

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.InstitutionSector one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.InstitutionSector) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.InstitutionSector five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.InstitutionSector) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.InstitutionSector ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.InstitutionSector) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## highest.cred.level

-   One year after graduating;
    -   Have meaningful employment in region
        -   Less likely: Bachelor degree or higher
        -   More likely: No college, Associate degree
    -   Meaningful emp in MN
        -   Less likely: Bachelor or higher
        -   More likely: No college
    -   Not meaningful
        -   Less likely: Less than associate degree or higher
        -   More likely: No college
    -   No MN employment record
        -   Less likely: Some college or higher
        -   More likely: No college
-   Five years after graduating;
    -   Have meaningful employment in region
        -   Less likely: Bachelor degree or higher
        -   More likely: Less than associate degree to associate degree
    -   Meaningful emp in MN
        -   Less likely: Bachelor degree
        -   More likely: Master degree or higher
    -   Not meaningful
        -   Less likely: Master degree or higher
        -   More likely: No college or some college
    -   No MN employment record
        -   Less likely: Associate degree or higher
        -   More likely: No college
-   Ten years after graduating;
    -   Have meaningful employment in region
        -   Less likely: Master degree, No college
        -   More likely: Associate degree, Less than associate degree
    -   Meaningful emp in MN
        -   Less likely: No college, some college, Associate degree, Less than associate degree
        -   More likely: Bachelor degree or higher
    -   Not meaningful
        -   Less likely: Associate degree or higher
        -   More likely: Some college
    -   No MN employment record
        -   Less likely: Less than Associate degree
        -   More likely:

::: panel-tabset
### One year after graduating

```{r post-secondary highest.cred.level one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, highest.cred.level) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary highest.cred.level five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, highest.cred.level) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary highest.cred.level ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, highest.cred.level) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

## ps.grad.location

-   One year after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside MN
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: Outside Minnesota
        -   More likely:
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota
        -   More likely: Outside Minnesota
    -   No MN employment record
        -   Less likely: Inside and outside region
        -   More likely: Inside and outside Minnesota, outside Minnesota

-   Five years after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside MN
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: In region, inside and outside region, inside and outside MN, outside MN
        -   More likely: In Minnesota
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota
        -   More likely:
    -   No MN employment record
        -   Less likely: Inside and outside region, Inside and outside MN
        -   More likely: outside Minnesota

-   Ten years after graduating;

    -   Have meaningful employment in region
        -   Less likely: In MN, Inside and outside MN, Outside Minnesota
        -   More likely: In region
    -   Meaningful emp in MN
        -   Less likely: In region
        -   More likely: In Minnesota, Inside and outside Minnesota
    -   Not meaningful
        -   Less likely: Inside and outside region, Inside and outside Minnesota, outside Minnesota
        -   More likely:
    -   No MN employment record
        -   Less likely: In region
        -   More likely: Inside and outside Minnesota, Outside minnesota

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.location one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.1 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.location five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.5 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.location ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.in.same.edr, ps.grad.in.MN) %>%
  filter(grad.year.10 != "After 2023") %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  droplevels() %>%
  filter(!ps.grad.location %in% c("Did not grad", "Never attended ps")) %>%
  mutate(ps.grad.location = fct_relevel(ps.grad.location, "In region", "Inside and outside region", "In MN", "Inside and outside MN", "Outside Minnesota"))

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

<br>

# Variables to keep

Okay, after this analysis, here is what we will move forward with;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   english.learner
-   High school characteristics
    -   Dem_Desc
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ap.exam
    -   ACTCompositeScore
    -   MCA.M
    -   MCA.R
    -   MCA.S
    -   cte.achievement
    -   n.cte.courses.taken
-   Post-secondary
    -   ps.grad
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   ps.grad.location

```{r create new master after analysis}
master.after.ct <- meaningful.emp.central %>%
  mutate(ps.grad.location = ifelse(ps.grad.in.same.edr == "In same EDR", "In region",
                                   ifelse(ps.grad.in.same.edr == "Outside EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                          ifelse(ps.grad.in.same.edr == "Inside and outside same EDR" & ps.grad.in.MN == "In Minnesota", "In state",
                                                 ifelse(ps.grad.in.MN == "Outside MN", "Outside Minnesota",
                                                        ifelse(ps.grad.in.same.edr == "Inside and outside same EDR", "Inside and outside region",
                                                        ifelse(ps.grad.in.MN == "Inside and outside MN", "Inside and outside MN", as.character(ps.grad.in.MN)))))))) %>%
  select(PersonID, grad.year.covid, economic.status, , pseo.participant, SpecialEdStatus, RaceEthnicity, Dem_Desc, took.ACT, cte.achievement, english.learner, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location, ACTCompositeScore, MCA.M, MCA.R, MCA.S, total.cte.courses.taken,grad.year.1, grad.year.5, grad.year.10)
  
write_csv(master.after.ct, "Data/SLEDS/Masters/After analysis/Master-after-ct.csv")
```
