---
title: "Post-secondary Paths CART"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
library(rpart)
library(rpart.plot)
library(tidymodels)
library(caret)
library(themis)
library(ranger)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-after-ct.csv") %>%
  mutate(ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "1", "Public, 4-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "4", "Public, 2-year", as.character(ps.grad.InstitutionSector)),
         ps.grad.InstitutionSector = ifelse(ps.grad.InstitutionSector == "11", "Other", as.character(ps.grad.InstitutionSector)),
         ps.grad.location = ifelse(ps.grad.location == "PS grad EDR", "PS grad region", ps.grad.location),
         highest.cred.level = as.factor(highest.cred.level),
         ps.grad.location = as.factor(ps.grad.location))

kable(names(master.original))

kable(head(master.original))
```

# Post-secondary Paths

In the previous analyses, post-secondary variables rose to the top again and again as a primary factor in whether an individual had meaningful employment in the region or not. In particular, these two variables were the highest and broke in the following ways;

## **Postsecondary Graduation Location (`ps.grad.location`)**

Across all three modeling scenarios we’ve explored — comparing meaningful employment in the region to employment in Minnesota, to not meaningful employment, and to no Minnesota employment record — `ps.grad.location` consistently emerged as one of the most important predictors.

Graduating from a postsecondary institution **within the region** was strongly associated with a higher likelihood of being meaningfully employed in the region, suggesting that geographic alignment between educational attainment and local labor markets fosters stronger employment ties.

Graduating outside the region, particularly outside Minnesota, was linked with an increased likelihood of either securing meaningful employment elsewhere or having no Minnesota employment record at all.

These patterns point to the potential “stickiness” of local graduates in the regional workforce, and the challenges of retaining graduates who leave to complete their education.

## **Highest Credential Level (`highest.cred.level`)**

`highest.cred.level` was a dominant predictor in multiple models, particularly when distinguishing between individuals who were meaningfully employed in the region and those who were not.

Earning an **associate degree or higher** — and especially a **bachelor’s degree or higher** — was linked with greater odds of being meaningfully employed in the region.

Lower levels of attainment, including “some college” without a credential, “less than associate degree,” or “no college,” were associated with higher rates of disconnectedness (no Minnesota employment record, not attending postsecondary) or employment outside the region.

This reinforces the central role of credential attainment in improving employment stability and alignment with regional labor needs.

------------------------------------------------------------------------

For both of these variables, we will combine classifications in order to have a smaller set to model against;

-   ps.grad.location
    1.  PS grad region
    2.  PS grad outside region: includes all individuals that graduated from college outside of the region.
    3.  Did not graduate ps
-   highest.cred.level
    1.  Associate degree or lower
    2.  Bachelors or higher
    3.  Did not graduate: includes individuals with some college and no college experience.

```{r}
master <- master.original %>%
  mutate(ps.grad.location.new = ifelse(ps.grad.location %in% c("PS grad MN", "PS grad outside MN"), "PS grad outside region", as.character(ps.grad.location)),
         highest.cred.level.new = ifelse(highest.cred.level %in% c("Associate degree", "Less than Associate Degree"), "Associate degree or lower", as.character(highest.cred.level)),
         highest.cred.level.new = ifelse(highest.cred.level.new %in% c("No college", "Some college"), "Did not graduate ps", as.character(highest.cred.level.new)))

kable(names(master))

kable(head(master))
```

<br>

# Methodology

This study investigates the factors that influence what secondary education pathway a high school graduate follows. To isolate and understand the predictors of this specific outcome, the analysis compares it separately against three alternative post-secondary variables:

1.  Post-secondary graduation location
2.  Highest Credential Earned in Post-secondary

Graduates classified as *Attending post-secondary* for either of these are excluded from the analysis.

The modeling process follows these steps:

1.  **Data Preparation** All predictor variables include demographic characteristics, high school characteristics, and high school achievement indicators. Records with missing target outcome categories are excluded.

2.  **Training and Test Split** For each post-secondary variable, the data is randomly split into a training set (70–80%) and a testing set (20–30%) to ensure that model evaluation is conducted on previously unseen data.

3.  **Modeling**

    We are going to use the following method to each comparison:

    -   **CART (Classification and Regression Tree)** for interpretability and identification of clear decision rules. We will start by developing the base model, and then do pre- and post-pruning to determine which model provides the best accuracy.

4.  **Evaluation** Models are evaluated on the testing set using accuracy, Cohen’s kappa, balanced accuracy, sensitivity, and specificity. Variable importance is examined to identify the most influential predictors for each comparison and time point.

This approach emphasizes clarity in interpretation by focusing each mode three post-secondary distinctions, making it easier to identify which factors most strongly differentiate graduates who follow pathways that lead to secondary pathways that benefit regional workforce employment.

<br>

# Analysis - Post secondary location

```{r}
data <- master %>%
    select(economic.status, RaceEthnicity, SpecialEdStatus, Dem_Desc, avg.unemp.rate, avg.wages.pct.state, ACTCompositeScore, pseo.participant, cte.achievement, total.cte.courses.taken, MCA.M, MCA.R, MCA.S, ps.grad.location.new, highest.cred.level.new)


kable(names(data))

kable(head(data))
```

The table below shows that 58% of the individuals did not graduate college, 30% graduated outside their high school region, and only 12% graduated from a college within their region.

```{r}
ps.grad.location <- data %>%
  select(-highest.cred.level.new) %>%
  filter(ps.grad.location.new != "Attending ps") %>%
  mutate(ps.grad.location.new = as.factor(ps.grad.location.new))

comma(nrow(ps.grad.location))

datatable(data = ps.grad.location %>%
            group_by(ps.grad.location.new) %>%
            summarize(n = n()) %>%
            ungroup() %>%
            mutate(pct = n / sum(n)) %>%
            ungroup() %>%
            select(1:3),
          class = "cell-border stripe", 
          filter = "top", 
          rownames = FALSE,
          options = list(
            columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)
```

<br>

## Base Model

I know the base model is going to overfit and provide an overly complex tree due to having a cost penalty of 0. So I'm just going to print the complexity plot in order to see what the appropriate CP would be to minimize xerror.

```{r}
set.seed(1234)

sample_ind <- sample(nrow(ps.grad.location), nrow(ps.grad.location) *.7)

train <- ps.grad.location[sample_ind,]

test <- ps.grad.location[-sample_ind,]

#Base model

ps.grad.location.model <- rpart(ps.grad.location.new ~ ., data = train, method = "class", control = rpart.control(cp = 0))

# Accuracy

test$pred <- predict(ps.grad.location.model, test, type = "class")

accuracy_base <- mean(test$pred == test$ps.grad.location.new)

percent(accuracy_base, accuracy = .1)

# Examine the complexity plot
printcp(ps.grad.location.model)

plotcp(ps.grad.location.model)
```

**Key takeaways from base tree results**

-   **Accuracy**: 63.6%
-   **Variables used:** 13 total, spanning demographics, high school characteristics, and high school accomplishments (e.g., ACTCompositeScore, avg.unemp.rate, avg.wages.pct.state, CTE achievement, economic status, MCA scores, pseo.participant, RaceEthnicity, SpecialEdStatus, total CTE courses taken).
-   **Lowest cross-validation error:** 0.79443 at cp ≈ 0.000097101 (nsplit = 277).

<br>

## Pre-pruning

Next, we need to prune. We can either pre-prune or post-prune. We will start with pre-pruning and use each method - max depth, min depth, and min bucket. It's recommended that I start with the following parameters;

-   maxdepth: 6
-   minsplit: round(2% of nrow(train)) which is 1617.
-   minbucket: round(minsplit / 3) which is 557
-   cp = 0

<br>

```{r, echo=TRUE}
maxdepth = 6

minsplit = round(.02*nrow(train))

minbucket = round(minsplit / 3)

ps.grad.location.model.preprune <- rpart(ps.grad.location.new ~ ., data = train, method = "class", 
                   control = rpart.control(cp = 0, maxdepth = maxdepth, minsplit = minsplit, minbucket = minbucket))

# Accuracy

test$pred <- predict(ps.grad.location.model.preprune, test, type = "class")

accuracy_preprun <- mean(test$pred == test$ps.grad.location.new)

percent(accuracy_preprun, accuracy = .1)

# Examine the complexity plot
printcp(ps.grad.location.model.preprune)

plotcp(ps.grad.location.model.preprune)
```

**Key Takeaways**

-   **Accuracy:** 65.9% — notable improvement over the base model’s 63.6%.
-   **Variables used (8 total):** A mix of high school accomplishments (ACTCompositeScore, pseo.participant, cte.achievement), Demographics (economic.status, RaceEthnicity, SpecialEdStatus) and high school characteristics (avg.wages.pct.state)
-   **Tree complexity:** Pre-pruning capped growth to **maxdepth**, resulting in 15 splits vs. 1000+ in the base model.
-   **Lowest cross-validation error (xerror):** ≈ 0.81799 at **cp ≈** 0.00045695 (nsplit = 12).

<br>

## Post-pruning

The idea here is to allow the decision tree to grow fully and observe the CP value. There are two methods for identifying the CP code;

1.  Use the cp value with the lowest xerror from the base model.
2.  Use the cp value with the next lowest xerror that is within the standard error.

<br>

```{r , echo=TRUE}

cp_table <- ps.grad.location.model.preprune$cptable

min_xerror <- min(cp_table[, "xerror"])
min_xerror_se <- cp_table[which.min(cp_table[,"xerror"]), "xstd"]

threshold <- min_xerror + min_xerror_se

one_se_index <- which(cp_table[, "xerror"] <= threshold)[1] 

cp_1se <- cp_table[one_se_index, "CP"]

optimal_cp <- ps.grad.location.model.preprune$cptable[which.min(ps.grad.location.model.preprune$cptable[, "xerror"]), "CP"]


# Prune the hr_base_model based on the optimal cp value
ps.grad.location.model.postprune <- rpart::prune(ps.grad.location.model.preprune, cp = cp_1se)

# Compute the accuracy of the pruned tree

test$pred <- predict(ps.grad.location.model.postprune, test, type = "class")

accuracy_postprun <- mean(test$pred == test$ps.grad.location.new)

percent(accuracy_postprun, accuracy = .1)

# Examine the complexity plot
printcp(ps.grad.location.model.postprune)

plotcp(ps.grad.location.model.postprune)
```

**Key Takeaways**

-   **Accuracy:** 65.8% — same as pre-pruned.
-   **Variables used (6 total):** HIgh school achievements (ACTCompositeScore, pseo.participant), High school characteristics (avg.wages.pct.state), Demographics (economic.status, RaceEthnicity, SpecialEdStatus)
-   **Tree complexity:** We simplified the tree to just 10 splits, compared to 15 for the pre-pruned tree.
-   **Lowest cross-validation error (xerror):** ≈ 0.82011 at **cp ≈** 0.0019849 (nsplit = 10).

So, the accuracy of these are clear;

-   Base model: `r percent(accuracy_base, accuracy = .1)`
-   Pre-pruned model: `r percent(accuracy_preprun, accuracy = .1)`
-   Post-pruned model: `r percent(accuracy_postprun, accuracy = .1)`

The post-pruned model provides a much simpler model - 10 splits instead of pre-pruned's 15 splits - while also giving the same accuracy. I will dive into that one below.

```{r}
#Plot tree
rpart.plot(ps.grad.location.model.postprune, nn = TRUE)
```

## **Main takeaway**

**Main Splits**\

The first and most influential split is **PSEO participation**. Individuals who did not participate in PSEO are much more likely to fall into the “Did not graduate PS” category. Among non-PSEO participants, the next split is **Race/Ethnicity** (AI/Black/Hispanic/Unknown vs. other groups), followed by **Special Education status** and **Economic status**. For PSEO participants, the main split is also **Race/Ethnicity** (AI/Unknown vs. other groups), followed by **Economic status**, and then **ACT Composite Score** thresholds (\< 21, \< 22, \< 26) along with **average wages as a percent of state average**. These combinations differentiate between “Did not graduate PS,” “PS grad outside region,” and “PS grad region.”

**Variable Importance**

1.  **PSEO participation** – Primary driver of the first split, strongest influence overall.
2.  **Race/Ethnicity** – Consistent second-level split on both sides of the tree.
3.  **Economic status** – Influential in both PSEO and non-PSEO branches.
4.  **ACT Composite Score** – Multiple threshold splits influence postsecondary graduation location.
5.  **Special Education status** – Key split for non-PSEO students from certain racial/ethnic groups.
6.  **Average wages (relative to state)** – Smaller role, only relevant in a limited branch.

**Interpretation** Participation in PSEO strongly predicts likelihood of postsecondary graduation and whether it occurs within the high school region. Students without PSEO experience tend to have lower postsecondary graduation rates, especially those from underrepresented racial/ethnic groups, with special education needs, or from lower economic backgrounds. For students with PSEO experience, ACT score thresholds and economic status further refine the likelihood of regional versus non-regional graduation. Higher ACT scores are associated with greater odds of graduating outside the region, while economic disadvantages tilt toward non-completion. Wage levels relative to the state average only appear as a predictor in a narrow subgroup, suggesting it’s not a major factor overall.

**Brief Conclusion** The model shows that PSEO participation is the most significant factor in predicting postsecondary graduation location, with race/ethnicity, economic status, and ACT performance providing additional differentiation. Non-PSEO participants—especially from underrepresented racial/ethnic groups and lower economic backgrounds—are much more likely not to graduate. Among PSEO participants, academic performance and economic status determine whether graduation is more likely to occur within or outside the high school region.

<br>

# Analysis - Highest Credential Earned

The table below shows that 58% of the individuals did not graduate college, 28% graduated with a Bachelor degree or higher, and 14% graduated with an Associate degree or lower.

```{r}
highest.cred.level <- data %>%
  select(-ps.grad.location.new) %>%
  filter(highest.cred.level.new != "Attending ps") %>%
  mutate(highest.cred.level.new = as.factor(highest.cred.level.new))

comma(nrow(highest.cred.level))

datatable(data = highest.cred.level %>%
            group_by(highest.cred.level.new) %>%
            summarize(n = n()) %>%
            ungroup() %>%
            mutate(pct = n / sum(n)) %>%
            ungroup() %>%
            select(1:3),
          class = "cell-border stripe", 
          filter = "top", 
          rownames = FALSE,
          options = list(
            columnDefs = list(list(className = "dt-center", targets = 1:2)))) %>%
  formatCurrency(2, "", digits = 0) %>%
  formatPercentage(3, digits = 1)
```

<br>

## Base Model

I know the base model is going to overfit and provide an overly complex tree due to having a cost penalty of 0. So I'm just going to print the complexity plot in order to see what the appropriate CP would be to minimize xerror.

```{r}
set.seed(1234)

sample_ind <- sample(nrow(highest.cred.level), nrow(highest.cred.level) *.7)

train <- highest.cred.level[sample_ind,]

test <- highest.cred.level[-sample_ind,]

#Base model

highest.cred.level.model <- rpart(highest.cred.level.new ~ ., data = train, method = "class", control = rpart.control(cp = 0))

# Accuracy

test$pred <- predict(highest.cred.level.model, test, type = "class")

accuracy_base <- mean(test$pred == test$highest.cred.level.new)

percent(accuracy_base, accuracy = .1)

# Examine the complexity plot
printcp(highest.cred.level.model)

plotcp(highest.cred.level.model)
```

**Key takeaways from base tree results**

-   **Accuracy**: 65.2%
-   **Variables used:** 13 total, spanning demographics, high school characteristics, and high school accomplishments (e.g., ACTCompositeScore, avg.unemp.rate, avg.wages.pct.state, CTE achievement, economic status, MCA scores, pseo.participant, RaceEthnicity, SpecialEdStatus, total CTE courses taken).
-   **Lowest cross-validation error:** 0.75361 at cp ≈ 0.00012866 (nsplit = 148).

<br>

## Pre-pruning

Next, we need to prune. We can either pre-prune or post-prune. We will start with pre-pruning and use each method - max depth, min depth, and min bucket. It's recommended that I start with the following parameters;

-   maxdepth: 6
-   minsplit: round(2% of nrow(train)) which is 1617.
-   minbucket: round(minsplit / 3) which is 557
-   cp = 0

<br>

```{r, echo=TRUE}
maxdepth = 6

minsplit = round(.02*nrow(train))

minbucket = round(minsplit / 3)

highest.cred.level.model.preprune <- rpart(highest.cred.level.new ~ ., data = train, method = "class", 
                   control = rpart.control(cp = 0, maxdepth = maxdepth, minsplit = minsplit, minbucket = minbucket))

# Accuracy

test$pred <- predict(highest.cred.level.model.preprune, test, type = "class")

accuracy_preprun <- mean(test$pred == test$highest.cred.level.new)

percent(accuracy_preprun, accuracy = .1)

# Examine the complexity plot
printcp(highest.cred.level.model.preprune)

plotcp(highest.cred.level.model.preprune)
```

**Key Takeaways**

-   **Accuracy:** 67.4% — notable improvement over the base model’s 65.2%.
-   **Variables used (6 total):**
    -   High school accomplishments: ACTCompositeScore, cte.achievement, pseo.participant
    -   Demographics: economic.status, RaceEthnicity, SpecialEdStatus
-   **Tree complexity:** Pre-pruning capped growth to **maxdepth**, resulting in 14 splits vs. 1000+ in the base model.
-   **Lowest cross-validation error (xerror):** ≈ 0.79172 at **cp ≈** 0.00211568 (nsplit = 10).

<br>

## Post-pruning

The idea here is to allow the decision tree to grow fully and observe the CP value. There are two methods for identifying the CP code;

1.  Use the cp value with the lowest xerror from the base model.
2.  Use the cp value with the next lowest xerror that is within the standard error.

<br>

```{r , echo=TRUE}

cp_table <- highest.cred.level.model.preprune$cptable

min_xerror <- min(cp_table[, "xerror"])
min_xerror_se <- cp_table[which.min(cp_table[,"xerror"]), "xstd"]

threshold <- min_xerror + min_xerror_se

one_se_index <- which(cp_table[, "xerror"] <= threshold)[1] 

cp_1se <- cp_table[one_se_index, "CP"]

optimal_cp <- highest.cred.level.model.preprune$cptable[which.min(highest.cred.level.model.preprune$cptable[, "xerror"]), "CP"]


# Prune the hr_base_model based on the optimal cp value
highest.cred.level.model.postprune <- rpart::prune(highest.cred.level.model.preprune, cp = cp_1se)

# Compute the accuracy of the pruned tree

test$pred <- predict(highest.cred.level.model.postprune, test, type = "class")

accuracy_postprun <- mean(test$pred == test$highest.cred.level.new)

percent(accuracy_postprun, accuracy = .1)

# Examine the complexity plot
printcp(highest.cred.level.model.postprune)

plotcp(highest.cred.level.model.postprune)
```

**Key Takeaways**

-   **Accuracy:** 67.4% — same as pre-pruned.
-   **Variables used (6 total):**
    -   High school accomplishments: ACTCompositeScore, cte.achievement, pseo.participant
    -   Demographics: economic.status, RaceEthnicity, SpecialEdStatus
-   **Tree complexity:** We simplified the tree to just 13 splits, compared to 14 for the pre-pruned tree.
-   **Lowest cross-validation error (xerror):** ≈ 0.78317 at **cp ≈** 0.00040026 (nsplit = 13).

So, the accuracy of these are clear;

-   Base model: `r percent(accuracy_base, accuracy = .1)`
-   Pre-pruned model: `r percent(accuracy_preprun, accuracy = .1)`
-   Post-pruned model: `r percent(accuracy_postprun, accuracy = .1)`

The post-pruned model provides a bit of a simpler model - 13 splits instead of pre-pruned's 14 splits - while also giving the same accuracy. I will dive into that one below.

```{r}
#Plot tree
rpart.plot(highest.cred.level.model.postprune, nn = TRUE)
```

## **Main takeaway**

**Main Splits** The most influential split is **PSEO participation**. Students without PSEO experience are more likely not to graduate from postsecondary education, while those with PSEO experience show higher rates of earning a bachelor’s degree or higher. For non-PSEO participants, the next split is **Race/Ethnicity** (Asian/PI/Black/Hispanic/White vs. others), followed by **ACT Composite Score** thresholds (≥ 21 and ≥ 26), with **economic status** and **CTE achievement** further refining outcomes. For PSEO participants, **ACT Composite Score** (≥ 20) is the initial determinant, followed by **Race/Ethnicity**, **economic status**, and **Special Education status**. In some branches, lower ACT scores (\< 24) and special education status predict much higher rates of non-completion.

**Variable Importance**

1.  **PSEO participation** – Strongest driver of overall classification, key root split.

2.  **ACT Composite Score** – Appears multiple times with different thresholds (18, 20, 21, 24, 26), strongly differentiating educational outcomes.

3.  **Race/Ethnicity** – Consistent second-level split, shaping degree level expectations.

4.  **Economic status** – Important in both PSEO and non-PSEO paths, associated with lower completion at higher credential levels.

5.  **CTE achievement** – Plays a smaller but notable role for high-scoring non-PSEO students.

6.  **Special Education status** – Particularly important for lower ACT PSEO students.

**Interpretation** PSEO participation sharply divides the likelihood of postsecondary credential attainment. Non-PSEO students—particularly from certain racial/ethnic groups and with lower ACT scores—are far more likely to not graduate, while higher ACT scores and favorable economic conditions increase the odds of earning bachelor’s degrees. For PSEO students, ACT performance remains a strong determinant, with high scorers more likely to reach bachelor’s or higher credentials, and low scorers—especially those from disadvantaged economic backgrounds or with special education status—more likely not to complete. CTE achievement provides a modest boost in certain high-performing, non-PSEO cases.

**Brief Conclusion** The model underscores PSEO participation and ACT performance as the two most critical predictors of highest credential earned, with race/ethnicity, economic status, and—more selectively—CTE achievement and special education status providing additional nuance. High ACT scores combined with PSEO participation strongly favor bachelor’s-level attainment, while low scores, economic disadvantage, and lack of PSEO participation correspond with non-completion.

<br>

# Comparison

Both models identify **PSEO participation** as the strongest predictor, but the way subsequent variables shape outcomes differs.

For **`ps.grad.location.new`**, the primary narrative is about *where* postsecondary graduation occurs—or if it happens at all. After PSEO participation, **Race/Ethnicity** is the next most influential factor in both branches, followed by **Economic status** and multiple **ACT Composite Score** thresholds. Special Education status plays a role mainly for non-PSEO students, while **average wages (relative to state)** only appears in a small branch. The outcome paths focus on the geographic context of graduation (within region, outside region, or no graduation).

For **`highest.cred.level`**, the emphasis is on *how far* a student progresses academically. After PSEO participation, **ACT Composite Score** takes a larger role in differentiating degree levels, with several score thresholds driving major branches. **Race/Ethnicity** still appears early but is sometimes secondary to ACT performance. **Economic status** consistently lowers the likelihood of earning bachelor’s degrees, while **CTE achievement** and **Special Education status** appear in smaller but notable branches. The outcome paths here are more strongly tied to academic performance and degree completion level.

**Key Differences**

-   **Second Split**: Location model → Race/Ethnicity; Credential model → ACT Composite Score.

-   **Geography vs. Attainment**: Location model’s later splits focus on regional vs. non-regional graduation; Credential model’s later splits focus on associate vs. bachelor’s degree attainment.

-   **Unique Predictors**: Location model includes average wages; Credential model includes CTE achievement.

-   **Special Education Influence**: Appears earlier and more prominently in Credential model’s PSEO branch, but mostly in the non-PSEO branch for Location model.

**Brief Conclusion** While both models agree that PSEO participation is the critical starting point, the determinants that follow diverge according to the outcome of interest. Geographic graduation patterns are more closely tied to demographic and economic factors, whereas credential level is more directly tied to academic performance and CTE involvement.
