---
title: "Local Employment Cross-tabs"
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
library(gmodels)
library(fastDummies)
library(car)
library(glmnet)
library(glmnetUtils)
library(pscl)
library(sjPlot)
library(ggforce)
```

```{r themes and shapefiles, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))

counties.regions.1 <- counties.regions %>%
  mutate(statefp = "27",
         project.pr = ifelse(edr %in% c("EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 4 - West Central"), "Northwest",
                             ifelse(edr == "EDR 3 - Arrowhead", "Northeast",
                                    ifelse(edr %in% c("EDR 5 - North Central", "EDR 7E- East Central", "EDR 7W- Central"), "Central",
                                           ifelse(edr %in% c("EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 8 - Southwest"), "Southwest",
                                                  ifelse(edr %in% c("EDR 9 - South Central", "EDR 10 - Southeast"), "Southern", as.character(planning.region)))))),
         project.pr = fct_relevel(project.pr, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southern"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)
```

```{r master dataset}
master.original <- read_csv("Data/SLEDS/Masters/After analysis/Master-final-after-mca.csv")

master.factor.names <- master.original %>%
  select(PersonID, Dem_Desc, edr, grad.year.covid, Gender, economic.status, pseo.participant, SpecialEdStatus, non.english.home, RaceEthnicity, took.ACT, ap.exam, cte.achievement, english.learner, sat.taken, attended.ps, attended.ps.within.first.year.hsgrad, ps.grad, ps.grad.InstitutionSector, highest.cred.level, ps.grad.location) 

master.states <- read_csv("Data/SLEDS/Masters/States/annual-seven-states.csv") %>%
  filter(grad.year %in% c("grad.year.1", "grad.year.5", "grad.year.10")) %>%
  select(PersonID, grad.year, states) %>%
  pivot_wider(names_from = grad.year, values_from = states)

master <- master.states %>%
  left_join(master.factor.names, by = "PersonID")

kable(head(master))

kable(names(master))
```

```{r independent and dependent variables}
updated.factor.var.names <-  master.factor.names %>%
  select(-PersonID) %>%
  names() %>%
  c("ps.grad.in.same.ruca", "ps.grad.in.same.edr", "ps.grad.in.same.pr", "ps.grad.in.MN")

numeric.var.names <- master.original %>%
  select(ACTCompositeScore, total.cte.courses.taken, `ACTE.SPED`:Youth.Apprenticeship,  MCA.M:MCA.S, n.institutions, avg.unemp.rate, wages.3year.avg, avg.wages.pct.state) %>%
  names()
  
updated.independent.var.names <- c(updated.factor.var.names, numeric.var.names)

cte.cc.var.names <- master.original %>%
  select(ACTE.SPED:Youth.Apprenticeship) %>%
  names()

dep.var.names <- master %>%
  select(PersonID, grad.year.1, grad.year.5, grad.year.10)

updated.master <- master.original %>%
  select(PersonID, all_of(updated.independent.var.names)) %>%
  left_join(dep.var.names, by = "PersonID") %>%
  mutate_at(vars(Dem_Desc:ps.grad.in.MN), as.factor) %>%
  mutate_at(vars(grad.year.1:grad.year.10), as.factor) %>%
  mutate(grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp County", "Meaningful emp EDR", "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps"),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Did not grad", "Some college", as.character(highest.cred.level)),
         highest.cred.level = ifelse(ps.grad.InstitutionSector == "Never attended ps", "No college", as.character(highest.cred.level)),
         highest.cred.level = as.factor(highest.cred.level),
         Dem_Desc = fct_relevel(Dem_Desc, "Town/rural mix", "Urban/town/rural mix", "Entirely urban")) %>%
  droplevels()

kable(head(updated.master))

kable(names(updated.master))
```

<br>

There are a number of "states" in which an individual in this dataset are assigned;

1.  Has meaningful employment
    1.  In the same county as their high school.
    2.  In the same EDR as their high school, but not county.
    3.  In Minnesota, but not in Central Minnesota.
2.  Is attending post-secondary.
3.  Has a MN employment record but it's not meaningful, and not attending post-secondary.
4.  Does not have a MN employment record and is not attending post-secondary.

There are three "times" in which we checked to see if the individual had meaningful employment within these geographies;

1.  one year after graduating high school
2.  five years after graduation, and
3.  ten years after graduation.

Meaningful employment is determined by whether an individual worked 1,000 hours for an employer during time X. In addition, if the individual is working for an employer but not meaningful at time X but worked 1,000 hours for that employer during another year it's still considered "meaningful".

Due to time x potentially being passed the date of the latest data (2023) for some individuals, the analysis below will filter out all individuals where time x is after 2023.

From the multiple correspondence analysis and the principle component analysis we determined that the following independent variables will be included in the analysis.

```{r list of ind var}
updated.master %>%
  select(-PersonID, -grad.year.1, -grad.year.5, -grad.year.10) %>%
  lapply(class) 

```

# Methodology

For this analysis we are keeping it simple. For categorical variables I’ll be using the Cramer’s v-square measure of association. The V\^2 value is from 0 to 1 with 0 meaning there’s no association and a value closer to 1 indicates a strong association.

| **Cramér’s V Value** | **Interpretation**               |
|----------------------|----------------------------------|
| 0.00–0.10            | Negligible association           |
| 0.10–0.20            | Weak                             |
| 0.20–0.40            | Moderate                         |
| 0.40–0.60            | Relatively strong                |
| 0.60–0.80            | Strong                           |
| 0.80–1.00            | Very strong (possibly redundant) |

For the variables with continuous values I’ll be using the Wilcoxon two paired groups test. P-value will be used to determine whether we can reject the NULL hypothesis at a .01 level. I’ll also include boxplots to see in which direction there might be influence from the variables.

We are going to split up the analysis into two different datasets. The first dataset will combine 3 of the previous states into one.

-   Has meaningful employment anywhere in Central Minnesota, which combines the following states;
    -   Meaningful emp County
    -   Meaningful emp EDR
    -   Meaningful emp region

The remaining states will stay the same.

-   Has meaningful employment in Minnesota, but not in Central Minnesota
-   Is attending a post-secondary institution.
-   Has a MN employment record but it's not meaningful, and they are not attending post-secondary.
-   Does not have a MN employment record and is not attending post-secondary.

We have quite a few variables remaining, even after MCA analysis. So let's take a look at how we are going to proceed and group all of this together to see how these relationships work.

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner
-   High school characteristics
    -   Dem_Desc
    -   edr
    -   avg.unemp.rate
    -   avg.wage.pct.state
-   High school accomplishments
    -   pseo.participant
    -   took.ACT
    -   ACTCompositeScore
    -   ap.exam
    -   sat.taken
    -   MCA scores
        -   MCA.M
        -   MCA.R
        -   MCA.S
    -   cte.achievement
    -   total.cte.courses.taken
-   Post-secondary
    -   attended.ps.within.first.year.hsgrad
    -   attended.ps
    -   ps.grad.
    -   ps.grad.InstitutionSector
    -   highest.cred.level
    -   Combine post-secondary grad location
        -   ps.grad.in.same.ruca
        -   ps.grad.in.same.edr
        -   ps.grad.in.same.pr
        -   ps.grad.in.MN

<br>

# Meaningful employment in Central Minnesota

```{r prep meaningful employment in Central MN}
meaningful.emp.central <- updated.master %>%
  mutate(grad.year.1 = ifelse(grad.year.1 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.1)),
         grad.year.1 = as_factor(grad.year.1),
         grad.year.1 = fct_relevel(grad.year.1, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.5 = ifelse(grad.year.5 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.5)),
         grad.year.5 = as_factor(grad.year.5),
         grad.year.5 = fct_relevel(grad.year.5, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         grad.year.10 = ifelse(grad.year.10 %in% c("Meaningful emp County", "Meaningful emp EDR"), "Meaningful emp region", as.character(grad.year.10)),
         grad.year.10 = as_factor(grad.year.10),
         grad.year.10 = fct_relevel(grad.year.10, "Meaningful emp region", "Meaningful emp MN", "Attending ps", "Not meaningful, not attending ps", "No MN emp record, not attending ps", "After 2023"),
         highest.cred.level = fct_relevel(highest.cred.level, "No college", "Some college", "Less than Associate Degree", "Associate degree", "Bachelor degree or higher"))

write_csv(meaningful.emp.central, "Data/SLEDS/Masters/After analysis/Master-meaningful-emp-central.csv")
```

Let's first take a look at proportion of states in each of the time.X. The chart below shows that half of the graduates are attending post-secondary (50%) one year after graduating high school. The further away from high school graduation results in a bit of an increase in the percentage of individuals who have meaningful employment in the region and in Minnesota, while about a quarter of graduates do not have a MN employment record and are not going to post-secondary.

```{r proportion of states meaningful emp SW}
prop.states.meaningful.emp.central <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10) %>%
  pivot_longer(names_to = "grad.year", values_to = "states", 1:3) %>%
  group_by(grad.year, states) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  filter(states != "After 2023") %>%
  group_by(grad.year) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         grad.year = fct_relevel(grad.year, "grad.year.1", "grad.year.5", "grad.year.10"))

prop.states.meaningful.emp.central.plot <- ggplot(prop.states.meaningful.emp.central, aes(grad.year, pct, fill = states, group = states)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(grad.year, "\nState of individuals: ", states, "\nNumber of people in states: ", comma(n), "\nPercent of people in state: ", percent(pct, accuracy = .1), sep = ""))) +
   labs(x="", y = "", color="", title = "Percent of people in each state")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu"),
                    guide = guide_legend(ncol = 2)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = prop.states.meaningful.emp.central.plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

Let's examine the variables to see if there are any significant differences in the proportion of individuals in these states across the variable categories or values.

<br>

# Demographics

In the demographics section we will examine the relationship between that paths taken by individuals and the following variables;

-   Demographics
    -   Gender
    -   grad.year.covid
    -   economic.status
    -   SpecialEdStatus
    -   RaceEthnicity
    -   non.english.home
    -   englisher.learner

## Summary of all variables

**Strength of Association (Cramer's V by Year)**

| Variable | Year 1 V | Year 5 V | Year 10 V | Peak Strength | Pattern of Change |
|------------|:----------:|:----------:|:----------:|------------|------------|
| **Gender** | 0.127 | 0.081 | 0.073 | Moderate (Year 1) | Declines over time |
| **Grad.year.covid** | 0.052 | — | — | Weak (Year 1) | Single-year data only |
| **economic.status** | 0.235 | 0.122 | 0.085 | Strong (Year 1) | Gradual decline |
| **SpecialEdStatus** | 0.248 | 0.145 | 0.084 | Strong (Year 1) | Declines but still meaningful |
| **RaceEthnicity** | 0.059 | 0.041 | 0.038 | Weak (Year 1) | Small but persistent |
| **non.english.home** | 0.101 | 0.073 | 0.066 | Weak–Moderate | Modest decline |
| **english.learner** | 0.139 | 0.084 | 0.068 | Moderate (Year 1) | Persistent, narrowing gap |

**Key Patterns & Comparative Insights**

**1. Strongest Predictors of Outcome Differences**

-   **SpecialEdStatus** and **economic.status** had the **largest effect sizes**, especially one year after graduation. Both were associated with **lower postsecondary participation** and **greater disconnection from meaningful work**, with disparities that persisted across a decade.
-   These two variables reflect **structural barriers** — economic and learning support — that strongly shape early and long-term outcomes.

**2. Language-Related Variables**

-   Both **english.learner** and **non.english.home** showed **moderate effects early on**.
-   Students with these language indicators were **less likely to attend postsecondary** and **more likely to appear in disconnected categories**.
-   While **gaps narrowed over time**, persistent differences suggest long-term barriers related to language acquisition and access.

**3. Gender**

-   Gender had a **moderate short-term effect** that decreased over time.
-   Males were **more likely to take local employment**, while females were **more likely to attend postsecondary**.
-   The gap gradually narrowed but **did not fully disappear** even after 10 years.

**4. RaceEthnicity**

-   Despite its importance, race/ethnicity had **weaker statistical associations** than other factors.
-   However, group-level disparities — especially for **AI, Black, and Hispanic** graduates — were **consistent**.
-   **Asian/PI** students showed unique patterns of **early postsecondary participation** and **underrepresentation in local employment**, possibly due to out-of-state or non-traditional career paths.

**5. COVID-era Graduation**

-   Graduation year relative to COVID had a **statistically significant but weak** effect.
-   Post-COVID grads were **less likely to attend PS** and **more likely to work locally**, likely due to **pandemic-era constraints**.

**Conclusion**

Among all the demographic variables, **economic disadvantage and special education status** emerged as the **most powerful and persistent predictors** of postsecondary and employment outcomes. **Language and gender** also played meaningful roles, particularly in the short term, while **race/ethnicity** and **COVID-era graduation** had weaker but still important associations.

These findings suggest that interventions aiming to improve post-high school pathways should prioritize support for **low-income students**, **students receiving special education**, and **multilingual or English learner populations**, especially in the critical first years after graduation.

<br>

## Gender

### **Key Findings**

-   There is a statistically significant relationship between gender and post-graduation outcomes in all three timeframes:
    -   **1 year after graduation**: χ² = 2083.575, df = 4, Cramer's V = **0.127**, *p* \< .001
    -   **5 years after graduation**: χ² = 612.448, df = 4, Cramer's V = **0.081**, *p* \< .001
    -   **10 years after graduation**: χ² = 278.976, df = 4, Cramer's V = **0.073**, *p* \< .001

The strength of association is **moderate in the first year** and diminishes over time to a **weak but still significant** relationship by year ten.

### **Interpretation Over Time**

**1 Year After Graduation**: Females are more likely to be **attending postsecondary education** (55.3% vs. 43.7%), while males are more likely to be employed — especially **within their high school region** (15.3% vs. 10.9%). Males also have higher rates of **no MN employment record** (11.9% vs. 7.7%). These early differences reflect a pattern in which **males enter the local workforce more often**, while **females pursue longer-term educational pathways**.

**5 Years After Graduation**: Postsecondary enrollment declines across the board, but differences remain. Males continue to have higher rates of **regional employment** and are slightly more likely to be employed meaningfully somewhere in MN. The gender gap in disconnected outcomes narrows, suggesting some convergence over time.

**10 Years After Graduation**: Differences persist but are smaller. Males still maintain an edge in meaningful employment — both **in-region and statewide** — while females are slightly more likely to appear in the **"no MN record"** group, possibly due to moving out of state or untracked employment.

### **Conclusion**

Gender-based differences are **most pronounced in the first year**, with males more likely to take **local employment routes** and females more likely to pursue postsecondary education. Over time, the differences **narrow but remain**, especially in terms of where employment is located and whether individuals are tracked in the MN employment system.

<br>

::: panel-tabset
### One year after graduating

```{r demographics gender one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Gender) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)

```

### Five years after graduating

```{r demographics gender five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Gender) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)

```

### Ten years after graduating

```{r demographics gender ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Gender) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Gender, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)

```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, Gender) %>%
  write_csv(., "Data/ChatGPT/Demographics/Gender.csv")
```

<br>

## Grad.year.covid

### **Key Findings**

There is a statistically significant relationship between graduation timing and outcomes **one year after high school** (χ² = 349.128, df = 4, *p* \< .001). The strength of association is **weak**, with a Cramer's V of **0.052**.

### **Interpretation (1 Year After Graduation)**

Graduates who finished high school **after the onset of COVID-19 (2020 or later)** were **less likely to be attending postsecondary education** (45.5% vs. 50.5%) and **more likely to be employed**, particularly in-region (15.0% vs. 12.6%) or statewide (9.2% vs. 7.1%). They also had slightly higher rates of being untracked or disconnected (10.6% with no MN record vs. 9.6%).

This pattern suggests that **COVID-era graduates may have faced barriers to postsecondary access** or opted more often for immediate employment, perhaps due to economic pressures, remote learning challenges, or changes in job availability during the pandemic.

<br>

::: panel-tabset
### One year after graduating

```{r demographics grad.year.covid one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.covid) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$grad.year.covid, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, grad.year.covid) %>%
  write_csv(., "Data/ChatGPT/Demographics/grad-year-covid.csv")
```

<br>

## economic.status

### **Key Findings**

There is a statistically significant relationship between economic status and post-graduation outcomes:

-   **1 year after graduation**: χ² = 7116.172, df = 4, Cramer's V = **0.235**, *p* \< .001
-   **5 years after graduation**: χ² = 1409.165, df = 4, Cramer's V = **0.122**, *p* \< .001
-   **10 years after graduation**: χ² = 373.783, df = 4, Cramer's V = **0.085**, *p* \< .001

The strength of association is **moderate to strong in the first year** and gradually weakens over time, though differences persist.

### **Interpretation Over Time**

**1 Year After Graduation**: Students not eligible for free/reduced lunch were **far more likely to attend postsecondary education** (57.4% vs. 32.6%). Students from low-income backgrounds were more likely to be employed **locally in their high school region** (16.2% vs. 11.7%) and more likely to be disconnected — either in non-meaningful roles (28.5%) or lacking MN employment records (13.5%).

**5 Years After Graduation**: Gaps narrow slightly but remain. Higher-income students continue to show **higher rates of meaningful MN employment**, while low-income students maintain **elevated disconnection** and are more concentrated in **regional jobs**, likely reflecting limited mobility or economic necessity.

**10 Years After Graduation**: Regional employment rates are similar, but higher-income students are still more likely to have **meaningful jobs elsewhere in MN** and **less likely to be disconnected**. The long-term trend suggests that **early disparities in education and employment compound over time**, even as the gap narrows.

### **Conclusion**

Economic status is a **powerful predictor** of early postsecondary and employment outcomes. Students from low-income backgrounds are more likely to take **local jobs out of necessity**, less likely to pursue postsecondary education, and more vulnerable to **long-term disconnection from the workforce**. While disparities decrease by year ten, they remain meaningful.

<br>

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a higher rate of meaningful employment in the region.
-   Have a higher rate of meaningful employment in Minnesota.
-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, economic.status) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Attend post-secondary at a significantly lower rate.
-   Have a higher rate of not having meaningful employment.
-   Have a higher rate of having no MN employment record.

```{r demographics economic.status five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, economic.status) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, weak association between economic status and the employment state.

Individuals that qualify for free or reduced lunch;

-   Have a lower rate of having meaningful employment in Minnesota.
-   Have a higher rate of not having meaningful employment.

```{r demographics economic.status ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, economic.status) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$economic.status, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, economic.status) %>%
  write_csv(., "Data/ChatGPT/Demographics/economic-status.csv")

```

<br>

## SpecialEdStatus

### **Key Findings**

There is a statistically significant relationship between SpecialEdStatus and post-graduation outcomes across all three years:

-   1 **year after graduation**: χ² = 3177.82, df = 4, Cramer's V = **0.248**, *p* \< .001
-   **5 years after graduation**: χ² = 1091.37, df = 4, Cramer's V = **0.145**, *p* \< .001
-   **10 years after graduation**: χ² = 366.90, df = 4, Cramer's V = **0.084**, *p* \< .001

The strength of association is **strongest one year after graduation** and diminishes over time, though remains statistically significant even ten years later.

### **Interpretation Over Time**

**1 Year After Graduation**: Students who received special education services in high school were substantially less likely to be **attending postsecondary education** (only 1,375 compared to 24,769 peers without special ed status) and more likely to be in **non-meaningful or untracked employment categories**, including **“no MN employment record”** and **“not meaningful”** employment. While some were employed locally (838 in-region), the disparities suggest substantial barriers to immediate postsecondary success.

**5 Years After Graduation**: The gap narrows but persists. Special education students still trail their peers in **postsecondary enrollment** and are **less likely to hold meaningful employment in MN** (1,068 vs. 11,204) or **in their local region** (1,529 vs. 9,866). They are overrepresented in **non-meaningful** and **no-record** categories, though many do gain some form of employment by this point.

**10 Years After Graduation**: By year ten, special education students continue to be **underrepresented in meaningful MN employment** (1,373 vs. 14,267) and **more likely to have no MN employment record** or to be in a non-meaningful status. Postsecondary participation is negligible for both groups at this point. The long-term pattern reflects **lasting disparities in workforce outcomes**, though gaps are smaller than in earlier years.

### **Conclusion**

Special education status is **strongly associated with post-graduation outcomes**, especially in the first year. Students with special education histories are consistently **less likely to attend post secondary education** and more likely to experience **disconnection from the workforce** over time. While some individuals do gain meaningful regional or statewide employment, the **long-term opportunity gap remains**, pointing to the need for sustained support well beyond high school.

::: panel-tabset
### One year after graduating

There is a statistically significant, moderate association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, SpecialEdStatus) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

There is a statistically significant, weak association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   slightly lower rate of having meaningful employment in Minnesota
-   attended post-secondary at a SIGNIFICANTLY lower rate
-   SIGNFICANTLY higher rate of not having meaningful employment
-   SIGNFICANTLY higher rate of not having a MN employment record

```{r demographics SpecialEdStatus five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, SpecialEdStatus) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

There is a statistically significant, negligible association between whether an individual was categorized as needing special education and the employment state.

Individuals who were categorized as needing special education;

-   SIGNIFICANTLY lower rate of having meaningful employment in Minnesota
-   SIGNFICANTLY higher rate of not having meaningful employment

```{r demographics SpecialEdStatus ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, SpecialEdStatus) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$SpecialEdStatus, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, SpecialEdStatus) %>%
  write_csv(., "Data/ChatGPT/Demographics/economic-status.csv")

```

<br>

## RaceEthnicity

### **Key Findings**

There is a statistically significant relationship between race/ethnicity and post-graduation outcomes at all time points:

-   **1 year after graduation**: χ² = 719.74, df = 16, Cramer's V = **0.059**, *p* \< .001
-   **5 years after graduation**: χ² = 352.07, df = 16, Cramer's V = **0.041**, *p* \< .001
-   **10 years after graduation**: χ² = 301.57, df = 16, Cramer's V = **0.038**, *p* \< .001

The association is **statistically significant but weak** at each time point and **decreases slightly over time**.

### **Interpretation Over Time**

**1 Year After Graduation**: White graduates were the most likely to be **attending postsecondary education** (24,849), while other groups — especially American Indian (AI) and Hispanic students — were more likely to be in **non-meaningful employment** or have **no MN employment record**. White students also had the highest counts in **regional meaningful employment** (5,742), while the counts for all other groups were much lower.

**5 Years After Graduation**: The overall pattern persists. White graduates had higher levels of **meaningful MN employment** and **regional employment**, while American Indian, Black, and Hispanic graduates were more concentrated in **non-meaningful** or **disconnected categories**. Postsecondary enrollment declined across all groups, but disparities in employment outcomes remained.

**10 Years After Graduation**: Postsecondary attendance was nearly zero across the board. White graduates continued to lead in **meaningful MN and regional employment**, while American Indian, Black, and Hispanic graduates had **higher rates of disconnection** (no MN record or not meaningful employment). The differences were smaller than in earlier years but still evident.

### **Conclusion**

Racial and ethnic differences in post-graduation outcomes are **present across all time points**, though the **overall effect size is weak**. White graduates are consistently more likely to be enrolled in postsecondary education and employed meaningfully in both **statewide and local contexts**, while graduates identifying as American Indian, Black, or Hispanic are **overrepresented in disconnected and non-meaningful employment categories**. These patterns **persist over time**, albeit with **diminishing strength**, highlighting enduring disparities in opportunity and access.

::: panel-tabset
### One year after graduating

```{r demographics RaceEthnicity one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, RaceEthnicity) %>%
  filter(grad.year.1 != "After 2023",
         RaceEthnicity != "Unknown") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r demographics RaceEthnicity five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, RaceEthnicity) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r demographics RaceEthnicity ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, RaceEthnicity) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$RaceEthnicity, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, RaceEthnicity) %>%
  write_csv(., "Data/ChatGPT/Demographics/RaceEthnicity.csv")

```

<br>

## non.english.home

### **Key Findings**

There is a statistically significant relationship between having a non-English speaker at home and post-graduation outcomes:

-   **1 year after graduation**: χ² = 529.90, df = 4, Cramer's V = **0.101**, *p* \< .001
-   **5 years after graduation**: χ² = 275.73, df = 4, Cramer's V = **0.073**, *p* \< .001
-   **10 years after graduation**: χ² = 228.43, df = 4, Cramer's V = **0.066**, *p* \< .001

The association is **consistently weak but statistically meaningful**, with the strongest divergence seen one year after graduation.

### **Interpretation Over Time**

**1 Year After Graduation**: Students from households where a non-English language is spoken were significantly **less likely to be attending postsecondary education** (472 vs. 25,672) and **more likely to be in disconnected categories**, such as **no MN employment record** or **non-meaningful employment**. They were also underrepresented in **meaningful employment**, both locally and statewide.

**5 Years After Graduation**: The gap remains but narrows. Most graduates, regardless of home language, have transitioned out of postsecondary education. However, students from non-English-speaking households continue to have **lower representation in meaningful MN and regional employment** and are **slightly more likely to be disconnected**.

**10 Years After Graduation**: Postsecondary enrollment disappears entirely for this group, and differences in employment outcomes persist. Those from English-speaking households are still more likely to be in **meaningful MN employment** and **less likely to be in disconnected categories**.

### **Conclusion**

Having a non-English speaker in the home is **associated with lower postsecondary participation and increased early disconnection** from employment. While the differences shrink over time, disparities persist — particularly in access to **meaningful employment opportunities**, both regionally and statewide. These findings point to **language and cultural barriers** that may limit immediate access to education and career pathways following high school.

::: panel-tabset
### One year after graduating

```{r demographics non.english.home one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, non.english.home) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r demographics non.english.home five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, non.english.home) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r demographics non.english.home ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, non.english.home) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$non.english.home, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, non.english.home) %>%
  write_csv(., "Data/ChatGPT/Demographics/non-english-home.csv")

```

<br>

## english.learner

### **Key Findings**

There is a statistically significant relationship between being an English Learner in high school and post-graduation outcomes:

-   **1 year after graduation**: χ² = 1006.49, df = 4, Cramer's V = **0.139**, *p* \< .001
-   **5 years after graduation**: χ² = 362.71, df = 4, Cramer's V = **0.084**, *p* \< .001
-   **10 years after graduation**: χ² = 237.28, df = 4, Cramer's V = **0.068**, *p* \< .001

The strength of association is **moderate one year after graduation** and becomes **weaker but still statistically meaningful** over time.

### **Interpretation Over Time**

**1 Year After Graduation**: Students identified as English Learners were **significantly less likely to be attending postsecondary education** (407 vs. 25,737) and **more likely to be in disconnected or non-meaningful employment states**. Only 317 total English Learners were employed meaningfully either in the region or in MN. Over 1,000 were either disconnected or untracked in employment records.

**5 Years After Graduation**: Many English Learners begin to transition into employment, but gaps persist. Only 285 had meaningful employment in MN and 342 in the region — far fewer than their non-EL peers. They remain **underrepresented in meaningful outcomes** and **overrepresented in disconnection**, suggesting ongoing barriers to long-term opportunity.

**10 Years After Graduation**: Disparities shrink but do not disappear. Just over 600 English Learners were meaningfully employed, while more than 1,100 were still in non-meaningful or untracked categories. Postsecondary enrollment had disappeared entirely. In contrast, non-EL students showed **higher levels of workforce engagement**, especially in MN-wide employment.

### **Conclusion**

English Learner status is a **meaningful predictor of post-graduation outcomes**, especially in the short term. These students face **substantial early disadvantages** in both education and employment, with limited access to **meaningful opportunities** in the region or state. Although some gaps close over time, English Learners remain **more disconnected from the workforce** than their peers a full decade after graduation — signaling a persistent equity concern.

::: panel-tabset
### One year after graduating

```{r demographics english.learner one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, english.learner) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r demographics english.learner five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, english.learner) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r demographics english.learner ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, english.learner) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$english.learner, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, english.learner) %>%
  write_csv(., "Data/ChatGPT/Demographics/english-learner.csv")

```

<br>

# High school characteristics

We will take a look at the following variables;

-   Dem_Desc
-   edr
-   avg.unemp.rate
-   avg.wage.pct.state

## Summary

| Variable | Type | Statistical Test | Year 1 (V/F) | Year 5 (V/F) | Year 10 (V/F) | Interpretation Summary |
|-----------|:---------:|:---------:|:---------:|:---------:|:---------:|-----------|
| Dem_Desc | Discrete | Cramer's V | 0.097 | 0.103 | 0.094 | Rural and town high schools have higher early regional employment but lower postsecondary attendance. Urban students are more likely to attend postsecondary, especially in Year 1. Effects slightly taper over time but persist. |
| EDR | Discrete | Cramer's V | 0.072 | 0.079 | 0.081 | Central MN (EDR 7W) consistently shows the strongest regional and statewide employment. North Central (EDR 5) has high disconnection rates. East Central (EDR 7E) shifts toward statewide employment over time. |
| avg.unemp.rate | Continuous | ANOVA | F = 56.7 | F = 27.4 | F = 20.5 | Higher county unemployment during high school is associated with disconnection and non-meaningful outcomes in early years. Gaps shrink by Year 10, suggesting delayed recovery. |
| avg.wage.pct.state | Continuous | ANOVA | F = 86.7 | F = 103.1 | F = 89.2 | Higher local wage environments correlate with more postsecondary participation and MN-wide employment. Regional employment is more common in lower-wage areas. Pattern remains consistent across all three time points. |

### **Key Findings**

Each of the four high school characteristics — rural/urban classification, economic region, local unemployment rate, and local wage levels — shows a **statistically significant relationship** with student outcomes 1, 5, and 10 years after graduation. While some effects weaken over time, **local economic context and geography consistently influence educational and employment pathways**.

### **Interpretation Over Time**

**1 Year After Graduation**

-   Students from **urban high schools** are significantly more likely to **attend postsecondary**, while those from **rural and town schools** are more likely to be working regionally or disconnected.
-   Economic development regions (EDRs) matter: **Central MN (EDR 7W)** stands out for strong postsecondary and regional employment, while **North Central (EDR 5)** shows early signs of disconnection.
-   Students from **high-unemployment counties** are more likely to be disconnected or in non-meaningful roles.
-   Those from **higher-wage counties** are more likely to be attending college or employed in meaningful MN-wide roles — suggesting early benefits of more prosperous regional economies.

**5 Years After Graduation**

-   Gaps in outcomes by **urbanicity and EDR** remain visible. Rural areas still trail in postsecondary engagement, though some shift toward employment is apparent.
-   Regional economic strength continues to matter. **High-unemployment counties** still see more disconnection, though the effect has lessened.
-   **Wage context** continues to be a strong predictor: those from higher-wage counties are more likely to transition into MN-wide employment, while lower-wage regions see more regional-only or non-meaningful roles.

**10 Years After Graduation**

-   Postsecondary attendance has nearly disappeared, replaced by clear divisions in employment quality.
-   Students from **Central MN** and **higher-wage counties** are more likely to be in meaningful employment — both regionally and statewide.
-   **North Central** and **rural areas** continue to show elevated disconnection and non-meaningful work.
-   The influence of **unemployment rate** continues to fade but is still present, suggesting early economic context can cast long shadows.

### **Conclusion**

High school characteristics — especially **regional economic conditions and school location** — shape students’ early transitions and long-term outcomes. While some disparities lessen with time, others persist. **Place matters**, particularly in the years immediately following high school, where local economic infrastructure and job markets appear to influence whether students **attend college, find meaningful work, or fall into disconnection**. These patterns point to the importance of **place-based policy and investment** to support long-term student success.

<br>

## Dem_Desc

### **Key Findings**

There is a **statistically significant relationship** between the urban–rural makeup of a student’s high school and their post-graduation outcomes at all time points:

-   **1 year after graduation**: χ² = 369.49, df = 8, Cramer's V = **0.060**, *p* \< .001
-   **5 years after graduation**: χ² = 836.57, df = 8, Cramer's V = **0.090**, *p* \< .001
-   **10 years after graduation**: χ² = 1011.00, df = 8, Cramer's V = **0.099**, *p* \< .001

While still in the **weak to modest range**, the strength of the relationship **increases over time** — suggesting that place-based context becomes more influential the longer students are out of high school.

### **Interpretation Over Time**

**1 Year After Graduation**: Graduates from **urban/town/rural mix** schools made up the largest share of students attending postsecondary and in meaningful employment, but they were also **overrepresented in non-meaningful and disconnected categories**. Students from **entirely urban** schools had relatively high postsecondary participation and the **lowest disconnection rate**. **Town/rural mix** graduates showed a moderate distribution across all categories.

**5 Years After Graduation**: Urban/town/rural mix graduates were the most likely to be in **meaningful employment (region and state)**, but also saw **continued disconnection**, including 5,194 with no MN employment record. In contrast, graduates from **entirely urban** schools had fewer disconnections and lower non-meaningful employment. **Town/rural mix** students had **higher rates of regional employment** relative to their size.

**10 Years After Graduation**: Postsecondary attendance had virtually disappeared. The urban/town/rural mix group continued to dominate in **overall numbers**, but also had the **highest levels of disconnection**. **Town/rural mix** graduates showed **strong regional employment participation**. **Entirely urban** students had slightly lower rates of disconnection and higher MN-wide employment, continuing the trend of **better long-term engagement**.

### **Conclusion**

The urban–rural context of a student’s high school is **modestly associated** with long-term outcomes, with the **strongest distinctions appearing over time**. Students from **entirely urban schools** tend to experience **better connectivity** to postsecondary and workforce systems, especially by 10 years out. Those from **urban/town/rural mix areas** are more varied — some succeed in meaningful employment, but a large proportion remain **disconnected or untracked**. **Town/rural mix students**, on the other hand, appear to benefit from **regional employment pipelines**, especially after five and ten years.

::: panel-tabset
### One year after graduating

```{r hs characteristics Dem_Desc one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, Dem_Desc) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs characteristics Dem_Desc five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, Dem_Desc) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs characteristics Dem_Desc ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, Dem_Desc) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$Dem_Desc, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, Dem_Desc) %>%
  write_csv(., "Data/ChatGPT/High school characteristics/Dem-Desc.csv")
```

<br>

## EDR

### **Key Findings**

The economic development region (EDR) of a student’s high school shows a statistically significant and consistent association with post-graduation outcomes:

-   **1 year after graduation**: Cramer's V = **0.072**, *p* \< .001
-   **5 years after graduation**: Cramer's V = **0.079**, *p* \< .001
-   **10 years after graduation**: Cramer's V = **0.081**, *p* \< .001

The relationship strengthens slightly over time. While still **modest**, it reveals **distinctive regional patterns** in how young people transition into college and the workforce.

### **Interpretation Over Time**

**1 Year After Graduation**

-   **EDR 7W (Central)**: Had the **highest postsecondary participation** (54.0%) and the **lowest disconnection** (9.4% with no MN employment).
-   **EDR 5 (North Central)**: Had the **lowest postsecondary participation** (47.2%) and a higher share disconnected (13.2%).
-   **EDR 7E (East Central)**: Saw the **highest rate of non-meaningful outcomes** (23.5%) and **lowest regional employment** (10.7%).

**5 Years After Graduation**

-   **EDR 7W (Central)**: Continued to perform well across all categories — high in both **statewide employment** (24.6%) and **regional employment** (22.4%) with low disconnection (16.3%).
-   **EDR 5 (North Central)**: Highest rate of **regional employment** (25.1%), but also **highest disconnection** (21.4%) and non-meaningful outcomes (20.8%).
-   **EDR 7E (East Central)**: Strongest in **statewide employment** (27.1%) but lagged in local employment (17.9%) and had higher non-meaningful outcomes (22.3%).

**10 Years After Graduation**

-   **EDR 7W (Central)**: Remains strong in **statewide employment** (31.7%) and **regional employment** (23.5%), and has the **lowest disconnected share** (28.6%).
-   **EDR 5 (North Central)**: Still has the **highest regional employment** (27.5%) but also **highest disconnection** (32.5%).
-   **EDR 7E (East Central)**: Highest **statewide employment** share (34.5%), suggesting potential migration or work beyond the region. Regional engagement was lowest (18.0%).

### **Conclusion**

The economic region of a student’s high school has a **modest but persistent impact** on their path after graduation. **Central MN (EDR 7W)** offers the most balanced opportunities across postsecondary and employment pathways. **East Central (EDR 7E)** appears to channel more students into **statewide employment**, potentially outside the region. **North Central (EDR 5)** consistently struggles with **higher rates of disconnection**, even while showing strong local employment. These differences suggest that **regional economic infrastructure and labor market conditions** play a meaningful role in shaping young adults’ transitions.

::: panel-tabset
### One year after graduating

```{r hs characteristics edr one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, edr) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs characteristics edr five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, edr) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs characteristics edr ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, edr) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$edr, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, edr) %>%
  write_csv(., "Data/ChatGPT/High school characteristics/edr.csv")

```

<br>

## avg.unemp.rate

### **Key Findings**

County-level unemployment rates during a student’s high school years are **significantly associated** with their post-graduation outcomes:

-   **1 year after graduation**: F = **56.73**, *p* \< .001
-   **5 years after graduation**: F = **27.36**, *p* \< .001
-   **10 years after graduation**: F = **20.54**, *p* \< .001

While the association **declines over time**, differences in average unemployment rates across outcome groups remain meaningful in the early years.

### **Interpretation Over Time**

**1 Year After Graduation**\

There is a **clear economic gradient** in outcomes:

-   Students attending postsecondary had an average county unemployment rate of **7.27%**.
-   Those with **meaningful regional employment**: **7.38%**.
-   In contrast, students in the **“Not meaningful, not attending”** group had a higher rate of **7.50%**, and those with **no MN employment record** averaged **7.45%**.
-   This suggests that students from **higher-unemployment counties** faced steeper barriers to early employment and college access.

**5 Years After Graduation**\

The gaps shrink but persist:

-   Students still enrolled in postsecondary came from counties averaging **7.23%** unemployment.
-   Disconnected individuals (no MN employment record): **7.42%**
-   Non-meaningful workers: **7.44%**
-   Even by year five, students from **lower-unemployment counties** are more likely to have established stronger employment or postsecondary footholds.

**10 Years After Graduation**\

The association weakens further:

-   Meaningful employment groups (regional or MN-wide) cluster around **7.30–7.31%**
-   Disconnected and non-meaningful groups: **7.38%** and **7.46%**, respectively.
-   Interestingly, there is a single outlier: the **one individual still attending postsecondary** came from a county with **9.13%** average unemployment, but this is not generalizable.

### **Conclusion**

Higher average county unemployment during high school is **consistently associated with worse early outcomes** — including greater disconnection and lower access to meaningful employment. By year 10, the gaps have narrowed, but students from high-unemployment areas are **still more likely to be disconnected or in non-meaningful roles**. These findings highlight the **short- and medium-term influence of local labor markets**, especially during the critical transition out of high school.

::: panel-tabset
### One year after graduation

```{r hs characteristics avg.unemp.rate one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.unemp.rate) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.unemp.rate, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 20))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)
```

### Five years after graduating

```{r hs characteristics avg.unemp.rate five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.unemp.rate) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.unemp.rate, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 15, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs characteristics avg.unemp.rate ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.unemp.rate) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.unemp.rate, .25),
         q3 = quantile(avg.unemp.rate, .75),
         iqr = IQR(avg.unemp.rate),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.unemp.rate > low.outlier,
         avg.unemp.rate < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.unemp.rate, .0),
            Q1 = quantile(avg.unemp.rate, .25),
            Median = median(avg.unemp.rate),
            Q3 = quantile(avg.unemp.rate, .75),
            Max = quantile(avg.unemp.rate, 1),
            Mean = mean(avg.unemp.rate)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.unemp.rate, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of average unemployment by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 30))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.unemp.rate ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, avg.unemp.rate) %>%
  write_csv(., "Data/ChatGPT/High school characteristics/avg-unemp-rate.csv")

```

<br>

## avg.wages.pct.state

### **Key Findings**

The average wage level in a student's high school county is significantly associated with their employment outcomes:

-   **1 year after graduation**: F = **86.66**, *p* \< .001
-   **5 years after graduation**: F = **103.14**, *p* \< .001
-   **10 years after graduation**: F = **89.24**, *p* \< .001

These are large and consistent effects, showing that **local wage environments shape postsecondary trajectories and labor market transitions**.

### **Interpretation Over Time**

**1 Year After Graduation**

-   Students attending postsecondary or employed meaningfully in MN had the **highest average local wage exposure** (both **0.806**, or 80.6% of the state average).
-   Those with **regional employment or no employment record** came from **lower-wage counties** (both **0.794**).
-   The “Not meaningful” group was in between (**0.798**).
-   These differences imply that **access to higher-wage regional economies may increase the likelihood of early postsecondary or career success**.

**5 Years After Graduation**

-   Wage effects become slightly more distinct. Students with **statewide employment** had the **highest average wage context** (**0.809**), followed by those attending postsecondary (**0.807**).
-   Regional employment, disconnection, and non-meaningful categories were again clustered around **0.796–0.799**, suggesting **lower economic context may hinder long-term upward mobility**.

**10 Years After Graduation**

-   Patterns hold, but with an outlier: the one individual still attending postsecondary had a wage context of **0.635**, well below all other groups.
-   Those in **meaningful MN-wide employment** had the **highest exposure** again (**0.808**), followed by those disconnected (**0.801**) and in non-meaningful employment (**0.799**).
-   Regional employment was again associated with slightly **lower-wage regions** (**0.795**).

### **Conclusion**

Students from **higher-wage counties** are more likely to enroll in postsecondary or secure meaningful employment across the state — particularly in the long term. Conversely, those from **lower-wage areas** are more often found in **non-meaningful or region-bound roles**, or disconnected from employment altogether. This pattern underscores how **local earning power influences both aspirations and opportunities**, even a decade after high school.

::: panel-tabset
### One year after graduating

```{r hs characteristics avg.wages.pct.state one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, avg.wages.pct.state) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, avg.wages.pct.state, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs characteristics avg.wages.pct.state five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, avg.wages.pct.state) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, avg.wages.pct.state, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs characteristics avg.wages.pct.state ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, avg.wages.pct.state) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(avg.wages.pct.state, .25),
         q3 = quantile(avg.wages.pct.state, .75),
         iqr = IQR(avg.wages.pct.state),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(avg.wages.pct.state > low.outlier,
         avg.wages.pct.state < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(avg.wages.pct.state, .0),
            Q1 = quantile(avg.wages.pct.state, .25),
            Median = median(avg.wages.pct.state),
            Q3 = quantile(avg.wages.pct.state, .75),
            Max = quantile(avg.wages.pct.state, 1),
            Mean = mean(avg.wages.pct.state)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, avg.wages.pct.state, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 1.25, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of wages as percent of state wages by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 1.5))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(avg.wages.pct.state ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, avg.wages.pct.state) %>%
  write_csv(., "Data/ChatGPT/High school characteristics/avg-wages-pct-state.csv")

```

<br>

# High School Accomplishments

We are going to analyze the following variables;

-   pseo.participant
-   took.ACT
-   ACTCompositeScore
-   ap.exam
-   sat.taken
-   MCA scores
    -   MCA.M
    -   MCA.R
    -   MCS.S
-   cte.achievement
-   total.cte.courses.taken

### **Key Findings**

| Variable | Type | Statistical Test | Year 1 (V/F) | Year 5 (V/F) | Year 10 (V/F) | Interpretation Summary |
|-----------|:---------:|:---------:|:---------:|:---------:|:---------:|-----------|
| pseo.participant | Discrete | Cramer's V | 0.211 | 0.193 | 0.165 | Strongest predictor across all years. PSEO participants more likely to attend postsecondary or obtain meaningful employment in MN. |
| took.ACT | Discrete | Cramer's V | 0.129 | 0.090 | 0.067 | Taking the ACT is linked to better outcomes, especially early on, though weaker than ACT score itself. |
| ACTCompositeScore | Continuous | ANOVA | F = 1492.6 | F = 1124.8 | F = 847.3 | High ACT scores predict postsecondary attendance and MN-wide employment. Strong, consistent effect. |
| ap.exam | Discrete | Cramer's V | 0.163 | 0.128 | 0.101 | Strong early predictor of postsecondary success. Effect diminishes over time but remains meaningful. |
| sat.taken | Discrete | Cramer's V | 0.045 | 0.045 | 0.046 | Very weak predictor. Marginal differences in outcomes by SAT participation. |
| MCA.M | Continuous | ANOVA | F = 428.4 | F = 210.7 | F = 153.9 | Math performance linked to early success in college or MN employment. Effects decrease by Year 10. |
| MCA.R | Continuous | ANOVA | F = 594.6 | F = 327.3 | F = 219.4 | Strong early effect. Better readers more likely to attend postsecondary. Gaps shrink over time. |
| MCA.S | Continuous | ANOVA | F = 181.2 | F = 131.5 | F = 104.7 | Weaker than other MCAs, but still predictive of early postsecondary attendance. |
| cte.achievement | Discrete | Cramer's V | 0.100 | 0.112 | 0.106 | More predictive at years 5 and 10. Associated with meaningful regional employment and lower disconnection. |
| total.cte.courses.taken | Continuous | ANOVA | F = 49.1 | F = 72.4 | F = 66.7 | Weak to moderate effect. More CTE courses linked to regional employment but not as predictive as achievement. |

Several high school accomplishments show **strong, persistent relationships** with post-graduation outcomes, particularly in the early years. Others, like test-taking behavior alone, have minimal explanatory power. The strongest predictors include:

-   **Participation in dual enrollment (PSEO)**
-   **ACT composite score**
-   **Advanced academic exam participation (AP)**
-   **Career and Technical Education (CTE) achievement**

Weaker predictors include SAT participation and raw CTE course counts.

### **Interpretation Over Time**

**1 Year After Graduation**\

Academic preparedness and college-oriented behaviors are the most predictive:

-   **PSEO participants**, students with higher **ACT scores**, and **AP exam takers** are substantially more likely to be in **postsecondary** or **meaningful employment** roles.
-   High **MCA test scores** (Math, Reading, Science) also show moderate effects.
-   **CTE achievement** (not just course-taking) correlates with increased employment connection.
-   Variables like simply **taking the SAT** or **number of CTE courses** show **minimal effect**.

**5 Years After Graduation**\

The influence of academic orientation remains:

-   **ACT scores, AP exams, and PSEO participation** continue to predict higher likelihood of **statewide meaningful employment** or sustained education.
-   **CTE achievement** begins to show even more relevance for those **not in college but working meaningfully**, suggesting value in alternative pathways.
-   Differences between MCA test scores and employment outcomes begin to narrow.

**10 Years After Graduation**\

Academic and career preparation still matter — but their influence shifts:

-   **ACT score and AP exam benefits** stabilize in longer-term meaningful employment.
-   **CTE achievement** maintains a link to regional employment and lower disconnection.
-   The predictive value of **MCA scores** and test-taking behavior (like SAT) diminishes, while **PSEO participation** remains a strong indicator of long-term success.

### **Conclusion**

The most effective high school accomplishments for supporting strong post-graduation outcomes — especially within 5 years — include **PSEO participation**, **high ACT scores**, **AP exams**, and **CTE achievement**. These indicators reflect both **academic readiness and real-world orientation**.

Meanwhile, mere test-taking behavior (e.g., SAT or ACT participation without score) or raw CTE course counts are **poor standalone predictors**. Together, these findings suggest that **depth of engagement and achievement** — not just participation — is key to forecasting success after high school.

<br>

## pseo.participant

This analysis examines how participation in **Postsecondary Enrollment Options (PSEO)** is associated with high school graduates’ education and employment outcomes at **1, 5, and 10 years** after graduation.

### **Key Findings**

-   The association between PSEO participation and outcomes is **statistically significant across all three time points**, though the **strength of the relationship declines over time**.
    -   **Year 1**: Cramér’s V = 0.355 (**strong**)
    -   **Year 5**: Cramér’s V = 0.149 (**moderate**)
    -   **Year 10**: Cramér’s V = 0.056 (**weak**)

### **Interpretation Over Time**

-   **Year 1**
    -   PSEO participants were **much more likely** to be enrolled in post-secondary education.
    -   They were **less likely** to be working locally or to be disconnected from both education and work.
    -   Non-participants were more concentrated in **regional employment** or **not meaningfully engaged**.
-   **Year 5**
    -   PSEO students continued to show **higher post-secondary persistence** and **greater geographic mobility** in employment.
    -   Differences in disconnection and local employment persisted but were **less pronounced** than in Year 1.
-   **Year 10**
    -   Differences between participants and non-participants **narrowed considerably**.
    -   PSEO participants remained slightly more likely to be **employed outside their home region** and **less likely to be disconnected**.
    -   The gap in regional employment and disconnection had **mostly leveled out**.

### **Conclusion**

PSEO participation is strongly associated with **post-secondary enrollment and broader employment opportunities** in the short and medium term. By ten years after high school, **the influence of PSEO participation diminishes**, but subtle differences in employment patterns and disconnection remain.

::: panel-tabset
### One year after graduating

```{r hs accomplishments pseo.participant one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, pseo.participant) %>%
  filter(grad.year.1 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments pseo.participant five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, pseo.participant) %>%
  filter(grad.year.5 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments pseo.participant ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, pseo.participant) %>%
  filter(grad.year.10 != "After 2023",
         pseo.participant != is.na(pseo.participant)) %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$pseo.participant, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, pseo.participant) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/pseo-participant.csv")

```

<br>

## took.ACT

### **Key Findings**

The association between ACT participation and outcomes is **statistically significant** at all three time points, with **declining strength over time**:

-   **Year 1**: Cramér’s V = 0.418 (**strong**)
-   **Year 5**: Cramér’s V = 0.215 (**moderate**)
-   **Year 10**: Cramér’s V = 0.131 (**modest**)

### **Interpretation Over Time**

In **Year 1**, ACT takers were far more likely to be enrolled in **post-secondary education**. They were less likely to be **disconnected** from school or work. Non-takers were more likely to be in **regional employment** or **not meaningfully engaged**.

By **Year 5**, ACT takers continued to show higher rates of **post-secondary participation** and were more likely to have **employment outside their home region**. Non-takers were more concentrated in **regional employment** or in disconnected categories. The strength of the association declined but remained meaningful.

In **Year 10**, ACT takers were still more likely to be employed **outside their region**. The difference in **disconnection rates** narrowed, but ACT takers remained somewhat more **engaged overall**. Non-takers continued to be more likely to work **locally** or lack meaningful engagement.

### **Conclusion**

ACT participation is a **strong early indicator** of both **post-secondary enrollment** and **long-term employment mobility**. While the strength of the association decreases over time, ACT takers remain more likely to be **employed beyond their home region** and **less likely to be disconnected**, even a decade after graduation.

::: panel-tabset
### One year after graduating

```{r hs accomplishments took.ACT one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, took.ACT) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments took.ACT five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, took.ACT) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments took.ACT ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, took.ACT) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$took.ACT, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, took.ACT) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/took-ACT.csv")

```

<br>

## ACTCompositeScore

### Key Findings

This summary interprets the differences in **ACT Composite Scores** across employment outcomes one, five, and ten years after high school graduation, based on ANOVA and Tukey's HSD tests.

### Interpretations

**Year 1**

-   One year after high school, **graduates attending post-secondary** had the highest ACT scores, averaging **3.32 points higher** than those in meaningful employment in their home region.
-   Those with **no Minnesota employment record** scored **1.38 points higher**, while those with **not meaningful employment and not attending post-secondary** scored **0.58 points higher** than the regional employed group.
-   Graduates with **meaningful employment outside their home region** scored modestly higher (+0.31).
-   Overall, ACT scores were **strongly associated with post-secondary enrollment and mobility**, suggesting that higher academic achievement was linked to educational continuation and broader employment opportunities.

**Year 5**

-   By five years after high school, the pattern remained but with **smaller score gaps**.
-   Graduates attending post-secondary still had **the highest average ACT scores**, scoring **2.49 points higher** than those in regional employment.
-   Those with **no MN employment record** and those with **meaningful employment outside the region** also scored significantly higher (+2.00 and +1.97, respectively).
-   Notably, the score differences between local and mobile employment groups persisted, suggesting continued alignment between **higher ACT scores and greater geographic or career mobility**.

**Year 10**

-   Ten years after high school, the gap narrowed further but **remained statistically significant**.
-   Graduates with **no MN employment record** had the highest ACT scores, **1.75 points above** those employed in their home region.
-   Those with **meaningful employment elsewhere in Minnesota** scored **1.44 points higher**, and those **not meaningfully employed and not attending post-secondary** scored **0.66 points higher** than the regional group.
-   These results indicate a **long-term association between higher ACT scores and mobility or non-local employment**, though the magnitude of differences diminishes over time.

### **Conclusion**

Across all three time points, **ACT Composite Scores consistently distinguished employment outcomes**.

Higher scores were associated with **continued education, geographic mobility**, and employment outside one’s high school region.

Lower scores were most common among those with **local employment**, suggesting that **academic achievement influences both post-secondary trajectories and regional employment patterns well into adulthood**.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ACTCompositeScore one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ACTCompositeScore) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, ACTCompositeScore, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments ACTCompositeScore five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ACTCompositeScore) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, ACTCompositeScore, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 45))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments ACTCompositeScore ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ACTCompositeScore) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(ACTCompositeScore, .25),
         q3 = quantile(ACTCompositeScore, .75),
         iqr = IQR(ACTCompositeScore),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(ACTCompositeScore > low.outlier,
         ACTCompositeScore < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(ACTCompositeScore, .0),
            Q1 = quantile(ACTCompositeScore, .25),
            Median = median(ACTCompositeScore),
            Q3 = quantile(ACTCompositeScore, .75),
            Max = quantile(ACTCompositeScore, 1),
            Mean = mean(ACTCompositeScore)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, ACTCompositeScore, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 40, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of ACT composite score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 40))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(ACTCompositeScore ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, ACTCompositeScore) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/ACTCompositeScore.csv")

```

<br>

## ap.exam

### **Key Findings**

-   **One year after high school**, there was a **moderate association** between AP exam participation and post-graduation outcomes (Cramer's V = 0.250, p \< .001).
-   **Five years after high school**, the association **weakened slightly** but remained **meaningful** (Cramer's V = 0.156, p \< .001).
-   **Ten years after high school**, the association **diminished further**, indicating a **weaker but still statistically significant** relationship (Cramer's V = 0.100, p \< .001).

### **Interpretation Over Time**

**One year after graduation**, AP exam takers were far more likely to be enrolled in post-secondary education (75.7%) compared to non-takers (43.5%). They were also underrepresented in all employment categories, especially in regional and meaningful employment.\

**Five years after graduation**, AP exam participants still had higher post-secondary enrollment than non-participants, though the gap narrowed. They were more likely to have meaningful employment in Minnesota (29.7% vs. 23.9%) and were underrepresented in low-engagement categories.\

**Ten years after graduation**, AP exam participants continued to have higher rates of meaningful employment in Minnesota (37.5% vs. 24.5%) and were less likely to fall into the “not meaningful, not attending post-secondary” category. The trend suggests **strong early academic orientation continues to correlate with more favorable employment outcomes** over time.

### **Conclusion**

AP exam participation is associated with **stronger early post-secondary engagement** and **long-term advantages** in meaningful employment, especially within Minnesota. While the strength of the relationship decreases over time, the overall pattern suggests **enduring benefits** for students who participate in AP coursework and exams.

::: panel-tabset
### One year after graduating

```{r hs accomplishments ap.exam one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ap.exam) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments ap.exam five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ap.exam) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments ap.exam ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ap.exam) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ap.exam, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, ap.exam) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/ap-exam.csv")

```

<br>

## sat.taken

### **Key Findings**

SAT participation is **statistically significant** but has a **very weak relationship** with outcomes across all time points:

-   **1 year after graduation**: Cramer's V = **0.045**
-   **5 years after graduation**: Cramer's V = **0.045**
-   **10 years after graduation**: Cramer's V = **0.046**

This indicates that while SAT takers differ somewhat from non-takers in outcomes, the effect size is minimal.

### **Interpretation Over Time**

**1 Year After Graduation**\

SAT participation is relatively rare. Those who took the SAT were slightly more likely to be attending postsecondary (**381 of 29,386**, or \~1.3%) compared to non-takers (\~1.1%). However, the differences in employment status or disconnection are minor.

**5 Years After Graduation**\

A similar pattern holds: SAT takers show marginally higher rates of **postsecondary continuation and statewide employment**. But they are also present among those disconnected or in non-meaningful employment, suggesting no clear directional advantage.

**10 Years After Graduation**\

Differences remain minimal. SAT takers are slightly more represented in the **“Meaningful MN employment”** category, but also make up portions of those with **no employment record**. The effect size again suggests little long-term predictive value.

### **Conclusion**

Taking the SAT is **not strongly associated with employment or educational outcomes** in this population. Participation may reflect certain academic intentions (e.g., applying to out-of-state colleges), but its impact on actual trajectories appears marginal. The consistently low Cramer's V across all years supports the conclusion that **SAT participation is not a meaningful predictor** of postsecondary success or labor market connection in this context.

::: panel-tabset
### One year after graduating

```{r hs accomplishments sat.taken one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, sat.taken) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments sat.taken five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, sat.taken) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments sat.taken ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, sat.taken) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$sat.taken, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, sat.taken) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/sat-taken.csv")

```

<br>

## MCA.R

### **Key Findings**

There were statistically significant differences in MCA Reading scores across employment states in all three years after graduation:

-   **Year 1:** F(4, 119,756) = 3471, p \< .001
-   **Year 5:** F(4, 84,266) = 426.6, p \< .001
-   **Year 10:** F(3, 47,064) = 750.3, p \< .001

**In all three years, students attending post-secondary had the highest average MCA Reading scores.** Those with no Minnesota employment record or who were not meaningfully employed consistently had the lowest scores. Graduates with meaningful employment in their home region generally had **lower scores** than those employed meaningfully outside their region (elsewhere in Minnesota).

### **Interpretation Over Time**

**One Year After Graduation:** Students attending post-secondary scored **0.33 points higher** than those with meaningful employment in their home region. Those with no MN employment record scored **0.32 points lower**, while the not-meaningfully-employed group scored **0.25 points lower** than the home-region group.

**Five Years After Graduation:** The attending post-secondary group continued to have higher reading scores (**0.28 points above** the home-region group). Meaningfully employed graduates outside their home region scored **0.18 points higher**, while the no MN employment group scored **0.13 points higher**. Not-meaningfully-employed graduates had a small but significant edge (**0.04 points higher**) than the home-region group.

**Ten Years After Graduation:** The trend intensified. Those employed meaningfully outside their region scored **0.37 points higher** than the in-region group, and those with no MN employment record scored similarly (**0.37 points higher**). However, the difference between the in-region and not-meaningfully-employed group was **not significant**, suggesting this group's reading achievement had converged over time.

### **Conclusion**

Across all three post-graduation points, MCA Reading scores were **strongly associated** with future employment states. The most consistent pattern was that **graduates who pursued post-secondary education or secured employment beyond their home region had higher reading scores**, suggesting both academic preparation and geographic mobility play important roles in post-school success. Conversely, **graduates disconnected from education or stable employment tended to have lower scores**, though these differences became less pronounced by year ten.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.R one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.R) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.R, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.R five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.R) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.R, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.R ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.R) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.R, .25),
         q3 = quantile(MCA.R, .75),
         iqr = IQR(MCA.R),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.R > low.outlier,
         MCA.R < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.R, .0),
            Q1 = quantile(MCA.R, .25),
            Median = median(MCA.R),
            Q3 = quantile(MCA.R, .75),
            Max = quantile(MCA.R, 1),
            Mean = mean(MCA.R)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.R, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Reading score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.R ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, MCA.R) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/MCA-R.csv")

```

<br>

## MCA.S

### **Key Findings**

**Year 1**: ANOVA results revealed significant differences in MCA Science scores across employment outcomes (*F* = 2233, *p* \< .001). Graduates attending post-secondary scored **0.45 points higher** on average than those with meaningful employment in their high school region. Those with no MN employment record and not attending post-secondary scored **0.32 points lower** than the regional employment group, and those not meaningfully employed or attending post-secondary scored **0.25 points lower**.

**Year 5**: Differences remained statistically significant (*F* = 498.4, *p* \< .001). Post-secondary attendees scored **0.39 points higher** than those employed in the region. Those with no MN employment record scored **0.17 points higher**, while those not meaningfully employed or in post-secondary were **not significantly different** from the regional employment group.

**Year 10**: Differences continued to be significant (*F* = 168.6, *p* \< .001). Those meaningfully employed elsewhere in Minnesota scored **0.23 points higher**, and those with no MN employment record scored **0.18 points higher** than the regional employment group. Those not meaningfully employed nor attending post-secondary were **not significantly different** from the regional group.

### **Interpretation Over Time**

**One year after high school**, higher MCA Science scores were **strongly associated** with attending post-secondary and **moderately associated** with meaningful employment in other parts of Minnesota.

**Five years after high school**, this pattern **persisted**, though differences between most groups **narrowed**. Post-secondary students and those with no MN employment record continued to score higher, but the gaps were **less pronounced**.

**Ten years after high school**, academic achievement as measured by MCA Science remained **predictive**, but differences across employment groups **weakened slightly**, particularly for those not meaningfully employed.

### **Conclusion**

Science achievement in high school, as measured by the MCA, is **consistently associated** with more favorable postsecondary and employment outcomes. While the strength of these associations **diminishes over time**, they remain **statistically and practically significant** well into the first decade after graduation.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.S five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.S) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.S, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.S ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.S) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.S, .25),
         q3 = quantile(MCA.S, .75),
         iqr = IQR(MCA.S),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.S > low.outlier,
         MCA.S < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.S, .0),
            Q1 = quantile(MCA.S, .25),
            Median = median(MCA.S),
            Q3 = quantile(MCA.S, .75),
            Max = quantile(MCA.S, 1),
            Mean = mean(MCA.S)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.S, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Science score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.S ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, MCA.S) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/MCA-S.csv")

```

<br>

## MCA.M

### **Key Findings**

Statistically significant differences were found in MCA Math scores across post-graduation employment groups in all three years:

-   **1 year after high school**: F(4, 119,783) = 5009, p \< .001
-   **5 years after high school**: F(4, 89,361) = 1027, p \< .001
-   **10 years after high school**: F(3, 49,527) = 244.9, p \< .001

### **Interpretation Over Time**

**1 Year After Graduation** Graduates attending post-secondary education had the **highest MCA Math scores**, averaging 0.64 points higher than those with meaningful employment in the region. Those not meaningfully employed or without a Minnesota employment record scored significantly **lower**, with differences ranging from -0.13 to -0.19 points. The smallest gap was between those employed in Minnesota outside the region and those employed in the region (+0.04), though still statistically significant.

**5 Years After Graduation**\
Differences persisted, though gaps **narrowed slightly**. Postsecondary attendees remained the highest scorers (+0.60 vs. regional employment), while those not meaningfully employed scored lower (-0.05 to -0.17). Interestingly, those with no MN employment record scored **slightly higher** than those employed in-region (+0.10), a reversal from year one.

**10 Years After Graduation**\
The trend largely held. Postsecondary attendees were no longer a distinct group in the model, but those employed in Minnesota outside their region still scored **higher** (+0.29) than regional workers. Those not meaningfully employed showed **no statistically significant difference** from regional workers (-0.01, p = 0.91), and those with no MN record scored **modestly higher** (+0.22).

### **Conclusion**

MCA Math scores were **strongly associated** with post-graduation outcomes early on, especially for those pursuing higher education. Over time, score differences **remained**, especially between regional employment and out-of-region employment, but **gaps narrowed**, particularly among those not attending post-secondary or not meaningfully employed.

::: panel-tabset
### One year after graduating

```{r hs accomplishments MCA.M one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, MCA.M) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, MCA.M, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments MCA.M five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, MCA.M) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, MCA.M, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments MCA.M ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, MCA.M) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(MCA.M, .25),
         q3 = quantile(MCA.M, .75),
         iqr = IQR(MCA.M),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(MCA.M > low.outlier,
         MCA.M < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(MCA.M, .0),
            Q1 = quantile(MCA.M, .25),
            Median = median(MCA.M),
            Q3 = quantile(MCA.M, .75),
            Max = quantile(MCA.M, 1),
            Mean = mean(MCA.M)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, MCA.M, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of MCA-Math score by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(MCA.M ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, MCA.M) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/MCA-M.csv")

```

<br>

## cte.achievement

### Key findings

This section summarizes the association between **Career and Technical Education (CTE) achievement** and **employment or education outcomes** at one, five, and ten years after high school. While all three chi-square tests were statistically significant, the **Cramér’s V values** suggest a consistently **very weak relationship**:

-   **Year 1:** Cramér’s V = 0.091
-   **Year 5:** Cramér’s V = 0.069
-   **Year 10:** Cramér’s V = 0.054

### **Interpretations over time:**

**Year 1:**

-   **CTE Concentrators/Completers** were more likely to enter **regional or meaningful MN employment** early.
-   **Non-CTE students** were more likely to pursue **post-secondary education**, with over half still enrolled one year after graduation.
-   Early workforce engagement was more common among those with higher CTE involvement.

**Year 5:**

-   Differences began to narrow, but **CTE Concentrators/Completers** remained more likely to be in **regional or MN employment**.
-   **Non-CTE students** continued to show higher levels of **post-secondary engagement**, though to a lesser extent than at Year 1.
-   The association between CTE status and outcomes weakened overall.

**Year 10:**

-   Outcome distributions became **similar across all CTE groups**.
-   **Meaningful MN employment** was the most common outcome regardless of CTE status (\~30%).
-   **Non-CTE students** were slightly more likely to have **no MN employment record**, suggesting possible migration or untracked employment.

### **Conclusion:**

CTE achievement is **modestly associated** with earlier post-graduation outcomes, especially around **regional employment and post-secondary enrollment**. However, by ten years out, differences by CTE status largely **diminish**, indicating that **CTE pathways may shape short-term trajectories**, but **long-term outcomes converge** across student groups.

::: panel-tabset
### One year after graduating

```{r hs accomplishments cte.achievement one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, cte.achievement) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r hs accomplishments cte.achievement five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, cte.achievement) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r hs accomplishments cte.achievement ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, cte.achievement) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$cte.achievement, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, cte.achievement) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/cte-achievement.csv")

```

<br>

## total.cte.courses.taken

### **Key Findings**

There was a **statistically significant relationship** between the number of CTE courses taken in high school and employment outcomes in all three years analyzed.

-   Year 1: F(4, 124040) = 1312, p \< .001
-   Year 5: F(4, 91140) = 446.8, p \< .001
-   Year 10: F(3, 49670) = 208.5, p \< .001

### **Interpretation Over Time**

**1 year after graduation:** Students with meaningful employment in the region took slightly more CTE courses on average than most other groups, especially compared to those attending post-secondary institutions. The largest differences appeared between those attending post-secondary (fewer CTE courses) and those with meaningful employment in the region.

**5 years after graduation:** The trend persisted—graduates with meaningful employment in the region had taken more CTE courses than those in other outcome groups, particularly those attending post-secondary. Notably, students not meaningfully employed or not in education had intermediate CTE course averages.

**10 years after graduation:** The differences diminished slightly but remained statistically significant. Graduates with meaningful employment in the region continued to show higher average CTE course completion than those not in the workforce or in meaningful employment. However, differences between the non-engaged groups (not employed, not in school) were narrower.

### **Conclusion**

Graduates with **meaningful employment in their home region** consistently took **more CTE courses** than their peers, especially compared to those who pursued post-secondary education. This pattern supports the notion that CTE coursework is more strongly associated with **early and regional employment outcomes**, while those pursuing college took fewer CTE courses overall. The strength of the association appears to **weaken modestly over time**.

::: panel-tabset
### One year after graduating

```{r hs accomplishments total.cte.courses.taken one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, total.cte.courses.taken) %>%
  filter(grad.year.1 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.1) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.1) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.1, total.cte.courses.taken, fill = grad.year.1, group = grad.year.1)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.1, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.1, data = data)

summary(summary)

TukeyHSD(summary)

```

### Five years after graduating

```{r hs accomplishments total.cte.courses.taken five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, total.cte.courses.taken) %>%
  filter(grad.year.5 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.5) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.5) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.5, total.cte.courses.taken, fill = grad.year.5, group = grad.year.5)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.5, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.5, data = data)

summary(summary)

TukeyHSD(summary)

```

### Ten years after graduating

```{r hs accomplishments total.cte.courses.taken ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, total.cte.courses.taken) %>%
  filter(grad.year.10 != "After 2023") %>%
  drop_na() %>%
  group_by(grad.year.10) %>%
  mutate(q1 = quantile(total.cte.courses.taken, .25),
         q3 = quantile(total.cte.courses.taken, .75),
         iqr = IQR(total.cte.courses.taken),
         low.outlier = q1 - 1.5*iqr,
         high.outlier = q3 + 1.5*iqr) %>%
  filter(total.cte.courses.taken > low.outlier,
         total.cte.courses.taken < high.outlier) %>%
  ungroup()

labels <- data %>%
  group_by(grad.year.10) %>%
  summarize(Min = quantile(total.cte.courses.taken, .0),
            Q1 = quantile(total.cte.courses.taken, .25),
            Median = median(total.cte.courses.taken),
            Q3 = quantile(total.cte.courses.taken, .75),
            Max = quantile(total.cte.courses.taken, 1),
            Mean = mean(total.cte.courses.taken)) %>%
  ungroup()

plot <- ggplot(data, aes(grad.year.10, total.cte.courses.taken, fill = grad.year.10, group = grad.year.10)) +
  geom_boxplot_interactive() +
  geom_label(data = labels, aes(x = grad.year.10, y = 5, label = paste("Min: ", comma(Min, accuracy = .1), "\nQ1: ", comma(Q1, accuracy = .1), "\nMedian: ", comma(Median, accuracy = .1), "\nQ3: ", comma(Q3, accuracy = .1), "\nMax: ", comma(Max, accuracy = ), "\nMean: ", comma(Mean, accuracy = .1))), color = "white") +
  labs(x="", y = "", color="", title = "Boxplots of number of CTE courses by employment state")+
  scale_y_continuous(labels=scales::comma,
                     limits = c(0, 6))+
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  theme_bar+
  theme(legend.position = "none",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = plot, width_svg = 10, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE)) 


summary <- aov(total.cte.courses.taken ~ grad.year.10, data = data)

summary(summary)

TukeyHSD(summary)

```
:::

```{r echo = FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, total.cte.courses.taken) %>%
  write_csv(., "Data/ChatGPT/High school accomplishments/total-cte-courses-taken.csv")

```

<br>

# Post-secondary

We will analyze the following variables;

-   attended.ps.within.first.year.hsgrad
-   attended.ps
-   ps.grad
-   ps.grad.InstitutionSector
-   highest.cred.level
-   ps.grad.location

### **Key Findings**

The strongest and most consistent predictors of **meaningful regional employment** and labor market engagement overall are:

-   **Whether an individual graduated** from a postsecondary institution
-   **Where** they graduated (especially regionally)
-   **What type of institution** they graduated from
-   **What level of credential** they earned

All variables in this section showed at least moderate associations with employment outcomes **within the first year** after high school, though the strength of these relationships generally **declined over time**.

### **Interpretation Over Time**

**1 Year After Graduation**\

Postsecondary **participation dominates** the employment landscape this early:

-   Individuals still attending college make up the majority in nearly every group, especially for those who would go on to earn **bachelor’s or higher degrees**, or graduate from **public 4-year institutions**.
-   **Graduates from 2-year institutions** and those with **associate degrees** were **more likely to already be in meaningful regional employment** than their peers.
-   Individuals who **never attended** or had **not yet graduated** were disproportionately in **non-meaningful employment or disconnected** from the labor market (e.g., over 40% not meaningfully employed or with no MN record).

**5 Years After Graduation**\

Graduation status and credential type take center stage:

-   Those with **associate degrees** or graduates from **2-year institutions** had the **highest rates of meaningful regional employment**, often around **30–33%**.
-   In contrast, individuals with **bachelor’s degrees or higher**, or who graduated from **4-year institutions**, were more likely to be in **meaningful employment outside their high school region**, signaling greater **geographic mobility** or **broader employment markets**.
-   The **location** of graduation mattered significantly: those who **graduated from a college in their high school region** had **twice the rate** of meaningful regional employment compared to graduates from MN colleges outside the region.

**10 Years After Graduation**\

Although patterns narrowed, **educational attainment and geography still mattered**:

-   Individuals with **postsecondary degrees** remained far more engaged in the workforce, especially in **meaningful employment**.
-   Those who **graduated locally** still had the **highest rates of meaningful employment in the region**, often exceeding **35%**.
-   People who **never attended or didn’t complete college** were consistently more disconnected, with **\~30–35%** showing **no MN employment record** a full decade later.

### **Conclusion**

The postsecondary journey plays a **central role in shaping long-term employment outcomes**. Specifically:

-   **Completion matters**: individuals who earn degrees—especially associate or higher—are much more likely to be employed meaningfully and remain engaged in the MN workforce.
-   **Location matters**: graduating from a postsecondary institution **in the same region** as high school strongly predicts **regional employment**, suggesting local educational pipelines support local labor market retention.
-   **Mobility varies by institution type**: graduates of **4-year and out-of-region institutions** are more likely to find employment elsewhere in MN, while **2-year grads** tend to stay local.
-   **Disconnection is concentrated** among those who did **not graduate** from college or **never attended**, who are persistently overrepresented among the not meaningfully employed and those with **no MN employment record**.

<br>

## attended.ps.within.first.year.hsgrad

### **Key Findings**

Attending postsecondary immediately after high school is **strongly associated** with future outcomes, especially early on:

-   **1 year after graduation**: Cramer's V = **0.548** → Very strong relationship
-   **5 years after graduation**: Cramer's V = **0.229** → Moderate relationship
-   **10 years after graduation**: Cramer's V = **0.088** → Weak but still statistically meaningful

### **Interpretation Over Time**

**1 Year After Graduation**

This is the **strongest relationship observed** among any postsecondary variables.

-   Nearly all individuals attending postsecondary at year 1 had enrolled immediately after high school.

-   Those who did **not** attend immediately were much more likely to be **disconnected or in non-meaningful employment**.

-   This sharp contrast highlights the early importance of immediate postsecondary transition.

**5 Years After Graduation**

The relationship weakens but remains meaningful.

-   Those who attended college immediately are still **more likely to be in meaningful employment (especially MN-wide)** or **still pursuing postsecondary education**.
-   Those who delayed or never attended show higher rates of **disconnection and non-meaningful employment**.

**10 Years After Graduation**

The gap narrows significantly but still favors early enrollees.

-   Individuals who attended right away are **more concentrated in meaningful employment**, especially statewide.
-   Those who never attended are overrepresented in **disconnected and non-meaningful employment categories**.
-   The fading strength suggests that other life experiences may catch up or intervene over time.

### **Conclusion**

Attending college within the first year after high school is one of the **strongest predictors of positive post-graduation outcomes**, especially in the short term. While the effect diminishes over time, early college enrollment remains a **powerful signal of future labor market success and educational persistence**. This reinforces the importance of **early access, readiness, and support for postsecondary transitions**.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps.within.first.year.hsgrad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps.within.first.year.hsgrad) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps.within.first.year.hsgrad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, attended.ps.within.first.year.hsgrad) %>%
  write_csv(., "Data/ChatGPT/Post-secondary/attended-ps-within-first-year-hsgrad.csv")

```

<br>

## attended.ps

### **Key Findings**

Attending postsecondary is **strongly associated** with better employment and educational outcomes, particularly in terms of **meaningful employment in the region**:

-   **1 year after graduation**: Cramer's V = **0.655** → Very strong association
-   **5 years after graduation**: Cramer's V = **0.321** → Moderate association
-   **10 years after graduation**: Cramer's V = **0.114** → Weaker but still relevant

### **Interpretation Over Time**

**1 Year After Graduation**\

Among those who attended college:

-   **11.4%** were in **meaningful regional employment**
-   **63.4%** were **attending postsecondary**
-   Only **6.9%** were **disconnected** (no MN record or not meaningful)

Among non-attendees:

-   Only **5.2%** were in **meaningful regional employment**
-   **0%** were attending postsecondary
-   Over **40%** were **disconnected**
-   This sharp contrast shows that **early college attendance is closely tied to both educational persistence and labor market connection**.

**5 Years After Graduation**\

Among college attendees:

-   **18.1%** were in **meaningful regional employment**
-   **16.6%** were **still in postsecondary**
-   About **31%** were **disconnected**

Among non-attendees:

-   Only **11.6%** had **meaningful regional employment**
-   Less than **2%** were still in postsecondary
-   Nearly **50%** were **disconnected**
-   The gap in **regional employment** remains substantial — nearly **7 percentage points** higher among college attendees.

**10 Years After Graduation**\

Among those who attended college:

-   **24.1%** had **meaningful employment in the region**
-   Nearly **33%** had **meaningful employment elsewhere in MN**
-   Just **26.4%** were **disconnected**

Among those who never attended:

-   Only **14.1%** had **meaningful regional employment**
-   Over **41%** were **disconnected**

The **10-point advantage in regional employment** persists even a decade later.

### **Conclusion**

Postsecondary attendance is a **critical predictor of meaningful employment in the region** across all time points. Those who attend are:

-   **More likely** to connect with their regional labor market
-   **Less likely** to be disconnected
-   Still showing benefits even a decade later

This underscores the regional workforce value of expanding and supporting **postsecondary access** for high school graduates.

::: panel-tabset
### One year after graduating

```{r post-secondary attended.ps one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, attended.ps) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary attended.ps five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, attended.ps) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary attended.ps ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, attended.ps) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$attended.ps, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, attended.ps) %>%
  write_csv(., "Data/ChatGPT/Post-secondary/attended-ps.csv")


```

<br>

## ps.grad

### **Key Findings**

Graduating from a postsecondary institution is **strongly associated** with better post-graduation outcomes, particularly **regional and statewide employment**:

-   **1 year after graduation**: Cramer's V = **0.393** → Strong relationship
-   **5 years after graduation**: Cramer's V = **0.252** → Moderate relationship
-   **10 years after graduation**: Cramer's V = **0.099** → Weak but still notable

### **Interpretation Over Time**

**1 Year After Graduation**\

Among those who had already graduated college (Y):

-   Only **7.6%** had meaningful employment in the region
-   But **78.9%** were still attending postsecondary — unsurprising this early
-   Those who had not graduated (No) were **over 3x more likely** to be disconnected (e.g., **18.3%** had no MN employment record vs. only **2.8%** among graduates)

Graduates were **more anchored in education or employment**, with much lower rates of disconnection.

**5 Years After Graduation**\

Postsecondary graduates had:

-   **19.4%** in **meaningful regional employment** (vs. 24.4% of non-graduates — a reversal)
-   **27.0%** in **meaningful MN employment** (vs. 20.7% of non-graduates)
-   Much lower disconnection: **11.5%** had no MN employment record (vs. 23.7% of non-graduates)

Graduates were **less regionally concentrated** but had **higher statewide engagement** and were **less disconnected**.

**10 Years After Graduation**\

At this point:

-   **22.4%** of graduates had meaningful employment in the region (vs. 23.7% of non-graduates)
-   But **36.5%** had meaningful employment statewide (vs. 24.5% for non-graduates)
-   Fewer were disconnected: **26.9%** of graduates had no MN record, vs. **31.5%** of non-graduates

Again, postsecondary graduation seems to **shift outcomes from regional to statewide success** — and is associated with **lower long-term disconnection**.

### **Conclusion**

Graduating from a postsecondary institution is consistently linked to **lower disconnection** and **higher rates of meaningful employment statewide**, though not always regionally. In fact, non-graduates slightly exceed graduates in regional employment at years 5 and 10 — suggesting that **graduates may be more mobile or drawn to employment outside their home region**.\

Still, graduation remains a **strong signal of economic integration**, especially **statewide**.

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, ps.grad) %>%
  write_csv(., "Data/ChatGPT/Post-secondary/ps-grad.csv")

```

<br>

## ps.grad.InstitutionSector

### **Key Findings**

The type of postsecondary institution is **strongly related to early outcomes** and **moderately predictive by year 5**, but only **weakly associated** by year 10:

-   **1 year after graduation**: Cramer's V = **0.356** → Strong
-   **5 years after graduation**: Cramer's V = **0.211** → Moderate
-   **10 years after graduation**: Cramer's V = **0.093** → Weak

### **Interpretation Over Time**

**1 Year After Graduation**\

Outcomes are dominated by whether the person is still in school:

-   Among those who eventually graduated from **public 4-year institutions**, **90%** were still attending school.
-   Among **public 2-year** and **other institutions**, around **65–77%** were still enrolled.
-   In contrast, those who **did not graduate** or **never attended** were far more likely to fall into **non-meaningful** or **no MN record** categories.
-   **Meaningful regional employment** was most common (15.7%) among public 2-year graduates — but generally low across all sectors due to ongoing schooling.

**5 Years After Graduation**\

Institutional differences become more apparent:

-   Those who graduated from **public 2-year institutions** had the **highest rate of meaningful regional employment** at **33.4%**, followed by other institutions (15.4%) and public 4-years (13.9%).
-   Graduates from **public 4-year institutions** were more likely to be in **meaningful employment in MN more broadly** (**26.4%**).
-   Non-completers (did not grad, never attended) had **much higher rates of disconnection**, with **28.8%** of "never attended" having no MN employment record.

**10 Years After Graduation**\

Differences persist but narrow:

-   Meaningful regional employment was **highest for public 2-year completers** (33.2%), followed by other institutions (19.1%) and public 4-year grads (18.5%).
-   Public 4-year grads were **most likely to have meaningful MN employment overall** (39%).
-   Disconnection remained highest among those who **never attended college** (**34.3%** had no MN employment record).

### **Conclusion**

Graduates of **public 2-year institutions** consistently show the **highest levels of meaningful regional employment** — especially by 5 and 10 years after high school. In contrast, **public 4-year graduates** are more likely to be employed **elsewhere in the state**, suggesting greater mobility or different job placement patterns.\

Those who never attended or did not complete postsecondary are **far more likely to be disconnected** from the labor market at all time points.\

The data supports the view that **2-year institutions play a unique role in anchoring talent locally**, while 4-year institutions may prepare students for **broader, statewide opportunities**.

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.InstitutionSector one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.InstitutionSector) %>%
  filter(grad.year.1 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.InstitutionSector five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.InstitutionSector) %>%
  filter(grad.year.5 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.InstitutionSector ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.InstitutionSector) %>%
  filter(grad.year.10 != "After 2023") %>%
  filter(ps.grad.InstitutionSector != "Did not grad") %>%
  filter(ps.grad.InstitutionSector != "Never attended ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.InstitutionSector, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, ps.grad.InstitutionSector) %>%
  write_csv(., "Data/ChatGPT/Post-secondary/ps-grad-InstitutionSector.csv")

```

<br>

## highest.cred.level

### **Key Findings**

Educational attainment is a **strong early predictor** of labor market and education outcomes. The strength of this association diminishes over time, but notable patterns persist.

-   **1 year after graduation**: Cramer's V = **0.357** → Strong relationship
-   **5 years after graduation**: Cramer's V = **0.214** → Moderate
-   **10 years after graduation**: Cramer's V = **0.093** → Weak but present

### **Interpretation Over Time**

**1 Year After Graduation**\

Most individuals are still attending postsecondary:

-   **Attending ps** rates were highest among those pursuing a **bachelor’s or higher (87%)**, followed by **associate degrees (67%)**.
-   For those who had **not attended college**, **46%** were already in the **not meaningful** category, and **28%** had **no MN employment record**.
-   Among associate degree seekers, **14.5%** had already achieved **meaningful regional employment**, a higher share than most other groups at this stage.

**5 Years After Graduation**\

Clear employment patterns emerge:

-   Those with **associate degrees** had the **highest proportion** in **meaningful regional employment** (**29.5%**).
-   **Bachelor’s+ degree holders** were more likely to be employed **somewhere in MN** (**27.4%** in meaningful MN employment).
-   Individuals with **no college** or **less than an associate degree** were **far more likely** to be disconnected, with **\~28–29%** having no MN employment record.

**10 Years After Graduation**\

Gaps narrow, but degree level still matters:

-   **Meaningful regional employment** was highest among **associate degree earners (30.7%)** and **those with less than an associate (31.7%)**.
-   Those with a **bachelor’s or higher** were more often employed **elsewhere in MN** (**39%**), rather than regionally (**18.3%**).
-   Individuals with **no college** had a disconnection rate of **34.3%**, nearly **10 percentage points higher** than those with degrees.

### **Conclusion**

Educational attainment — especially an **associate degree** — is consistently associated with **stronger regional labor market outcomes**, particularly by 5 and 10 years.\

Individuals with **bachelor’s degrees or higher** are more likely to be employed **outside their home region**, reflecting broader career mobility.\

Those with **no college experience** remain at high risk of labor market **disconnection**, reinforcing the value of postsecondary pathways for regional and statewide workforce integration.

::: panel-tabset
### One year after graduating

```{r post-secondary highest.cred.level one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, highest.cred.level) %>%
  filter(grad.year.1 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary highest.cred.level five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, highest.cred.level) %>%
  filter(grad.year.5 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary highest.cred.level ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, highest.cred.level) %>%
  filter(grad.year.10 != "After 2023") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$highest.cred.level, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, highest.cred.level) %>%
  write_csv(., "Data/ChatGPT/Post-secondary/highest-cred-level.csv")

```

<br>

## ps.grad.location

### **Key Findings**

Where an individual graduates from college has a **moderate relationship** with their employment outcomes shortly after high school, which **persists but weakens** over time:

-   **1 year after graduation**: Cramer's V = **0.159**
-   **5 years after graduation**: Cramer's V = **0.186**
-   **10 years after graduation**: Cramer's V = **0.147**

### **Interpretation Over Time**

**1 Year After Graduation**

-   Graduates from institutions in the **same region as their high school** had the **highest rate** of **meaningful regional employment (48.7%)**.
-   Those graduating from MN institutions **outside their region** were more likely to have **meaningful MN employment (30.5%)**, but not regionally anchored.
-   Non-completers had high rates of disconnection: **24.3%** had no MN employment record, and **44.4%** were not meaningfully employed nor in school.

**5 Years After Graduation**

-   The regional advantage **intensifies**: those who graduated in the same region had **47.8%** in meaningful regional employment — over **twice as high** as MN grads outside the region (19.6%).
-   Graduates from outside MN were far more likely to be working in MN (26.2%) than regionally (11.9%), suggesting possible **return migration** without regional anchoring.
-   Again, non-completers had high disconnection: **25%** had no MN employment record.

**10 Years After Graduation**

-   Regional graduates continue to show the **strongest ties to regional employment (37.5%)**.
-   Those who graduated from MN colleges outside their region have **higher statewide employment (43%)** but relatively lower regional outcomes (17.8%).
-   Out-of-state grads and non-completers are more often **disconnected**, with **48%** and **31.5%**, respectively, having no MN employment record.

### **Conclusion**

Graduating from a **college in the same region as one’s high school** is consistently associated with **greater regional employment integration**, especially in the first five years after high school.\

Those who graduate from **MN institutions outside their region** or from **out-of-state** are more likely to be employed in Minnesota broadly, but **less likely to be regionally anchored**.\

Meanwhile, those who do not complete college show **higher rates of disconnection**, reinforcing the importance of both **completion** and **local institutional pathways** for regional workforce development.

::: panel-tabset
### One year after graduating

```{r post-secondary ps.grad.location one year}
data <- meaningful.emp.central %>%
  select(grad.year.1, ps.grad.location) %>%
  filter(grad.year.1 != "After 2023",
         grad.year.1 != "Attending ps",
         ps.grad.location != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.1, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Five years after graduating

```{r post-secondary ps.grad.location five year}
data <- meaningful.emp.central %>%
  select(grad.year.5, ps.grad.location) %>%
  filter(grad.year.5 != "After 2023",
         grad.year.5 != "Attending ps",
         ps.grad.location != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.5, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```

### Ten years after graduating

```{r post-secondary ps.grad.location ten year}
data <- meaningful.emp.central %>%
  select(grad.year.10, ps.grad.location) %>%
  filter(grad.year.10 != "After 2023",
         grad.year.10 != "Attending ps",
         ps.grad.location != "Attending ps") %>%
  droplevels()

tab_xtab(var.row = data$grad.year.10, var.col = data$ps.grad.location, 
         title = "", 
         show.exp = TRUE,
         show.col.prc = TRUE)
```
:::

```{r echo=FALSE}
data <- meaningful.emp.central %>%
  select(grad.year.1, grad.year.5, grad.year.10, ps.grad.location) %>%
  write_csv(., "Data/ChatGPT/Post-secondary/ps-grad-location.csv")

```

<br>

# Total Summary

## **1. Strong Predictors of Regional Employment and Disconnection**

These variables **consistently show strong or moderate Cramer's V values** and **clear differences in meaningful employment in the region**, making them strong candidates for inclusion in modeling.

| Variable | Type | Key Insights |
|------------------|------------------|-------------------------------------|
| ps.grad.location | Postsecondary | Strongest predictor of regional employment. Local grads more likely to stay. |
| ps.grad | Postsecondary | Graduation itself reduces disconnection and improves employment outcomes. |
| highest.cred.level | Postsecondary | Clear stepwise value: higher credential = better outcomes. |
| ps.grad.InstitutionSector | Postsecondary | 2-year grads more likely to stay regional; 4-year grads more mobile. |
| economic.status | Demographic | Low-income students consistently more disconnected, especially early. |
| RaceEthnicity | Demographic | Persistent racial gaps. Black, AI, Hispanic students more likely disconnected. |
| Dem_Desc | High School Characteristic | Rural grads more regionally tied; metro students more mobile/disconnected. |
| ACTCompositeScore | Accomplishment (cont.) | Higher scores = more statewide mobility, but not always regional. |
| pseo.participant | Accomplishment | Linked to early employment or enrollment; reduces disconnection. |

## **2. Moderate Predictors — Some Association, Less Consistency**

These variables show **moderate associations** with outcomes. Their predictive value may depend on how they interact with other variables and could be useful in modeling, especially when looking for cumulative effects.

| Variable | Type | Key Insights |
|------------------|------------------|-------------------------------------|
| SpecialEdStatus | Demographic | Somewhat lower employment, higher disconnection. |
| Grad.year.covid | Demographic | Post-COVID grads slightly more disconnected early on. |
| cte.achievement | Accomplishment | Boosts regional employment, especially early. |
| total.cte.courses.taken | Accomplishment (cont.) | More courses = slight boost in employment, especially regionally. |
| avg.unemp.rate | HS Characteristic (cont.) | Higher unemployment → more disconnection. Small effect. |
| avg.wage.pct.state | HS Characteristic (cont.) | Higher wage counties → less disconnection. Small but consistent. |
| MCA Math & Reading | Accomplishment (cont.) | Slight association with higher employment, mostly MN-wide. |

## **3. Weak or Inconsistent Predictors — Candidates for Dropping**

These variables showed **very weak associations** (low Cramer's V, minimal group differences) and could be considered for **dropping from modeling**, unless interaction effects suggest otherwise.

| Variable | Type | Notes |
|------------------|------------------|-------------------------------------|
| english.learner | Demographic | Minimal differences, especially when controlling for other variables. |
| non.english.home | Demographic | Weak signal, may overlap with other demographics. |
| sat.taken | Accomplishment | Tiny population, not informative. |
| took.ACT | Accomplishment | Near-universal participation, little discriminatory value. |
| ap.exam | Accomplishment | Modest effect, likely overlapping with other academic indicators. |

```{r create new master after analysis}
master.after.ct <- meaningful.emp.central %>%
  select(-english.learner, -non.english.home, -sat.taken, -took.ACT, -ap.exam)
  
write_csv(master.after.ct, "Data/SLEDS/Masters/After analysis/Master-after-ct.csv")
```
